---
title: "Rarefaction Analysis - Bacterial Diversity & Resistance Genes (All Conditions)"
author: "Alice Bradbury"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    fig_width: 14
    fig_height: 8
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center"
)
```

# Overview

Rarefaction analyses for **6 samples (CH01–CH06)** across **6 conditions**:

| Condition | Sampling point | Enrichment | Kraken2 | Abricate |
|-----------|---------------|------------|---------|----------|
| ASC       | Post-ASC      | None (direct) | ✓ 6 samples | — no data |
| ASC_CAT   | Post-ASC      | BF Bolton + CAT | ✓ 6 samples | ✓ 6 samples |
| ASC_RVS   | Post-ASC      | BPW + RVS | ✓ 6 samples | ✓ 6 samples |
| EVI       | Post-EVI      | None (direct) | ✓ 6 samples | ⚠ partial (CH01, CH04; card/ncbi only) |
| EVI_CAT   | Post-EVI      | BF Bolton + CAT | ✓ 6 samples | ✓ 6 samples |
| EVI_RVS   | Post-EVI      | BPW + RVS | ✓ 6 samples | ✓ 6 samples |

Curves are interpolated (solid opaque line) and extrapolated to 2× the largest
observed sample size within each condition (long-dashed, semi-transparent line),
with 95% CIs (shaded ribbon, lighter in the extrapolated region), using `iNEXT`. Kraken2 matrices are subsampled to ≤ 50,000 reads per sample
before iNEXT to keep compute time manageable.

---

# 1. Setup

## 1.1 Libraries

```{r libraries}
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
library(vegan)
library(iNEXT)
library(ggplot2)
library(patchwork)
```

## 1.2 Paths & Conditions

All paths and filename patterns are defined here — edit this block to add/remove
conditions or fix directory names.

```{r config}
base_reads <- "/Volumes/Alice/2_POULTRY_PROCESSING/Analysis/Reads"
output_dir <- file.path(base_reads, "rarefaction_output")
dir.create(output_dir, showWarnings = FALSE)

# Named list: one entry per condition.
# abricate_dir = NULL means no abricate data available for that condition.
conditions <- list(
  ASC = list(
    label          = "Post-ASC: No enrichment",
    sampling_point = "Post-ASC",
    enrichment     = "None",
    kraken_dir     = file.path(base_reads, "Kraken2_reads", "ASC_Kraken2_reads"),
    kraken_pat     = "_ASC_reads_kraken2\\.report$",
    abricate_dir   = NULL,
    abricate_pat   = NULL
  ),
  ASC_CAT = list(
    label          = "Post-ASC: Blood-free Bolton base + CAT",
    sampling_point = "Post-ASC",
    enrichment     = "CAT",
    kraken_dir     = file.path(base_reads, "Kraken2_reads", "ASC_CAT_Kraken2_reads"),
    kraken_pat     = "_ASC_CAT_reads_kraken2\\.report$",
    abricate_dir   = file.path(base_reads, "ASC_CAT_reads_abricate"),
    abricate_pat   = "_ASC_CAT_reads_(card|resfinder|ncbi)\\.tsv$"
  ),
  ASC_RVS = list(
    label          = "Post-ASC: BPW then RVS",
    sampling_point = "Post-ASC",
    enrichment     = "RVS",
    kraken_dir     = file.path(base_reads, "Kraken2_reads", "ASC_RVS"),
    kraken_pat     = "_ASC_RVS_reads_kraken2\\.report$",
    abricate_dir   = file.path(base_reads, "ASC_RVS_reads_abrciate"),  # NB: typo in folder name
    abricate_pat   = "_ASC_RVS_reads_(card|resfinder|ncbi)\\.tsv$"
  ),
  EVI = list(
    label          = "Post-EVI: No enrichment",
    sampling_point = "Post-EVI",
    enrichment     = "None",
    kraken_dir     = file.path(base_reads, "Kraken2_reads", "EVI_Kraken2_reads"),
    kraken_pat     = "_EVI_reads_kraken2\\.report$",
    abricate_dir   = file.path(base_reads, "EVI_reads_abricate"),  # partial: CH01+CH04 only
    abricate_pat   = "_EVI_reads_(card|resfinder|ncbi)\\.tsv$"
  ),
  EVI_CAT = list(
    label          = "Post-EVI: Blood-free Bolton base + CAT",
    sampling_point = "Post-EVI",
    enrichment     = "CAT",
    kraken_dir     = file.path(base_reads, "Kraken2_reads", "EVI_CAT_Kraken2_reads"),
    kraken_pat     = "_EVI_CAT_reads_kraken2\\.report$",
    abricate_dir   = file.path(base_reads, "EVI_CAT_reads_abricate"),
    abricate_pat   = "_EVI_CAT_reads_(card|resfinder|ncbi)\\.tsv$"
  ),
  EVI_RVS = list(
    label          = "Post-EVI: BPW then RVS",
    sampling_point = "Post-EVI",
    enrichment     = "RVS",
    kraken_dir     = file.path(base_reads, "Kraken2_reads", "EVI_RVS"),
    kraken_pat     = "_EVI_RVS_reads_kraken2\\.report$",
    abricate_dir   = file.path(base_reads, "EVI_RVS_reads_abricate"),
    abricate_pat   = "_EVI_RVS_reads_(card|resfinder|ncbi)\\.tsv$"
  )
)

# Ordered factor levels so facets appear in the right order (ASC row, then EVI row)
condition_label_order <- map_chr(conditions, "label")
```

## 1.3 Colour Palettes

```{r colours}
# One colour per sample — used consistently across all plots
sample_colours <- c(
  CH01 = "#E63946",
  CH02 = "#F4A261",
  CH03 = "#2A9D8F",
  CH04 = "#457B9D",
  CH05 = "#8338EC",
  CH06 = "#FB5607"
)

# Blues for post-ASC, oranges for post-EVI (dark = enriched)
condition_colours <- c(
  ASC     = "#A6CEE3",
  ASC_CAT = "#1F78B4",
  ASC_RVS = "#08519C",
  EVI     = "#FDBE85",
  EVI_CAT = "#E6550D",
  EVI_RVS = "#7F2704"
)
```

---

# 2. Kraken2 - Bacterial Diversity

## 2.1 Load, Subsample & Run iNEXT

Genus-level (`rank == "G"`) rows are extracted from each `.report` file using
`reads_covered` (total reads in the genus clade, including all species beneath it).
`reads_direct` is not used as it only counts reads assigned to the genus node itself
and is near zero for most genera — species-classified reads dominate.
Each condition's matrix is subsampled to ≤ 50,000 reads (or the condition minimum,
whichever is lower) before iNEXT — raw metagenomic depths are too large for iNEXT
to handle efficiently. iNEXT is then run per condition; results are combined into a
single tidy data frame for plotting.

```{r kraken-load, cache=TRUE}
read_kraken_report <- function(filepath, sample_name) {
  df <- read.table(filepath, sep = "\t", header = FALSE,
                   col.names = c("pct", "reads_covered", "reads_direct",
                                 "rank", "taxid", "name"),
                   fill = TRUE, quote = "")
  df %>%
    filter(rank == "G", reads_covered > 0) %>%
    mutate(name = trimws(name), sample = sample_name) %>%
    select(sample, name, reads_covered)
}

build_kraken_matrix <- function(cfg) {
  files <- list.files(cfg$kraken_dir, pattern = "\\.report$", full.names = TRUE)
  stopifnot("No .report files found" = length(files) > 0)

  samples <- sub(cfg$kraken_pat, "", basename(files))
  long    <- bind_rows(mapply(read_kraken_report, files, samples, SIMPLIFY = FALSE))

  mat <- long %>%
    pivot_wider(names_from = name, values_from = reads_covered, values_fill = 0L) %>%
    as.data.frame()
  rownames(mat) <- mat$sample
  mat$sample    <- NULL
  round(mat)
}

kraken_inext_df <- map_dfr(names(conditions), function(cond) {
  cfg <- conditions[[cond]]
  message("Kraken2 — ", cfg$label)

  mat <- build_kraken_matrix(cfg)
  mat <- mat[rowSums(mat) > 0, , drop = FALSE]

  min_depth       <- min(rowSums(mat))
  subsample_depth <- min(min_depth, 50000L)
  message("  subsampling to ", subsample_depth, " reads (min = ", min_depth, ")")

  set.seed(42)
  sub <- rrarefy(mat, sample = subsample_depth)
  sub <- sub[, colSums(sub) > 0, drop = FALSE]

  inext_list <- lapply(rownames(sub), function(s) as.numeric(sub[s, sub[s, ] > 0]))
  names(inext_list) <- rownames(sub)

  out <- iNEXT(inext_list, q = 0, datatype = "abundance",
               endpoint = 2L * subsample_depth, nboot = 20)
  saveRDS(out, file.path(output_dir, paste0("inext_kraken_", cond, ".rds")))

  out$iNextEst$size_based %>%
    mutate(
      condition       = cond,
      condition_label = factor(cfg$label, levels = condition_label_order),
      sample          = Assemblage
    )
})
```

## 2.2 Summary Table

```{r kraken-summary}
kraken_inext_df %>%
  filter(grepl("observed", Method, ignore.case = TRUE)) %>%
  select(condition_label, sample, m, qD) %>%
  rename(Condition = condition_label, Sample = sample,
         Reads_subsampled = m, Genera_observed = qD) %>%
  arrange(Condition, Sample)
```

## 2.3 Rarefaction Curves — All Conditions

```{r kraken-plot, fig.width=16, fig.height=10, fig.cap="Bacterial genus richness. Solid opaque = interpolated; long-dashed transparent = extrapolated (2x subsampled depth); ribbon = 95% CI. Top row: post-ASC; bottom row: post-EVI."}
p_kraken <- ggplot(kraken_inext_df,
                   aes(x = m, y = qD, colour = sample, fill = sample,
                       group = sample)) +
  geom_ribbon(data = \(d) filter(d, Method %in% c("Interpolated", "Rarefaction")),
              aes(ymin = qD.LCL, ymax = qD.UCL), alpha = 0.18, colour = NA) +
  geom_ribbon(data = \(d) filter(d, Method %in% c("Extrapolated", "Extrapolation")),
              aes(ymin = qD.LCL, ymax = qD.UCL), alpha = 0.07, colour = NA) +
  geom_line(data = \(d) filter(d, Method %in% c("Interpolated", "Rarefaction")),
            linewidth = 0.9) +
  geom_line(data = \(d) filter(d, Method %in% c("Extrapolated", "Extrapolation")),
            linewidth = 0.8, linetype = "longdash", alpha = 0.45) +
  geom_point(data = \(d) filter(d, Method == "Observed" | grepl("observed", Method, ignore.case = TRUE)),
             size = 2.5, shape = 21, colour = "white", stroke = 0.6) +
  facet_wrap(~ condition_label, scales = "free_x", ncol = 3) +
  scale_colour_manual(values = sample_colours) +
  scale_fill_manual(values = sample_colours) +
  labs(
    title  = "Bacterial Genus Diversity (Kraken2, genus level)",
    x      = "Reads subsampled",
    y      = "Genus richness",
    colour = "Sample",
    fill   = "Sample"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    panel.grid.minor = element_blank(),
    legend.position  = "right"
  )

p_kraken

ggsave(file.path(output_dir, "kraken2_diversity_all_conditions.png"),
       p_kraken, width = 16, height = 10, dpi = 600)
message("Saved: kraken2_diversity_all_conditions.png")
```

---

# 3. Abricate — Resistance Genes & Classes

> **Note:** The post-ASC no-enrichment condition has no Abricate data.
> The post-EVI no-enrichment condition has partial data (CH01 and CH04 only;
> CARD and NCBI databases only). These samples are included where data exists
> but the curves should be interpreted cautiously.

## 3.1 Load All Conditions

Each `.tsv` may contain repeated `#FILE` header lines (Abricate concatenation
artefact); these are stripped before parsing.

```{r abricate-load}
read_abricate_file <- function(filepath, sample_name) {
  lines <- readLines(filepath, warn = FALSE)
  header_idx <- grep("^#FILE", lines)
  if (length(header_idx) == 0) return(NULL)

  clean <- c(lines[header_idx[1]], lines[!grepl("^#FILE", lines)])
  df <- tryCatch(
    read.table(text = paste(clean, collapse = "\n"),
               sep = "\t", header = TRUE, quote = "",
               comment.char = "", fill = TRUE),
    error = function(e) NULL
  )
  if (is.null(df) || nrow(df) == 0) return(NULL)

  colnames(df)[1] <- "FILE"
  df %>%
    filter(!is.na(GENE), GENE != "") %>%
    mutate(sample = sample_name) %>%
    select(sample, GENE, RESISTANCE, DATABASE)
}

abricate_long <- map_dfr(names(conditions), function(cond) {
  cfg <- conditions[[cond]]
  if (is.null(cfg$abricate_dir)) return(NULL)

  files <- list.files(cfg$abricate_dir, pattern = "\\.tsv$", full.names = TRUE)
  if (length(files) == 0) { message("No .tsv files for: ", cond); return(NULL) }
  message("Abricate — ", cfg$label, " (", length(files), " files)")

  map_dfr(files, function(f) {
    sname <- sub(cfg$abricate_pat, "", basename(f))
    df    <- read_abricate_file(f, sname)
    if (!is.null(df))
      mutate(df, condition = cond, condition_label = cfg$label)
  })
})
```

## 3.2 Summary

```{r abricate-summary}
abricate_long %>%
  count(condition_label, sample, DATABASE) %>%
  pivot_wider(names_from = DATABASE, values_from = n, values_fill = 0L) %>%
  rename(Condition = condition_label, Sample = sample) %>%
  arrange(Condition, Sample)
```

## 3.3 Shared Helpers

```{r helpers}
make_matrix <- function(df, feature_col) {
  df %>%
    count(sample, .data[[feature_col]]) %>%
    pivot_wider(names_from = all_of(feature_col), values_from = n,
                values_fill = 0L) %>%
    as.data.frame() %>%
    { rownames(.) <- .$sample; .$sample <- NULL; . }
}

run_inext <- function(mat, cond, cfg) {
  mat <- mat[rowSums(mat) > 0, , drop = FALSE]
  if (nrow(mat) == 0) return(NULL)

  inext_list <- lapply(rownames(mat), function(s) as.numeric(mat[s, mat[s, ] > 0]))
  names(inext_list) <- rownames(mat)

  out <- iNEXT(inext_list, q = 0, datatype = "abundance",
               endpoint = 2L * max(rowSums(mat)), nboot = 20)

  out$iNextEst$size_based %>%
    mutate(
      condition       = cond,
      condition_label = factor(cfg$label, levels = condition_label_order),
      sample          = Assemblage
    )
}

# Conditions that have abricate data (in label order for facets)
abricate_conds <- names(conditions)[
  !map_lgl(conditions, ~ is.null(.x$abricate_dir)) &
  map_chr(conditions, "enrichment") != "None"
]
abricate_label_order <- map_chr(conditions[abricate_conds], "label")

plot_rarefaction <- function(df, title, xlab, ylab) {
  df <- df %>%
    mutate(condition_label = factor(condition_label, levels = abricate_label_order))

  ggplot(df, aes(x = m, y = qD, colour = sample, fill = sample, group = sample)) +
    geom_ribbon(data = \(d) filter(d, Method %in% c("Interpolated", "Rarefaction")),
                aes(ymin = qD.LCL, ymax = qD.UCL), alpha = 0.18, colour = NA) +
    geom_ribbon(data = \(d) filter(d, Method %in% c("Extrapolated", "Extrapolation")),
                aes(ymin = qD.LCL, ymax = qD.UCL), alpha = 0.07, colour = NA) +
    geom_line(data = \(d) filter(d, Method %in% c("Interpolated", "Rarefaction")),
              linewidth = 0.9) +
    geom_line(data = \(d) filter(d, Method %in% c("Extrapolated", "Extrapolation")),
              linewidth = 0.8, linetype = "longdash", alpha = 0.45) +
    geom_point(data = \(d) filter(d, Method == "Observed" | grepl("observed", Method, ignore.case = TRUE)),
               size = 2.5, shape = 21, colour = "white", stroke = 0.6) +
    facet_wrap(~ condition_label, scales = "free_x", ncol = 3) +
    scale_colour_manual(values = sample_colours) +
    scale_fill_manual(values = sample_colours) +
    labs(title = title, x = xlab, y = ylab, colour = "Sample", fill = "Sample") +
    theme_bw(base_size = 12) +
    theme(
      plot.title       = element_text(face = "bold", size = 13),
      panel.grid.minor = element_blank(),
      legend.position  = "right"
    )
}
```

---

## 3.4 Resistance Gene Richness — All Databases Combined

```{r genes-all, fig.width=16, fig.height=10, fig.cap="Resistance gene richness (CARD + ResFinder + NCBI combined). Top row: post-ASC; bottom row: post-EVI. EVI no-enrichment shown with partial data only."}
gene_df_all <- map_dfr(abricate_conds, function(cond) {
  sub_df <- filter(abricate_long, condition == cond)
  if (nrow(sub_df) == 0) return(NULL)
  run_inext(make_matrix(sub_df, "GENE"), cond, conditions[[cond]])
})

p_genes_all <- plot_rarefaction(
  gene_df_all,
  title = "Resistance Gene Richness - All Databases (CARD + ResFinder + NCBI)",
  xlab  = "Hits sampled",
  ylab  = "Unique resistance genes"
)
p_genes_all

ggsave(file.path(output_dir, "resistance_genes_all_databases_all_conditions.png"),
       p_genes_all, width = 16, height = 10, dpi = 300)
message("Saved: resistance_genes_all_databases_all_conditions.png")
```

---

## 3.5 Resistance Class Richness — All Databases Combined

```{r classes-all, fig.width=16, fig.height=10, fig.cap="Resistance class richness across all conditions."}
class_df_all <- map_dfr(abricate_conds, function(cond) {
  sub_df <- filter(abricate_long, condition == cond,
                   !is.na(RESISTANCE), RESISTANCE != "")
  if (nrow(sub_df) == 0) return(NULL)
  run_inext(make_matrix(sub_df, "RESISTANCE"), cond, conditions[[cond]])
})

p_classes_all <- plot_rarefaction(
  class_df_all,
  title = "Resistance Class Richness - All Databases",
  xlab  = "Hits sampled",
  ylab  = "Unique resistance classes"
)
p_classes_all

ggsave(file.path(output_dir, "resistance_classes_all_databases_all_conditions.png"),
       p_classes_all, width = 16, height = 10, dpi = 300)
message("Saved: resistance_classes_all_databases_all_conditions.png")
```

---

## 3.6 Per-Database Breakdown

### Resistance Gene Richness

```{r genes-per-db, fig.width=16, fig.height=10}
databases <- sort(unique(abricate_long$DATABASE))

for (db in databases) {
  df_db <- map_dfr(abricate_conds, function(cond) {
    sub_df <- filter(abricate_long, condition == cond, DATABASE == db)
    if (nrow(sub_df) == 0) return(NULL)
    run_inext(make_matrix(sub_df, "GENE"), cond, conditions[[cond]])
  })
  if (is.null(df_db) || nrow(df_db) == 0) next

  p <- plot_rarefaction(df_db,
    title = paste0("Resistance Gene Richness - ", toupper(db)),
    xlab  = "Hits sampled", ylab = "Unique resistance genes")
  print(p)
  fname <- paste0("resistance_genes_", tolower(db), "_all_conditions.png")
  ggsave(file.path(output_dir, fname), p, width = 16, height = 10, dpi = 300)
  message("Saved: ", fname)
}
```

### Resistance Class Richness

```{r classes-per-db, fig.width=16, fig.height=10}
for (db in databases) {
  df_db <- map_dfr(abricate_conds, function(cond) {
    sub_df <- filter(abricate_long, condition == cond, DATABASE == db,
                     !is.na(RESISTANCE), RESISTANCE != "")
    if (nrow(sub_df) == 0) return(NULL)
    run_inext(make_matrix(sub_df, "RESISTANCE"), cond, conditions[[cond]])
  })
  if (is.null(df_db) || nrow(df_db) == 0) next

  p <- plot_rarefaction(df_db,
    title = paste0("Resistance Class Richness - ", toupper(db)),
    xlab  = "Hits sampled", ylab = "Unique resistance classes")
  print(p)
  fname <- paste0("resistance_classes_", tolower(db), "_all_conditions.png")
  ggsave(file.path(output_dir, fname), p, width = 16, height = 10, dpi = 300)
  message("Saved: ", fname)
}
```

---

# 4. MEGARes — Resistance Genes & Drug Classes

MEGARes data comes from ABRicate screening against the MEGARes v3 database.
`all_samples_BPKG_long.tsv` holds one row per gene per sample with two distinct
count columns:

- **`n_reads`** — raw integer count of reads mapping to that gene (one row per
  read in the original ABRicate TSV). This is confirmed: 28 raw ABRicate
  alignments for CAT in barcode01 ASC_BFBB = `n_reads = 28` in this file.
- **`BPKG`** — the normalised value (reads per kilobase per genome equivalent).
  **This is NOT used here.** iNEXT requires integer abundance counts.

Only enriched conditions are present (BF Bolton + CAT and BPW + RVS, at both
sampling points); no-enrichment samples were not screened against MEGARes.

The MEGARes sample naming uses `BFBB` (BF Bolton Broth Base) for the CAT-enriched
condition. Conditions are mapped to the same labels used in sections 2 and 3 above.
Drug class is parsed from the hierarchical `product` annotation
(`Type:Class:Mechanism:Gene` — e.g. `Drugs:Phenicol:...:CAT`).

## 4.1 Load & Parse

```{r megares-load}
megares_raw <- read.delim(
  file.path(base_reads, "all_samples_BPKG_long.tsv"),
  sep = "\t", header = TRUE, quote = "", stringsAsFactors = FALSE
) %>%
  filter(database == "megares", n_reads > 0)
# n_reads = raw read count; BPKG column (normalised) is intentionally not used

# Map BFBB → CAT to match the condition keys used throughout this document
megares_cond_map <- c(
  ASC_BFBB = "ASC_CAT",
  ASC_RVS  = "ASC_RVS",
  EVI_BFBB = "EVI_CAT",
  EVI_RVS  = "EVI_RVS"
)

megares_long <- megares_raw %>%
  mutate(
    sample_id    = sub("_(ASC|EVI)_(BFBB|RVS)$", "", sample),
    cond_raw     = sub("^CH\\d+_", "", sample),
    condition    = megares_cond_map[cond_raw],
    drug_class   = sapply(strsplit(product, ":"), function(x)
                     if (length(x) >= 2) x[2] else x[1])
  ) %>%
  filter(!is.na(condition)) %>%
  select(sample = sample_id, condition, GENE, drug_class, n_reads)

megares_conds <- sort(unique(megares_long$condition)) %>%
  .[map_chr(conditions[.], "enrichment") != "None"]
megares_label_order <- map_chr(conditions[megares_conds], "label")
```

## 4.2 Summary

```{r megares-summary}
megares_long %>%
  group_by(condition, sample) %>%
  summarise(total_reads = sum(n_reads), unique_genes = n_distinct(GENE),
            .groups = "drop") %>%
  mutate(condition_label = map_chr(condition, ~ conditions[[.x]]$label)) %>%
  select(Condition = condition_label, Sample = sample,
         Total_reads = total_reads, Unique_genes = unique_genes) %>%
  arrange(Condition, Sample)
```

## 4.3 Shared iNEXT Helper (MEGARes)

iNEXT treats `n_reads` as abundance counts — the x-axis reflects reads subsampled
from the MEGARes-mapped read pool, and the y-axis is unique genes detected.

```{r megares-helper}
run_inext_megares <- function(mat, cond) {
  mat <- mat[rowSums(mat) > 0, , drop = FALSE]
  if (nrow(mat) == 0) return(NULL)

  inext_list <- lapply(rownames(mat), function(s) as.numeric(mat[s, mat[s, ] > 0]))
  names(inext_list) <- rownames(mat)

  out <- iNEXT(inext_list, q = 0, datatype = "abundance",
               endpoint = 2L * max(rowSums(mat)), nboot = 20)

  out$iNextEst$size_based %>%
    mutate(
      condition       = cond,
      condition_label = factor(conditions[[cond]]$label,
                               levels = megares_label_order),
      sample          = Assemblage
    )
}

plot_rarefaction_megares <- function(df, title, xlab, ylab) {
  ggplot(df, aes(x = m, y = qD, colour = sample, fill = sample, group = sample)) +
    geom_ribbon(data = \(d) filter(d, Method %in% c("Interpolated", "Rarefaction")),
                aes(ymin = qD.LCL, ymax = qD.UCL), alpha = 0.18, colour = NA) +
    geom_ribbon(data = \(d) filter(d, Method %in% c("Extrapolated", "Extrapolation")),
                aes(ymin = qD.LCL, ymax = qD.UCL), alpha = 0.07, colour = NA) +
    geom_line(data = \(d) filter(d, Method %in% c("Interpolated", "Rarefaction")),
              linewidth = 0.9) +
    geom_line(data = \(d) filter(d, Method %in% c("Extrapolated", "Extrapolation")),
              linewidth = 0.8, linetype = "longdash", alpha = 0.45) +
    geom_point(data = \(d) filter(d, Method == "Observed" | grepl("observed", Method, ignore.case = TRUE)),
               size = 2.5, shape = 21, colour = "white", stroke = 0.6) +
    facet_wrap(~ condition_label, scales = "free_x", ncol = 2) +
    scale_colour_manual(values = sample_colours) +
    scale_fill_manual(values = sample_colours) +
    labs(title = title, x = xlab, y = ylab, colour = "Sample", fill = "Sample") +
    theme_bw(base_size = 12) +
    theme(
      plot.title       = element_text(face = "bold", size = 13),
      panel.grid.minor = element_blank(),
      legend.position  = "right"
    )
}
```

## 4.4 Resistance Gene Richness

```{r megares-genes, fig.width=14, fig.height=9, fig.cap="MEGARes resistance gene richness. Enriched conditions only (no-enrichment samples not available). 2x2 grid: top row post-ASC, bottom row post-EVI."}
megares_gene_df <- map_dfr(megares_conds, function(cond) {
  mat <- megares_long %>%
    filter(condition == cond) %>%
    select(sample, GENE, n_reads) %>%
    pivot_wider(names_from = GENE, values_from = n_reads,
                values_fn = sum, values_fill = 0L) %>%
    as.data.frame() %>%
    { rownames(.) <- .$sample; .$sample <- NULL; . }
  run_inext_megares(mat, cond)
})

p_megares_genes <- plot_rarefaction_megares(
  megares_gene_df,
  title = "Resistance Gene Richness - MEGARes",
  xlab  = "Reads subsampled (MEGARes-mapped)",
  ylab  = "Unique resistance genes"
)
p_megares_genes

ggsave(file.path(output_dir, "megares_resistance_genes_all_conditions.png"),
       p_megares_genes, width = 14, height = 9, dpi = 600)
message("Saved: megares_resistance_genes_all_conditions.png")
```

## 4.5 Drug Class Richness

```{r megares-classes, fig.width=14, fig.height=9, fig.cap="MEGARes drug class richness (second level of MEGARes hierarchy)."}
megares_class_df <- map_dfr(megares_conds, function(cond) {
  mat <- megares_long %>%
    filter(condition == cond) %>%
    select(sample, drug_class, n_reads) %>%
    pivot_wider(names_from = drug_class, values_from = n_reads,
                values_fn = sum, values_fill = 0L) %>%
    as.data.frame() %>%
    { rownames(.) <- .$sample; .$sample <- NULL; . }
  run_inext_megares(mat, cond)
})

p_megares_classes <- plot_rarefaction_megares(
  megares_class_df,
  title = "Drug Class Richness - MEGARes",
  xlab  = "Reads subsampled (MEGARes-mapped)",
  ylab  = "Unique drug classes"
)
p_megares_classes

ggsave(file.path(output_dir, "megares_drug_classes_all_conditions.png"),
       p_megares_classes, width = 14, height = 9, dpi = 600)
message("Saved: megares_drug_classes_all_conditions.png")
```

---

# 5. Combined Summary Panel

```{r combined-panel, fig.width=16, fig.height=20, fig.cap="Top: bacterial diversity (all 6 conditions). Middle: Abricate resistance gene richness. Bottom: MEGARes resistance gene richness."}
p_combined <- p_kraken / p_megares_genes / p_megares_classes +
  plot_annotation(
    title = "Rarefaction Analysis",
    theme = theme(plot.title = element_text(face = "bold", size = 15, hjust = 0.5))
  )

p_combined

ggsave(file.path(output_dir, "combined_rarefaction_all_conditions.png"),
       p_combined, width = 16, height = 20, dpi = 600)
message("Saved: combined_rarefaction_all_conditions.png")
```

---

# 6. Session Info

```{r session-info}
sessionInfo()
```
