---
title: "AMR Resistance Genes: ASC vs EVI — CAT & RVS Contigs"
author: "Alice Bradbury"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 12, fig.height = 7)
```

```{r packages}
library(tidyverse)
library(pheatmap)
library(RColorBrewer)
library(ggplot2)
library(patchwork)
library(scales)
```

---

# Data Loading

```{r load-data}
read_abricate_contigs_dir <- function(dir_path) {
  files <- list.files(dir_path, pattern = "\\.tsv$", full.names = TRUE)

  map_dfr(files, function(f) {
    df <- read_tsv(f, col_types = cols(.default = "c"), comment = "",
                   show_col_types = FALSE)
    if (nrow(df) == 0) return(NULL)

    fname          <- basename(f)
    sample_id      <- str_extract(fname, "^CH\\d+")
    sample_type_raw <- str_extract(fname, "(ASC|EVI)")
    sample_type    <- ifelse(sample_type_raw == "ASC", "Post-ASC", "Post-EVI")
    database       <- str_extract(fname, "(card|ncbi|resfinder)(?=\\.tsv)")

    df %>%
      rename_with(~ str_remove(., "^#")) %>%
      mutate(
        sample_type = sample_type,
        sample_id   = sample_id,
        database    = database,
        `%COVERAGE` = as.numeric(`%COVERAGE`),
        `%IDENTITY` = as.numeric(`%IDENTITY`)
      )
  })
}

filter_abricate <- function(data) {
  data %>% filter(`%COVERAGE` >= 80, `%IDENTITY` >= 90)
}

cat_data <- filter_abricate(bind_rows(
  read_abricate_contigs_dir("ABRicate_contigs/ASC_CAT"),
  read_abricate_contigs_dir("ABRicate_contigs/EVI_CAT")
))

rvs_data <- filter_abricate(bind_rows(
  read_abricate_contigs_dir("ABRicate_contigs/ASC_RVS"),
  read_abricate_contigs_dir("ABRicate_contigs/EVI_RVS")
))

cat("CAT hits (post-filter):", nrow(cat_data), "\n")
cat("RVS hits (post-filter):", nrow(rvs_data), "\n")

# Factor order: Post-EVI first throughout all plots
evi_first <- c("Post-EVI", "Post-ASC")
cat_data  <- cat_data  %>% mutate(sample_type = factor(sample_type, levels = evi_first))
rvs_data  <- rvs_data  %>% mutate(sample_type = factor(sample_type, levels = evi_first))

# Host-depleted read counts (post-QC + host DNA removal) — assemblies built from these reads
read_counts <- tribble(
  ~sample_type, ~sample_id, ~enrich_method,          ~filtered_reads,
  "Post-ASC",   "CH01",     "BF Bolton base + CAT",  1009220,
  "Post-ASC",   "CH02",     "BF Bolton base + CAT",  2074979,
  "Post-ASC",   "CH03",     "BF Bolton base + CAT",  2558130,
  "Post-ASC",   "CH04",     "BF Bolton base + CAT",    13862,
  "Post-ASC",   "CH05",     "BF Bolton base + CAT",  3809387,
  "Post-ASC",   "CH06",     "BF Bolton base + CAT",   763614,
  "Post-EVI",   "CH01",     "BF Bolton base + CAT",  1720176,
  "Post-EVI",   "CH02",     "BF Bolton base + CAT",  5286510,
  "Post-EVI",   "CH03",     "BF Bolton base + CAT",  2483850,
  "Post-EVI",   "CH04",     "BF Bolton base + CAT",  4470527,
  "Post-EVI",   "CH05",     "BF Bolton base + CAT",  1684200,
  "Post-EVI",   "CH06",     "BF Bolton base + CAT",  3775897,
  "Post-ASC",   "CH01",     "BPW + RVS",             4627620,
  "Post-ASC",   "CH02",     "BPW + RVS",             2624242,
  "Post-ASC",   "CH03",     "BPW + RVS",               4251,
  "Post-ASC",   "CH04",     "BPW + RVS",               5585,
  "Post-ASC",   "CH05",     "BPW + RVS",             2184262,
  "Post-ASC",   "CH06",     "BPW + RVS",             1845128,
  "Post-EVI",   "CH01",     "BPW + RVS",             1041538,
  "Post-EVI",   "CH02",     "BPW + RVS",             3083490,
  "Post-EVI",   "CH03",     "BPW + RVS",             1596293,
  "Post-EVI",   "CH04",     "BPW + RVS",             1406725,
  "Post-EVI",   "CH05",     "BPW + RVS",             1932407,
  "Post-EVI",   "CH06",     "BPW + RVS",              949873
)

cat_data <- cat_data %>%
  left_join(read_counts %>% filter(enrich_method == "BF Bolton base + CAT"),
            by = c("sample_type", "sample_id"))

rvs_data <- rvs_data %>%
  left_join(read_counts %>% filter(enrich_method == "BPW + RVS"),
            by = c("sample_type", "sample_id"))

# Minimum read threshold — excludes low-depth samples from normalised analyses
min_reads    <- 100000
cat_data_rpm <- cat_data %>% filter(filtered_reads >= min_reads)
rvs_data_rpm <- rvs_data %>% filter(filtered_reads >= min_reads)

cat("Samples excluded from normalised analyses (< 100,000 filtered reads):\n")
print(read_counts %>% filter(filtered_reads < min_reads) %>%
        select(sample_type, sample_id, enrich_method, filtered_reads))
```

---

# Helper Functions

```{r helpers}
dir.create("figures_contigs", showWarnings = FALSE)

# Colours
db_colours    <- c(card = "#E64B35", ncbi = "#4DBBD5", resfinder = "#00A087")
db_labels     <- c(card = "CARD", ncbi = "NCBI", resfinder = "ResFinder")
type_colours  <- c(`Post-ASC` = "#F39B7F", `Post-EVI` = "#8491B4")
annot_colours <- list(Sample_Type = c(`Post-ASC` = "#F39B7F", `Post-EVI` = "#8491B4"))

pub_theme <- theme_classic(base_size = 14) +
  theme(
    strip.background  = element_rect(fill = "#3C5488", colour = NA),
    strip.text        = element_text(colour = "white", face = "bold", size = 13),
    axis.text         = element_text(colour = "black"),
    axis.title        = element_text(face = "bold"),
    legend.title      = element_text(face = "bold"),
    legend.position   = "bottom",
    plot.title        = element_text(face = "bold", size = 14),
    panel.border      = element_rect(colour = "black", fill = NA, linewidth = 0.6)
  )

save_gg <- function(p, filename, w = 12, h = 7) {
  ggsave(file.path("figures_contigs", filename), plot = p,
         width = w, height = h, units = "in", dpi = 600)
}

save_pheatmap <- function(hm, filename, w, h) {
  png(file.path("figures_contigs", filename), width = w, height = h,
      units = "in", res = 600)
  grid::grid.newpage()
  grid::grid.draw(hm$gtable)
  dev.off()
}

# ── Bar charts ────────────────────────────────────────────────────────────────
plot_gene_counts <- function(data, title_prefix) {
  count_data <- data %>%
    group_by(sample_type, sample_id, database) %>%
    summarise(n_hits = n(), n_unique_genes = n_distinct(GENE), .groups = "drop") %>%
    mutate(database = recode(database, !!!db_labels))

  p_hits <- ggplot(count_data,
                   aes(x = sample_id, y = n_hits, fill = database)) +
    geom_col(position = "dodge", colour = "white", linewidth = 0.3) +
    facet_wrap(~ sample_type) +
    scale_fill_manual(values = setNames(db_colours, db_labels), name = "Database") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
    labs(title = paste0("(A) ", title_prefix, " — Total AMR Gene Hits per Sample"),
         subtitle = "Coverage \u2265 80%, Identity \u2265 90%",
         x = NULL, y = "Number of Hits") +
    pub_theme +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  p_unique <- ggplot(count_data,
                     aes(x = sample_id, y = n_unique_genes, fill = database)) +
    geom_col(position = "dodge", colour = "white", linewidth = 0.3) +
    facet_wrap(~ sample_type) +
    scale_fill_manual(values = setNames(db_colours, db_labels), name = "Database") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
    labs(title = paste0("(B) ", title_prefix, " — Unique AMR Genes per Sample"),
         x = "Sample", y = "Number of Unique Genes") +
    pub_theme +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  combined <- p_hits / p_unique
  print(combined)
  save_gg(combined, paste0(title_prefix, "_gene_counts.png"), w = 12, h = 10)
}

plot_comparison <- function(data, title_prefix) {
  compare_data <- data %>%
    group_by(sample_type, database) %>%
    summarise(n_unique_genes = n_distinct(GENE), .groups = "drop") %>%
    mutate(database = recode(database, !!!db_labels))

  p <- ggplot(compare_data,
              aes(x = database, y = n_unique_genes, fill = sample_type)) +
    geom_col(position = "dodge", colour = "white", linewidth = 0.4) +
    geom_text(aes(label = n_unique_genes),
              position = position_dodge(width = 0.9), vjust = -0.5, size = 4.5,
              fontface = "bold") +
    scale_fill_manual(values = type_colours, name = "Sample Type") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
    labs(title = paste0(title_prefix, " — Unique AMR Genes: Post-ASC vs Post-EVI by Database"),
         x = "Database", y = "Number of Unique Genes") +
    pub_theme

  print(p)
  save_gg(p, paste0(title_prefix, "_comparison.png"))
}

plot_resistance_classes <- function(data, title_prefix) {
  res_count <- data %>%
    filter(!is.na(RESISTANCE), RESISTANCE != "") %>%
    mutate(RESISTANCE = str_split(RESISTANCE, ";")) %>%
    unnest(RESISTANCE) %>%
    mutate(RESISTANCE = str_trim(str_to_title(RESISTANCE))) %>%
    filter(RESISTANCE != "") %>%
    group_by(sample_type, RESISTANCE) %>%
    summarise(total = n(), .groups = "drop") %>%
    group_by(sample_type) %>%
    slice_max(order_by = total, n = 15) %>%
    ungroup()

  p <- ggplot(res_count,
              aes(x = reorder_within(RESISTANCE, total, sample_type),
                  y = total, fill = sample_type)) +
    geom_col(show.legend = FALSE) +
    facet_wrap(~ sample_type, scales = "free_y") +
    scale_x_reordered() +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
    scale_fill_manual(values = type_colours) +
    coord_flip() +
    labs(title = paste0(title_prefix, " — Top 15 AMR Resistance Classes"),
         x = NULL, y = "Total Hits") +
    pub_theme +
    theme(axis.text.y = element_text(size = 12))

  print(p)
  save_gg(p, paste0(title_prefix, "_resistance_classes.png"), w = 13, h = 7)
}

# ── Presence / absence heatmap ────────────────────────────────────────────────
sort_cols <- function(mat) {
  col_order <- colnames(mat) %>%
    tibble(label = .) %>%
    mutate(
      stype = str_extract(label, "^(EVI|ASC)"),
      snum  = as.integer(str_extract(label, "(?<=CH)\\d+"))
    ) %>%
    arrange(desc(stype), snum) %>%
    pull(label)
  mat[, col_order, drop = FALSE]
}

make_heatmap_matrix <- function(data, db) {
  mat <- data %>%
    filter(database == db) %>%
    mutate(label = paste0(str_extract(as.character(sample_type), "(ASC|EVI)"),
                          "-", sample_id)) %>%
    group_by(GENE, label) %>%
    summarise(n = n(), .groups = "drop") %>%
    pivot_wider(names_from = label, values_from = n, values_fill = 0) %>%
    column_to_rownames("GENE")

  gene_totals <- rowSums(mat)
  mat <- mat[order(gene_totals, decreasing = TRUE)[1:min(40, nrow(mat))], ]
  sort_cols(mat)
}

make_col_annot <- function(mat) {
  data.frame(
    Sample_Type = ifelse(str_starts(colnames(mat), "ASC"), "Post-ASC", "Post-EVI"),
    row.names = colnames(mat)
  )
}

db_display <- c(card = "CARD", ncbi = "NCBI", resfinder = "ResFinder")

# Normalised (RPM) gene hit bar chart
plot_gene_counts_rpm <- function(data, title_prefix) {
  rpm_data <- data %>%
    group_by(sample_type, sample_id, database, filtered_reads) %>%
    summarise(n_hits = n(), .groups = "drop") %>%
    mutate(rpm      = n_hits / filtered_reads * 1e6,
           database = recode(database, !!!db_labels))

  p <- ggplot(rpm_data, aes(x = sample_id, y = rpm, fill = database)) +
    geom_col(position = "dodge", colour = "white", linewidth = 0.3) +
    facet_wrap(~ sample_type) +
    scale_fill_manual(values = setNames(db_colours, db_labels), name = "Database") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05)),
                       labels = scales::comma) +
    labs(title = paste0(title_prefix,
                        " — AMR Gene Hits per Million Reads (RPM) per Sample"),
         subtitle = "Normalised by host-depleted read count",
         x = "Sample", y = "Hits per Million Reads") +
    pub_theme +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  print(p)
  save_gg(p, paste0(title_prefix, "_gene_counts_rpm.png"), w = 12, h = 6)
}

# Unique genes per million reads (UGPMR) per sample
plot_unique_rpm <- function(data, title_prefix) {
  ugpmr <- data %>%
    group_by(sample_type, sample_id, database, filtered_reads) %>%
    summarise(n_unique = n_distinct(GENE), .groups = "drop") %>%
    mutate(ugpmr    = n_unique / filtered_reads * 1e6,
           database = recode(database, !!!db_labels))

  p <- ggplot(ugpmr, aes(x = sample_id, y = ugpmr, fill = database)) +
    geom_col(position = "dodge", colour = "white", linewidth = 0.3) +
    facet_wrap(~ sample_type) +
    scale_fill_manual(values = setNames(db_colours, db_labels), name = "Database") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05)),
                       labels = scales::comma) +
    labs(title = paste0(title_prefix,
                        " — Unique AMR Genes per Million Reads (UGPMR) per Sample"),
         subtitle = "Normalised by host-depleted read count; samples with <100,000 reads excluded",
         x = "Sample", y = "Unique Genes per Million Reads") +
    pub_theme +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  print(p)
  save_gg(p, paste0(title_prefix, "_unique_rpm.png"), w = 12, h = 6)
}

plot_heatmaps <- function(data, title_prefix) {
  cols <- list(
    card      = c("#F7F7F7", "#FED976", "#E31A1C"),
    ncbi      = c("#F7F7F7", "#A6CEE3", "#1F78B4"),
    resfinder = c("#F7F7F7", "#B2DF8A", "#33A02C")
  )

  for (db in names(cols)) {
    mat <- make_heatmap_matrix(data, db)
    hm <- pheatmap(
      mat,
      color             = colorRampPalette(cols[[db]])(100),
      annotation_col    = make_col_annot(mat),
      annotation_colors = annot_colours,
      cluster_rows      = FALSE,
      cluster_cols      = FALSE,
      fontsize_row      = 10,
      fontsize_col      = 11,
      fontsize          = 11,
      angle_col         = 45,
      cellwidth         = 38,
      cellheight        = 14,
      main              = paste0(title_prefix, " \u2014 ", db_display[db],
                                 " Database \u2014 Top 40 AMR Genes (Hit Count)"),
      border_color      = "grey90",
      na_col            = "grey95",
      silent            = TRUE
    )
    print(hm)
    save_pheatmap(hm,
                  paste0(title_prefix, "_heatmap_", db, ".png"),
                  w = 13, h = 11)
  }

  # Resistance class heatmap (all databases combined)
  res_mat <- data %>%
    filter(!is.na(RESISTANCE), RESISTANCE != "") %>%
    mutate(RESISTANCE = str_split(RESISTANCE, ";")) %>%
    unnest(RESISTANCE) %>%
    mutate(
      RESISTANCE = str_trim(str_to_title(RESISTANCE)),
      label = paste0(str_extract(as.character(sample_type), "(ASC|EVI)"),
                     "-", sample_id)
    ) %>%
    filter(RESISTANCE != "") %>%
    group_by(RESISTANCE, label) %>%
    summarise(n = n(), .groups = "drop") %>%
    pivot_wider(names_from = label, values_from = n, values_fill = 0) %>%
    column_to_rownames("RESISTANCE")

  res_mat <- res_mat[apply(res_mat, 1, max) >= 2, ]
  res_mat <- sort_cols(res_mat)

  hm_res <- pheatmap(
    res_mat,
    color             = colorRampPalette(c("#F7F7F7", "#FEE08B", "#9E0142"))(100),
    annotation_col    = make_col_annot(res_mat),
    annotation_colors = annot_colours,
    cluster_rows      = FALSE,
    cluster_cols      = FALSE,
    fontsize_row      = 10,
    fontsize_col      = 11,
    fontsize          = 11,
    angle_col         = 45,
    cellwidth         = 38,
    cellheight        = 14,
    main              = paste0(title_prefix,
                               " \u2014 Resistance Classes Across All Databases (Hit Count)"),
    border_color      = "grey90",
    silent            = TRUE
  )
  print(hm_res)
  save_pheatmap(hm_res,
                paste0(title_prefix, "_heatmap_resistance_classes.png"),
                w = 13, h = 14)
}

# ── Gene presence / absence heatmap ──────────────────────────────────────────
plot_presence_absence <- function(data, title_prefix, db) {
  mat_pa <- data %>%
    filter(database == db) %>%
    mutate(label = paste0(str_extract(as.character(sample_type), "(ASC|EVI)"),
                          "-", sample_id)) %>%
    distinct(GENE, label) %>%
    mutate(present = 1L) %>%
    pivot_wider(names_from = label, values_from = present, values_fill = 0L) %>%
    column_to_rownames("GENE")

  # Keep genes present in >= 2 samples
  mat_pa <- mat_pa[rowSums(mat_pa) >= 2, , drop = FALSE]
  if (nrow(mat_pa) == 0) {
    message("No genes present in >= 2 samples for ", db); return(invisible(NULL))
  }
  mat_pa <- sort_cols(mat_pa)

  hm <- pheatmap(
    mat_pa,
    color             = c("#F7F7F7", "#3C5488"),
    annotation_col    = make_col_annot(mat_pa),
    annotation_colors = annot_colours,
    cluster_rows      = FALSE,
    cluster_cols      = FALSE,
    fontsize_row      = 9,
    fontsize_col      = 11,
    angle_col         = 45,
    cellwidth         = 38,
    cellheight        = 13,
    main              = paste0(title_prefix, " \u2014 ", db_display[db],
                               " — Gene Presence / Absence"),
    border_color      = "grey90",
    legend            = FALSE,
    silent            = TRUE
  )
  print(hm)
  save_pheatmap(hm,
                paste0(title_prefix, "_presence_absence_", db, ".png"),
                w = 13, h = max(6, nrow(mat_pa) * 0.3 + 3))
}
```

---

# CAT Contigs

## Gene Counts

```{r cat-counts, fig.height=10}
plot_gene_counts(cat_data, "CAT")
```

## Normalised Gene Counts (RPM)

```{r cat-rpm, fig.height=6}
plot_gene_counts_rpm(cat_data_rpm, "CAT")
```

## Unique Genes per Million Reads (UGPMR)

```{r cat-unique-rpm, fig.height=6}
plot_unique_rpm(cat_data_rpm, "CAT")
```

## ASC vs EVI Comparison

```{r cat-comparison, fig.height=5}
plot_comparison(cat_data, "CAT")
```

## Resistance Classes

```{r cat-resistance, fig.height=6, fig.width=13}
plot_resistance_classes(cat_data, "CAT")
```

## Heatmaps (Hit Count)

```{r cat-heatmaps, fig.height=9, fig.width=13}
plot_heatmaps(cat_data, "CAT")
```

## Gene Presence / Absence

```{r cat-pa-card, fig.height=9, fig.width=13}
plot_presence_absence(cat_data, "CAT", "card")
```

```{r cat-pa-ncbi, fig.height=9, fig.width=13}
plot_presence_absence(cat_data, "CAT", "ncbi")
```

```{r cat-pa-resfinder, fig.height=9, fig.width=13}
plot_presence_absence(cat_data, "CAT", "resfinder")
```

---

# RVS Contigs

## Gene Counts

```{r rvs-counts, fig.height=10}
plot_gene_counts(rvs_data, "RVS")
```

## Normalised Gene Counts (RPM)

```{r rvs-rpm, fig.height=6}
plot_gene_counts_rpm(rvs_data_rpm, "RVS")
```

## Unique Genes per Million Reads (UGPMR)

```{r rvs-unique-rpm, fig.height=6}
plot_unique_rpm(rvs_data_rpm, "RVS")
```

## ASC vs EVI Comparison

```{r rvs-comparison, fig.height=5}
plot_comparison(rvs_data, "RVS")
```

## Resistance Classes

```{r rvs-resistance, fig.height=6, fig.width=13}
plot_resistance_classes(rvs_data, "RVS")
```

## Heatmaps (Hit Count)

```{r rvs-heatmaps, fig.height=9, fig.width=13}
plot_heatmaps(rvs_data, "RVS")
```

## Gene Presence / Absence

```{r rvs-pa-card, fig.height=9, fig.width=13}
plot_presence_absence(rvs_data, "RVS", "card")
```

```{r rvs-pa-ncbi, fig.height=9, fig.width=13}
plot_presence_absence(rvs_data, "RVS", "ncbi")
```

```{r rvs-pa-resfinder, fig.height=9, fig.width=13}
plot_presence_absence(rvs_data, "RVS", "resfinder")
```

---

# CAT vs RVS: Cross-Enrichment Comparison

```{r combined-data}
all_combined <- bind_rows(
  cat_data %>% mutate(enrich_method = "BF Bolton base + CAT"),
  rvs_data %>% mutate(enrich_method = "BPW + RVS")
) %>%
  mutate(
    sample_type   = factor(sample_type, levels = evi_first),
    enrich_method = factor(enrich_method,
                           levels = c("BF Bolton base + CAT", "BPW + RVS"))
  )

all_combined_rpm <- all_combined %>% filter(filtered_reads >= min_reads)
```

## Unique Resistance Genes: Post-ASC vs Post-EVI across Enrichment Methods

```{r cross-unique-genes, fig.height=6, fig.width=11}
unique_genes <- all_combined %>%
  group_by(sample_type, enrich_method, database) %>%
  summarise(n_unique = n_distinct(GENE), .groups = "drop") %>%
  mutate(database = recode(database, !!!db_labels))

p_cross <- ggplot(unique_genes,
                  aes(x = enrich_method, y = n_unique, fill = sample_type)) +
  geom_col(position = "dodge", colour = "white", linewidth = 0.4) +
  geom_text(aes(label = n_unique),
            position = position_dodge(width = 0.9),
            vjust = -0.5, size = 4, fontface = "bold") +
  facet_wrap(~ database) +
  scale_fill_manual(values = type_colours, name = "Sample Type") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Unique AMR Genes: Post-ASC vs Post-EVI by Enrichment Method and Database",
       x = "Enrichment Method", y = "Number of Unique Genes") +
  pub_theme +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

print(p_cross)
save_gg(p_cross, "cross_unique_genes.png", w = 11, h = 6)
```

## Resistance Class Profiles: Do Classes Change Between Groups?

```{r cross-resistance-classes, fig.height=8, fig.width=13}
res_combined <- all_combined %>%
  filter(!is.na(RESISTANCE), RESISTANCE != "") %>%
  mutate(RESISTANCE = str_split(RESISTANCE, ";")) %>%
  unnest(RESISTANCE) %>%
  mutate(RESISTANCE = str_trim(str_to_title(RESISTANCE))) %>%
  filter(RESISTANCE != "")

top_classes <- res_combined %>%
  mutate(RESISTANCE = str_replace_all(RESISTANCE, "_", " ")) %>%
  count(RESISTANCE, sort = TRUE) %>%
  slice_head(n = 15) %>%
  pull(RESISTANCE)

res_profile <- res_combined %>%
  mutate(RESISTANCE = str_replace_all(RESISTANCE, "_", " ")) %>%
  filter(RESISTANCE %in% top_classes) %>%
  group_by(sample_type, enrich_method, RESISTANCE) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(sample_type, enrich_method) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup() %>%
  mutate(RESISTANCE = factor(RESISTANCE, levels = rev(top_classes)))

p_stack <- ggplot(res_profile,
                  aes(x = sample_type, y = pct, fill = RESISTANCE)) +
  geom_col(colour = "white", linewidth = 0.25, width = 0.65) +
  facet_wrap(~ enrich_method) +
  scale_fill_manual(
    values = colorRampPalette(brewer.pal(11, "Spectral"))(15),
    name = "Resistance Class"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.01)),
                     labels = function(x) paste0(x, "%")) +
  labs(title = "Resistance Class Composition by Sample Type and Enrichment Method",
       subtitle = "Top 15 resistance classes as % of total hits",
       x = "Sample Type", y = "Proportion of Hits (%)") +
  pub_theme +
  guides(fill = guide_legend(ncol = 1, keyheight = 0.75, keywidth = 0.75,
                              override.aes = list(colour = NA)))

print(p_stack)
save_gg(p_stack, "cross_resistance_profile.png", w = 11, h = 8)
```

```{r cross-resistance-dotplot, fig.height=8, fig.width=13}
res_dot <- res_combined %>%
  mutate(RESISTANCE = str_replace_all(RESISTANCE, "_", " ")) %>%
  filter(RESISTANCE %in% top_classes) %>%
  group_by(sample_type, enrich_method, RESISTANCE) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(RESISTANCE = factor(RESISTANCE, levels = rev(top_classes)))

p_dot <- ggplot(res_dot,
                aes(x = n, y = RESISTANCE, colour = sample_type)) +
  geom_line(aes(group = RESISTANCE), colour = "grey75", linewidth = 0.7) +
  geom_point(size = 3.5, alpha = 0.9) +
  facet_wrap(~ enrich_method, scales = "free_x") +
  scale_colour_manual(values = type_colours, name = "Sample Type") +
  scale_x_continuous(expand = expansion(mult = c(0.05, 0.1)),
                     labels = scales::comma) +
  labs(title = "Resistance Class Hit Counts: Post-ASC vs Post-EVI by Enrichment Method",
       subtitle = "Lines connect Post-ASC and Post-EVI for each resistance class",
       x = "Total Hits", y = NULL) +
  pub_theme +
  theme(panel.grid.major.x = element_line(colour = "grey92", linewidth = 0.4),
        axis.text.y = element_text(size = 11))

print(p_dot)
save_gg(p_dot, "cross_resistance_dotplot.png", w = 13, h = 8)
```

## Normalised Cleveland Dot Plot (RPM)

```{r cross-resistance-dotplot-rpm, fig.height=8, fig.width=13}
res_dot_rpm <- all_combined_rpm %>%
  filter(!is.na(RESISTANCE), RESISTANCE != "") %>%
  mutate(RESISTANCE = str_split(RESISTANCE, ";")) %>%
  unnest(RESISTANCE) %>%
  mutate(RESISTANCE = str_trim(str_to_title(str_replace_all(RESISTANCE, "_", " ")))) %>%
  filter(RESISTANCE %in% top_classes) %>%
  group_by(sample_type, enrich_method, sample_id, filtered_reads, RESISTANCE) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(rpm = n / filtered_reads * 1e6) %>%
  group_by(sample_type, enrich_method, RESISTANCE) %>%
  summarise(mean_rpm = mean(rpm), se_rpm = sd(rpm) / sqrt(n()),
            .groups = "drop") %>%
  mutate(RESISTANCE = factor(RESISTANCE, levels = rev(top_classes)))

p_dot_rpm <- ggplot(res_dot_rpm,
                    aes(x = mean_rpm, y = RESISTANCE, colour = sample_type)) +
  geom_line(aes(group = RESISTANCE), colour = "grey75", linewidth = 0.7) +
  geom_errorbarh(aes(xmin = mean_rpm - se_rpm, xmax = mean_rpm + se_rpm),
                 height = 0.25, linewidth = 0.5, alpha = 0.7) +
  geom_point(size = 3.5, alpha = 0.9) +
  facet_wrap(~ enrich_method, scales = "free_x") +
  scale_colour_manual(values = type_colours, name = "Sample Type") +
  scale_x_continuous(expand = expansion(mult = c(0.05, 0.15)),
                     labels = scales::comma) +
  labs(title = "Resistance Class Prevalence (RPM): Post-ASC vs Post-EVI by Enrichment Method",
       subtitle = "Mean \u00b1 SE across replicates; normalised by host-depleted read count",
       x = "Mean Hits per Million Reads (RPM)", y = NULL) +
  pub_theme +
  theme(panel.grid.major.x = element_line(colour = "grey92", linewidth = 0.4),
        axis.text.y = element_text(size = 11))

print(p_dot_rpm)
save_gg(p_dot_rpm, "cross_resistance_dotplot_rpm.png", w = 13, h = 8)
```

---

# Statistical Analysis

```{r stats-setup}
# Per-sample hit count per resistance class (zeros filled for missing combinations)
hits_per_class <- all_combined %>%
  filter(!is.na(RESISTANCE), RESISTANCE != "") %>%
  mutate(RESISTANCE = str_split(RESISTANCE, ";")) %>%
  unnest(RESISTANCE) %>%
  mutate(RESISTANCE = str_trim(str_to_title(str_replace_all(RESISTANCE, "_", " ")))) %>%
  filter(RESISTANCE %in% top_classes) %>%
  group_by(enrich_method, sample_type, sample_id, RESISTANCE) %>%
  summarise(n = n(), .groups = "drop") %>%
  complete(nesting(enrich_method, sample_type, sample_id),
           RESISTANCE, fill = list(n = 0))

# Per-sample RPM per resistance class (zeros filled; low-depth samples excluded)
rpm_per_class <- all_combined_rpm %>%
  filter(!is.na(RESISTANCE), RESISTANCE != "") %>%
  mutate(RESISTANCE = str_split(RESISTANCE, ";")) %>%
  unnest(RESISTANCE) %>%
  mutate(RESISTANCE = str_trim(str_to_title(str_replace_all(RESISTANCE, "_", " ")))) %>%
  filter(RESISTANCE %in% top_classes) %>%
  group_by(enrich_method, sample_type, sample_id, filtered_reads, RESISTANCE) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(rpm = n / filtered_reads * 1e6) %>%
  complete(nesting(enrich_method, sample_type, sample_id, filtered_reads),
           RESISTANCE, fill = list(n = 0, rpm = 0))

run_wilcox <- function(data, group_var, level_a, level_b, split_var, value_col = "n") {
  data %>%
    group_by(across(all_of(c(split_var, "RESISTANCE")))) %>%
    filter(n_distinct(.data[[group_var]]) == 2) %>%
    summarise(
      p_value  = wilcox.test(.data[[value_col]][.data[[group_var]] == level_a],
                             .data[[value_col]][.data[[group_var]] == level_b],
                             exact = FALSE)$p.value,
      median_a = median(.data[[value_col]][.data[[group_var]] == level_a]),
      median_b = median(.data[[value_col]][.data[[group_var]] == level_b]),
      .groups  = "drop"
    ) %>%
    rename(!!paste0("median_", level_a) := median_a,
           !!paste0("median_", level_b) := median_b) %>%
    mutate(sig = case_when(p_value < 0.001 ~ "***",
                           p_value < 0.01  ~ "**",
                           p_value < 0.05  ~ "*",
                           TRUE            ~ "ns"))
}
```

## Post-EVI vs Post-ASC (within each enrichment method)

```{r stats-evi-vs-asc}
wilcox_evi_asc <- run_wilcox(hits_per_class,
                              group_var = "sample_type",
                              level_a   = "Post-EVI",
                              level_b   = "Post-ASC",
                              split_var = "enrich_method")

wilcox_evi_asc %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  arrange(enrich_method, p_value) %>%
  knitr::kable(caption = "Wilcoxon rank-sum test: Post-EVI vs Post-ASC per resistance class")
```

## CAT vs RVS (within each sample type)

```{r stats-cat-vs-rvs}
wilcox_cat_rvs <- run_wilcox(hits_per_class,
                              group_var = "enrich_method",
                              level_a   = "BF Bolton base + CAT",
                              level_b   = "BPW + RVS",
                              split_var = "sample_type")

wilcox_cat_rvs %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  arrange(sample_type, p_value) %>%
  knitr::kable(caption = "Wilcoxon rank-sum test: BF Bolton base + CAT vs BPW + RVS per resistance class")
```

## Overall AMR Burden: Total Hits per Sample

```{r stats-total-hits, fig.height=5, fig.width=10}
total_hits <- all_combined %>%
  group_by(enrich_method, sample_type, sample_id) %>%
  summarise(n = n(), .groups = "drop")

total_wilcox <- total_hits %>%
  group_by(enrich_method) %>%
  summarise(
    p_value        = wilcox.test(n[sample_type == "Post-EVI"],
                                 n[sample_type == "Post-ASC"],
                                 exact = FALSE)$p.value,
    median_PostEVI = median(n[sample_type == "Post-EVI"]),
    median_PostASC = median(n[sample_type == "Post-ASC"]),
    .groups        = "drop"
  ) %>%
  mutate(sig = case_when(p_value < 0.001 ~ "***",
                         p_value < 0.01  ~ "**",
                         p_value < 0.05  ~ "*",
                         TRUE            ~ "ns"))

knitr::kable(total_wilcox %>% mutate(across(where(is.numeric), ~ round(.x, 3))),
             caption = "Wilcoxon rank-sum test: overall AMR burden (total hits) Post-EVI vs Post-ASC")

p_total <- ggplot(total_hits,
                  aes(x = sample_type, y = n, fill = sample_type)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA, width = 0.5) +
  geom_jitter(width = 0.1, size = 2.5, alpha = 0.8, aes(colour = sample_type)) +
  geom_text(data = total_wilcox,
            aes(x = 1.5, y = Inf, label = paste0("p = ", round(p_value, 3),
                                                   "  ", sig)),
            vjust = 1.5, size = 4, fontface = "italic", inherit.aes = FALSE) +
  facet_wrap(~ enrich_method) +
  scale_fill_manual(values   = type_colours, guide = "none") +
  scale_colour_manual(values = type_colours, guide = "none") +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Total AMR Gene Hits per Sample (Contigs)",
       subtitle = "Wilcoxon rank-sum test",
       x = NULL, y = "Total Hits") +
  pub_theme

print(p_total)
save_gg(p_total, "stats_total_hits_boxplot.png", w = 10, h = 5)
```

## Unique Gene Diversity: Post-EVI vs Post-ASC

```{r stats-unique-genes, fig.height=9, fig.width=11}
unique_per_sample <- all_combined %>%
  group_by(enrich_method, sample_type, sample_id, database) %>%
  summarise(n_unique = n_distinct(GENE), .groups = "drop") %>%
  mutate(database_label = recode(database, !!!db_labels))

unique_wilcox <- unique_per_sample %>%
  group_by(enrich_method, database_label) %>%
  summarise(
    p_value        = wilcox.test(n_unique[sample_type == "Post-EVI"],
                                 n_unique[sample_type == "Post-ASC"],
                                 exact = FALSE)$p.value,
    median_PostEVI = median(n_unique[sample_type == "Post-EVI"]),
    median_PostASC = median(n_unique[sample_type == "Post-ASC"]),
    .groups        = "drop"
  ) %>%
  mutate(sig = case_when(p_value < 0.001 ~ "***",
                         p_value < 0.01  ~ "**",
                         p_value < 0.05  ~ "*",
                         TRUE            ~ "ns"))

knitr::kable(unique_wilcox %>% mutate(across(where(is.numeric), ~ round(.x, 3))),
             caption = "Wilcoxon rank-sum test: unique AMR genes per sample, Post-EVI vs Post-ASC")

p_unique_box <- ggplot(unique_per_sample,
                       aes(x = sample_type, y = n_unique, fill = sample_type)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA, width = 0.5) +
  geom_jitter(width = 0.1, size = 2.5, alpha = 0.8,
              aes(colour = sample_type)) +
  geom_text(data = unique_wilcox,
            aes(x = 1.5, y = Inf,
                label = paste0("p = ", round(p_value, 3), "  ", sig)),
            vjust = 1.5, size = 3.8, fontface = "italic", inherit.aes = FALSE) +
  facet_grid(database_label ~ enrich_method) +
  scale_fill_manual(values = type_colours, guide = "none") +
  scale_colour_manual(values = type_colours, guide = "none") +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Unique AMR Genes per Sample: Post-EVI vs Post-ASC (Contigs)",
       subtitle = "Wilcoxon rank-sum test; each point = one replicate",
       x = NULL, y = "Number of Unique Genes") +
  pub_theme

print(p_unique_box)
save_gg(p_unique_box, "stats_unique_genes_boxplot.png", w = 11, h = 9)
```

---

# RPM-Normalised Statistical Analysis

## Post-EVI vs Post-ASC (within each enrichment method) — RPM

```{r stats-rpm-evi-vs-asc}
wilcox_rpm_evi_asc <- run_wilcox(rpm_per_class,
                                  group_var  = "sample_type",
                                  level_a    = "Post-EVI",
                                  level_b    = "Post-ASC",
                                  split_var  = "enrich_method",
                                  value_col  = "rpm")

wilcox_rpm_evi_asc %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  arrange(enrich_method, p_value) %>%
  knitr::kable(caption = "Wilcoxon rank-sum test (RPM): Post-EVI vs Post-ASC per resistance class")
```

```{r stats-rpm-evi-vs-asc-plot, fig.height=8, fig.width=14}
sig_annot_rpm <- wilcox_rpm_evi_asc %>%
  filter(sig != "ns") %>%
  mutate(RESISTANCE = factor(RESISTANCE, levels = rev(top_classes)))

star_pos_rpm <- res_dot_rpm %>%
  group_by(enrich_method, RESISTANCE) %>%
  summarise(x_pos = max(mean_rpm + se_rpm, na.rm = TRUE), .groups = "drop") %>%
  inner_join(sig_annot_rpm, by = c("enrich_method", "RESISTANCE"))

p_dot_rpm_sig <- p_dot_rpm +
  geom_text(data = star_pos_rpm,
            aes(x = x_pos, y = RESISTANCE, label = sig),
            colour = "black", size = 4.5, hjust = -0.2,
            fontface = "bold", inherit.aes = FALSE) +
  labs(title = "Resistance Class Prevalence (RPM) with Significance Annotations",
       subtitle = "Stars indicate Wilcoxon p-value: * <0.05  ** <0.01  *** <0.001\nSamples with <100,000 reads excluded") +
  scale_x_continuous(expand = expansion(mult = c(0.05, 0.25)),
                     labels = scales::comma)

print(p_dot_rpm_sig)
save_gg(p_dot_rpm_sig, "cross_resistance_dotplot_rpm_sig.png", w = 14, h = 8)
```

## CAT vs RVS (within each sample type) — RPM

```{r stats-rpm-cat-vs-rvs}
wilcox_rpm_cat_rvs <- run_wilcox(rpm_per_class,
                                  group_var = "enrich_method",
                                  level_a   = "BF Bolton base + CAT",
                                  level_b   = "BPW + RVS",
                                  split_var = "sample_type",
                                  value_col = "rpm")

wilcox_rpm_cat_rvs %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  arrange(sample_type, p_value) %>%
  knitr::kable(caption = "Wilcoxon rank-sum test (RPM): BF Bolton base + CAT vs BPW + RVS per resistance class")
```

## Overall AMR Burden: Total RPM per Sample

```{r stats-total-rpm, fig.height=5, fig.width=10}
total_rpm <- all_combined_rpm %>%
  group_by(enrich_method, sample_type, sample_id, filtered_reads) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(rpm = n / filtered_reads * 1e6)

total_rpm_wilcox <- total_rpm %>%
  group_by(enrich_method) %>%
  summarise(
    p_value        = wilcox.test(rpm[sample_type == "Post-EVI"],
                                 rpm[sample_type == "Post-ASC"],
                                 exact = FALSE)$p.value,
    median_PostEVI = median(rpm[sample_type == "Post-EVI"]),
    median_PostASC = median(rpm[sample_type == "Post-ASC"]),
    .groups        = "drop"
  ) %>%
  mutate(sig = case_when(p_value < 0.001 ~ "***",
                         p_value < 0.01  ~ "**",
                         p_value < 0.05  ~ "*",
                         TRUE            ~ "ns"))

knitr::kable(total_rpm_wilcox %>% mutate(across(where(is.numeric), ~ round(.x, 3))),
             caption = "Wilcoxon rank-sum test: overall AMR burden (total RPM) Post-EVI vs Post-ASC")

p_total_rpm <- ggplot(total_rpm,
                      aes(x = sample_type, y = rpm, fill = sample_type)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA, width = 0.5) +
  geom_jitter(width = 0.1, size = 2.5, alpha = 0.8, aes(colour = sample_type)) +
  geom_text(data = total_rpm_wilcox,
            aes(x = 1.5, y = Inf, label = paste0("p = ", round(p_value, 3),
                                                   "  ", sig)),
            vjust = 1.5, size = 4, fontface = "italic", inherit.aes = FALSE) +
  facet_wrap(~ enrich_method) +
  scale_fill_manual(values   = type_colours, guide = "none") +
  scale_colour_manual(values = type_colours, guide = "none") +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Total AMR Gene Hits per Million Reads (RPM) per Sample (Contigs)",
       subtitle = "Wilcoxon rank-sum test",
       x = NULL, y = "Total Hits per Million Reads") +
  pub_theme

print(p_total_rpm)
save_gg(p_total_rpm, "stats_total_rpm_boxplot.png", w = 10, h = 5)
```

## Unique Gene Diversity (UGPMR): Post-EVI vs Post-ASC

```{r stats-ugpmr, fig.height=9, fig.width=11}
ugpmr_per_sample <- all_combined_rpm %>%
  group_by(enrich_method, sample_type, sample_id, filtered_reads, database) %>%
  summarise(n_unique = n_distinct(GENE), .groups = "drop") %>%
  mutate(ugpmr         = n_unique / filtered_reads * 1e6,
         database_label = recode(database, !!!db_labels))

ugpmr_wilcox <- ugpmr_per_sample %>%
  group_by(enrich_method, database_label) %>%
  summarise(
    p_value        = wilcox.test(ugpmr[sample_type == "Post-EVI"],
                                 ugpmr[sample_type == "Post-ASC"],
                                 exact = FALSE)$p.value,
    median_PostEVI = median(ugpmr[sample_type == "Post-EVI"]),
    median_PostASC = median(ugpmr[sample_type == "Post-ASC"]),
    .groups        = "drop"
  ) %>%
  mutate(sig = case_when(p_value < 0.001 ~ "***",
                         p_value < 0.01  ~ "**",
                         p_value < 0.05  ~ "*",
                         TRUE            ~ "ns"))

knitr::kable(ugpmr_wilcox %>% mutate(across(where(is.numeric), ~ round(.x, 3))),
             caption = "Wilcoxon rank-sum test: unique AMR genes per million reads (UGPMR), Post-EVI vs Post-ASC")

p_ugpmr_box <- ggplot(ugpmr_per_sample,
                      aes(x = sample_type, y = ugpmr, fill = sample_type)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA, width = 0.5) +
  geom_jitter(width = 0.1, size = 2.5, alpha = 0.8,
              aes(colour = sample_type)) +
  geom_text(data = ugpmr_wilcox,
            aes(x = 1.5, y = Inf,
                label = paste0("p = ", round(p_value, 3), "  ", sig)),
            vjust = 1.5, size = 3.8, fontface = "italic", inherit.aes = FALSE) +
  facet_grid(database_label ~ enrich_method) +
  scale_fill_manual(values = type_colours, guide = "none") +
  scale_colour_manual(values = type_colours, guide = "none") +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Unique AMR Genes per Million Reads (UGPMR): Post-EVI vs Post-ASC (Contigs)",
       subtitle = "Wilcoxon rank-sum test; each point = one replicate",
       x = NULL, y = "Unique Genes per Million Reads") +
  pub_theme

print(p_ugpmr_box)
save_gg(p_ugpmr_box, "stats_ugpmr_boxplot.png", w = 11, h = 9)
```

---

# Session Info

```{r session-info}
sessionInfo()
```
