---
title: "Bacterial Diversity (Contigs) — EVI vs. ASC"
author: "Alice Bradbury"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    fig_width: 12
    fig_height: 10
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE
)

required_pkgs <- c("tidyverse", "RColorBrewer", "patchwork", "scales",
                   "fs", "pheatmap", "gtable", "DT", "gridExtra", "vegan")
to_install    <- required_pkgs[!sapply(required_pkgs, requireNamespace, quietly = TRUE)]
if (length(to_install)) install.packages(to_install)

library(tidyverse)
library(RColorBrewer)
library(patchwork)
library(scales)
library(fs)
library(pheatmap)
library(gtable)
library(DT)
library(gridExtra)
library(vegan)

fig_dir <- "/Volumes/Alice/2_POULTRY_PROCESSING/Desktop/figures_bacterial_diversity_contigs"
dir_create(fig_dir)

sample_labels <- c(
  "CH01_EVI_CAT" = "Chicken 1 BF Bolton base + CAT",
  "CH02_EVI_CAT" = "Chicken 2 BF Bolton base + CAT",
  "CH03_EVI_CAT" = "Chicken 3 BF Bolton base + CAT",
  "CH04_EVI_CAT" = "Chicken 4 BF Bolton base + CAT",
  "CH05_EVI_CAT" = "Chicken 5 BF Bolton base + CAT",
  "CH06_EVI_CAT" = "Chicken 6 BF Bolton base + CAT",
  "CH01_ASC_CAT" = "Chicken 1 BF Bolton base + CAT",
  "CH02_ASC_CAT" = "Chicken 2 BF Bolton base + CAT",
  "CH03_ASC_CAT" = "Chicken 3 BF Bolton base + CAT",
  "CH04_ASC_CAT" = "Chicken 4 BF Bolton base + CAT",
  "CH05_ASC_CAT" = "Chicken 5 BF Bolton base + CAT",
  "CH06_ASC_CAT" = "Chicken 6 BF Bolton base + CAT",
  "CH01_EVI_RVS" = "Chicken 1 BPW & RVS",
  "CH02_EVI_RVS" = "Chicken 2 BPW & RVS",
  "CH03_EVI_RVS" = "Chicken 3 BPW & RVS",
  "CH04_EVI_RVS" = "Chicken 4 BPW & RVS",
  "CH05_EVI_RVS" = "Chicken 5 BPW & RVS",
  "CH06_EVI_RVS" = "Chicken 6 BPW & RVS",
  "CH01_ASC_RVS" = "Chicken 1 BPW & RVS",
  "CH02_ASC_RVS" = "Chicken 2 BPW & RVS",
  "CH03_ASC_RVS" = "Chicken 3 BPW & RVS",
  "CH04_ASC_RVS" = "Chicken 4 BPW & RVS",
  "CH05_ASC_RVS" = "Chicken 5 BPW & RVS",
  "CH06_ASC_RVS" = "Chicken 6 BPW & RVS"
)
```

```{r shared_palette}
custom_pal <- brewer.pal(12, "Paired")

# Fixed colors matching the read-level figure so taxa are comparable across plots
family_colors_fixed <- c(
  "Selenomonadaceae"   = "#1F78B4",
  "Bacteroidaceae"     = "#33A02C",
  "Campylobacteraceae" = "#E31A1C",
  "Lactobacillaceae"   = "#FF7F00",
  "Comamonadaceae"     = "#6A3D9A",
  "Enterobacteriaceae" = "#B15928",
  "Tannerellaceae"     = "#A6CEE3",
  "Veillonellaceae"    = "#B2DF8A",
  "Pasteurellaceae"    = "#FB9A99",
  "Pseudomonadaceae"   = "#FDBF6F",
  "Morganellaceae"     = "#CAB2D6",
  "Gallionellaceae"    = "#FFFF99"
)

genus_colors_fixed <- c(
  "Campylobacter"  = "#E31A1C",
  "Bacteroides"    = "#33A02C",
  "Lactobacillus"  = "#FF7F00",
  "Pseudomonas"    = "#FDBF6F",
  "Veillonella"    = "#B2DF8A",
  "Tannerella"     = "#A6CEE3",
  "Escherichia"    = "#B15928",
  "Salmonella"     = "#B15928",
  "Comamonas"      = "#6A3D9A"
)

# Fixed species colours — same species always gets same colour across all plots
species_colors_escherichia <- c(
  "Escherichia albertii"      = "#66C2A5",
  "Escherichia fergusonii"    = "#FC8D62",
  "Escherichia marmotae"      = "#8DA0CB",
  "Escherichia ruysiae"       = "#E78AC3",
  "Escherichia coli"          = "#A6D854"
)

species_colors_salmonella <- c(
  "Salmonella enterica"       = "#E41A1C",
  "Salmonella bongori"        = "#377EB8"
)

species_colors_campylobacter <- c(
  "Campylobacter lari"            = "#1B9E77",
  "Campylobacter upsaliensis"     = "#D95F02",
  "Campylobacter hyointestinalis" = "#7570B3",
  "Campylobacter helveticus"      = "#E7298A",
  "Campylobacter concisus"        = "#66A61E",
  "Campylobacter fetus"           = "#E6AB02",
  "Campylobacter showae"          = "#A6761D",
  "Campylobacter ureolyticus"     = "#666666"
)

species_colors_klebsiella <- c(
  "Klebsiella pneumoniae"         = "#66C2A5",
  "Klebsiella oxytoca"            = "#FC8D62",
  "Klebsiella variicola"          = "#8DA0CB",
  "Klebsiella quasipneumoniae"    = "#E78AC3",
  "Klebsiella aerogenes"          = "#A6D854"
)

build_fill_vals <- function(taxa, fixed_colors = NULL, pal = custom_pal) {
  if (is.null(fixed_colors) || length(fixed_colors) == 0) {
    return(c(setNames(pal[seq_len(length(taxa))], taxa), Other = "grey80"))
  }
  known   <- intersect(taxa, names(fixed_colors))
  unknown <- setdiff(taxa, names(fixed_colors))
  used    <- fixed_colors[known]
  avail   <- pal[!pal %in% used]
  extra   <- setNames(avail[seq_len(length(unknown))], unknown)
  c(used, extra, Other = "grey80")
}
```

```{r shared_functions}
# ── Data loading & normalisation ───────────────────────────────────────────────

# Genus → Family lookup for common poultry microbiome bacteria.
# Used to roll genus-/species-level contig classifications up to family level.
genus_to_family <- c(
  # Bacteroidota
  Bacteroides       = "Bacteroidaceae",
  Parabacteroides   = "Tannerellaceae",
  Tannerella        = "Tannerellaceae",
  Alistipes         = "Rikenellaceae",
  Barnesiella       = "Porphyromonadaceae",
  Butyricimonas     = "Odoribacteraceae",
  Paraprevotella    = "Prevotellaceae",
  # Campylobacterota
  Campylobacter     = "Campylobacteraceae",
  # Pseudomonadota (Proteobacteria)
  Comamonas         = "Comamonadaceae",
  Sutterella        = "Sutterellaceae",
  Kerstersia        = "Alcaligenaceae",
  Bordetella        = "Alcaligenaceae",
  Acinetobacter     = "Moraxellaceae",
  Pseudomonas       = "Pseudomonadaceae",
  Escherichia       = "Enterobacteriaceae",
  Shigella          = "Enterobacteriaceae",
  Klebsiella        = "Enterobacteriaceae",
  Salmonella        = "Enterobacteriaceae",
  Proteus           = "Morganellaceae",
  Avibacterium      = "Pasteurellaceae",
  Gallibacterium    = "Pasteurellaceae",
  Fusobacterium     = "Fusobacteriaceae",
  # Bacillota (Firmicutes)
  Lactobacillus     = "Lactobacillaceae",
  Megamonas         = "Selenomonadaceae",
  Megasphaera       = "Selenomonadaceae",
  Cloacibacillus    = "Synergistaceae",
  Christensenella   = "Christensenellaceae"
)

# Noise terms to drop (too high in taxonomy to be informative)
noise_taxa <- c(
  "root", "cellular organisms", "Bacteria", "Archaea",
  "Firmicutes", "Bacteroidetes", "Proteobacteria", "Campylobacterota",
  "Bacteroidota", "Pseudomonadota", "Bacillota",
  "Gammaproteobacteria", "Betaproteobacteria", "Alphaproteobacteria",
  "Negativicutes", "Bacilli", "Clostridia",
  "Terrabacteria group", "Bacteroidetes/Chlorobi group",
  "environmental samples", "uncultured bacterium", "uncultured organism",
  "uncultured prokaryote", "unidentified plasmid"
)

parse_kraken2 <- function(path, rank_code = "G",
                           strip_suffix = "_m1000_kraken2\\.out$") {
  # Kraken2 output format: C/U | seq_id | "Name (taxid N)" | length | kmer_str
  raw <- read.delim(path, header = FALSE, sep = "\t",
                    col.names = c("status", "seq_id", "taxon", "length", "kmer"),
                    stringsAsFactors = FALSE)

  human_re <- regex("Homo|sapiens|Hominidae|Homininae|phage|virus|Virus|viridae|
                     plasmid|prophage|unclassified|uncultured|environmental|
                     unidentified|unidentified plasmid",
                    ignore_case = TRUE)

  parsed <- raw %>%
    as_tibble() %>%
    filter(status == "C") %>%
    mutate(
      name  = str_remove(taxon, "\\s*\\(taxid \\d+\\)$") %>% str_trim(),
      taxid = as.integer(str_extract(taxon, "(?<=taxid )\\d+"))
    ) %>%
    filter(taxid != 0L,
           !name %in% noise_taxa,
           !str_detect(name, human_re)) %>%
    mutate(
      # Infer rank from name structure
      inferred_rank = case_when(
        str_detect(name, "\\s")      ~ "S",          # multi-word = species
        str_detect(name, "aceae$")   ~ "F",           # family
        str_detect(name, "ales$")    ~ "O",           # order
        str_detect(name, "ia$|etes$|bacteria$|ota$|group$|complex$") ~ "X",
        TRUE                         ~ "G"            # genus
      )
    )

  sample_id <- path_file(path) %>% str_remove(strip_suffix)

  if (rank_code == "G") {
    # Genus: use G-rank directly; extract genus (first word) from species
    out <- parsed %>%
      filter(inferred_rank %in% c("G", "S")) %>%
      mutate(name = if_else(inferred_rank == "S",
                            str_extract(name, "^[A-Z][a-z]+"),
                            name)) %>%
      filter(!is.na(name))

  } else if (rank_code == "F") {
    # Family: use F-rank directly; map genus/species to family via lookup
    fam_direct <- parsed %>% filter(inferred_rank == "F")
    fam_mapped <- parsed %>%
      filter(inferred_rank %in% c("G", "S")) %>%
      mutate(genus = if_else(inferred_rank == "S",
                             str_extract(name, "^[A-Z][a-z]+"),
                             name)) %>%
      filter(!is.na(genus), genus %in% names(genus_to_family)) %>%
      mutate(name = genus_to_family[genus])
    out <- bind_rows(fam_direct, fam_mapped)

  } else if (rank_code == "S") {
    out <- parsed %>% filter(inferred_rank == "S")

  } else {
    out <- parsed %>% filter(inferred_rank == rank_code)
  }

  out %>%
    mutate(sample = sample_id) %>%
    group_by(sample, name) %>%
    summarise(reads = sum(as.numeric(length)), .groups = "drop")  # sum contig base pairs
}

load_group <- function(dir_path, rank_code = "G",
                       strip_suffix = "_m1000_kraken2\\.out$") {
  files <- dir_ls(dir_path, regexp = "_m1000_kraken2\\.out$")
  if (length(files) == 0) stop("No Kraken2 .out files found in: ", dir_path)
  map_dfr(files, parse_kraken2, rank_code = rank_code, strip_suffix = strip_suffix)
}

# Relative abundance (%) AND base pairs per million (BPM) per sample
normalise_all <- function(df) {
  df %>%
    group_by(sample) %>%
    mutate(total_reads = sum(reads, na.rm = TRUE),
           rel_abund   = reads / total_reads * 100,
           rpm         = reads / total_reads * 1e6) %>%  # rpm col kept for compatibility; unit is now BPM
    ungroup() %>%
    select(-total_reads)
}

rename_vec <- function(x, labels) {
  if (is.null(labels) || length(labels) == 0) return(x)
  idx <- match(x, names(labels))
  ifelse(!is.na(idx), labels[idx], x)
}

top_n_collapse <- function(df, top_taxa) {
  df %>%
    mutate(name = if_else(name %in% top_taxa, name, "Other")) %>%
    group_by(sample, name) %>%
    summarise(rel_abund = sum(rel_abund), .groups = "drop")
}

# ── Stacked bar ────────────────────────────────────────────────────────────────
plot_grouped <- function(df, sp_order, fill_vals, title_str, caption_str = NULL,
                         sample_labels = NULL) {
  df %>%
    mutate(name   = factor(name, levels = sp_order),
           sample = rename_vec(sample, sample_labels)) %>%
    ggplot(aes(x = sample, y = rel_abund, fill = name)) +
    geom_col(width = 0.75, colour = "white", linewidth = 0.2) +
    scale_fill_manual(values = fill_vals, name = "Taxon", drop = FALSE) +
    scale_x_discrete(labels = label_wrap(12)) +
    scale_y_continuous(labels = label_percent(scale = 1),
                       expand = expansion(mult = c(0, 0.02)), limits = c(0, 101)) +
    facet_wrap(~ group, scales = "free_x", ncol = 1) +
    guides(fill = guide_legend(nrow = 2)) +
    labs(title = str_wrap(title_str, 80), subtitle = caption_str, x = NULL, y = "Relative Abundance (%)") +
    theme_light(base_size = 13) +
    theme(axis.text.x      = element_text(angle = 0, hjust = 0.5, size = 12),
          axis.text.y      = element_text(size = 12),
          legend.text      = element_text(size = 10),
          legend.key.size  = unit(0.45, "cm"),
          plot.title       = element_text(face = "bold", size = 14),
          plot.subtitle    = element_text(size = 10, colour = "grey40"),
          strip.text       = element_text(face = "bold", size = 13, colour = "black"),
          strip.background = element_rect(fill = "lightgrey"),
          legend.position  = "bottom")
}

# ── Heatmap colour scales ──────────────────────────────────────────────────────
hm_colors  <- colorRampPalette(c("#FFFFFF", "#FEE0D2", "#FCBBA1", "#FB6A4A", "#CB181D", "#67000D"))(100)
hm_breaks  <- seq(0, 100, length.out = 101)
rpm_colors <- colorRampPalette(c("#FFFFFF", "#EFF3FF", "#BDD7E7", "#6BAED6", "#2171B5", "#084594"))(100)

add_legend_title <- function(gt, title) {
  leg_idx <- which(gt$layout$name == "legend")
  if (length(leg_idx) == 0) return(gt)
  leg_row <- gt$layout$t[leg_idx[1]]
  leg_l   <- gt$layout$l[leg_idx[1]]
  leg_r   <- gt$layout$r[leg_idx[1]]
  gt <- gtable_add_rows(gt, grid::unit(0.6, "cm"), pos = leg_row - 1)
  gt <- gtable_add_grob(gt,
    grid::textGrob(title, x = 0.5, hjust = 0.5,
                   gp = grid::gpar(fontsize = 8, fontface = "bold")),
    t = leg_row, b = leg_row, l = leg_l, r = leg_r,
    name = "legend_title", clip = "off")
  gt
}

make_heatmap <- function(all_data, group_a_pattern, group_b_pattern,
                          group_a_label, group_b_label, title_str,
                          min_mean = 0.05, min_samples = 2,
                          keep_taxa = NULL,
                          legend_title = "Relative Abundance (%)",
                          cellheight = 11, cellwidth = 55,
                          save_path = NULL, save_width = 14, save_height = 14,
                          sample_labels = NULL) {
  mat <- all_data %>%
    select(sample, name, rel_abund) %>%
    pivot_wider(names_from = sample, values_from = rel_abund, values_fill = 0) %>%
    column_to_rownames("name") %>% as.matrix()
  if (!is.null(keep_taxa)) {
    mat <- mat[rownames(mat) %in% keep_taxa, , drop = FALSE]
    mat <- mat[match(keep_taxa[keep_taxa %in% rownames(mat)], rownames(mat)), , drop = FALSE]
  } else {
    mat <- mat[rowMeans(mat) >= min_mean & rowSums(mat > 0) >= min_samples, , drop = FALSE]
    mat <- mat[order(rowMeans(mat), decreasing = TRUE), , drop = FALSE]
  }
  cols_a <- sort(colnames(mat)[str_detect(colnames(mat), group_a_pattern)])
  cols_b <- sort(colnames(mat)[str_detect(colnames(mat), group_b_pattern)])
  mat_a  <- mat[, cols_a, drop = FALSE]
  mat_b  <- mat[, cols_b, drop = FALSE]
  colnames(mat_a) <- rename_vec(cols_a, sample_labels)
  colnames(mat_b) <- rename_vec(cols_b, sample_labels)
  ph_a <- pheatmap(mat_a, color = hm_colors, breaks = hm_breaks,
                   cluster_rows = FALSE, cluster_cols = FALSE,
                   show_rownames = TRUE, show_colnames = TRUE, legend = FALSE,
                   fontsize_row = 8, fontsize_col = 10, angle_col = 90,
                   main = group_a_label,
                   border_color = NA, cellheight = cellheight, cellwidth = cellwidth,
                   silent = TRUE)
  ph_b <- pheatmap(mat_b, color = hm_colors, breaks = hm_breaks,
                   cluster_rows = FALSE, cluster_cols = FALSE,
                   show_rownames = TRUE, show_colnames = TRUE, legend = TRUE,
                   fontsize_row = 8, fontsize_col = 10, angle_col = 90,
                   main = group_b_label,
                   legend_breaks = c(0, 25, 50, 75, 100),
                   legend_labels = c("0%", "25%", "50%", "75%", "100%"),
                   border_color = NA, cellheight = cellheight, cellwidth = cellwidth,
                   silent = TRUE)
  gt_a <- ph_a$gtable
  gt_b <- add_legend_title(ph_b$gtable, legend_title)
  w_a  <- sum(gt_a$widths)
  w_b  <- sum(gt_b$widths)
  combined <- gridExtra::arrangeGrob(gt_a, gt_b, ncol = 2,
               widths = grid::unit.c(w_a, w_b),
               top = grid::textGrob(str_wrap(title_str, 80),
                                    gp = grid::gpar(fontsize = 12, fontface = "bold")))
  grid::grid.newpage(); grid::grid.draw(combined)
  if (!is.null(save_path)) {
    png(save_path, width = save_width * 600, height = save_height * 600, res = 600)
    grid::grid.newpage(); grid::grid.draw(combined)
    dev.off()
  }
  invisible(combined)
}

make_heatmap_rpm <- function(all_data, group_a_pattern, group_b_pattern,
                              group_a_label, group_b_label, title_str,
                              min_mean = 0.05, min_samples = 2,
                              keep_taxa = NULL,
                              legend_title = "BPM (log\u2081\u2080)",
                              cellheight = 11, cellwidth = 55,
                              save_path = NULL, save_width = 14, save_height = 14,
                              sample_labels = NULL) {
  pct_mat <- all_data %>%
    select(sample, name, rel_abund) %>%
    pivot_wider(names_from = sample, values_from = rel_abund, values_fill = 0) %>%
    column_to_rownames("name") %>% as.matrix()
  if (!is.null(keep_taxa)) {
    taxa <- keep_taxa[keep_taxa %in% rownames(pct_mat)]
  } else {
    taxa <- rownames(pct_mat)[rowMeans(pct_mat) >= min_mean & rowSums(pct_mat > 0) >= min_samples]
    taxa <- taxa[order(rowMeans(pct_mat[taxa, , drop = FALSE]), decreasing = TRUE)]
  }
  mat_rpm <- all_data %>%
    select(sample, name, rpm) %>%
    pivot_wider(names_from = sample, values_from = rpm, values_fill = 0) %>%
    column_to_rownames("name") %>% as.matrix()
  mat_log <- log10(mat_rpm[taxa, , drop = FALSE] + 1)
  cols_a  <- sort(colnames(mat_log)[str_detect(colnames(mat_log), group_a_pattern)])
  cols_b  <- sort(colnames(mat_log)[str_detect(colnames(mat_log), group_b_pattern)])
  mat_a   <- mat_log[, cols_a, drop = FALSE]
  mat_b   <- mat_log[, cols_b, drop = FALSE]
  colnames(mat_a) <- rename_vec(cols_a, sample_labels)
  colnames(mat_b) <- rename_vec(cols_b, sample_labels)
  max_log    <- max(mat_log, na.rm = TRUE)
  if (!is.finite(max_log) || max_log <= 0) max_log <- 1
  brk        <- seq(0, ceiling(max_log * 10) / 10, length.out = 101)
  leg_log    <- seq(0, max(brk), length.out = 5)
  leg_rpm    <- round(10^leg_log - 1); leg_rpm[1] <- 0
  leg_labels <- paste0(formatC(leg_rpm, format = "d", big.mark = ","), " BPM")
  ph_a <- pheatmap(mat_a, color = rpm_colors, breaks = brk,
                   cluster_rows = FALSE, cluster_cols = FALSE,
                   show_rownames = TRUE, show_colnames = TRUE, legend = FALSE,
                   fontsize_row = 8, fontsize_col = 10, angle_col = 90,
                   main = group_a_label,
                   border_color = NA, cellheight = cellheight, cellwidth = cellwidth,
                   silent = TRUE)
  ph_b <- pheatmap(mat_b, color = rpm_colors, breaks = brk,
                   cluster_rows = FALSE, cluster_cols = FALSE,
                   show_rownames = TRUE, show_colnames = TRUE, legend = TRUE,
                   fontsize_row = 8, fontsize_col = 10, angle_col = 90,
                   main = group_b_label,
                   legend_breaks = leg_log, legend_labels = leg_labels,
                   border_color = NA, cellheight = cellheight, cellwidth = cellwidth,
                   silent = TRUE)
  gt_a <- ph_a$gtable
  gt_b <- add_legend_title(ph_b$gtable, legend_title)
  w_a  <- sum(gt_a$widths)
  w_b  <- sum(gt_b$widths)
  combined <- gridExtra::arrangeGrob(gt_a, gt_b, ncol = 2,
               widths = grid::unit.c(w_a, w_b),
               top = grid::textGrob(str_wrap(title_str, 80),
                                    gp = grid::gpar(fontsize = 12, fontface = "bold")))
  grid::grid.newpage(); grid::grid.draw(combined)
  if (!is.null(save_path)) {
    png(save_path, width = save_width * 600, height = save_height * 600, res = 600)
    grid::grid.newpage(); grid::grid.draw(combined)
    dev.off()
  }
  invisible(combined)
}

# ── Statistical testing ────────────────────────────────────────────────────────
run_taxon_tests <- function(data_a, data_b, label_a, label_b,
                             min_mean = 0.1, min_samples = 2) {
  combined <- bind_rows(
    data_a %>% select(sample, name, rel_abund, rpm) %>% mutate(group = label_a),
    data_b %>% select(sample, name, rel_abund, rpm) %>% mutate(group = label_b)
  ) %>% complete(nesting(sample, group), name, fill = list(rel_abund = 0, rpm = 0))

  taxa_keep <- combined %>%
    group_by(name) %>%
    summarise(mean_abund = mean(rel_abund), n_nonzero = sum(rel_abund > 0), .groups = "drop") %>%
    filter(mean_abund >= min_mean, n_nonzero >= min_samples) %>% pull(name)

  results <- combined %>%
    filter(name %in% taxa_keep) %>%
    group_by(name) %>%
    summarise(
      mean_pct_a = mean(rel_abund[group == label_a]),
      mean_pct_b = mean(rel_abund[group == label_b]),
      mean_rpm_a = mean(rpm[group == label_a]),
      mean_rpm_b = mean(rpm[group == label_b]),
      p_value    = tryCatch(
        wilcox.test(rel_abund[group == label_a], rel_abund[group == label_b],
                    exact = FALSE)$p.value,
        error = function(e) NA_real_),
      .groups = "drop"
    ) %>%
    mutate(
      p_adj     = p.adjust(p_value, method = "BH"),
      sig_unadj = case_when(p_value < 0.001 ~ "***", p_value < 0.01 ~ "**",
                             p_value < 0.05  ~ "*",   TRUE           ~ "ns"),
      sig       = case_when(p_adj   < 0.001 ~ "***", p_adj   < 0.01 ~ "**",
                             p_adj   < 0.05  ~ "*",   TRUE           ~ "ns")
    ) %>%
    arrange(p_value)

  names(results)[names(results) == "mean_pct_a"] <- paste0("Mean ", label_a, " (%)")
  names(results)[names(results) == "mean_pct_b"] <- paste0("Mean ", label_b, " (%)")
  names(results)[names(results) == "mean_rpm_a"] <- paste0("Mean ", label_a, " (BPM)")
  names(results)[names(results) == "mean_rpm_b"] <- paste0("Mean ", label_b, " (BPM)")
  results
}

format_stats_display <- function(results, adjusted = TRUE) {
  pct_cols <- grep("\\(%\\)$",   names(results), value = TRUE)
  rpm_cols <- grep("\\(BPM\\)$", names(results), value = TRUE)
  df <- results
  df[pct_cols] <- lapply(df[pct_cols], round, 2)
  df[rpm_cols] <- lapply(df[rpm_cols], round, 0)
  df$p_value   <- formatC(df$p_value, format = "e", digits = 2)

  if (adjusted) {
    df$p_adj <- formatC(df$p_adj, format = "e", digits = 2)
    df <- df %>% select(name, all_of(pct_cols), all_of(rpm_cols), p_value, p_adj, sig)
    names(df)[names(df) == "sig"]     <- "Sig. (BH)"
    names(df)[names(df) == "p_adj"]   <- "p-adj (BH)"
  } else {
    df <- df %>% select(name, all_of(pct_cols), all_of(rpm_cols), p_value, sig_unadj)
    names(df)[names(df) == "sig_unadj"] <- "Sig."
  }
  names(df)[names(df) == "name"]    <- "Taxon"
  names(df)[names(df) == "p_value"] <- "p-value"
  df
}

show_stats_table <- function(results, adjusted = TRUE, caption_text = NULL) {
  df      <- format_stats_display(results, adjusted)
  sig_col <- if (adjusted) "Sig. (BH)" else "Sig."
  datatable(df, rownames = FALSE, caption = caption_text,
            options = list(pageLength = 25, scrollX = TRUE),
            class = "stripe hover compact") %>%
    formatStyle(sig_col,
                backgroundColor = styleEqual(c("*", "**", "***"),
                                             c("#fff3cd", "#ffc107", "#fd7e14")))
}

draw_stats_table <- function(df, title_str) {
  title_str <- gsub("\u2014|\u2013", "-", title_str)
  sig_col  <- if ("Sig. (BH)" %in% names(df)) "Sig. (BH)" else "Sig."
  sig_vals <- df[[sig_col]]
  n_rows   <- nrow(df)
  n_cols   <- ncol(df)

  row_fill <- case_when(sig_vals == "***" ~ "#fd7e14",
                        sig_vals == "**"  ~ "#ffc107",
                        sig_vals == "*"   ~ "#fff3cd",
                        TRUE              ~ "#ffffff")
  fill_mat <- matrix(rep(row_fill, n_cols), nrow = n_rows, ncol = n_cols)

  tbl <- tableGrob(
    df, rows = NULL,
    theme = ttheme_default(
      core    = list(bg_params = list(fill = fill_mat, col = "#dee2e6"),
                     fg_params = list(fontsize = 7)),
      colhead = list(bg_params = list(fill = "#495057", col = "#dee2e6"),
                     fg_params = list(fontsize = 7.5, col = "white", fontface = "bold"))
    )
  )
  title_grob <- grid::textGrob(title_str,
                                gp   = grid::gpar(fontsize = 10, fontface = "bold"),
                                hjust = 0.5, x = 0.5)
  combined <- arrangeGrob(title_grob, tbl, nrow = 2, heights = c(0.06, 0.94))
  grid::grid.draw(combined)
}

save_stats_pdf <- function(results, title_str, path) {
  unadj_df <- format_stats_display(results, adjusted = FALSE)
  adj_df   <- format_stats_display(results, adjusted = TRUE)
  page_h   <- max(5, nrow(results) * 0.26 + 1.8)
  pdf(path, width = 16, height = page_h)
  draw_stats_table(unadj_df, paste0(title_str, "  |  Unadjusted p-values"))
  grid::grid.newpage()
  draw_stats_table(adj_df,   paste0(title_str, "  |  BH-adjusted p-values (FDR)"))
  dev.off()
  invisible(path)
}

# ── Alpha diversity ────────────────────────────────────────────────────────────
plot_alpha <- function(data_a, data_b, label_a, label_b, title_str,
                       col_a = "#377EB8", col_b = "#E41A1C",
                       save_path = NULL) {
  make_tbl <- function(df, label) {
    wide <- df %>%
      select(sample, name, reads) %>%
      pivot_wider(names_from = name, values_from = reads, values_fill = 0L)
    mat <- wide %>% column_to_rownames("sample") %>% as.matrix()
    tibble(sample   = rownames(mat),
           group    = label,
           Shannon  = diversity(mat, index = "shannon"),
           Simpson  = diversity(mat, index = "simpson"),
           Richness = rowSums(mat > 0))
  }
  alpha_df <- bind_rows(make_tbl(data_a, label_a), make_tbl(data_b, label_b)) %>%
    mutate(group = factor(group, levels = c(label_a, label_b)))
  cols <- setNames(c(col_a, col_b), c(label_a, label_b))

  alpha_long <- alpha_df %>%
    pivot_longer(c(Shannon, Simpson, Richness), names_to = "Metric", values_to = "Value") %>%
    mutate(Metric = factor(Metric, levels = c("Shannon", "Simpson", "Richness")))

  pvals <- alpha_long %>%
    group_by(Metric) %>%
    summarise(
      p = tryCatch(wilcox.test(Value[group == label_a], Value[group == label_b],
                               exact = FALSE)$p.value, error = function(e) NA_real_),
      .groups = "drop"
    ) %>%
    mutate(label = case_when(p < 0.001 ~ "***", p < 0.01 ~ "**", p < 0.05 ~ "*",
                             !is.na(p) ~ paste0("p = ", round(p, 3)), TRUE ~ ""))

  p <- ggplot(alpha_long, aes(x = group, y = Value, fill = group, colour = group)) +
    geom_boxplot(width = 0.45, alpha = 0.6, outlier.shape = NA, colour = "grey30") +
    geom_jitter(width = 0.1, size = 2.2, alpha = 0.85) +
    geom_text(data = pvals, aes(x = 1.5, y = Inf, label = label),
              inherit.aes = FALSE, vjust = 1.6, size = 3.8, colour = "black") +
    facet_wrap(~ Metric, scales = "free_y", ncol = 3) +
    scale_fill_manual(values = cols) +
    scale_colour_manual(values = cols) +
    labs(title    = str_wrap(title_str, 80),
         subtitle = "Wilcoxon rank-sum: *** p < 0.001, ** < 0.01, * < 0.05",
         x = NULL, y = NULL) +
    theme_bw(base_size = 13) +
    theme(legend.position  = "none",
          axis.text        = element_text(size = 11),
          axis.text.x      = element_text(size = 10, angle = 45, hjust = 1),
          strip.text       = element_text(face = "bold", size = 13),
          plot.title       = element_text(face = "bold", size = 14),
          plot.subtitle    = element_text(size = 10, colour = "grey40"),
          strip.background = element_rect(fill = "lightgrey"))
  print(p)
  if (!is.null(save_path)) ggsave(save_path, p, width = 10, height = 5, dpi = 600)
  invisible(p)
}

# ── Beta diversity (Bray-Curtis PCoA + PERMANOVA) ─────────────────────────────
plot_genus_linear_presence <- function(df, sample_labels, title, subtitle = NULL,
                                       layout = "side_by_side", species_colors = NULL,
                                       save_path = NULL) {
  if (nrow(df) == 0) { cat("No data - combined plot skipped.\n"); return(invisible(NULL)) }

  species_order <- df %>%
    group_by(name) %>% summarise(m = mean(rel_abund), .groups = "drop") %>%
    arrange(desc(m)) %>% pull(name)
  lvls   <- rev(species_order)
  n_spp  <- length(species_order)
  n_samp <- length(unique(df$sample))

  if (!is.null(species_colors)) {
    spp_fill <- build_fill_vals(species_order, species_colors, pal = brewer.pal(8, "Dark2"))
    spp_fill <- spp_fill[names(spp_fill) != "Other"]
  } else {
    spp_cols <- if (n_spp <= 8) brewer.pal(max(3, n_spp), "Dark2")[seq_len(n_spp)] else
                  colorRampPalette(brewer.pal(8, "Dark2"))(n_spp)
    spp_fill <- setNames(spp_cols, species_order)
  }

  max_ra <- max(df$rel_abund, na.rm = TRUE)
  x_acc  <- if (max_ra >= 1) 0.1 else if (max_ra >= 0.1) 0.01 else 0.001

  if (layout == "stacked") {
    fig_h <- max(10, n_spp * 0.65 + 6)
    fig_w <- max(14, n_samp * 2.2 + 4)
  } else {
    fig_h <- max(5,  n_spp * 0.35 + 3)
    fig_w <- max(14, n_samp * 2.2 + 4)
  }

  p_bar <- df %>%
    mutate(name   = factor(name, levels = lvls),
           sample = factor(rename_vec(sample, sample_labels))) %>%
    ggplot(aes(y = name, x = rel_abund, fill = name)) +
    geom_col(width = 0.65) +
    facet_wrap(~sample, nrow = 1, scales = "free_x") +
    scale_x_continuous(labels = label_percent(scale = 1, accuracy = x_acc),
                       n.breaks = 3,
                       expand   = expansion(mult = c(0, 0.10))) +
    scale_y_discrete(expand = expansion(add = 0.5)) +
    scale_fill_manual(values = spp_fill, guide = "none") +
    labs(title    = title,
         subtitle = subtitle,
         x = "Relative Abundance (%)", y = NULL) +
    theme_bw(base_size = 12) +
    theme(axis.text.y         = element_text(size = 9, face = "italic"),
          axis.text.x         = element_text(size = 7, angle = 45, hjust = 1),
          panel.grid.major.y  = element_blank(),
          panel.grid.minor    = element_blank(),
          strip.background    = element_rect(fill = "grey92", colour = "grey70"),
          strip.text          = element_text(size = 9, face = "bold"),
          plot.title          = element_text(face = "bold", size = 13),
          plot.subtitle       = element_text(size = 10, colour = "grey40"),
          plot.margin         = if (layout == "stacked") margin(6, 6, 2, 6)
                                else                     margin(6, 2, 6, 6))

  all_samples <- unique(df$sample)
  grid_df <- expand.grid(name = species_order, sample = all_samples,
                         stringsAsFactors = FALSE) %>%
    left_join(df %>% select(sample, name, rel_abund), by = c("sample", "name")) %>%
    mutate(detected = !is.na(rel_abund) & rel_abund > 0,
           name     = factor(name, levels = lvls),
           sample   = rename_vec(sample, sample_labels))

  p_pres <- grid_df %>%
    ggplot(aes(x = sample, y = name, fill = detected)) +
    geom_tile(colour = "white", linewidth = 0.5) +
    scale_fill_manual(values = c("TRUE" = "#1B9E77", "FALSE" = "#EEEEEE"),
                      labels = c("TRUE" = "Detected", "FALSE" = "Not detected"),
                      name   = NULL,
                      guide  = guide_legend(
                        override.aes = list(linewidth = 0),
                        keywidth  = unit(0.8, "lines"),
                        keyheight = unit(0.8, "lines"))) +
    scale_y_discrete(expand = expansion(add = 0.5)) +
    labs(x = NULL, y = NULL, title = "Presence / Absence") +
    theme_bw(base_size = 12) +
    theme(axis.text.x  = element_text(angle = 45, hjust = 1, size = 9),
          plot.title   = element_text(size = 10, colour = "grey30",
                                      face = "bold", hjust = 0),
          legend.position = "right",
          panel.grid   = element_blank(),
          plot.margin  = if (layout == "stacked") margin(2, 6, 6, 6)
                         else                     margin(6, 6, 6, 2))

  if (layout == "side_by_side") {
    p_pres <- p_pres +
      theme(axis.text.y  = element_blank(),
            axis.ticks.y = element_blank())
  } else {
    p_pres <- p_pres +
      theme(axis.text.y = element_text(size = 9, face = "italic"))
  }

  if (layout == "stacked") {
    combined <- (p_bar / p_pres) +
      plot_layout(heights = c(1, 1), guides = "collect") &
      theme(legend.position = "right")
  } else {
    combined <- (p_bar | p_pres) +
      plot_layout(widths = c(n_samp, 1), guides = "collect") &
      theme(legend.position = "right")
  }

  print(combined)
  if (!is.null(save_path))
    ggsave(save_path, combined,
           width     = fig_w,
           height    = fig_h,
           dpi       = 600,
           limitsize = FALSE)
  invisible(combined)
}

plot_genus_presence <- function(df, sample_labels, title, subtitle = NULL, save_path = NULL) {
  if (nrow(df) == 0) { cat("No data - presence/absence plot skipped.\n"); return(invisible(NULL)) }
  species_order <- df %>%
    group_by(name) %>% summarise(m = mean(rel_abund), .groups = "drop") %>%
    arrange(m) %>% pull(name)
  all_samples <- unique(df$sample)
  grid <- expand.grid(name = species_order, sample = all_samples,
                      stringsAsFactors = FALSE) %>%
    left_join(df %>% select(sample, name, rel_abund), by = c("sample", "name")) %>%
    mutate(detected = !is.na(rel_abund) & rel_abund > 0,
           name     = factor(name, levels = species_order),
           sample   = rename_vec(sample, sample_labels))
  p <- grid %>%
    ggplot(aes(x = sample, y = name, fill = detected)) +
    geom_tile(colour = "white", linewidth = 0.6) +
    scale_fill_manual(values = c("TRUE" = "#1B9E77", "FALSE" = "#EEEEEE"),
                      labels = c("TRUE" = "Detected", "FALSE" = "Not detected"),
                      name   = NULL) +
    scale_y_discrete(expand = expansion(add = 0.5)) +
    labs(title = title, subtitle = subtitle, x = NULL, y = NULL) +
    theme_light(base_size = 13) +
    theme(axis.text.x    = element_text(angle = 30, hjust = 1, size = 11),
          axis.text.y    = element_text(size = 10, face = "italic"),
          plot.title     = element_text(face = "bold", size = 14),
          plot.subtitle  = element_text(size = 10, colour = "grey40"),
          legend.position = "right",
          panel.grid     = element_blank())
  print(p)
  if (!is.null(save_path)) ggsave(save_path, p, width = 10, height = 8, dpi = 600)
  invisible(p)
}

plot_beta_pcoa <- function(data_a, data_b, label_a, label_b, title_str,
                            col_a = "#377EB8", col_b = "#E41A1C",
                            save_path = NULL) {
  combined <- bind_rows(data_a %>% mutate(group = label_a),
                        data_b %>% mutate(group = label_b))
  wide <- combined %>%
    select(sample, name, rel_abund) %>%
    pivot_wider(names_from = name, values_from = rel_abund, values_fill = 0)
  mat <- wide %>% column_to_rownames("sample") %>% as.matrix()

  sample_groups <- combined %>% distinct(sample, group)
  group_vec     <- sample_groups$group[match(rownames(mat), sample_groups$sample)]

  dist_mat <- vegdist(mat, method = "bray")
  pcoa     <- cmdscale(dist_mat, k = 2, eig = TRUE)
  var_expl <- abs(pcoa$eig) / sum(abs(pcoa$eig)) * 100

  coords <- as.data.frame(pcoa$points) %>%
    setNames(c("PC1", "PC2")) %>%
    rownames_to_column("sample") %>%
    mutate(group = factor(group_vec, levels = c(label_a, label_b)))

  perm <- adonis2(dist_mat ~ group_vec, permutations = 999)
  r2   <- round(perm$R2[1], 3)
  p_v  <- perm$`Pr(>F)`[1]

  p <- ggplot(coords, aes(x = PC1, y = PC2, colour = group)) +
    geom_point(size = 3.5, alpha = 0.85) +
    stat_ellipse(level = 0.95, linewidth = 0.8) +
    scale_colour_manual(values = setNames(c(col_a, col_b), c(label_a, label_b))) +
    guides(colour = guide_legend(title = NULL, override.aes = list(size = 4))) +
    labs(title    = str_wrap(title_str, 70),
         subtitle = paste0("Bray-Curtis PCoA | PERMANOVA: R\u00b2 = ", r2,
                           ", p = ", round(p_v, 3), " (999 permutations)"),
         x = paste0("PC1 (", round(var_expl[1], 1), "% variance)"),
         y = paste0("PC2 (", round(var_expl[2], 1), "% variance)")) +
    theme_bw(base_size = 13) +
    theme(plot.title       = element_text(face = "bold", size = 15),
          plot.subtitle    = element_text(size = 11, colour = "grey40"),
          axis.text        = element_text(size = 12),
          axis.title       = element_text(size = 14),
          strip.background = element_rect(fill = "lightgrey"),
          legend.text      = element_text(size = 12),
          legend.position  = "bottom")
  print(p)
  if (!is.null(save_path)) ggsave(save_path, p, width = 8, height = 6, dpi = 600)
  invisible(p)
}
```

---

# Contig Classification Overview

Number of assembled contigs (≥ 1,000 bp) classified by Kraken2 at each taxonomic rank, shown per sample and summarised by group. Rank is inferred directly from the Kraken2 assignment (multi-word name = species; name ending in *aceae* = family; single-word name = genus). Higher-level assignments (order, class, phylum) and noise terms (unclassified, phage, environmental) are excluded from rank counts but included in totals where noted.

```{r contig_overview_fn}
# Raw contig-count summary — no base-pair weighting, no rollup
summarise_ranks <- function(dir_path, group_label,
                             strip_suffix = "_m1000_kraken2\\.out$") {
  files <- dir_ls(dir_path, regexp = "_m1000_kraken2\\.out$")
  noise_re <- regex("phage|virus|Virus|viridae|plasmid|prophage|
                     unclassified|uncultured|environmental|unidentified",
                    ignore_case = TRUE)

  map_dfr(files, function(path) {
    raw <- read.delim(path, header = FALSE, sep = "\t",
                      col.names = c("status", "seq_id", "taxon", "length", "kmer"),
                      stringsAsFactors = FALSE)
    sample_id <- path_file(path) %>% str_remove(strip_suffix)

    total_contigs      <- nrow(raw)
    total_unclassified <- sum(raw$status == "U")

    classified <- raw %>%
      filter(status == "C") %>%
      mutate(
        name  = str_remove(taxon, "\\s*\\(taxid \\d+\\)$") %>% str_trim(),
        taxid = as.integer(str_extract(taxon, "(?<=taxid )\\d+"))
      ) %>%
      filter(taxid != 0L, !str_detect(name, noise_re)) %>%
      mutate(rank = case_when(
        str_detect(name, "\\s")                                         ~ "Species",
        str_detect(name, "aceae$")                                      ~ "Family",
        str_detect(name, "ales$")                                       ~ "Order",
        str_detect(name, "ia$|etes$|bacteria$|ota$|group$|complex$")   ~ "Higher",
        name %in% noise_taxa                                            ~ "Higher",
        TRUE                                                            ~ "Genus"
      ))

    tibble(
      Group              = group_label,
      Sample             = sample_id,
      `Total contigs`    = total_contigs,
      Unclassified       = total_unclassified,
      `Classified`       = total_contigs - total_unclassified,
      Species            = sum(classified$rank == "Species"),
      Genus              = sum(classified$rank == "Genus"),
      Family             = sum(classified$rank == "Family"),
      `Order/Higher`     = sum(classified$rank %in% c("Order", "Higher"))
    )
  })
}
```

```{r contig_overview_table}
base_path <- "/Volumes/Alice/2_POULTRY_PROCESSING/Desktop/Kraken2_contigs"

overview_tbl <- bind_rows(
  summarise_ranks(file.path(base_path, "EVI_CAT"), "Post-EVI (BFBB+CAT)"),
  summarise_ranks(file.path(base_path, "ASC_CAT"), "Post-ASC (BFBB+CAT)"),
  summarise_ranks(file.path(base_path, "EVI_RVS"), "Post-EVI (BPW&RVS)"),
  summarise_ranks(file.path(base_path, "ASC_RVS"), "Post-ASC (BPW&RVS)")
) %>%
  arrange(Group, Sample)

datatable(
  overview_tbl,
  rownames  = FALSE,
  caption   = "Contig counts per sample by Kraken2 classification rank. Species = multi-word name; Genus = single-word name (no family suffix); Family = name ending in 'aceae'.",
  options   = list(pageLength = 24, scrollX = TRUE),
  class     = "stripe hover compact"
) %>%
  formatRound(c("Total contigs", "Unclassified", "Classified",
                "Species", "Genus", "Family", "Order/Higher"),
              digits = 0) %>%
  formatStyle("Group",
              backgroundColor = styleEqual(
                c("Post-EVI (BFBB+CAT)", "Post-ASC (BFBB+CAT)",
                  "Post-EVI (BPW&RVS)",  "Post-ASC (BPW&RVS)"),
                c("#dbeafe", "#fee2e2", "#dcfce7", "#fef9c3")
              ))
```

```{r contig_overview_group_summary}
# Group-level summary (mean ± SD across 6 chickens per group)
group_summary <- overview_tbl %>%
  group_by(Group) %>%
  summarise(across(c(`Total contigs`, Unclassified, Classified,
                     Species, Genus, Family, `Order/Higher`),
                   list(mean = ~ round(mean(.)), sd = ~ round(sd(.))),
                   .names = "{.col}__{.fn}"),
            .groups = "drop") %>%
  pivot_longer(-Group, names_to = c("Metric", "stat"), names_sep = "__") %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  mutate(`Mean (SD)` = paste0(mean, " (±", sd, ")")) %>%
  select(Group, Metric, `Mean (SD)`) %>%
  pivot_wider(names_from = Group, values_from = `Mean (SD)`)

datatable(
  group_summary,
  rownames  = FALSE,
  caption   = "Mean (±SD) contig counts per group across 6 samples.",
  options   = list(pageLength = 10, dom = "t"),
  class     = "stripe hover compact"
)
```

---

# Post-EVI (BFBB+CAT) vs. Post-ASC (BFBB+CAT) (BF Bolton base + CAT) — Contigs

```{r cat_paths}
base_path   <- "/Volumes/Alice/2_POULTRY_PROCESSING/Desktop/Kraken2_contigs"
asc_cat_dir <- file.path(base_path, "ASC_CAT")
evi_cat_dir <- file.path(base_path, "EVI_CAT")
```

```{r cat_load}
evi_CAT_data <- load_group(evi_cat_dir) %>% normalise_all()
asc_CAT_data <- load_group(asc_cat_dir) %>% normalise_all()
all_CAT_data <- bind_rows(evi_CAT_data, asc_CAT_data)
evi_CAT_fam  <- load_group(evi_cat_dir, rank_code = "F") %>% normalise_all()
asc_CAT_fam  <- load_group(asc_cat_dir, rank_code = "F") %>% normalise_all()
all_CAT_fam  <- bind_rows(evi_CAT_fam, asc_CAT_fam)
```

```{r cat_top12}
top12_CAT <- all_CAT_data %>%
  group_by(name) %>% summarise(mean_abund = mean(rel_abund), .groups = "drop") %>%
  slice_max(mean_abund, n = 12, with_ties = FALSE) %>% arrange(desc(mean_abund)) %>% pull(name)
sp_CAT_order  <- c(top12_CAT, "Other")
fill_vals_CAT <- build_fill_vals(top12_CAT, genus_colors_fixed)
```

## Top 12 Genera — Stacked Bar

```{r cat_plots, fig.height=10, fig.width=12}
combined_cat <- bind_rows(
  top_n_collapse(evi_CAT_data, top12_CAT) %>% mutate(group = "Post-EVI (BFBB+CAT)"),
  top_n_collapse(asc_CAT_data, top12_CAT) %>% mutate(group = "Post-ASC (BFBB+CAT)")
) %>% mutate(group = factor(group, levels = c("Post-EVI (BFBB+CAT)", "Post-ASC (BFBB+CAT)")))
p <- plot_grouped(combined_cat, sp_CAT_order, fill_vals_CAT,
                  title_str   = "Bacterial Diversity in Post-EVI and Post-ASC Samples Enriched in BF Bolton Base Plus CAT of Contigs (Relative Abundance (%) of the Top 12 Genera)",
                  caption_str = "Remaining taxa grouped as 'Other'",
                  sample_labels = sample_labels)
print(p)
ggsave(file.path(fig_dir, "01_EVI_CAT_ASC_CAT_genus_top12_bar.png"), p, width = 12, height = 10, dpi = 600)
```

## Genus — Heatmap (%)

```{r cat_heatmap_genus, fig.height=14, fig.width=14}
make_heatmap(all_CAT_data, "EVI_CAT", "ASC_CAT", "Post-EVI (BFBB+CAT)", "Post-ASC (BFBB+CAT)",
             "Relative Abundance (%) of Bacterial Genera: Post-EVI (BFBB+CAT) vs. Post-ASC (BFBB+CAT) (Contigs)",
             save_path = file.path(fig_dir, "02_EVI_CAT_ASC_CAT_genus_all_heatmap_pct.png"),
             save_width = 14, save_height = 14, sample_labels = sample_labels)
```

## Genus — Heatmap (BPM, log₁₀)

```{r cat_heatmap_genus_rpm, fig.height=14, fig.width=14}
make_heatmap_rpm(all_CAT_data, "EVI_CAT", "ASC_CAT", "Post-EVI (BFBB+CAT)", "Post-ASC (BFBB+CAT)",
                 "Bacterial Genera — Base Pairs Per Million (log₁₀): Post-EVI (BFBB+CAT) vs. Post-ASC (BFBB+CAT) (Contigs)",
                 save_path = file.path(fig_dir, "02r_EVI_CAT_ASC_CAT_genus_all_heatmap_rpm.png"),
                 save_width = 14, save_height = 14, sample_labels = sample_labels)
```

## Family — Top 12 Stacked Bar

```{r cat_family_top12}
top12_CAT_fam <- all_CAT_fam %>%
  group_by(name) %>% summarise(mean_abund = mean(rel_abund), .groups = "drop") %>%
  slice_max(mean_abund, n = 12, with_ties = FALSE) %>% arrange(desc(mean_abund)) %>% pull(name)
sp_CAT_fam_order  <- c(top12_CAT_fam, "Other")
fill_vals_CAT_fam <- build_fill_vals(top12_CAT_fam, family_colors_fixed)
```

```{r cat_family_top12_plot, fig.height=10, fig.width=12}
combined_cat_f12 <- bind_rows(
  top_n_collapse(evi_CAT_fam, top12_CAT_fam) %>% mutate(group = "Post-EVI (BFBB+CAT)"),
  top_n_collapse(asc_CAT_fam, top12_CAT_fam) %>% mutate(group = "Post-ASC (BFBB+CAT)")
) %>% mutate(group = factor(group, levels = c("Post-EVI (BFBB+CAT)", "Post-ASC (BFBB+CAT)")))
p <- plot_grouped(combined_cat_f12, sp_CAT_fam_order, fill_vals_CAT_fam,
                  title_str   = "Bacterial Relative Abundance (Top 12 Families): Post-EVI (BFBB+CAT) vs. Post-ASC (BFBB+CAT) (Contigs)",
                  caption_str = "Families ranked by mean relative abundance. Remaining taxa grouped as 'Other'.",
                  sample_labels = sample_labels)
print(p)
ggsave(file.path(fig_dir, "03_EVI_CAT_ASC_CAT_family_top12_bar.png"), p, width = 12, height = 10, dpi = 600)
```

## Family — Heatmap (%)

```{r cat_heatmap_family, fig.height=14, fig.width=14}
make_heatmap(all_CAT_fam, "EVI_CAT", "ASC_CAT", "Post-EVI (BFBB+CAT)", "Post-ASC (BFBB+CAT)",
             "Relative Abundance (%) of Bacterial Families: Post-EVI (BFBB+CAT) vs. Post-ASC (BFBB+CAT) (Contigs)",
             save_path = file.path(fig_dir, "04_EVI_CAT_ASC_CAT_family_all_heatmap_pct.png"),
             save_width = 14, save_height = 14, sample_labels = sample_labels)
```

## Family — Heatmap (BPM, log₁₀)

```{r cat_heatmap_family_rpm, fig.height=14, fig.width=14}
make_heatmap_rpm(all_CAT_fam, "EVI_CAT", "ASC_CAT", "Post-EVI (BFBB+CAT)", "Post-ASC (BFBB+CAT)",
                 "Bacterial Families — Base Pairs Per Million (log₁₀): Post-EVI (BFBB+CAT) vs. Post-ASC (BFBB+CAT) (Contigs)",
                 save_path = file.path(fig_dir, "04r_EVI_CAT_ASC_CAT_family_all_heatmap_rpm.png"),
                 save_width = 14, save_height = 14, sample_labels = sample_labels)
```

## Genus-Level — Statistical Tests {.tabset}

Wilcoxon rank-sum test (two-sided, continuity correction) per genus. Mean relative abundance (%) and BPM shown for each group. Only genera with mean ≥ 0.1% in ≥ 2 samples tested.

```{r cat_genus_stats}
stats_cat_genus <- run_taxon_tests(evi_CAT_data, asc_CAT_data, "Post-EVI (BFBB+CAT)", "Post-ASC (BFBB+CAT)")
save_stats_pdf(stats_cat_genus, "Post-EVI (BFBB+CAT) vs. Post-ASC (BFBB+CAT) — Genus Level (Contigs)",
               file.path(fig_dir, "stats_EVI_CAT_ASC_CAT_genus.pdf"))
```

### Unadjusted p-values

```{r cat_genus_unadj}
show_stats_table(stats_cat_genus, adjusted = FALSE,
                 caption_text = "Post-EVI (BFBB+CAT) vs. Post-ASC (BFBB+CAT) — Genus (Contigs). Unadjusted: * p < 0.05, ** < 0.01, *** < 0.001.")
```

### BH-adjusted p-values

```{r cat_genus_adj}
show_stats_table(stats_cat_genus, adjusted = TRUE,
                 caption_text = "Post-EVI (BFBB+CAT) vs. Post-ASC (BFBB+CAT) — Genus (Contigs). BH-corrected: * p-adj < 0.05, ** < 0.01, *** < 0.001.")
```

### Significant only (BH-adjusted)

```{r cat_genus_sig}
sig <- stats_cat_genus %>% filter(sig != "ns")
if (nrow(sig) == 0) cat("No genera reached significance after BH correction.") else
  show_stats_table(sig, adjusted = TRUE,
                   caption_text = "Post-EVI (BFBB+CAT) vs. Post-ASC (BFBB+CAT) — Significant genera only (p-adj < 0.05).")
```

## Family-Level — Statistical Tests {.tabset}

```{r cat_family_stats}
stats_cat_fam <- run_taxon_tests(evi_CAT_fam, asc_CAT_fam, "Post-EVI (BFBB+CAT)", "Post-ASC (BFBB+CAT)")
save_stats_pdf(stats_cat_fam, "Post-EVI (BFBB+CAT) vs. Post-ASC (BFBB+CAT) — Family Level (Contigs)",
               file.path(fig_dir, "stats_EVI_CAT_ASC_CAT_family.pdf"))
```

### Unadjusted p-values

```{r cat_family_unadj}
show_stats_table(stats_cat_fam, adjusted = FALSE,
                 caption_text = "Post-EVI (BFBB+CAT) vs. Post-ASC (BFBB+CAT) — Family (Contigs). Unadjusted: * p < 0.05, ** < 0.01, *** < 0.001.")
```

### BH-adjusted p-values

```{r cat_family_adj}
show_stats_table(stats_cat_fam, adjusted = TRUE,
                 caption_text = "Post-EVI (BFBB+CAT) vs. Post-ASC (BFBB+CAT) — Family (Contigs). BH-corrected: * p-adj < 0.05, ** < 0.01, *** < 0.001.")
```

### Significant only (BH-adjusted)

```{r cat_family_sig}
sig <- stats_cat_fam %>% filter(sig != "ns")
if (nrow(sig) == 0) cat("No families reached significance after BH correction.") else
  show_stats_table(sig, adjusted = TRUE,
                   caption_text = "Post-EVI (BFBB+CAT) vs. Post-ASC (BFBB+CAT) — Significant families only (p-adj < 0.05).")
```

## Species-Level — Statistical Tests {.tabset}

Wilcoxon rank-sum test (two-sided, continuity correction) per species. Only species with mean ≥ 0.1% in ≥ 2 samples tested.

```{r cat_species_load}
evi_CAT_spp <- load_group(evi_cat_dir, rank_code = "S") %>% normalise_all()
asc_CAT_spp <- load_group(asc_cat_dir, rank_code = "S") %>% normalise_all()
```

```{r cat_species_stats}
stats_cat_spp <- run_taxon_tests(evi_CAT_spp, asc_CAT_spp, "Post-EVI (BFBB+CAT)", "Post-ASC (BFBB+CAT)")
save_stats_pdf(stats_cat_spp, "Post-EVI (BFBB+CAT) vs. Post-ASC (BFBB+CAT) — Species Level (Contigs)",
               file.path(fig_dir, "stats_EVI_CAT_ASC_CAT_species.pdf"))
```

### Unadjusted p-values

```{r cat_species_unadj}
show_stats_table(stats_cat_spp, adjusted = FALSE,
                 caption_text = "Post-EVI (BFBB+CAT) vs. Post-ASC (BFBB+CAT) — Species (Contigs). Unadjusted: * p < 0.05, ** < 0.01, *** < 0.001.")
```

### BH-adjusted p-values

```{r cat_species_adj}
show_stats_table(stats_cat_spp, adjusted = TRUE,
                 caption_text = "Post-EVI (BFBB+CAT) vs. Post-ASC (BFBB+CAT) — Species (Contigs). BH-corrected: * p-adj < 0.05, ** < 0.01, *** < 0.001.")
```

### Significant only (BH-adjusted)

```{r cat_species_sig}
sig <- stats_cat_spp %>% filter(sig != "ns")
if (nrow(sig) == 0) cat("No species reached significance after BH correction.") else
  show_stats_table(sig, adjusted = TRUE,
                   caption_text = "Post-EVI (BFBB+CAT) vs. Post-ASC (BFBB+CAT) — Significant species only (p-adj < 0.05).")
```

## Alpha Diversity

Observed richness (number of genera detected), Shannon index (richness and evenness), and Simpson index (dominance) calculated per sample from genus-level contig counts. Wilcoxon rank-sum p-value shown above each panel.

```{r cat_alpha_div, fig.height=5, fig.width=10}
plot_alpha(evi_CAT_data, asc_CAT_data, "Post-EVI (BFBB+CAT)", "Post-ASC (BFBB+CAT)",
           title_str = "Alpha Diversity (Genus Level): Post-EVI (BFBB+CAT) vs. Post-ASC (BFBB+CAT) (Contigs)",
           save_path = file.path(fig_dir, "13_EVI_CAT_ASC_CAT_alpha_diversity.png"))
```

## Beta Diversity — Bray-Curtis PCoA

Principal Coordinates Analysis (PCoA) of Bray-Curtis dissimilarity calculated from genus-level relative abundances. Ellipses represent 95% confidence intervals. PERMANOVA (999 permutations) tests for significant community composition differences between groups.

```{r cat_beta_div, fig.height=6, fig.width=8}
plot_beta_pcoa(evi_CAT_data, asc_CAT_data, "Post-EVI (BFBB+CAT)", "Post-ASC (BFBB+CAT)",
               title_str = "Beta Diversity — Bray-Curtis PCoA (Genus Level): Post-EVI (BFBB+CAT) vs. Post-ASC (BFBB+CAT) (Contigs)",
               save_path = file.path(fig_dir, "14_EVI_CAT_ASC_CAT_beta_diversity_pcoa.png"))
```

---

# Post-EVI (BPW&RVS) vs. Post-ASC (BPW&RVS) (BPW + RVS) — Contigs

```{r rvs_paths}
asc_rvs_dir <- file.path(base_path, "ASC_RVS")
evi_rvs_dir <- file.path(base_path, "EVI_RVS")
```

```{r rvs_genus_load}
evi_RVS_genus <- load_group(evi_rvs_dir) %>% normalise_all()
asc_RVS_genus <- load_group(asc_rvs_dir) %>% normalise_all()
all_RVS_genus <- bind_rows(evi_RVS_genus, asc_RVS_genus)
```

## Top 12 Genera — Stacked Bar

```{r rvs_genus_top12}
top12_RVS_genus <- all_RVS_genus %>%
  group_by(name) %>% summarise(mean_abund = mean(rel_abund), .groups = "drop") %>%
  slice_max(mean_abund, n = 12, with_ties = FALSE) %>% arrange(desc(mean_abund)) %>% pull(name)
sp_RVS_genus_order  <- c(top12_RVS_genus, "Other")
fill_vals_RVS_genus <- build_fill_vals(top12_RVS_genus, genus_colors_fixed)
```

```{r rvs_genus_top12_plot, fig.height=10, fig.width=12}
combined_rvs_g12 <- bind_rows(
  top_n_collapse(evi_RVS_genus, top12_RVS_genus) %>% mutate(group = "Post-EVI (BPW&RVS)"),
  top_n_collapse(asc_RVS_genus, top12_RVS_genus) %>% mutate(group = "Post-ASC (BPW&RVS)")
) %>% mutate(group = factor(group, levels = c("Post-EVI (BPW&RVS)", "Post-ASC (BPW&RVS)")))
p <- plot_grouped(combined_rvs_g12, sp_RVS_genus_order, fill_vals_RVS_genus,
                  title_str   = "Bacterial Relative Abundance (Top 12 Genera): Post-EVI (BPW&RVS) vs. Post-ASC (BPW&RVS) (Contigs)",
                  caption_str = "Genera ranked by mean relative abundance. Remaining taxa grouped as 'Other'.",
                  sample_labels = sample_labels)
print(p)
ggsave(file.path(fig_dir, "05_EVI_RVS_ASC_RVS_genus_top12_bar.png"), p, width = 12, height = 10, dpi = 600)
```

## All Genera — Stacked Bar

```{r rvs_genus_all_taxa}
all_RVS_genera_keep <- all_RVS_genus %>%
  group_by(name) %>%
  summarise(mean_abund = mean(rel_abund), n_present = sum(rel_abund > 0), .groups = "drop") %>%
  filter(mean_abund >= 0.1, n_present >= 2) %>% arrange(desc(mean_abund)) %>% pull(name)
sp_all_g_order      <- c(all_RVS_genera_keep, "Other")
fill_vals_all_RVS_g <- build_fill_vals(all_RVS_genera_keep, genus_colors_fixed,
                                        pal = rep(custom_pal, length.out = length(all_RVS_genera_keep) + 12))
```

```{r rvs_genus_all_plot, fig.height=10, fig.width=14}
combined_rvs_g_all <- bind_rows(
  top_n_collapse(evi_RVS_genus, all_RVS_genera_keep) %>% mutate(group = "Post-EVI (BPW&RVS)"),
  top_n_collapse(asc_RVS_genus, all_RVS_genera_keep) %>% mutate(group = "Post-ASC (BPW&RVS)")
) %>% mutate(group = factor(group, levels = c("Post-EVI (BPW&RVS)", "Post-ASC (BPW&RVS)")))
p <- plot_grouped(combined_rvs_g_all, sp_all_g_order, fill_vals_all_RVS_g,
                  title_str   = "Bacterial Relative Abundance (All Genera, ≥0.1% mean, ≥2 samples): Post-EVI (BPW&RVS) vs. Post-ASC (BPW&RVS) (Contigs)",
                  caption_str = "Rare taxa (mean <0.1% or present in <2 samples) grouped as 'Other'.",
                  sample_labels = sample_labels)
print(p)
ggsave(file.path(fig_dir, "06_EVI_RVS_ASC_RVS_genus_all_bar.png"), p, width = 14, height = 10, dpi = 600)
```

## Top 12 Genera — Heatmap (%)

```{r rvs_genus_top12_heatmap, fig.height=10, fig.width=14}
make_heatmap(all_RVS_genus, "EVI_RVS", "ASC_RVS", "Post-EVI (BPW&RVS)", "Post-ASC (BPW&RVS)",
             "Top 12 Genera — Relative Abundance (%): Post-EVI (BPW&RVS) vs. Post-ASC (BPW&RVS) (Contigs)",
             keep_taxa = top12_RVS_genus, cellheight = 18, cellwidth = 55,
             save_path = file.path(fig_dir, "07_EVI_RVS_ASC_RVS_genus_top12_heatmap.png"),
             save_width = 14, save_height = 10, sample_labels = sample_labels)
```

## All Genera — Heatmap (%)

```{r rvs_genus_all_heatmap, fig.height=14, fig.width=14}
make_heatmap(all_RVS_genus, "EVI_RVS", "ASC_RVS", "Post-EVI (BPW&RVS)", "Post-ASC (BPW&RVS)",
             "All Genera (≥0.05% mean, ≥2 samples) — Relative Abundance (%): Post-EVI (BPW&RVS) vs. Post-ASC (BPW&RVS) (Contigs)",
             save_path = file.path(fig_dir, "08_EVI_RVS_ASC_RVS_genus_all_heatmap_pct.png"),
             save_width = 14, save_height = 14, sample_labels = sample_labels)
```

## All Genera — Heatmap (BPM, log₁₀)

```{r rvs_genus_all_heatmap_rpm, fig.height=14, fig.width=14}
make_heatmap_rpm(all_RVS_genus, "EVI_RVS", "ASC_RVS", "Post-EVI (BPW&RVS)", "Post-ASC (BPW&RVS)",
                 "All Genera — Base Pairs Per Million (log₁₀): Post-EVI (BPW&RVS) vs. Post-ASC (BPW&RVS) (Contigs)",
                 save_path = file.path(fig_dir, "08r_EVI_RVS_ASC_RVS_genus_all_heatmap_rpm.png"),
                 save_width = 14, save_height = 14, sample_labels = sample_labels)
```

## Genus-Level — Statistical Tests {.tabset}

Wilcoxon rank-sum test (two-sided, continuity correction) per genus. Mean relative abundance (%) and BPM shown for each group. Only genera with mean ≥ 0.1% in ≥ 2 samples tested.

```{r rvs_genus_stats}
stats_rvs_genus <- run_taxon_tests(evi_RVS_genus, asc_RVS_genus, "Post-EVI (BPW&RVS)", "Post-ASC (BPW&RVS)")
save_stats_pdf(stats_rvs_genus, "Post-EVI (BPW&RVS) vs. Post-ASC (BPW&RVS) — Genus Level (Contigs)",
               file.path(fig_dir, "stats_EVI_RVS_ASC_RVS_genus.pdf"))
```

### Unadjusted p-values

```{r rvs_genus_unadj}
show_stats_table(stats_rvs_genus, adjusted = FALSE,
                 caption_text = "Post-EVI (BPW&RVS) vs. Post-ASC (BPW&RVS) — Genus (Contigs). Unadjusted: * p < 0.05, ** < 0.01, *** < 0.001.")
```

### BH-adjusted p-values

```{r rvs_genus_adj}
show_stats_table(stats_rvs_genus, adjusted = TRUE,
                 caption_text = "Post-EVI (BPW&RVS) vs. Post-ASC (BPW&RVS) — Genus (Contigs). BH-corrected: * p-adj < 0.05, ** < 0.01, *** < 0.001.")
```

### Significant only (BH-adjusted)

```{r rvs_genus_sig}
sig <- stats_rvs_genus %>% filter(sig != "ns")
if (nrow(sig) == 0) cat("No genera reached significance after BH correction.") else
  show_stats_table(sig, adjusted = TRUE,
                   caption_text = "Post-EVI (BPW&RVS) vs. Post-ASC (BPW&RVS) — Significant genera only (p-adj < 0.05).")
```

---

# Post-EVI (BPW&RVS) vs. Post-ASC (BPW&RVS) — Family Level

```{r rvs_family_load}
evi_RVS_fam <- load_group(evi_rvs_dir, rank_code = "F") %>% normalise_all()
asc_RVS_fam <- load_group(asc_rvs_dir, rank_code = "F") %>% normalise_all()
all_RVS_fam <- bind_rows(evi_RVS_fam, asc_RVS_fam)
```

## Top 12 Families — Stacked Bar

```{r rvs_family_top12}
top12_RVS_fam <- all_RVS_fam %>%
  group_by(name) %>% summarise(mean_abund = mean(rel_abund), .groups = "drop") %>%
  slice_max(mean_abund, n = 12, with_ties = FALSE) %>% arrange(desc(mean_abund)) %>% pull(name)
sp_RVS_fam_order  <- c(top12_RVS_fam, "Other")
fill_vals_RVS_fam <- build_fill_vals(top12_RVS_fam, family_colors_fixed)
```

```{r rvs_family_top12_plot, fig.height=10, fig.width=12}
combined_rvs_f12 <- bind_rows(
  top_n_collapse(evi_RVS_fam, top12_RVS_fam) %>% mutate(group = "Post-EVI (BPW&RVS)"),
  top_n_collapse(asc_RVS_fam, top12_RVS_fam) %>% mutate(group = "Post-ASC (BPW&RVS)")
) %>% mutate(group = factor(group, levels = c("Post-EVI (BPW&RVS)", "Post-ASC (BPW&RVS)")))
p <- plot_grouped(combined_rvs_f12, sp_RVS_fam_order, fill_vals_RVS_fam,
                  title_str   = "Bacterial Relative Abundance (Top 12 Families): Post-EVI (BPW&RVS) vs. Post-ASC (BPW&RVS) (Contigs)",
                  caption_str = "Families ranked by mean relative abundance. Remaining taxa grouped as 'Other'.",
                  sample_labels = sample_labels)
print(p)
ggsave(file.path(fig_dir, "09_EVI_RVS_ASC_RVS_family_top12_bar.png"), p, width = 12, height = 10, dpi = 600)
```

## All Families — Stacked Bar

```{r rvs_family_all_taxa}
all_RVS_fam_keep <- all_RVS_fam %>%
  group_by(name) %>%
  summarise(mean_abund = mean(rel_abund), n_present = sum(rel_abund > 0), .groups = "drop") %>%
  filter(mean_abund >= 0.1, n_present >= 2) %>% arrange(desc(mean_abund)) %>% pull(name)
sp_all_f_order      <- c(all_RVS_fam_keep, "Other")
fill_vals_all_RVS_f <- build_fill_vals(all_RVS_fam_keep, family_colors_fixed,
                                        pal = rep(custom_pal, length.out = length(all_RVS_fam_keep) + 12))
```

```{r rvs_family_all_plot, fig.height=10, fig.width=14}
combined_rvs_f_all <- bind_rows(
  top_n_collapse(evi_RVS_fam, all_RVS_fam_keep) %>% mutate(group = "Post-EVI (BPW&RVS)"),
  top_n_collapse(asc_RVS_fam, all_RVS_fam_keep) %>% mutate(group = "Post-ASC (BPW&RVS)")
) %>% mutate(group = factor(group, levels = c("Post-EVI (BPW&RVS)", "Post-ASC (BPW&RVS)")))
p <- plot_grouped(combined_rvs_f_all, sp_all_f_order, fill_vals_all_RVS_f,
                  title_str   = "Bacterial Relative Abundance (All Families, ≥0.1% mean, ≥2 samples): Post-EVI (BPW&RVS) vs. Post-ASC (BPW&RVS) (Contigs)",
                  caption_str = "Rare taxa (mean <0.1% or present in <2 samples) grouped as 'Other'.",
                  sample_labels = sample_labels)
print(p)
ggsave(file.path(fig_dir, "10_EVI_RVS_ASC_RVS_family_all_bar.png"), p, width = 14, height = 10, dpi = 600)
```

## Top 12 Families — Heatmap (%)

```{r rvs_family_top12_heatmap, fig.height=10, fig.width=14}
make_heatmap(all_RVS_fam, "EVI_RVS", "ASC_RVS", "Post-EVI (BPW&RVS)", "Post-ASC (BPW&RVS)",
             "Top 12 Families — Relative Abundance (%): Post-EVI (BPW&RVS) vs. Post-ASC (BPW&RVS) (Contigs)",
             keep_taxa = top12_RVS_fam, cellheight = 18, cellwidth = 55,
             save_path = file.path(fig_dir, "11_EVI_RVS_ASC_RVS_family_top12_heatmap.png"),
             save_width = 14, save_height = 10, sample_labels = sample_labels)
```

## All Families — Heatmap (%)

```{r rvs_family_all_heatmap, fig.height=14, fig.width=14}
make_heatmap(all_RVS_fam, "EVI_RVS", "ASC_RVS", "Post-EVI (BPW&RVS)", "Post-ASC (BPW&RVS)",
             "All Families (≥0.05% mean, ≥2 samples) — Relative Abundance (%): Post-EVI (BPW&RVS) vs. Post-ASC (BPW&RVS) (Contigs)",
             save_path = file.path(fig_dir, "12_EVI_RVS_ASC_RVS_family_all_heatmap_pct.png"),
             save_width = 14, save_height = 14, sample_labels = sample_labels)
```

## All Families — Heatmap (BPM, log₁₀)

```{r rvs_family_all_heatmap_rpm, fig.height=14, fig.width=14}
make_heatmap_rpm(all_RVS_fam, "EVI_RVS", "ASC_RVS", "Post-EVI (BPW&RVS)", "Post-ASC (BPW&RVS)",
                 "All Families — Base Pairs Per Million (log₁₀): Post-EVI (BPW&RVS) vs. Post-ASC (BPW&RVS) (Contigs)",
                 save_path = file.path(fig_dir, "12r_EVI_RVS_ASC_RVS_family_all_heatmap_rpm.png"),
                 save_width = 14, save_height = 14, sample_labels = sample_labels)
```

## Family-Level — Statistical Tests {.tabset}

```{r rvs_family_stats}
stats_rvs_fam <- run_taxon_tests(evi_RVS_fam, asc_RVS_fam, "Post-EVI (BPW&RVS)", "Post-ASC (BPW&RVS)")
save_stats_pdf(stats_rvs_fam, "Post-EVI (BPW&RVS) vs. Post-ASC (BPW&RVS) — Family Level (Contigs)",
               file.path(fig_dir, "stats_EVI_RVS_ASC_RVS_family.pdf"))
```

### Unadjusted p-values

```{r rvs_family_unadj}
show_stats_table(stats_rvs_fam, adjusted = FALSE,
                 caption_text = "Post-EVI (BPW&RVS) vs. Post-ASC (BPW&RVS) — Family (Contigs). Unadjusted: * p < 0.05, ** < 0.01, *** < 0.001.")
```

### BH-adjusted p-values

```{r rvs_family_adj}
show_stats_table(stats_rvs_fam, adjusted = TRUE,
                 caption_text = "Post-EVI (BPW&RVS) vs. Post-ASC (BPW&RVS) — Family (Contigs). BH-corrected: * p-adj < 0.05, ** < 0.01, *** < 0.001.")
```

### Significant only (BH-adjusted)

```{r rvs_family_sig}
sig <- stats_rvs_fam %>% filter(sig != "ns")
if (nrow(sig) == 0) cat("No families reached significance after BH correction.") else
  show_stats_table(sig, adjusted = TRUE,
                   caption_text = "Post-EVI (BPW&RVS) vs. Post-ASC (BPW&RVS) — Significant families only (p-adj < 0.05).")
```

## Species-Level — Statistical Tests {.tabset}

Wilcoxon rank-sum test (two-sided, continuity correction) per species. Only species with mean ≥ 0.1% in ≥ 2 samples tested.

```{r rvs_species_load}
evi_RVS_spp <- load_group(evi_rvs_dir, rank_code = "S") %>% normalise_all()
asc_RVS_spp <- load_group(asc_rvs_dir, rank_code = "S") %>% normalise_all()
```

```{r rvs_species_stats}
stats_rvs_spp <- run_taxon_tests(evi_RVS_spp, asc_RVS_spp, "Post-EVI (BPW&RVS)", "Post-ASC (BPW&RVS)")
save_stats_pdf(stats_rvs_spp, "Post-EVI (BPW&RVS) vs. Post-ASC (BPW&RVS) — Species Level (Contigs)",
               file.path(fig_dir, "stats_EVI_RVS_ASC_RVS_species.pdf"))
```

### Unadjusted p-values

```{r rvs_species_unadj}
show_stats_table(stats_rvs_spp, adjusted = FALSE,
                 caption_text = "Post-EVI (BPW&RVS) vs. Post-ASC (BPW&RVS) — Species (Contigs). Unadjusted: * p < 0.05, ** < 0.01, *** < 0.001.")
```

### BH-adjusted p-values

```{r rvs_species_adj}
show_stats_table(stats_rvs_spp, adjusted = TRUE,
                 caption_text = "Post-EVI (BPW&RVS) vs. Post-ASC (BPW&RVS) — Species (Contigs). BH-corrected: * p-adj < 0.05, ** < 0.01, *** < 0.001.")
```

### Significant only (BH-adjusted)

```{r rvs_species_sig}
sig <- stats_rvs_spp %>% filter(sig != "ns")
if (nrow(sig) == 0) cat("No species reached significance after BH correction.") else
  show_stats_table(sig, adjusted = TRUE,
                   caption_text = "Post-EVI (BPW&RVS) vs. Post-ASC (BPW&RVS) — Significant species only (p-adj < 0.05).")
```

## Alpha Diversity

Observed richness (number of genera detected), Shannon index (richness and evenness), and Simpson index (dominance) calculated per sample from genus-level contig counts. Wilcoxon rank-sum p-value shown above each panel.

```{r rvs_alpha_div, fig.height=5, fig.width=10}
plot_alpha(evi_RVS_genus, asc_RVS_genus, "Post-EVI (BPW&RVS)", "Post-ASC (BPW&RVS)",
           title_str = "Alpha Diversity (Genus Level): Post-EVI (BPW&RVS) vs. Post-ASC (BPW&RVS) (Contigs)",
           save_path = file.path(fig_dir, "15_EVI_RVS_ASC_RVS_alpha_diversity.png"))
```

## Beta Diversity — Bray-Curtis PCoA

Principal Coordinates Analysis (PCoA) of Bray-Curtis dissimilarity calculated from genus-level relative abundances. Ellipses represent 95% confidence intervals. PERMANOVA (999 permutations) tests for significant community composition differences between groups.

```{r rvs_beta_div, fig.height=6, fig.width=8}
plot_beta_pcoa(evi_RVS_genus, asc_RVS_genus, "Post-EVI (BPW&RVS)", "Post-ASC (BPW&RVS)",
               title_str = "Beta Diversity — Bray-Curtis PCoA (Genus Level): Post-EVI (BPW&RVS) vs. Post-ASC (BPW&RVS) (Contigs)",
               save_path = file.path(fig_dir, "16_EVI_RVS_ASC_RVS_beta_diversity_pcoa.png"))
```

---

# *Escherichia* Species Diversity - CAT Enrichment (Contigs)

Species within the genus *Escherichia* detected in Post-EVI and Post-ASC samples enriched with BF Bolton Base + CAT. Relative abundance is expressed as a percentage of total classified contig base pairs per sample.

## Stacked Bar - Relative Abundance (%)

```{r cat_escherichia_bar, fig.height=7, fig.width=12}
cat_esch <- bind_rows(
  evi_CAT_spp %>% mutate(group = "Post-EVI (BFBB+CAT)"),
  asc_CAT_spp %>% mutate(group = "Post-ASC (BFBB+CAT)")
) %>%
  filter(str_detect(name, "^Escherichia")) %>%
  mutate(group = factor(group, levels = c("Post-EVI (BFBB+CAT)", "Post-ASC (BFBB+CAT)")))

esch_cat_spp <- cat_esch %>%
  group_by(name) %>%
  summarise(mean_ra = mean(rel_abund), .groups = "drop") %>%
  arrange(desc(mean_ra)) %>%
  pull(name)

n_esch_cat   <- length(esch_cat_spp)
esch_cat_fill <- build_fill_vals(esch_cat_species, species_colors_escherichia, pal = brewer.pal(8, "Set2"))

p_esch_cat_bar <- cat_esch %>%
  mutate(name   = factor(name, levels = esch_cat_spp),
         sample = rename_vec(sample, sample_labels)) %>%
  ggplot(aes(x = sample, y = rel_abund, fill = name)) +
  geom_col(width = 0.65, colour = "white", linewidth = 0.25) +
  facet_wrap(~ group, scales = "free_x", ncol = 1) +
  scale_fill_manual(values = esch_cat_fill, name = "Species") +
  scale_y_continuous(labels = label_percent(scale = 1),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Escherichia Species Relative Abundance - CAT Enrichment (Contigs)",
       x = NULL, y = "Relative Abundance (%)") +
  theme_light(base_size = 13) +
  theme(axis.text.x     = element_text(angle = 0, hjust = 0.5, size = 11),
        axis.text.y     = element_text(size = 12),
        legend.text     = element_text(size = 10, face = "italic"),
        legend.key.size = unit(0.5, "cm"),
        plot.title      = element_text(face = "bold", size = 14),
        strip.text      = element_text(face = "bold", size = 12),
        strip.background = element_rect(fill = "lightgrey"),
        legend.position = "right")
print(p_esch_cat_bar)
ggsave(file.path(fig_dir, "17_CAT_Escherichia_species_bar.png"),
       p_esch_cat_bar, width = 12, height = 7, dpi = 600)
```

## Linear Bar + Presence / Absence

```{r cat_escherichia_combined, fig.height=10, fig.width=18}
plot_genus_linear_presence(
  cat_esch, sample_labels,
  title     = "Escherichia Species Relative Abundance - CAT Enrichment (Contigs)",
  species_colors = species_colors_escherichia,
  save_path = file.path(fig_dir, "17e_CAT_Escherichia_linear_presence.png")
)
```

## Presence / Absence

```{r cat_escherichia_presence, fig.height=8, fig.width=10}
plot_genus_presence(
  cat_esch, sample_labels,
  title     = "Escherichia Species Presence / Absence - CAT Enrichment (Contigs)",
  save_path = file.path(fig_dir, "17p_CAT_Escherichia_presence.png")
)
```

---

# *Escherichia* Species Diversity - RVS Enrichment (Contigs)

Species within the genus *Escherichia* detected in Post-EVI and Post-ASC samples enriched with BPW + RVS. Relative abundance is expressed as a percentage of total classified contig base pairs per sample.

## Stacked Bar - Relative Abundance (%)

```{r rvs_escherichia_bar, fig.height=7, fig.width=12}
rvs_esch <- bind_rows(
  evi_RVS_spp %>% mutate(group = "Post-EVI (BPW&RVS)"),
  asc_RVS_spp %>% mutate(group = "Post-ASC (BPW&RVS)")
) %>%
  filter(str_detect(name, "^Escherichia")) %>%
  mutate(group = factor(group, levels = c("Post-EVI (BPW&RVS)", "Post-ASC (BPW&RVS)")))

esch_rvs_spp <- rvs_esch %>%
  group_by(name) %>%
  summarise(mean_ra = mean(rel_abund), .groups = "drop") %>%
  arrange(desc(mean_ra)) %>%
  pull(name)

n_esch_rvs   <- length(esch_rvs_spp)
esch_rvs_fill <- build_fill_vals(esch_rvs_species, species_colors_escherichia, pal = brewer.pal(8, "Set2"))

p_esch_rvs_bar <- rvs_esch %>%
  mutate(name   = factor(name, levels = esch_rvs_spp),
         sample = rename_vec(sample, sample_labels)) %>%
  ggplot(aes(x = sample, y = rel_abund, fill = name)) +
  geom_col(width = 0.65, colour = "white", linewidth = 0.25) +
  facet_wrap(~ group, scales = "free_x", ncol = 1) +
  scale_fill_manual(values = esch_rvs_fill, name = "Species") +
  scale_y_continuous(labels = label_percent(scale = 1),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Escherichia Species Relative Abundance - RVS Enrichment (Contigs)",
       x = NULL, y = "Relative Abundance (%)") +
  theme_light(base_size = 13) +
  theme(axis.text.x     = element_text(angle = 0, hjust = 0.5, size = 11),
        axis.text.y     = element_text(size = 12),
        legend.text     = element_text(size = 10, face = "italic"),
        legend.key.size = unit(0.5, "cm"),
        plot.title      = element_text(face = "bold", size = 14),
        strip.text      = element_text(face = "bold", size = 12),
        strip.background = element_rect(fill = "lightgrey"),
        legend.position = "right")
print(p_esch_rvs_bar)
ggsave(file.path(fig_dir, "18_RVS_Escherichia_species_bar.png"),
       p_esch_rvs_bar, width = 12, height = 7, dpi = 600)
```

## Linear Bar + Presence / Absence

```{r rvs_escherichia_combined, fig.height=10, fig.width=18}
plot_genus_linear_presence(
  rvs_esch, sample_labels,
  title     = "Escherichia Species Relative Abundance - RVS Enrichment (Contigs)",
  species_colors = species_colors_escherichia,
  save_path = file.path(fig_dir, "18e_RVS_Escherichia_linear_presence.png")
)
```

## Presence / Absence

```{r rvs_escherichia_presence, fig.height=8, fig.width=10}
plot_genus_presence(
  rvs_esch, sample_labels,
  title     = "Escherichia Species Presence / Absence - RVS Enrichment (Contigs)",
  save_path = file.path(fig_dir, "18p_RVS_Escherichia_presence.png")
)
```

---

# *Salmonella* Species Diversity - CAT Enrichment (Contigs)

Species within the genus *Salmonella* detected in Post-EVI and Post-ASC samples enriched with BF Bolton Base + CAT. Relative abundance is expressed as a percentage of total classified contig base pairs per sample.

## Stacked Bar - Relative Abundance (%)

```{r cat_salmonella_bar, fig.height=7, fig.width=12}
cat_salm <- bind_rows(
  evi_CAT_spp %>% mutate(group = "Post-EVI (BFBB+CAT)"),
  asc_CAT_spp %>% mutate(group = "Post-ASC (BFBB+CAT)")
) %>%
  filter(str_detect(name, "^Salmonella")) %>%
  mutate(group = factor(group, levels = c("Post-EVI (BFBB+CAT)", "Post-ASC (BFBB+CAT)")))

salm_cat_spp <- cat_salm %>%
  group_by(name) %>%
  summarise(mean_ra = mean(rel_abund), .groups = "drop") %>%
  arrange(desc(mean_ra)) %>%
  pull(name)

n_salm_cat   <- length(salm_cat_spp)
salm_cat_fill <- build_fill_vals(salm_cat_species, species_colors_salmonella, pal = brewer.pal(8, "Set1"))

p_salm_cat_bar <- cat_salm %>%
  mutate(name   = factor(name, levels = salm_cat_spp),
         sample = rename_vec(sample, sample_labels)) %>%
  ggplot(aes(x = sample, y = rel_abund, fill = name)) +
  geom_col(width = 0.65, colour = "white", linewidth = 0.25) +
  facet_wrap(~ group, scales = "free_x", ncol = 1) +
  scale_fill_manual(values = salm_cat_fill, name = "Species") +
  scale_y_continuous(labels = label_percent(scale = 1),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Salmonella Species Relative Abundance - CAT Enrichment (Contigs)",
       x = NULL, y = "Relative Abundance (%)") +
  theme_light(base_size = 13) +
  theme(axis.text.x     = element_text(angle = 0, hjust = 0.5, size = 11),
        axis.text.y     = element_text(size = 12),
        legend.text     = element_text(size = 10, face = "italic"),
        legend.key.size = unit(0.5, "cm"),
        plot.title      = element_text(face = "bold", size = 14),
        strip.text      = element_text(face = "bold", size = 12),
        strip.background = element_rect(fill = "lightgrey"),
        legend.position = "right")
print(p_salm_cat_bar)
ggsave(file.path(fig_dir, "19_CAT_Salmonella_species_bar.png"),
       p_salm_cat_bar, width = 12, height = 7, dpi = 600)
```

## Linear Bar + Presence / Absence

```{r cat_salmonella_combined, fig.height=10, fig.width=18}
plot_genus_linear_presence(
  cat_salm, sample_labels,
  title     = "Salmonella Species Relative Abundance - CAT Enrichment (Contigs)",
  species_colors = species_colors_salmonella,
  save_path = file.path(fig_dir, "19e_CAT_Salmonella_linear_presence.png")
)
```

## Presence / Absence

```{r cat_salmonella_presence, fig.height=8, fig.width=10}
plot_genus_presence(
  cat_salm, sample_labels,
  title     = "Salmonella Species Presence / Absence - CAT Enrichment (Contigs)",
  save_path = file.path(fig_dir, "19p_CAT_Salmonella_presence.png")
)
```

---

# *Salmonella* Species Diversity - RVS Enrichment (Contigs)

Species within the genus *Salmonella* detected in Post-EVI and Post-ASC samples enriched with BPW + RVS. Relative abundance is expressed as a percentage of total classified contig base pairs per sample.

## Stacked Bar - Relative Abundance (%)

```{r rvs_salmonella_bar, fig.height=7, fig.width=12}
rvs_salm <- bind_rows(
  evi_RVS_spp %>% mutate(group = "Post-EVI (BPW&RVS)"),
  asc_RVS_spp %>% mutate(group = "Post-ASC (BPW&RVS)")
) %>%
  filter(str_detect(name, "^Salmonella")) %>%
  mutate(group = factor(group, levels = c("Post-EVI (BPW&RVS)", "Post-ASC (BPW&RVS)")))

salm_rvs_spp <- rvs_salm %>%
  group_by(name) %>%
  summarise(mean_ra = mean(rel_abund), .groups = "drop") %>%
  arrange(desc(mean_ra)) %>%
  pull(name)

n_salm_rvs   <- length(salm_rvs_spp)
salm_rvs_fill <- build_fill_vals(salm_rvs_species, species_colors_salmonella, pal = brewer.pal(8, "Set1"))

p_salm_rvs_bar <- rvs_salm %>%
  mutate(name   = factor(name, levels = salm_rvs_spp),
         sample = rename_vec(sample, sample_labels)) %>%
  ggplot(aes(x = sample, y = rel_abund, fill = name)) +
  geom_col(width = 0.65, colour = "white", linewidth = 0.25) +
  facet_wrap(~ group, scales = "free_x", ncol = 1) +
  scale_fill_manual(values = salm_rvs_fill, name = "Species") +
  scale_y_continuous(labels = label_percent(scale = 1),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Salmonella Species Relative Abundance - RVS Enrichment (Contigs)",
       x = NULL, y = "Relative Abundance (%)") +
  theme_light(base_size = 13) +
  theme(axis.text.x     = element_text(angle = 0, hjust = 0.5, size = 11),
        axis.text.y     = element_text(size = 12),
        legend.text     = element_text(size = 10, face = "italic"),
        legend.key.size = unit(0.5, "cm"),
        plot.title      = element_text(face = "bold", size = 14),
        strip.text      = element_text(face = "bold", size = 12),
        strip.background = element_rect(fill = "lightgrey"),
        legend.position = "right")
print(p_salm_rvs_bar)
ggsave(file.path(fig_dir, "20_RVS_Salmonella_species_bar.png"),
       p_salm_rvs_bar, width = 12, height = 7, dpi = 600)
```

## Linear Bar + Presence / Absence

```{r rvs_salmonella_combined, fig.height=10, fig.width=18}
plot_genus_linear_presence(
  rvs_salm, sample_labels,
  title     = "Salmonella Species Relative Abundance - RVS Enrichment (Contigs)",
  species_colors = species_colors_salmonella,
  save_path = file.path(fig_dir, "20e_RVS_Salmonella_linear_presence.png")
)
```

## Presence / Absence

```{r rvs_salmonella_presence, fig.height=8, fig.width=10}
plot_genus_presence(
  rvs_salm, sample_labels,
  title     = "Salmonella Species Presence / Absence - RVS Enrichment (Contigs)",
  save_path = file.path(fig_dir, "20p_RVS_Salmonella_presence.png")
)
```

---

# *Campylobacter* Species Diversity - CAT Enrichment (Contigs)

Species within the genus *Campylobacter* detected in Post-EVI and Post-ASC samples enriched with BF Bolton Base + CAT. Relative abundance is expressed as a percentage of total classified contig base pairs per sample.

## Stacked Bar - Relative Abundance (%)

```{r cat_campylobacter_bar, fig.height=7, fig.width=12}
cat_camp <- bind_rows(
  evi_CAT_spp %>% mutate(group = "Post-EVI (BFBB+CAT)"),
  asc_CAT_spp %>% mutate(group = "Post-ASC (BFBB+CAT)")
) %>%
  filter(str_detect(name, "^Campylobacter")) %>%
  mutate(group = factor(group, levels = c("Post-EVI (BFBB+CAT)", "Post-ASC (BFBB+CAT)")))

camp_cat_spp <- cat_camp %>%
  group_by(name) %>%
  summarise(mean_ra = mean(rel_abund), .groups = "drop") %>%
  arrange(desc(mean_ra)) %>%
  pull(name)

camp_cat_fill <- build_fill_vals(camp_cat_spp, species_colors_campylobacter, pal = brewer.pal(8, "Dark2"))

p_camp_cat_bar <- cat_camp %>%
  mutate(name   = factor(name, levels = camp_cat_spp),
         sample = rename_vec(sample, sample_labels)) %>%
  ggplot(aes(x = sample, y = rel_abund, fill = name)) +
  geom_col(width = 0.65, colour = "white", linewidth = 0.25) +
  facet_wrap(~ group, scales = "free_x", ncol = 1) +
  scale_fill_manual(values = camp_cat_fill, name = "Species") +
  scale_y_continuous(labels = label_percent(scale = 1),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Campylobacter Species Relative Abundance - CAT Enrichment (Contigs)",
       x = NULL, y = "Relative Abundance (%)") +
  theme_light(base_size = 13) +
  theme(axis.text.x     = element_text(angle = 0, hjust = 0.5, size = 11),
        axis.text.y     = element_text(size = 12),
        legend.text     = element_text(size = 10, face = "italic"),
        legend.key.size = unit(0.5, "cm"),
        plot.title      = element_text(face = "bold", size = 14),
        strip.text      = element_text(face = "bold", size = 12),
        strip.background = element_rect(fill = "lightgrey"),
        legend.position = "right")
print(p_camp_cat_bar)
ggsave(file.path(fig_dir, "21_CAT_Campylobacter_species_bar.png"),
       p_camp_cat_bar, width = 12, height = 7, dpi = 600)
```

## Linear Bar + Presence / Absence

```{r cat_campylobacter_combined, fig.height=10, fig.width=18}
plot_genus_linear_presence(
  cat_camp, sample_labels,
  title          = "Campylobacter Species Relative Abundance - CAT Enrichment (Contigs)",
  species_colors = species_colors_campylobacter,
  save_path      = file.path(fig_dir, "21e_CAT_Campylobacter_linear_presence.png")
)
```

## Presence / Absence

```{r cat_campylobacter_presence, fig.height=8, fig.width=10}
plot_genus_presence(
  cat_camp, sample_labels,
  title     = "Campylobacter Species Presence / Absence - CAT Enrichment (Contigs)",
  save_path = file.path(fig_dir, "21p_CAT_Campylobacter_presence.png")
)
```

---

# *Campylobacter* Species Diversity - RVS Enrichment (Contigs)

Species within the genus *Campylobacter* detected in Post-EVI and Post-ASC samples enriched with BPW + RVS. Relative abundance is expressed as a percentage of total classified contig base pairs per sample.

## Stacked Bar - Relative Abundance (%)

```{r rvs_campylobacter_bar, fig.height=7, fig.width=12}
rvs_camp <- bind_rows(
  evi_RVS_spp %>% mutate(group = "Post-EVI (BPW&RVS)"),
  asc_RVS_spp %>% mutate(group = "Post-ASC (BPW&RVS)")
) %>%
  filter(str_detect(name, "^Campylobacter")) %>%
  mutate(group = factor(group, levels = c("Post-EVI (BPW&RVS)", "Post-ASC (BPW&RVS)")))

camp_rvs_spp <- rvs_camp %>%
  group_by(name) %>%
  summarise(mean_ra = mean(rel_abund), .groups = "drop") %>%
  arrange(desc(mean_ra)) %>%
  pull(name)

camp_rvs_fill <- build_fill_vals(camp_rvs_spp, species_colors_campylobacter, pal = brewer.pal(8, "Dark2"))

p_camp_rvs_bar <- rvs_camp %>%
  mutate(name   = factor(name, levels = camp_rvs_spp),
         sample = rename_vec(sample, sample_labels)) %>%
  ggplot(aes(x = sample, y = rel_abund, fill = name)) +
  geom_col(width = 0.65, colour = "white", linewidth = 0.25) +
  facet_wrap(~ group, scales = "free_x", ncol = 1) +
  scale_fill_manual(values = camp_rvs_fill, name = "Species") +
  scale_y_continuous(labels = label_percent(scale = 1),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Campylobacter Species Relative Abundance - RVS Enrichment (Contigs)",
       x = NULL, y = "Relative Abundance (%)") +
  theme_light(base_size = 13) +
  theme(axis.text.x     = element_text(angle = 0, hjust = 0.5, size = 11),
        axis.text.y     = element_text(size = 12),
        legend.text     = element_text(size = 10, face = "italic"),
        legend.key.size = unit(0.5, "cm"),
        plot.title      = element_text(face = "bold", size = 14),
        strip.text      = element_text(face = "bold", size = 12),
        strip.background = element_rect(fill = "lightgrey"),
        legend.position = "right")
print(p_camp_rvs_bar)
ggsave(file.path(fig_dir, "22_RVS_Campylobacter_species_bar.png"),
       p_camp_rvs_bar, width = 12, height = 7, dpi = 600)
```

## Linear Bar + Presence / Absence

```{r rvs_campylobacter_combined, fig.height=10, fig.width=18}
plot_genus_linear_presence(
  rvs_camp, sample_labels,
  title          = "Campylobacter Species Relative Abundance - RVS Enrichment (Contigs)",
  species_colors = species_colors_campylobacter,
  save_path      = file.path(fig_dir, "22e_RVS_Campylobacter_linear_presence.png")
)
```

## Presence / Absence

```{r rvs_campylobacter_presence, fig.height=8, fig.width=10}
plot_genus_presence(
  rvs_camp, sample_labels,
  title     = "Campylobacter Species Presence / Absence - RVS Enrichment (Contigs)",
  save_path = file.path(fig_dir, "22p_RVS_Campylobacter_presence.png")
)
```
