---
title: "Read Level Normalised Bacterial Diversity - Chapter 3"
output: html_notebook
---

- EVI vs ASC samples 

*Sample normalisation:*
- Bases per million reads = (bases assigned to taxon / total bases in sample) × 1,000,000
- Then presented as relative abundance of classified samples

```{r Install packages}

library(tidyverse)
library(ggplot2)
library(patchwork)
library(scales)
library(vegan)
library(gridExtra)
library(RColorBrewer)

```

```{r Label data}
# Enrichment type labels 

enrichment_labels <- c(
  "No enrichment"        = "No enrichment",
  "BPW + RVS enrichment" = "BPW + RVS enrichment",
  "BF Bolton Base + CAT" = "BF Bolton Base + CAT"
)

# Sample labels 

sample_labels <- c(
  "CH01_ASC"      = "CH01",
  "CH02_ASC"      = "CH02",
  "CH03_ASC"      = "CH03",
  "CH04_ASC"      = "CH04",
  "CH05_ASC"      = "CH05",
  "CH06_ASC"      = "CH06",
  "CH01_ASC_RVS"  = "CH01",
  "CH02_ASC_RVS"  = "CH02",
  "CH03_ASC_RVS"  = "CH03",
  "CH04_ASC_RVS"  = "CH04",
  "CH05_ASC_RVS"  = "CH05",
  "CH06_ASC_RVS"  = "CH06",
  "CH01_ASC_BFBB" = "CH01",
  "CH02_ASC_BFBB" = "CH02",
  "CH03_ASC_BFBB" = "CH03",
  "CH04_ASC_BFBB" = "CH04",
  "CH05_ASC_BFBB" = "CH05",
  "CH06_ASC_BFBB" = "CH06",
  "CH01_EVI"      = "CH01",
  "CH02_EVI"      = "CH02",
  "CH03_EVI"      = "CH03",
  "CH04_EVI"      = "CH04",
  "CH05_EVI"      = "CH05",
  "CH06_EVI"      = "CH06",
  "CH01_EVI_RVS"  = "CH01",
  "CH02_EVI_RVS"  = "CH02",
  "CH03_EVI_RVS"  = "CH03",
  "CH04_EVI_RVS"  = "CH04",
  "CH05_EVI_RVS"  = "CH05",
  "CH06_EVI_RVS"  = "CH06",
  "CH01_EVI_BFBB" = "CH01",
  "CH02_EVI_BFBB" = "CH02",
  "CH03_EVI_BFBB" = "CH03",
  "CH04_EVI_BFBB" = "CH04",
  "CH05_EVI_BFBB" = "CH05",
  "CH06_EVI_BFBB" = "CH06"
)

```


```{r Data import and set-up}

bpm_file <- "/Volumes/Alice/2_POULTRY_PROCESSING/Analysis/all_samples_bpm.tsv"
out_dir  <- "/Volumes/Alice/2_POULTRY_PROCESSING/Analysis/figures_read"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# Map group -> enrichment type
enrich_map <- c(
  "ASC"      = "No enrichment",
  "EVI"      = "No enrichment",
  "ASC_RVS"  = "BPW + RVS enrichment",
  "EVI_RVS"  = "BPW + RVS enrichment",
  "ASC_BFBB" = "BF Bolton Base + CAT",
  "EVI_BFBB" = "BF Bolton Base + CAT"
)

add_factors <- function(d) {
  d %>% mutate(
    group           = factor(group, levels = c("ASC", "ASC_RVS", "ASC_BFBB",
                                               "EVI", "EVI_RVS", "EVI_BFBB")),
    enrichment_type = factor(enrich_map[as.character(group)],
                             levels = names(enrichment_labels)),
    sample_label    = factor(sample_labels[sample],
                             levels = unique(sample_labels[unique(sample[order(group, sample)])]))
  )
}

# Full dataset (all superkingdoms) - used for the host proportion chart
df_raw <- read.delim(bpm_file, header = TRUE, sep = "\t", stringsAsFactors = FALSE) %>%
  add_factors()

# Bacteria-only dataset - used for all composition plots
# If superkingdom column is present (requires re-running kraken2_bpm.py), filter by it.
# Otherwise fall back to name-based exclusion of known eukaryotes.
if ("superkingdom" %in% names(df_raw)) {
  df <- df_raw %>% filter(superkingdom == "Bacteria")
} else {
  message("NOTE: 'superkingdom' column not found in TSV. ",
          "Re-run kraken2_bpm.py to enable automatic eukaryote filtering. ",
          "Using name-based fallback filter for now.")
  df <- df_raw %>%
    filter(!grepl(paste0("^Homo|Hominidae|Homininae|Hominini|Primates|",
                         "^Gallus|Phasianidae|Galliformes|Aves|",
                         "^Meleagris|^Sus |^Bos |^Felis|^Canis"),
                  taxon))
}

# Per-sample totals (denominators for classified/unclassified chart)
# Drawn from df_raw so total_bases reflects all sequenced bases
sample_totals <- df_raw %>%
  filter(tax_level == "genus") %>%
  group_by(enrichment_type, group, sample, sample_label) %>%
  summarise(
    total_bases      = first(total_bases),
    classified_bases = first(classified_bases),
    .groups = "drop"
  )

palette_bar <- c("#1F78B4", "#33A02C", "#E31A1C", "#FF7F00",
                 "#6A3D9A", "#B15928", "#A6CEE3", "#B2DF8A",
                 "#FB9A99", "#FDBF6F", "#CAB2D6", "#FFFF99")

genus_colors_fixed <- c(
  "Campylobacter"  = "#E31A1C",
  "Bacteroides"    = "#33A02C",
  "Lactobacillus"  = "#FF7F00",
  "Pseudomonas"    = "#FDBF6F",
  "Veillonella"    = "#B2DF8A",
  "Tannerella"     = "#A6CEE3",
  "Escherichia"    = "#B15928",
  "Salmonella"     = "#B15928",
  "Comamonas"      = "#6A3D9A",
  "Klebsiella"     = "#1F78B4",
  "Enterobacter"   = "#FB9A99",
  "Proteus"        = "#CAB2D6",
  "Hafnia"         = "#FFFF99"
)

family_colors_fixed <- c(
  "Selenomonadaceae"   = "#1F78B4",
  "Bacteroidaceae"     = "#33A02C",
  "Campylobacteraceae" = "#E31A1C",
  "Lactobacillaceae"   = "#FF7F00",
  "Comamonadaceae"     = "#6A3D9A",
  "Enterobacteriaceae" = "#B15928",
  "Tannerellaceae"     = "#A6CEE3",
  "Veillonellaceae"    = "#B2DF8A",
  "Pasteurellaceae"    = "#FB9A99",
  "Pseudomonadaceae"   = "#FDBF6F",
  "Morganellaceae"     = "#CAB2D6",
  "Gallionellaceae"    = "#FFFF99"
)

build_fill_vals_main <- function(taxa, fixed_colors = NULL, pal = palette_bar) {
  out <- setNames(character(length(taxa)), taxa)
  if (!is.null(fixed_colors)) {
    matched    <- intersect(taxa, names(fixed_colors))
    out[matched] <- fixed_colors[matched]
  }
  unmatched <- taxa[out == ""]
  if (length(unmatched) > 0) {
    # Reserve ALL fixed colours from the fallback pool (not just those assigned
    # this call) so unmatched genera get the same colour regardless of which
    # fixed genera happen to appear in the top 12 for this enrichment/level.
    reserved  <- if (!is.null(fixed_colors)) unique(unname(fixed_colors)) else character(0)
    already   <- unique(c(out[out != ""], reserved))
    avail     <- pal[!pal %in% already]
    if (length(avail) < length(unmatched))
      avail <- colorRampPalette(pal)(length(unmatched))
    out[unmatched] <- avail[seq_len(length(unmatched))]
  }
  out
}

```


```{r Plots}

tax_level_labels <- c(genus = "Genus", family = "Family", species = "Species")

for (tax_lvl in c("genus", "family", "species")) {
for (enrich in levels(df$enrichment_type)) {

  enrich_df <- df %>% filter(enrichment_type == enrich, tax_level == tax_lvl)

  # Skip if no data for this combination
  if (nrow(enrich_df) == 0) {
    message("Skipping (no data): ", enrich, " | ", tax_lvl)
    next
  }

  evi_samples <- sort(unique(enrich_df$sample[grepl("^EVI", as.character(enrich_df$group))]))
  asc_samples <- sort(unique(enrich_df$sample[grepl("^ASC", as.character(enrich_df$group))]))
  ordered_labels <- unique(sample_labels[c(evi_samples, asc_samples)])

  enrich_df <- enrich_df %>%
    mutate(
      sample_label = factor(sample_labels[sample], levels = ordered_labels),
      sample_type  = factor(if_else(grepl("^EVI", as.character(group)), "Post-EVI", "Post-ASC"),
                            levels = c("Post-EVI", "Post-ASC"))
    ) %>%
    droplevels()

  # Skip if sample_type has no levels after droplevels (all filtered out)
  if (nlevels(enrich_df$sample_type) == 0 || nrow(enrich_df) == 0) {
    message("Skipping (empty after filter): ", enrich, " | ", tax_lvl)
    next
  }

  enrich_label <- enrichment_labels[enrich]
  lvl_label    <- tax_level_labels[tax_lvl]
  file_slug    <- paste0(gsub("[^A-Za-z0-9]", "_", enrich), "_", tax_lvl)

  # Top 12 taxa for this enrichment type × taxonomic level
  top12 <- enrich_df %>%
    group_by(taxon) %>%
    summarise(mean_ra = mean(rel_abund), .groups = "drop") %>%
    slice_max(mean_ra, n = 12) %>%
    arrange(desc(mean_ra)) %>%
    pull(taxon)

  # Rank-based colour assignment: most abundant taxon gets palette_bar[1], etc.
  stack_colours <- c(setNames(palette_bar[seq_len(length(top12))], top12), "Other" = "#CCCCCC")

  # Stacked bar
  other_df <- enrich_df %>%
    group_by(group, sample, sample_label) %>%
    summarise(
      rel_abund = 100 - sum(rel_abund[taxon %in% top12]),
      .groups   = "drop"
    ) %>%
    mutate(taxon = "Other")

  sample_type_map <- enrich_df %>% distinct(sample, sample_type)

  stack_df <- enrich_df %>%
    filter(taxon %in% top12) %>%
    select(group, sample, sample_label, sample_type, taxon, rel_abund) %>%
    bind_rows(other_df %>% left_join(sample_type_map, by = "sample")) %>%
    mutate(taxon = fct_reorder(taxon, rel_abund, .fun = mean))

  p_stack <- ggplot(stack_df, aes(x = sample_label, y = rel_abund, fill = taxon)) +
    geom_col(width = 0.7, colour = "white", linewidth = 0.3) +
    scale_fill_manual(values = stack_colours) +
    scale_y_continuous(labels = function(x) paste0(x, "%"),
                       expand = expansion(mult = c(0, 0.02))) +
    facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +
    labs(
      title    = paste0(enrich_label, " - ", lvl_label, "-level composition"),
      subtitle = paste0("Top 12 ", tolower(lvl_label), "s by mean relative abundance across samples in this group"),
      x        = NULL,
      y        = "Relative abundance (% of classified bases)",
      fill     = lvl_label
    ) +
    theme_bw(base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(size = 10, colour = "grey40"),
      axis.text.x   = element_text(angle = 45, hjust = 1),
      legend.text   = element_text(size = 12),
      strip.text    = element_text(face = "bold", size = 12),
      panel.spacing = unit(0.8, "lines")
    )

  # Classified vs Unclassified (same for all tax levels - use genus-level totals)
  class_df <- sample_totals %>%
    filter(enrichment_type == enrich) %>%
    left_join(enrich_df %>% distinct(sample, sample_type), by = "sample") %>%
    filter(!is.na(sample_type)) %>%   # drop samples absent from enrich_df
    mutate(
      sample_label = factor(sample_labels[sample], levels = levels(enrich_df$sample_label)),
      Classified   = classified_bases / total_bases * 100,
      Unclassified = 100 - Classified
    ) %>%
    pivot_longer(cols = c(Classified, Unclassified),
                 names_to = "status", values_to = "pct")

  p_class <- ggplot(class_df, aes(x = sample_label, y = pct, fill = status)) +
    geom_col(width = 0.6, colour = "white", linewidth = 0.4) +
    scale_fill_manual(values = c("Classified" = "#1F78B4", "Unclassified" = "#B2DF8A")) +
    scale_y_continuous(labels = function(x) paste0(x, "%"),
                       expand = expansion(mult = c(0, 0.02))) +
    facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +
    labs(
      title = paste0(enrich_label, " (", lvl_label, ") - Classified vs Unclassified bases"),
      x     = NULL,
      y     = "% of total bases",
      fill  = NULL
    ) +
    theme_bw(base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold", size = 13),
      axis.text.x   = element_text(angle = 45, hjust = 1),
      strip.text    = element_text(face = "bold", size = 12),
      panel.spacing = unit(0.8, "lines")
    )

  # Combined panel
  p_combined <- p_stack / p_class +
    plot_annotation(
      title = paste0("Read level bacterial composition - ", enrich_label, " (", lvl_label, " level)"),
      theme = theme(plot.title = element_text(face = "bold", size = 15, hjust = 0.5))
    ) +
    plot_layout(heights = c(2, 1))

  fname <- file.path(out_dir, paste0(file_slug, "_bpm_combined.png"))
  ggsave(fname, p_combined, width = 12, height = 10, dpi = 600)
  message("Saved: ", basename(fname))

  fname2 <- file.path(out_dir, paste0(file_slug, "_bpm_stacked.png"))
  ggsave(fname2, p_stack, width = 10, height = 7, dpi = 600)
  message("Saved: ", basename(fname2))

  # Extra unfiltered plot for no-enrichment group only
  if (enrich == "No enrichment") {

    all_taxa <- enrich_df %>%
      group_by(taxon) %>%
      summarise(mean_ra = mean(rel_abund), .groups = "drop") %>%
      arrange(desc(mean_ra)) %>%
      pull(taxon)

    n_taxa      <- length(all_taxa)
    palette_all <- colorRampPalette(palette_bar)(n_taxa)
    all_colours <- setNames(palette_all, all_taxa)

    all_stack_df <- enrich_df %>%
      filter(rel_abund > 0) %>%
      mutate(taxon = fct_reorder(taxon, rel_abund, .fun = mean))

    p_all <- ggplot(all_stack_df, aes(x = sample_label, y = rel_abund, fill = taxon)) +
      geom_col(width = 0.7, colour = "white", linewidth = 0.2) +
      scale_fill_manual(values = all_colours) +
      scale_y_continuous(labels = function(x) paste0(x, "%"),
                         expand = expansion(mult = c(0, 0.02))) +
      facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +
      labs(
        title    = paste0(enrich_label, " - All ", tolower(lvl_label), "s (unfiltered)"),
        subtitle = paste0(n_taxa, " ", tolower(lvl_label), "s detected across samples"),
        x        = NULL,
        y        = "Relative abundance (% of classified bases)",
        fill     = lvl_label
      ) +
      theme_bw(base_size = 12) +
      theme(
        plot.title      = element_text(face = "bold", size = 13),
        plot.subtitle   = element_text(size = 10, colour = "grey40"),
        axis.text.x     = element_text(angle = 45, hjust = 1),
        legend.position = "none",
        strip.text      = element_text(face = "bold", size = 11),
        panel.spacing   = unit(0.8, "lines")
      )

    fname3 <- file.path(out_dir, paste0("No_enrichment_bpm_all_", tax_lvl, ".png"))
    ggsave(fname3, p_all, width = 12, height = 8, dpi = 600)
    message("Saved: ", basename(fname3))
  }

} # end enrich loop
} # end tax_lvl loop

message("\nAll done. Figures saved to: ", out_dir)

```


---

# Host DNA Proportion - Bacteria vs Eukaryota vs Unclassified

Proportion of total sequenced bases attributable to host/eukaryotic DNA (e.g. *Gallus gallus*), bacterial reads, other classified (Archaea, Viruses), and unclassified reads. Calculated per sample at genus level.

```{r Host proportion chart, fig.height=6, fig.width=12}

if (!"superkingdom" %in% names(df_raw)) {
  message("NOTE: 'superkingdom' column missing - host proportion chart requires re-running kraken2_bpm.py. Skipping.")
  knitr::knit_exit()
}

host_df <- df_raw %>%
  filter(tax_level == "genus") %>%
  group_by(enrichment_type, group, sample) %>%
  summarise(
    total_bases     = first(total_bases),
    bacteria_bases  = sum(bases[superkingdom == "Bacteria"],                          na.rm = TRUE),
    gallus_bases    = sum(bases[taxon == "Gallus"],                                   na.rm = TRUE),
    other_euk_bases = sum(bases[superkingdom == "Eukaryota" & taxon != "Gallus"],     na.rm = TRUE),
    other_bases     = sum(bases[!superkingdom %in% c("Bacteria", "Eukaryota", "Unknown")],
                          na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    unclassified_bases = total_bases - bacteria_bases - gallus_bases -
                         other_euk_bases - other_bases,
    `Bacteria`         = bacteria_bases    / total_bases * 100,
    `Gallus gallus`    = gallus_bases      / total_bases * 100,
    `Other Eukaryota`  = other_euk_bases   / total_bases * 100,
    `Other classified` = other_bases       / total_bases * 100,
    `Unclassified`     = unclassified_bases / total_bases * 100,
    sample_label = factor(sample_labels[sample],
                          levels = unique(sample_labels[unique(sample[order(group, sample)])])),
    sample_type  = factor(if_else(grepl("^EVI", as.character(group)), "Post-EVI", "Post-ASC"),
                          levels = c("Post-EVI", "Post-ASC"))
  ) %>%
  pivot_longer(
    cols      = c(`Bacteria`, `Gallus gallus`, `Other Eukaryota`,
                  `Other classified`, `Unclassified`),
    names_to  = "category",
    values_to = "pct"
  ) %>%
  mutate(category = factor(category,
                           levels = c("Gallus gallus", "Other Eukaryota",
                                      "Bacteria", "Other classified", "Unclassified")))

host_colours <- c(
  "Gallus gallus"    = "#FF7F00",
  "Other Eukaryota"  = "#E31A1C",
  "Bacteria"         = "#1F78B4",
  "Other classified" = "#6A3D9A",
  "Unclassified"     = "#B2DF8A"
)

# Combined 3-row × 2-column plot: rows = enrichment type, columns = Post-EVI / Post-ASC
host_plot_df <- host_df %>%
  mutate(
    sample_label    = factor(sample_labels[sample], levels = paste0("CH0", 1:6)),
    sample_type     = factor(if_else(grepl("^EVI", as.character(group)),
                                     "Post-EVI", "Post-ASC"),
                             levels = c("Post-EVI", "Post-ASC")),
    enrichment_type = factor(as.character(enrichment_type),
                             levels = levels(df_raw$enrichment_type))
  )

p_host_all <- ggplot(host_plot_df, aes(x = sample_label, y = pct, fill = category)) +
  geom_col(width = 0.7, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = host_colours) +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.02))) +
  facet_grid(enrichment_type ~ sample_type, scales = "free_x", space = "free_x") +
  labs(
    title    = "Host vs Bacterial vs Unclassified bases - All enrichment types",
    subtitle = "Proportion of total sequenced bases; Gallus gallus shown separately from other Eukaryota",
    x        = NULL,
    y        = "% of total bases",
    fill     = NULL
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, colour = "grey40"),
    axis.text.x   = element_text(angle = 45, hjust = 1),
    strip.text.x  = element_text(face = "bold", size = 12),
    strip.text.y  = element_text(face = "bold", size = 10),
    panel.spacing = unit(0.6, "lines"),
    legend.position = "right"
  )

print(p_host_all)
ggsave(file.path(out_dir, "host_proportion_all_enrichments.png"),
       p_host_all, width = 12, height = 10, dpi = 600)
message("Saved: host_proportion_all_enrichments.png")

```


---

# Diversity Analysis - Alpha, Beta & PERMANOVA

```{r Diversity helper functions}

# ── Alpha diversity ────────────────────────────────────────────────────────────
# Builds Shannon, Simpson, and observed richness from base counts per sample,
# then plots boxplots + jitter with Wilcoxon p-values per metric.
# data_sub : a data.frame with columns sample, taxon, bases, group (factor)
# label_a/b: group labels to compare (Post-EVI vs Post-ASC)

plot_alpha_bpm <- function(data_sub, label_a, label_b, title_str,
                            col_a = "#377EB8", col_b = "#E41A1C",
                            save_path = NULL) {

  make_tbl <- function(df, label) {
    wide <- df %>%
      filter(as.character(group) == label) %>%
      select(sample, taxon, bases) %>%
      pivot_wider(names_from = taxon, values_from = bases, values_fill = 0L)
    mat <- wide %>% column_to_rownames("sample") %>% as.matrix()
    tibble(
      sample   = rownames(mat),
      group    = label,
      Shannon  = diversity(mat, index = "shannon"),
      Simpson  = diversity(mat, index = "simpson"),
      Richness = rowSums(mat > 0)
    )
  }

  alpha_df <- bind_rows(
    make_tbl(data_sub, label_a),
    make_tbl(data_sub, label_b)
  ) %>%
    mutate(group = factor(group, levels = c(label_a, label_b)))

  cols <- setNames(c(col_a, col_b), c(label_a, label_b))

  alpha_long <- alpha_df %>%
    pivot_longer(c(Shannon, Simpson, Richness),
                 names_to = "Metric", values_to = "Value") %>%
    mutate(Metric = factor(Metric, levels = c("Shannon", "Simpson", "Richness")))

  pvals <- alpha_long %>%
    group_by(Metric) %>%
    summarise(
      p = tryCatch(
        wilcox.test(Value[group == label_a], Value[group == label_b],
                    exact = FALSE)$p.value,
        error = function(e) NA_real_),
      .groups = "drop"
    ) %>%
    mutate(label = case_when(
      p < 0.001   ~ "***",
      p < 0.01    ~ "**",
      p < 0.05    ~ "*",
      !is.na(p)   ~ paste0("p = ", round(p, 3)),
      TRUE        ~ ""
    ))

  p <- ggplot(alpha_long, aes(x = group, y = Value, fill = group, colour = group)) +
    geom_boxplot(width = 0.45, alpha = 0.6, outlier.shape = NA, colour = "grey30") +
    geom_jitter(width = 0.1, size = 2.2, alpha = 0.85) +
    geom_text(data = pvals, aes(x = 1.5, y = Inf, label = label),
              inherit.aes = FALSE, vjust = 1.6, size = 3.8, colour = "black") +
    facet_wrap(~ Metric, scales = "free_y", ncol = 3) +
    scale_fill_manual(values = cols) +
    scale_colour_manual(values = cols) +
    labs(title    = title_str,
         subtitle = "Wilcoxon rank-sum: *** p < 0.001, ** < 0.01, * < 0.05\nAlpha diversity calculated from base counts assigned per taxon",
         x = NULL, y = NULL) +
    theme_bw(base_size = 13) +
    theme(legend.position  = "none",
          axis.text        = element_text(size = 11),
          strip.text       = element_text(face = "bold", size = 13),
          plot.title       = element_text(face = "bold", size = 14),
          plot.subtitle    = element_text(size = 10, colour = "grey40"),
          strip.background = element_rect(fill = "lightgrey"))

  print(p)
  if (!is.null(save_path)) ggsave(save_path, p, width = 10, height = 5, dpi = 600)
  invisible(p)
}


# ── Beta diversity - Bray-Curtis PCoA + PERMANOVA ─────────────────────────────
# Uses relative abundance for Bray-Curtis dissimilarities.
# Plots PCoA with 95% ellipses; PERMANOVA R² and p-value shown in subtitle.

plot_beta_pcoa_bpm <- function(data_sub, label_a, label_b, title_str,
                                col_a = "#377EB8", col_b = "#E41A1C",
                                save_path = NULL) {

  wide <- data_sub %>%
    select(sample, taxon, rel_abund) %>%
    pivot_wider(names_from = taxon, values_from = rel_abund, values_fill = 0)
  mat <- wide %>% column_to_rownames("sample") %>% as.matrix()

  # Group vector aligned with matrix row order
  sample_grp <- data_sub %>%
    distinct(sample, group) %>%
    mutate(group = as.character(group))
  group_vec <- sample_grp$group[match(rownames(mat), sample_grp$sample)]

  dist_mat <- vegdist(mat, method = "bray")
  pcoa     <- cmdscale(dist_mat, k = 2, eig = TRUE)
  var_expl <- abs(pcoa$eig) / sum(abs(pcoa$eig)) * 100

  coords <- as.data.frame(pcoa$points) %>%
    setNames(c("PC1", "PC2")) %>%
    rownames_to_column("sample") %>%
    mutate(group = factor(group_vec, levels = c(label_a, label_b)))

  perm <- adonis2(dist_mat ~ group_vec, permutations = 999)
  r2   <- round(perm$R2[1], 3)
  p_v  <- perm$`Pr(>F)`[1]

  p <- ggplot(coords, aes(x = PC1, y = PC2, colour = group)) +
    geom_point(size = 3.5, alpha = 0.85) +
    stat_ellipse(level = 0.95, linewidth = 0.8) +
    scale_colour_manual(values = setNames(c(col_a, col_b), c(label_a, label_b))) +
    guides(colour = guide_legend(title = NULL, override.aes = list(size = 4))) +
    labs(title    = title_str,
         subtitle = paste0("Bray-Curtis PCoA | PERMANOVA: R² = ", r2,
                           ", p = ", round(p_v, 3), " (999 permutations)"),
         x = paste0("PC1 (", round(var_expl[1], 1), "% variance)"),
         y = paste0("PC2 (", round(var_expl[2], 1), "% variance)")) +
    theme_bw(base_size = 13) +
    theme(plot.title      = element_text(face = "bold", size = 14),
          plot.subtitle   = element_text(size = 11, colour = "grey40"),
          axis.text       = element_text(size = 12),
          axis.title      = element_text(size = 13),
          legend.text     = element_text(size = 12),
          legend.position = "bottom")

  print(p)
  if (!is.null(save_path)) ggsave(save_path, p, width = 8, height = 6, dpi = 600)
  invisible(p)
}


# ── Per-taxon Wilcoxon tests (Post-EVI vs Post-ASC) ───────────────────────────
# Returns a table with mean rel_abund per group, raw p-value, BH-adjusted p-value,
# and significance stars. Only taxa with mean >= min_mean in >= min_samples tested.

run_taxon_tests_bpm <- function(data_sub, label_a, label_b,
                                 min_mean = 0.1, min_samples = 2) {
  combined <- bind_rows(
    data_sub %>% filter(group == label_a) %>%
      select(sample, taxon, rel_abund, bpm) %>% mutate(grp = label_a),
    data_sub %>% filter(group == label_b) %>%
      select(sample, taxon, rel_abund, bpm) %>% mutate(grp = label_b)
  ) %>%
    complete(nesting(sample, grp), taxon, fill = list(rel_abund = 0, bpm = 0))

  taxa_keep <- combined %>%
    group_by(taxon) %>%
    summarise(mean_abund  = mean(rel_abund),
              n_nonzero   = sum(rel_abund > 0),
              .groups     = "drop") %>%
    filter(mean_abund >= min_mean, n_nonzero >= min_samples) %>%
    pull(taxon)

  results <- combined %>%
    filter(taxon %in% taxa_keep) %>%
    group_by(taxon) %>%
    summarise(
      mean_pct_a = mean(rel_abund[grp == label_a]),
      mean_pct_b = mean(rel_abund[grp == label_b]),
      p_value    = tryCatch(
        wilcox.test(rel_abund[grp == label_a],
                    rel_abund[grp == label_b],
                    exact = FALSE)$p.value,
        error = function(e) NA_real_),
      .groups = "drop"
    ) %>%
    mutate(
      p_adj     = p.adjust(p_value, method = "BH"),
      sig_unadj = case_when(p_value < 0.001 ~ "***", p_value < 0.01 ~ "**",
                             p_value < 0.05  ~ "*",   TRUE           ~ "ns"),
      sig       = case_when(p_adj   < 0.001 ~ "***", p_adj   < 0.01 ~ "**",
                             p_adj   < 0.05  ~ "*",   TRUE           ~ "ns")
    ) %>%
    arrange(p_value)

  names(results)[names(results) == "mean_pct_a"] <- paste0("Mean ", label_a, " (%)")
  names(results)[names(results) == "mean_pct_b"] <- paste0("Mean ", label_b, " (%)")
  results
}


# Format results table for display / PDF output
format_stats_display_bpm <- function(results, adjusted = TRUE) {
  pct_cols <- grep("\\(%\\)$", names(results), value = TRUE)
  df <- results
  df[pct_cols]  <- lapply(df[pct_cols], round, 2)
  df$p_value    <- formatC(df$p_value, format = "e", digits = 2)

  if (adjusted) {
    df$p_adj <- formatC(df$p_adj, format = "e", digits = 2)
    df <- df %>% select(taxon, all_of(pct_cols), p_value, p_adj, sig)
    names(df)[names(df) == "sig"]   <- "Sig. (BH)"
    names(df)[names(df) == "p_adj"] <- "p-adj (BH)"
  } else {
    df <- df %>% select(taxon, all_of(pct_cols), p_value, sig_unadj)
    names(df)[names(df) == "sig_unadj"] <- "Sig."
  }
  names(df)[names(df) == "taxon"]   <- "Taxon"
  names(df)[names(df) == "p_value"] <- "p-value"
  df
}


# Draw one table page to the active PDF device
draw_stats_table_bpm <- function(df, title_str) {
  sig_col  <- if ("Sig. (BH)" %in% names(df)) "Sig. (BH)" else "Sig."
  sig_vals <- df[[sig_col]]
  n_rows   <- nrow(df)
  n_cols   <- ncol(df)

  row_fill <- case_when(sig_vals == "***" ~ "#fd7e14",
                        sig_vals == "**"  ~ "#ffc107",
                        sig_vals == "*"   ~ "#fff3cd",
                        TRUE              ~ "#ffffff")
  fill_mat <- matrix(rep(row_fill, n_cols), nrow = n_rows, ncol = n_cols)

  tbl <- tableGrob(
    df, rows = NULL,
    theme = ttheme_default(
      core    = list(bg_params = list(fill = fill_mat, col = "#dee2e6"),
                     fg_params = list(fontsize = 7)),
      colhead = list(bg_params = list(fill = "#495057", col = "#dee2e6"),
                     fg_params = list(fontsize = 7.5, col = "white",
                                      fontface = "bold"))
    )
  )

  title_grob <- grid::textGrob(title_str,
                                gp    = grid::gpar(fontsize = 10, fontface = "bold"),
                                hjust = 0.5, x = 0.5)
  combined <- arrangeGrob(title_grob, tbl, nrow = 2, heights = c(0.06, 0.94))
  grid::grid.draw(combined)
}


# Save a 2-page PDF: page 1 = unadjusted p-values, page 2 = BH-adjusted
save_stats_pdf_bpm <- function(results, title_str, path) {
  unadj_df <- format_stats_display_bpm(results, adjusted = FALSE)
  adj_df   <- format_stats_display_bpm(results, adjusted = TRUE)
  page_h   <- max(5, nrow(results) * 0.26 + 1.8)
  pdf(path, width = 16, height = page_h)
  draw_stats_table_bpm(unadj_df, paste0(title_str, "  |  Unadjusted p-values"))
  grid::grid.newpage()
  draw_stats_table_bpm(adj_df,   paste0(title_str, "  |  BH-adjusted p-values (FDR)"))
  dev.off()
  message("Saved: ", basename(path))
  invisible(path)
}

```


## Alpha Diversity - Post-EVI vs Post-ASC

Shannon entropy, Simpson index, and observed richness calculated from base counts assigned to each taxon per sample. Wilcoxon rank-sum p-value shown above each panel.

```{r Alpha diversity, fig.height=5, fig.width=10}

tax_level_labels <- c(genus = "Genus", family = "Family", species = "Species")

for (enrich in levels(df$enrichment_type)) {
  for (tax_lvl in c("genus", "family", "species")) {

    div_df <- df %>%
      filter(enrichment_type == enrich, tax_level == tax_lvl) %>%
      mutate(group = as.character(group),
             sample_type = if_else(grepl("^EVI", group), "Post-EVI", "Post-ASC"))

    # Need at least 2 samples per group
    n_evi <- n_distinct(div_df$sample[div_df$sample_type == "Post-EVI"])
    n_asc <- n_distinct(div_df$sample[div_df$sample_type == "Post-ASC"])
    if (n_evi < 2 || n_asc < 2) next

    enrich_lbl <- enrichment_labels[enrich]
    lvl_lbl    <- tax_level_labels[tax_lvl]
    file_slug  <- paste0(gsub("[^A-Za-z0-9]", "_", enrich), "_", tax_lvl)

    plot_alpha_bpm(
      data_sub   = div_df %>% select(-group) %>% rename(group = sample_type),
      label_a    = "Post-EVI",
      label_b    = "Post-ASC",
      title_str  = paste0("Alpha Diversity (", lvl_lbl, " level) - ", enrich_lbl),
      save_path  = file.path(out_dir, paste0(file_slug, "_alpha_diversity.png"))
    )
  }
}

```


## Beta Diversity - Bray-Curtis PCoA + PERMANOVA

Principal coordinates analysis of Bray-Curtis dissimilarities (relative abundance). Ellipses show 95% confidence intervals. PERMANOVA tests whether group membership explains community composition.

```{r Beta diversity PERMANOVA, fig.height=6, fig.width=8}

for (enrich in levels(df$enrichment_type)) {
  for (tax_lvl in c("genus", "family", "species")) {

    div_df <- df %>%
      filter(enrichment_type == enrich, tax_level == tax_lvl) %>%
      mutate(sample_type = if_else(grepl("^EVI", as.character(group)),
                                   "Post-EVI", "Post-ASC"))

    # Need at least 2 samples per group for PCoA
    n_evi <- n_distinct(div_df$sample[div_df$sample_type == "Post-EVI"])
    n_asc <- n_distinct(div_df$sample[div_df$sample_type == "Post-ASC"])
    if (n_evi < 2 || n_asc < 2) next

    enrich_lbl <- enrichment_labels[enrich]
    lvl_lbl    <- tax_level_labels[tax_lvl]
    file_slug  <- paste0(gsub("[^A-Za-z0-9]", "_", enrich), "_", tax_lvl)

    plot_beta_pcoa_bpm(
      data_sub  = div_df %>% select(-group) %>% rename(group = sample_type),
      label_a   = "Post-EVI",
      label_b   = "Post-ASC",
      title_str = paste0("Beta Diversity - Bray-Curtis PCoA (", lvl_lbl, " level)\n", enrich_lbl),
      save_path = file.path(out_dir, paste0(file_slug, "_beta_pcoa.png"))
    )
  }
}

```


## PERMANOVA Summary Table

```{r PERMANOVA summary table}

perm_results <- list()

for (enrich in levels(df$enrichment_type)) {
  for (tax_lvl in c("genus", "family", "species")) {

    div_df <- df %>%
      filter(enrichment_type == enrich, tax_level == tax_lvl) %>%
      mutate(sample_type = if_else(grepl("^EVI", as.character(group)),
                                   "Post-EVI", "Post-ASC"))

    n_evi <- n_distinct(div_df$sample[div_df$sample_type == "Post-EVI"])
    n_asc <- n_distinct(div_df$sample[div_df$sample_type == "Post-ASC"])
    if (n_evi < 2 || n_asc < 2) next

    wide <- div_df %>%
      select(sample, taxon, rel_abund) %>%
      pivot_wider(names_from = taxon, values_from = rel_abund, values_fill = 0)
    mat <- wide %>% column_to_rownames("sample") %>% as.matrix()

    sample_grp <- div_df %>% distinct(sample, sample_type)
    group_vec  <- sample_grp$sample_type[match(rownames(mat), sample_grp$sample)]

    dist_mat <- vegdist(mat, method = "bray")
    perm     <- adonis2(dist_mat ~ group_vec, permutations = 999)

    perm_results[[paste(enrich, tax_lvl, sep = " | ")]] <- tibble(
      Enrichment     = enrichment_labels[enrich],
      `Tax. level`   = tax_level_labels[tax_lvl],
      `n Post-EVI`   = n_evi,
      `n Post-ASC`   = n_asc,
      `R²`           = round(perm$R2[1], 3),
      `p-value`      = round(perm$`Pr(>F)`[1], 3),
      Significant    = ifelse(perm$`Pr(>F)`[1] < 0.05, "Yes", "No")
    )
  }
}

perm_table <- bind_rows(perm_results)
print(perm_table)

# Save as TSV
write.table(perm_table,
            file.path(out_dir, "permanova_summary.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)
message("PERMANOVA summary saved to: permanova_summary.tsv")

```


---

## Per-Taxon Statistical Tests - Post-EVI vs Post-ASC

Wilcoxon rank-sum test (two-sided) per taxon at genus, family, and species level, run separately within each enrichment type. Only taxa present with mean relative abundance ≥ 0.1% in ≥ 2 samples are tested. Page 1 of each PDF shows unadjusted p-values; page 2 shows Benjamini-Hochberg FDR-corrected p-values. Rows highlighted orange (***), yellow (**), or pale yellow (*).

```{r Statistical tests per taxon}

tax_level_labels <- c(genus = "Genus", family = "Family", species = "Species")

for (tax_lvl in c("genus", "family", "species")) {
  for (enrich in levels(df$enrichment_type)) {

    test_df <- df %>%
      filter(enrichment_type == enrich, tax_level == tax_lvl) %>%
      mutate(group = if_else(grepl("^EVI", as.character(group)),
                             "Post-EVI", "Post-ASC"))

    n_evi <- n_distinct(test_df$sample[test_df$group == "Post-EVI"])
    n_asc <- n_distinct(test_df$sample[test_df$group == "Post-ASC"])
    if (n_evi < 2 || n_asc < 2) next

    results <- run_taxon_tests_bpm(test_df, "Post-EVI", "Post-ASC")

    if (nrow(results) == 0) {
      message("No taxa passed filters for: ", enrich, " | ", tax_lvl)
      next
    }

    enrich_lbl <- enrichment_labels[enrich]
    lvl_lbl    <- tax_level_labels[tax_lvl]
    file_slug  <- paste0(gsub("[^A-Za-z0-9]", "_", enrich), "_", tax_lvl)
    title_str  <- paste0("Post-EVI vs Post-ASC - ", lvl_lbl, " level | ", enrich_lbl)

    pdf_path <- file.path(out_dir, paste0("stats_", file_slug, "_EVI_vs_ASC.pdf"))
    save_stats_pdf_bpm(results, title_str, pdf_path)
  }
}

message("\nAll statistical PDFs saved to: ", out_dir)

```




---

# Species Presence and Abundance - Escherichia, Salmonella, Campylobacter

```{r species_helper_functions_bpm}

rename_vec <- function(x, labels) {
  mapped <- labels[x]
  ifelse(is.na(mapped), x, mapped)
}

build_fill_vals <- function(taxa, fixed_colors = NULL,
                             pal = brewer.pal(8, "Dark2")) {
  out <- setNames(character(length(taxa)), taxa)
  if (!is.null(fixed_colors)) {
    matched <- intersect(taxa, names(fixed_colors))
    out[matched] <- fixed_colors[matched]
  }
  unmatched <- taxa[out == ""]
  if (length(unmatched) > 0) {
    extra <- colorRampPalette(pal)(length(unmatched))
    out[unmatched] <- extra
  }
  out
}

species_colors_escherichia <- c(
  "Escherichia coli"       = "#E31A1C",  # most abundant/important - prominent red
  "Escherichia fergusonii" = "#FF7F00",
  "Escherichia albertii"   = "#33A02C",
  "Escherichia marmotae"   = "#6A3D9A",
  "Escherichia ruysiae"    = "#1F78B4",
  "Escherichia dysenteriae" = "#B15928",
  "Escherichia vulneris"   = "#A6CEE3",
  "Escherichia hermannii"  = "#B2DF8A"
)

species_colors_salmonella <- c(
  "Salmonella enterica" = "#E41A1C",
  "Salmonella bongori"  = "#377EB8"
)

# Salmonella subspecies and common serovars
# Kraken2 may classify these at tax_level "subspecies" or below species
species_colors_salmonella_serotype <- c(
  # Subspecies of S. enterica
  "Salmonella enterica subsp. enterica"    = "#E31A1C",  # most common - red
  "Salmonella enterica subsp. salamae"     = "#FF7F00",
  "Salmonella enterica subsp. arizonae"    = "#33A02C",
  "Salmonella enterica subsp. diarizonae"  = "#6A3D9A",
  "Salmonella enterica subsp. houtenae"    = "#1F78B4",
  "Salmonella enterica subsp. indica"      = "#B15928",
  # Common poultry-associated serovars (appear as subspecies/strain level)
  "Salmonella Enteritidis"                 = "#E31A1C",
  "Salmonella Typhimurium"                 = "#FF7F00",
  "Salmonella Infantis"                    = "#33A02C",
  "Salmonella Kentucky"                    = "#1F78B4",
  "Salmonella Heidelberg"                  = "#6A3D9A",
  "Salmonella Newport"                     = "#B15928",
  "Salmonella Hadar"                       = "#A6CEE3",
  "Salmonella Mbandaka"                    = "#B2DF8A",
  "Salmonella Virchow"                     = "#FB9A99",
  "Salmonella Stanley"                     = "#FDBF6F",
  "Salmonella Thompson"                    = "#CAB2D6",
  "Salmonella Agona"                       = "#FFFF99"
)

species_colors_campylobacter <- c(
  "Campylobacter jejuni"          = "#E31A1C",  # most important - prominent red
  "Campylobacter coli"            = "#1F78B4",  # second most important - blue
  "Campylobacter lari"            = "#1B9E77",
  "Campylobacter upsaliensis"     = "#D95F02",
  "Campylobacter hyointestinalis" = "#7570B3",
  "Campylobacter helveticus"      = "#E7298A",
  "Campylobacter concisus"        = "#66A61E",
  "Campylobacter fetus"           = "#E6AB02",
  "Campylobacter showae"          = "#A6761D",
  "Campylobacter ureolyticus"     = "#666666",
  "Campylobacter insulaenigrae"   = "#A6CEE3",
  "Campylobacter pinnipediorum"   = "#FB9A99",
  "Campylobacter sputorum"        = "#B2DF8A",
  "Campylobacter mucosalis"       = "#FDBF6F"
)

species_colors_bacteroides <- c(
  "Bacteroides fragilis"         = "#A6CEE3",
  "Bacteroides thetaiotaomicron" = "#1F78B4",
  "Bacteroides uniformis"        = "#B2DF8A",
  "Bacteroides ovatus"           = "#33A02C",
  "Bacteroides vulgatus"         = "#FB9A99",
  "Bacteroides dorei"            = "#E31A1C",
  "Bacteroides stercoris"        = "#FDBF6F",
  "Bacteroides cellulosilyticus" = "#FF7F00"
)

species_colors_lactobacillus <- c(
  "Lactobacillus acidophilus"  = "#8DD3C7",
  "Lactobacillus reuteri"      = "#FFFFB3",
  "Lactobacillus helveticus"   = "#BEBADA",
  "Lactobacillus sakei"        = "#FB8072",
  "Lactobacillus gasseri"      = "#80B1D3",
  "Lactobacillus johnsonii"    = "#FDB462",
  "Lactobacillus mucosae"      = "#B3DE69",
  "Lactobacillus rhamnosus"    = "#FCCDE5"
)

species_colors_pseudomonas <- c(
  "Pseudomonas aeruginosa"   = "#1B9E77",
  "Pseudomonas fluorescens"  = "#D95F02",
  "Pseudomonas putida"       = "#7570B3",
  "Pseudomonas stutzeri"     = "#E7298A",
  "Pseudomonas fragi"        = "#66A61E"
)

species_colors_veillonella <- c(
  "Veillonella parvula"  = "#E41A1C",
  "Veillonella atypica"  = "#377EB8",
  "Veillonella dispar"   = "#4DAF4A",
  "Veillonella rogosae"  = "#984EA3"
)

species_colors_tannerella <- c(
  "Tannerella forsythia" = "#66C2A5"
)

species_colors_comamonas <- c(
  "Comamonas testosteroni" = "#7FC97F",
  "Comamonas aquatica"     = "#BEAED4",
  "Comamonas kerstersii"   = "#FDC086",
  "Comamonas badia"        = "#FFFF99"
)

# Append Post-EVI/Post-ASC to avoid duplicate "CH01" labels across both groups
spp_display_labels <- setNames(
  paste0(sample_labels[names(sample_labels)], " (",
         ifelse(grepl("EVI", names(sample_labels)), "Post-EVI", "Post-ASC"), ")"),
  names(sample_labels)
)

plot_genus_linear_presence <- function(df, disp_labels, title, subtitle = NULL,
                                       layout = "side_by_side",
                                       species_colors = NULL,
                                       spp_palette = brewer.pal(8, "Dark2"),
                                       save_path = NULL) {
  if (nrow(df) == 0) { cat("No data - combined plot skipped.\n"); return(invisible(NULL)) }

  species_order <- df %>%
    group_by(name) %>% summarise(m = mean(rel_abund), .groups = "drop") %>%
    filter(m > 0) %>%
    arrange(desc(m)) %>% pull(name)

  if (length(species_order) == 0) { cat("No present species - combined plot skipped.\n"); return(invisible(NULL)) }

  lvls   <- rev(species_order)
  n_spp  <- length(species_order)
  n_samp <- length(unique(df$sample))

  if (!is.null(species_colors)) {
    spp_fill <- build_fill_vals(species_order, species_colors, pal = spp_palette)
    spp_fill <- spp_fill[names(spp_fill) != "Other"]
  } else {
    spp_cols <- if (n_spp <= 8) spp_palette[seq_len(n_spp)] else
                  colorRampPalette(spp_palette)(n_spp)
    spp_fill <- setNames(spp_cols, species_order)
  }

  bar_df <- df %>%
    mutate(
      disp_label = factor(rename_vec(sample, disp_labels),
                          levels = unique(rename_vec(names(disp_labels), disp_labels))),
      name       = factor(name, levels = lvls)
    )

  p_bar <- ggplot(bar_df, aes(x = rel_abund, y = disp_label, fill = name)) +
    geom_col(width = 0.7, colour = "white", linewidth = 0.3) +
    scale_fill_manual(values = spp_fill, name = "Species") +
    scale_x_continuous(labels = function(x) paste0(x, "%"),
                       expand = expansion(mult = c(0, 0.02))) +
    facet_grid(. ~ sample_type, scales = "free_y", space = "free_y") +
    labs(
      title    = title,
      subtitle = subtitle,
      x        = "Relative abundance (% of classified bases)",
      y        = NULL
    ) +
    theme_bw(base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(size = 10, colour = "grey40"),
      axis.text.y   = element_text(size = 10),
      legend.text   = element_text(size = 11),
      strip.text    = element_text(face = "bold", size = 12),
      panel.spacing = unit(0.8, "lines")
    )

  grid_df <- df %>%
    mutate(
      disp_label = rename_vec(sample, disp_labels),
      present    = rel_abund > 0
    ) %>%
    group_by(name, disp_label) %>%
    summarise(present = any(present), .groups = "drop")

  all_combos <- expand.grid(
    name       = species_order,
    disp_label = unique(grid_df$disp_label),
    stringsAsFactors = FALSE
  )
  stype_map <- df %>%
    mutate(disp_label = rename_vec(sample, disp_labels)) %>%
    distinct(disp_label, sample_type)
  grid_df <- all_combos %>%
    left_join(grid_df, by = c("name", "disp_label")) %>%
    left_join(stype_map, by = "disp_label") %>%
    mutate(
      present    = replace_na(present, FALSE),
      disp_label = factor(disp_label, levels = rev(unique(rename_vec(names(disp_labels), disp_labels)))),
      name       = factor(name, levels = species_order)
    )

  p_tile <- ggplot(grid_df, aes(x = name, y = disp_label, fill = present)) +
    geom_tile(colour = "white", linewidth = 0.5) +
    scale_fill_manual(values = c("TRUE" = "#2166AC", "FALSE" = "#F7F7F7"),
                      labels = c("TRUE" = "Present", "FALSE" = "Absent"),
                      name   = NULL) +
    facet_grid(sample_type ~ ., scales = "free_y", space = "free_y") +
    labs(x = NULL, y = NULL) +
    theme_bw(base_size = 11) +
    theme(
      axis.text.x     = element_text(angle = 45, hjust = 1, size = 10, face = "italic"),
      axis.text.y     = element_text(size = 10),
      strip.text      = element_text(face = "bold", size = 11),
      panel.spacing   = unit(0.6, "lines"),
      legend.position = "bottom"
    )

  if (layout == "side_by_side") {
    p_out <- p_bar | p_tile
  } else {
    p_out <- p_bar / p_tile + plot_layout(heights = c(2, 1))
  }

  print(p_out)
  if (!is.null(save_path)) {
    ggsave(save_path, p_out,
           width  = if (layout == "side_by_side") 14 else 10,
           height = if (layout == "side_by_side") max(5, n_samp * 0.5 + 2) else
                      max(7, n_samp * 0.4 + 4),
           dpi    = 600)
    message("Saved: ", basename(save_path))
  }
  invisible(p_out)
}


plot_genus_presence <- function(df, disp_labels, title, subtitle = NULL,
                                present_colour = "#2166AC",
                                save_path = NULL) {
  if (nrow(df) == 0) { cat("No data - presence plot skipped.\n"); return(invisible(NULL)) }

  species_order <- df %>%
    group_by(name) %>% summarise(m = mean(rel_abund), .groups = "drop") %>%
    filter(m > 0) %>%
    arrange(desc(m)) %>% pull(name)

  if (length(species_order) == 0) { cat("No present species - presence plot skipped.\n"); return(invisible(NULL)) }

  grid_df <- df %>%
    mutate(
      disp_label = rename_vec(sample, disp_labels),
      present    = rel_abund > 0
    ) %>%
    group_by(name, disp_label) %>%
    summarise(present = any(present), .groups = "drop")

  all_combos <- expand.grid(
    name       = species_order,
    disp_label = unique(grid_df$disp_label),
    stringsAsFactors = FALSE
  )
  stype_map <- df %>%
    mutate(disp_label = rename_vec(sample, disp_labels)) %>%
    distinct(disp_label, sample_type)
  grid_df <- all_combos %>%
    left_join(grid_df, by = c("name", "disp_label")) %>%
    left_join(stype_map, by = "disp_label") %>%
    mutate(
      present    = replace_na(present, FALSE),
      # Strip "(Post-EVI)"/"(Post-ASC)" suffix - the facet strip already shows sample_type,
      # so the suffix is redundant and causes label crowding on the x-axis
      disp_label = factor(
        sub(" \\(Post-[A-Z]+\\)$", "", as.character(disp_label)),
        levels = unique(sub(" \\(Post-[A-Z]+\\)$", "",
                            unique(rename_vec(names(disp_labels), disp_labels))))
      ),
      name       = factor(name, levels = species_order)
    )

  # Species on y-axis, samples on x-axis: tall rather than wide
  p <- ggplot(grid_df, aes(x = disp_label, y = name, fill = present)) +
    geom_tile(colour = "white", linewidth = 0.6) +
    scale_fill_manual(
      values = c("TRUE" = present_colour, "FALSE" = "#F7F7F7"),
      labels = c("TRUE" = "Present",      "FALSE" = "Absent"),
      name   = NULL
    ) +
    facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +
    labs(title = title, subtitle = subtitle, x = NULL, y = NULL) +
    theme_bw(base_size = 12) +
    theme(
      plot.title      = element_text(face = "bold", size = 13),
      plot.subtitle   = element_text(size = 10, colour = "grey40"),
      axis.text.x     = element_text(size = 11),
      axis.text.y     = element_text(size = 11, face = "italic"),
      strip.text      = element_text(face = "bold", size = 12),
      panel.spacing   = unit(0.6, "lines"),
      legend.position = "bottom"
    )

  print(p)
  if (!is.null(save_path)) {
    n_samp  <- length(unique(df$sample))
    n_taxa  <- length(species_order)
    # Scale height-per-taxon down for large datasets to avoid > 50-inch limit
    h_per   <- if (n_taxa > 80) 0.22 else if (n_taxa > 40) 0.32 else 0.5
    ggsave(save_path, p,
           width     = max(6, min(20, n_samp * 0.55 + 3)),
           height    = max(5, n_taxa * h_per + 2),
           dpi       = 600,
           limitsize = FALSE)
    message("Saved: ", basename(save_path))
  }
  invisible(p)
}

```


## Escherichia Species

```{r bpm_esch_species, fig.height=7, fig.width=14}

for (enrich in levels(df$enrichment_type)) {

  spp_df <- df %>%
    filter(tax_level == "species",
           str_detect(taxon, "^Escherichia"),
           enrichment_type == enrich) %>%
    rename(name = taxon) %>%
    mutate(
      sample_type = factor(
        if_else(grepl("^EVI", as.character(group)), "Post-EVI", "Post-ASC"),
        levels = c("Post-EVI", "Post-ASC")
      )
    )

  if (nrow(spp_df) == 0) {
    message("No Escherichia species data for: ", enrich)
    next
  }

  enrich_lbl <- enrichment_labels[enrich]
  file_slug  <- paste0("esch_", gsub("[^A-Za-z0-9]", "_", enrich))

  esch_species <- sort(unique(spp_df$name))
  esch_fill    <- build_fill_vals(esch_species, species_colors_escherichia,
                                   pal = brewer.pal(8, "Set2"))

  # Stacked bar
  bar_df <- spp_df %>%
    mutate(
      disp_label = factor(rename_vec(sample, spp_display_labels),
                          levels = unique(rename_vec(names(spp_display_labels),
                                                     spp_display_labels))),
      name       = factor(name, levels = rev(esch_species))
    )

  p_bar <- ggplot(bar_df, aes(x = disp_label, y = rel_abund, fill = name)) +
    geom_col(width = 0.7, colour = "white", linewidth = 0.3) +
    scale_fill_manual(values = esch_fill, name = "Species") +
    scale_y_continuous(labels = function(x) paste0(x, "%"),
                       expand = expansion(mult = c(0, 0.02))) +
    facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +
    labs(
      title    = paste0("Escherichia Species - ", enrich_lbl, " (Reads)"),
      subtitle = "Relative abundance of classified bases",
      x        = NULL,
      y        = "Relative abundance (%)"
    ) +
    theme_bw(base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(size = 10, colour = "grey40"),
      axis.text.x   = element_text(angle = 45, hjust = 1),
      legend.text   = element_text(size = 11),
      strip.text    = element_text(face = "bold", size = 12),
      panel.spacing = unit(0.8, "lines")
    )

  print(p_bar)
  ggsave(file.path(out_dir, paste0(file_slug, "_species_bar.png")),
         p_bar, width = 10, height = 6, dpi = 600)
  message("Saved: ", file_slug, "_species_bar.png")

  plot_genus_linear_presence(
    spp_df, spp_display_labels,
    title          = paste0("Escherichia Species Relative Abundance & Presence - ", enrich_lbl, " (Reads)"),
    save_path      = file.path(out_dir, paste0(file_slug, "_linear_presence.png"))
  )

  plot_genus_presence(
    spp_df, spp_display_labels,
    title          = paste0("Escherichia Species Presence/Absence - ", enrich_lbl, " (Reads)"),
    save_path      = file.path(out_dir, paste0(file_slug, "_presence.png"))
  )

  # Virus/bacteriophage presence associated with this enrichment - genus and species
  if ("superkingdom" %in% names(df_raw)) {
    for (tax_lvl_v in c("genus", "species")) {
      lvl_lbl_v  <- c(genus = "Genus", species = "Species")[tax_lvl_v]
      virus_esch <- df_raw %>%
        filter(superkingdom == "Viruses", tax_level == tax_lvl_v,
               enrichment_type == enrich, rel_abund > 0) %>%
        rename(name = taxon) %>%
        mutate(sample_type = factor(
          if_else(grepl("^EVI", as.character(group)), "Post-EVI", "Post-ASC"),
          levels = c("Post-EVI", "Post-ASC")
        ))
      if (nrow(virus_esch) > 0) {
        plot_genus_presence(
          virus_esch, spp_display_labels,
          title          = paste0("Viruses/Bacteriophage Presence/Absence (", lvl_lbl_v,
                                  ") - ", enrich_lbl, " (Reads)"),
          subtitle       = paste0("Viral ", tolower(lvl_lbl_v),
                                  "s detected in Escherichia samples; purple = present"),
          present_colour = "#762A83",
          save_path      = file.path(out_dir, paste0(file_slug, "_virus_",
                                                      tax_lvl_v, "_presence.png"))
        )
      }
    }
  }
}

```


## Salmonella Species

```{r bpm_salm_species, fig.height=7, fig.width=14}

for (enrich in levels(df$enrichment_type)) {

  spp_df <- df %>%
    filter(tax_level == "species",
           str_detect(taxon, "^Salmonella"),
           enrichment_type == enrich) %>%
    rename(name = taxon) %>%
    mutate(
      sample_type = factor(
        if_else(grepl("^EVI", as.character(group)), "Post-EVI", "Post-ASC"),
        levels = c("Post-EVI", "Post-ASC")
      )
    )

  if (nrow(spp_df) == 0) {
    message("No Salmonella species data for: ", enrich)
    next
  }

  enrich_lbl <- enrichment_labels[enrich]
  file_slug  <- paste0("salm_", gsub("[^A-Za-z0-9]", "_", enrich))

  salm_species <- sort(unique(spp_df$name))
  salm_fill    <- build_fill_vals(salm_species, species_colors_salmonella,
                                   pal = brewer.pal(8, "Set1"))

  bar_df <- spp_df %>%
    mutate(
      disp_label = factor(rename_vec(sample, spp_display_labels),
                          levels = unique(rename_vec(names(spp_display_labels),
                                                     spp_display_labels))),
      name       = factor(name, levels = rev(salm_species))
    )

  p_bar <- ggplot(bar_df, aes(x = disp_label, y = rel_abund, fill = name)) +
    geom_col(width = 0.7, colour = "white", linewidth = 0.3) +
    scale_fill_manual(values = salm_fill, name = "Species") +
    scale_y_continuous(labels = function(x) paste0(x, "%"),
                       expand = expansion(mult = c(0, 0.02))) +
    facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +
    labs(
      title    = paste0("Salmonella Species - ", enrich_lbl, " (Reads)"),
      subtitle = "Relative abundance of classified bases",
      x        = NULL,
      y        = "Relative abundance (%)"
    ) +
    theme_bw(base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(size = 10, colour = "grey40"),
      axis.text.x   = element_text(angle = 45, hjust = 1),
      legend.text   = element_text(size = 11),
      strip.text    = element_text(face = "bold", size = 12),
      panel.spacing = unit(0.8, "lines")
    )

  print(p_bar)
  ggsave(file.path(out_dir, paste0(file_slug, "_species_bar.png")),
         p_bar, width = 10, height = 6, dpi = 600)
  message("Saved: ", file_slug, "_species_bar.png")

  plot_genus_linear_presence(
    spp_df, spp_display_labels,
    title          = paste0("Salmonella Species Relative Abundance & Presence - ", enrich_lbl, " (Reads)"),
    save_path      = file.path(out_dir, paste0(file_slug, "_linear_presence.png"))
  )

  plot_genus_presence(
    spp_df, spp_display_labels,
    title          = paste0("Salmonella Species Presence/Absence - ", enrich_lbl, " (Reads)"),
    save_path      = file.path(out_dir, paste0(file_slug, "_presence.png"))
  )

  # Virus/bacteriophage presence associated with this enrichment - genus and species
  if ("superkingdom" %in% names(df_raw)) {
    for (tax_lvl_v in c("genus", "species")) {
      lvl_lbl_v  <- c(genus = "Genus", species = "Species")[tax_lvl_v]
      virus_salm <- df_raw %>%
        filter(superkingdom == "Viruses", tax_level == tax_lvl_v,
               enrichment_type == enrich, rel_abund > 0) %>%
        rename(name = taxon) %>%
        mutate(sample_type = factor(
          if_else(grepl("^EVI", as.character(group)), "Post-EVI", "Post-ASC"),
          levels = c("Post-EVI", "Post-ASC")
        ))
      if (nrow(virus_salm) > 0) {
        plot_genus_presence(
          virus_salm, spp_display_labels,
          title          = paste0("Viruses/Bacteriophage Presence/Absence (", lvl_lbl_v,
                                  ") - ", enrich_lbl, " (Reads)"),
          subtitle       = paste0("Viral ", tolower(lvl_lbl_v),
                                  "s detected in Salmonella samples; purple = present"),
          present_colour = "#762A83",
          save_path      = file.path(out_dir, paste0(file_slug, "_virus_",
                                                      tax_lvl_v, "_presence.png"))
        )
      }
    }
  }
}

```


## Salmonella Subspecies and Serotypes

Kraken2 can classify Salmonella below species level, including subspecies of *S. enterica*
and individual serovars (e.g. Typhimurium, Enteritidis). Taxon names matching
`^Salmonella` at any taxonomic rank below genus are captured here.

```{r bpm_salm_serotype, fig.height=8, fig.width=16}

# Helper: shorten long serovar names for axis labels
shorten_salm <- function(x) {
  x %>%
    str_replace("Salmonella enterica subsp\\. enterica serovar ", "S. ser. ") %>%
    str_replace("Salmonella enterica subsp\\. ", "S. subsp. ") %>%
    str_replace("^Salmonella ", "S. ")
}

for (enrich in levels(df$enrichment_type)) {

  # Capture everything under Salmonella at ranks below genus
  sero_df <- df_raw %>%
    filter(
      superkingdom == "Bacteria" | !("superkingdom" %in% names(df_raw)),
      str_detect(taxon, "^Salmonella"),
      !tax_level %in% c("genus", "family", "order", "class", "phylum",
                        "superkingdom", "kingdom"),
      enrichment_type == enrich,
      rel_abund > 0
    ) %>%
    rename(name = taxon) %>%
    mutate(
      sample_type  = factor(
        if_else(grepl("^EVI", as.character(group)), "Post-EVI", "Post-ASC"),
        levels = c("Post-EVI", "Post-ASC")
      ),
      short_name = shorten_salm(name)
    )

  if (nrow(sero_df) == 0) {
    message("No Salmonella subspecies/serotype data for: ", enrich); next
  }

  enrich_lbl <- enrichment_labels[enrich]
  file_slug  <- paste0("salm_serotype_", gsub("[^A-Za-z0-9]", "_", enrich))

  sero_order <- sero_df %>%
    group_by(name) %>% summarise(m = mean(rel_abund), .groups = "drop") %>%
    filter(m > 0) %>%
    arrange(desc(m)) %>% pull(name)

  sero_fill  <- build_fill_vals(sero_order, species_colors_salmonella_serotype,
                                 pal = brewer.pal(12, "Paired"))
  sero_fill  <- sero_fill[names(sero_fill) != "Other"]

  # Stacked bar with shortened names
  bar_df <- sero_df %>%
    mutate(
      disp_label = factor(rename_vec(sample, spp_display_labels),
                          levels = unique(rename_vec(names(spp_display_labels),
                                                     spp_display_labels))),
      name       = factor(name, levels = rev(sero_order)),
      short_name = factor(shorten_salm(as.character(name)),
                          levels = rev(shorten_salm(sero_order)))
    )

  p_bar <- ggplot(bar_df, aes(x = disp_label, y = rel_abund, fill = name)) +
    geom_col(width = 0.7, colour = "white", linewidth = 0.3) +
    scale_fill_manual(
      values = sero_fill, name = "Subspecies / Serovar",
      labels = shorten_salm(names(sero_fill))
    ) +
    scale_y_continuous(labels = function(x) paste0(x, "%"),
                       expand = expansion(mult = c(0, 0.02))) +
    facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +
    labs(
      title    = paste0("Salmonella Subspecies & Serotypes - ", enrich_lbl, " (Reads)"),
      subtitle = "Relative abundance of sub-species level classifications",
      x = NULL, y = "Relative abundance (%)"
    ) +
    theme_bw(base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(size = 10, colour = "grey40"),
      axis.text.x   = element_text(angle = 45, hjust = 1),
      legend.text   = element_text(size = 10, face = "italic"),
      strip.text    = element_text(face = "bold", size = 12),
      panel.spacing = unit(0.8, "lines")
    )
  print(p_bar)
  ggsave(file.path(out_dir, paste0(file_slug, "_bar.png")),
         p_bar, width = 12, height = 6, dpi = 600)

  # Presence/absence tile with shortened axis labels
  sero_grid <- sero_df %>%
    mutate(disp_label = rename_vec(sample, spp_display_labels),
           present    = rel_abund > 0) %>%
    group_by(name, disp_label) %>%
    summarise(present = any(present), .groups = "drop")

  all_combos <- expand.grid(
    name       = sero_order,
    disp_label = unique(sero_grid$disp_label),
    stringsAsFactors = FALSE
  )
  stype_map <- sero_df %>%
    mutate(disp_label = rename_vec(sample, spp_display_labels)) %>%
    distinct(disp_label, sample_type)

  sero_grid <- all_combos %>%
    left_join(sero_grid, by = c("name", "disp_label")) %>%
    left_join(stype_map, by = "disp_label") %>%
    mutate(
      present     = replace_na(present, FALSE),
      disp_label  = factor(disp_label,
                            levels = rev(unique(rename_vec(names(spp_display_labels),
                                                           spp_display_labels)))),
      name        = factor(name, levels = sero_order),
      short_name  = factor(shorten_salm(as.character(name)),
                           levels = shorten_salm(sero_order)),
      fill_val    = factor(if_else(present, as.character(name), "Absent"),
                           levels = c(as.character(sero_order), "Absent"))
    )

  tile_vals <- c(sero_fill, "Absent" = "#F0F0F0")

  p_tile <- ggplot(sero_grid,
                   aes(x = short_name, y = disp_label, fill = fill_val)) +
    geom_tile(colour = "white", linewidth = 0.6) +
    scale_fill_manual(
      values = tile_vals,
      breaks = as.character(sero_order),
      labels = shorten_salm(sero_order),
      name   = NULL, drop = FALSE
    ) +
    facet_grid(sample_type ~ ., scales = "free_y", space = "free_y") +
    labs(
      title    = paste0("Salmonella Subspecies & Serotypes - Presence/Absence - ", enrich_lbl),
      subtitle = "Coloured by subspecies/serovar; grey = absent",
      x = NULL, y = NULL
    ) +
    theme_bw(base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(size = 10, colour = "grey40"),
      axis.text.x   = element_text(angle = 45, hjust = 1, size = 10, face = "italic"),
      axis.text.y   = element_text(size = 10),
      strip.text    = element_text(face = "bold", size = 12),
      panel.spacing = unit(0.6, "lines"),
      legend.position = "bottom"
    )
  print(p_tile)
  n_samp <- length(unique(sero_df$sample))
  ggsave(file.path(out_dir, paste0(file_slug, "_presence.png")),
         p_tile,
         width  = max(8, length(sero_order) * 1.1 + 2),
         height = max(5, n_samp * 0.45 + 3),
         dpi    = 600)
  message("Saved: ", file_slug)
}

```


## Campylobacter Species

```{r bpm_camp_species, fig.height=7, fig.width=14}

for (enrich in levels(df$enrichment_type)) {

  spp_df <- df %>%
    filter(tax_level == "species",
           str_detect(taxon, "^Campylobacter"),
           enrichment_type == enrich) %>%
    rename(name = taxon) %>%
    mutate(
      sample_type = factor(
        if_else(grepl("^EVI", as.character(group)), "Post-EVI", "Post-ASC"),
        levels = c("Post-EVI", "Post-ASC")
      )
    )

  if (nrow(spp_df) == 0) {
    message("No Campylobacter species data for: ", enrich)
    next
  }

  enrich_lbl <- enrichment_labels[enrich]
  file_slug  <- paste0("camp_", gsub("[^A-Za-z0-9]", "_", enrich))

  camp_species <- sort(unique(spp_df$name))
  camp_fill    <- build_fill_vals(camp_species, species_colors_campylobacter,
                                   pal = brewer.pal(8, "Dark2"))

  bar_df <- spp_df %>%
    mutate(
      disp_label = factor(rename_vec(sample, spp_display_labels),
                          levels = unique(rename_vec(names(spp_display_labels),
                                                     spp_display_labels))),
      name       = factor(name, levels = rev(camp_species))
    )

  p_bar <- ggplot(bar_df, aes(x = disp_label, y = rel_abund, fill = name)) +
    geom_col(width = 0.7, colour = "white", linewidth = 0.3) +
    scale_fill_manual(values = camp_fill, name = "Species") +
    scale_y_continuous(labels = function(x) paste0(x, "%"),
                       expand = expansion(mult = c(0, 0.02))) +
    facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +
    labs(
      title    = paste0("Campylobacter Species - ", enrich_lbl, " (Reads)"),
      subtitle = "Relative abundance of classified bases",
      x        = NULL,
      y        = "Relative abundance (%)"
    ) +
    theme_bw(base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(size = 10, colour = "grey40"),
      axis.text.x   = element_text(angle = 45, hjust = 1),
      legend.text   = element_text(size = 11),
      strip.text    = element_text(face = "bold", size = 12),
      panel.spacing = unit(0.8, "lines")
    )

  print(p_bar)
  ggsave(file.path(out_dir, paste0(file_slug, "_species_bar.png")),
         p_bar, width = 10, height = 6, dpi = 600)
  message("Saved: ", file_slug, "_species_bar.png")

  plot_genus_linear_presence(
    spp_df, spp_display_labels,
    title          = paste0("Campylobacter Species Relative Abundance & Presence - ", enrich_lbl, " (Reads)"),
    save_path      = file.path(out_dir, paste0(file_slug, "_linear_presence.png"))
  )

  plot_genus_presence(
    spp_df, spp_display_labels,
    title          = paste0("Campylobacter Species Presence/Absence - ", enrich_lbl, " (Reads)"),
    save_path      = file.path(out_dir, paste0(file_slug, "_presence.png"))
  )

  # Virus/bacteriophage presence associated with this enrichment - genus and species
  if ("superkingdom" %in% names(df_raw)) {
    for (tax_lvl_v in c("genus", "species")) {
      lvl_lbl_v  <- c(genus = "Genus", species = "Species")[tax_lvl_v]
      virus_camp <- df_raw %>%
        filter(superkingdom == "Viruses", tax_level == tax_lvl_v,
               enrichment_type == enrich, rel_abund > 0) %>%
        rename(name = taxon) %>%
        mutate(sample_type = factor(
          if_else(grepl("^EVI", as.character(group)), "Post-EVI", "Post-ASC"),
          levels = c("Post-EVI", "Post-ASC")
        ))
      if (nrow(virus_camp) > 0) {
        plot_genus_presence(
          virus_camp, spp_display_labels,
          title          = paste0("Viruses/Bacteriophage Presence/Absence (", lvl_lbl_v,
                                  ") - ", enrich_lbl, " (Reads)"),
          subtitle       = paste0("Viral ", tolower(lvl_lbl_v),
                                  "s detected in Campylobacter samples; purple = present"),
          present_colour = "#762A83",
          save_path      = file.path(out_dir, paste0(file_slug, "_virus_",
                                                      tax_lvl_v, "_presence.png"))
        )
      }
    }
  }
}

```


## Bacteroides Species

```{r bpm_bact_species, fig.height=7, fig.width=14}

for (enrich in levels(df$enrichment_type)) {
  spp_df <- df %>%
    filter(tax_level == "species",
           str_detect(taxon, "^Bacteroides"),
           enrichment_type == enrich) %>%
    rename(name = taxon) %>%
    mutate(sample_type = factor(
      if_else(grepl("^EVI", as.character(group)), "Post-EVI", "Post-ASC"),
      levels = c("Post-EVI", "Post-ASC")
    ))
  if (nrow(spp_df) == 0) { message("No Bacteroides data for: ", enrich); next }

  enrich_lbl  <- enrichment_labels[enrich]
  file_slug   <- paste0("bact_", gsub("[^A-Za-z0-9]", "_", enrich))
  bact_species <- sort(unique(spp_df$name))
  bact_fill    <- build_fill_vals(bact_species, species_colors_bacteroides, pal = brewer.pal(8, "Paired"))

  bar_df <- spp_df %>%
    mutate(disp_label = factor(rename_vec(sample, spp_display_labels),
                               levels = unique(rename_vec(names(spp_display_labels), spp_display_labels))),
           name = factor(name, levels = rev(bact_species)))
  p_bar <- ggplot(bar_df, aes(x = disp_label, y = rel_abund, fill = name)) +
    geom_col(width = 0.7, colour = "white", linewidth = 0.3) +
    scale_fill_manual(values = bact_fill, name = "Species") +
    scale_y_continuous(labels = function(x) paste0(x, "%"), expand = expansion(mult = c(0, 0.02))) +
    facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +
    labs(title = paste0("Bacteroides Species - ", enrich_lbl, " (Reads)"),
         subtitle = "Relative abundance of classified bases", x = NULL, y = "Relative abundance (%)") +
    theme_bw(base_size = 12) +
    theme(plot.title = element_text(face = "bold", size = 13),
          plot.subtitle = element_text(size = 10, colour = "grey40"),
          axis.text.x = element_text(angle = 45, hjust = 1),
          legend.text = element_text(size = 11), strip.text = element_text(face = "bold", size = 12),
          panel.spacing = unit(0.8, "lines"))
  print(p_bar)
  ggsave(file.path(out_dir, paste0(file_slug, "_species_bar.png")), p_bar, width = 10, height = 6, dpi = 600)
  message("Saved: ", file_slug, "_species_bar.png")

  plot_genus_linear_presence(spp_df, spp_display_labels,
    title = paste0("Bacteroides Species Relative Abundance & Presence - ", enrich_lbl, " (Reads)"),
    save_path = file.path(out_dir, paste0(file_slug, "_linear_presence.png")))
  plot_genus_presence(spp_df, spp_display_labels,
    title          = paste0("Bacteroides Species Presence/Absence - ", enrich_lbl, " (Reads)"),
    save_path      = file.path(out_dir, paste0(file_slug, "_presence.png")))
}

```


## Lactobacillus Species

```{r bpm_lact_species, fig.height=7, fig.width=14}

for (enrich in levels(df$enrichment_type)) {
  spp_df <- df %>%
    filter(tax_level == "species",
           str_detect(taxon, "^Lactobacillus"),
           enrichment_type == enrich) %>%
    rename(name = taxon) %>%
    mutate(sample_type = factor(
      if_else(grepl("^EVI", as.character(group)), "Post-EVI", "Post-ASC"),
      levels = c("Post-EVI", "Post-ASC")
    ))
  if (nrow(spp_df) == 0) { message("No Lactobacillus data for: ", enrich); next }

  enrich_lbl   <- enrichment_labels[enrich]
  file_slug    <- paste0("lact_", gsub("[^A-Za-z0-9]", "_", enrich))
  lact_species <- sort(unique(spp_df$name))
  lact_fill    <- build_fill_vals(lact_species, species_colors_lactobacillus, pal = brewer.pal(8, "Set3"))

  bar_df <- spp_df %>%
    mutate(disp_label = factor(rename_vec(sample, spp_display_labels),
                               levels = unique(rename_vec(names(spp_display_labels), spp_display_labels))),
           name = factor(name, levels = rev(lact_species)))
  p_bar <- ggplot(bar_df, aes(x = disp_label, y = rel_abund, fill = name)) +
    geom_col(width = 0.7, colour = "white", linewidth = 0.3) +
    scale_fill_manual(values = lact_fill, name = "Species") +
    scale_y_continuous(labels = function(x) paste0(x, "%"), expand = expansion(mult = c(0, 0.02))) +
    facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +
    labs(title = paste0("Lactobacillus Species - ", enrich_lbl, " (Reads)"),
         subtitle = "Relative abundance of classified bases", x = NULL, y = "Relative abundance (%)") +
    theme_bw(base_size = 12) +
    theme(plot.title = element_text(face = "bold", size = 13),
          plot.subtitle = element_text(size = 10, colour = "grey40"),
          axis.text.x = element_text(angle = 45, hjust = 1),
          legend.text = element_text(size = 11), strip.text = element_text(face = "bold", size = 12),
          panel.spacing = unit(0.8, "lines"))
  print(p_bar)
  ggsave(file.path(out_dir, paste0(file_slug, "_species_bar.png")), p_bar, width = 10, height = 6, dpi = 600)
  message("Saved: ", file_slug, "_species_bar.png")

  plot_genus_linear_presence(spp_df, spp_display_labels,
    title = paste0("Lactobacillus Species Relative Abundance & Presence - ", enrich_lbl, " (Reads)"),
    save_path = file.path(out_dir, paste0(file_slug, "_linear_presence.png")))
  plot_genus_presence(spp_df, spp_display_labels,
    title          = paste0("Lactobacillus Species Presence/Absence - ", enrich_lbl, " (Reads)"),
    save_path      = file.path(out_dir, paste0(file_slug, "_presence.png")))
}

```


## Pseudomonas Species

```{r bpm_pseu_species, fig.height=7, fig.width=14}

for (enrich in levels(df$enrichment_type)) {
  spp_df <- df %>%
    filter(tax_level == "species",
           str_detect(taxon, "^Pseudomonas"),
           enrichment_type == enrich) %>%
    rename(name = taxon) %>%
    mutate(sample_type = factor(
      if_else(grepl("^EVI", as.character(group)), "Post-EVI", "Post-ASC"),
      levels = c("Post-EVI", "Post-ASC")
    ))
  if (nrow(spp_df) == 0) { message("No Pseudomonas data for: ", enrich); next }

  enrich_lbl   <- enrichment_labels[enrich]
  file_slug    <- paste0("pseu_", gsub("[^A-Za-z0-9]", "_", enrich))
  pseu_species <- sort(unique(spp_df$name))
  pseu_fill    <- build_fill_vals(pseu_species, species_colors_pseudomonas, pal = brewer.pal(8, "Dark2"))

  bar_df <- spp_df %>%
    mutate(disp_label = factor(rename_vec(sample, spp_display_labels),
                               levels = unique(rename_vec(names(spp_display_labels), spp_display_labels))),
           name = factor(name, levels = rev(pseu_species)))
  p_bar <- ggplot(bar_df, aes(x = disp_label, y = rel_abund, fill = name)) +
    geom_col(width = 0.7, colour = "white", linewidth = 0.3) +
    scale_fill_manual(values = pseu_fill, name = "Species") +
    scale_y_continuous(labels = function(x) paste0(x, "%"), expand = expansion(mult = c(0, 0.02))) +
    facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +
    labs(title = paste0("Pseudomonas Species - ", enrich_lbl, " (Reads)"),
         subtitle = "Relative abundance of classified bases", x = NULL, y = "Relative abundance (%)") +
    theme_bw(base_size = 12) +
    theme(plot.title = element_text(face = "bold", size = 13),
          plot.subtitle = element_text(size = 10, colour = "grey40"),
          axis.text.x = element_text(angle = 45, hjust = 1),
          legend.text = element_text(size = 11), strip.text = element_text(face = "bold", size = 12),
          panel.spacing = unit(0.8, "lines"))
  print(p_bar)
  ggsave(file.path(out_dir, paste0(file_slug, "_species_bar.png")), p_bar, width = 10, height = 6, dpi = 600)
  message("Saved: ", file_slug, "_species_bar.png")

  plot_genus_linear_presence(spp_df, spp_display_labels,
    title = paste0("Pseudomonas Species Relative Abundance & Presence - ", enrich_lbl, " (Reads)"),
    save_path = file.path(out_dir, paste0(file_slug, "_linear_presence.png")))
  plot_genus_presence(spp_df, spp_display_labels,
    title          = paste0("Pseudomonas Species Presence/Absence - ", enrich_lbl, " (Reads)"),
    save_path      = file.path(out_dir, paste0(file_slug, "_presence.png")))
}

```


## Veillonella Species

```{r bpm_veil_species, fig.height=7, fig.width=14}

for (enrich in levels(df$enrichment_type)) {
  spp_df <- df %>%
    filter(tax_level == "species",
           str_detect(taxon, "^Veillonella"),
           enrichment_type == enrich) %>%
    rename(name = taxon) %>%
    mutate(sample_type = factor(
      if_else(grepl("^EVI", as.character(group)), "Post-EVI", "Post-ASC"),
      levels = c("Post-EVI", "Post-ASC")
    ))
  if (nrow(spp_df) == 0) { message("No Veillonella data for: ", enrich); next }

  enrich_lbl   <- enrichment_labels[enrich]
  file_slug    <- paste0("veil_", gsub("[^A-Za-z0-9]", "_", enrich))
  veil_species <- sort(unique(spp_df$name))
  veil_fill    <- build_fill_vals(veil_species, species_colors_veillonella, pal = brewer.pal(8, "Set1"))

  bar_df <- spp_df %>%
    mutate(disp_label = factor(rename_vec(sample, spp_display_labels),
                               levels = unique(rename_vec(names(spp_display_labels), spp_display_labels))),
           name = factor(name, levels = rev(veil_species)))
  p_bar <- ggplot(bar_df, aes(x = disp_label, y = rel_abund, fill = name)) +
    geom_col(width = 0.7, colour = "white", linewidth = 0.3) +
    scale_fill_manual(values = veil_fill, name = "Species") +
    scale_y_continuous(labels = function(x) paste0(x, "%"), expand = expansion(mult = c(0, 0.02))) +
    facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +
    labs(title = paste0("Veillonella Species - ", enrich_lbl, " (Reads)"),
         subtitle = "Relative abundance of classified bases", x = NULL, y = "Relative abundance (%)") +
    theme_bw(base_size = 12) +
    theme(plot.title = element_text(face = "bold", size = 13),
          plot.subtitle = element_text(size = 10, colour = "grey40"),
          axis.text.x = element_text(angle = 45, hjust = 1),
          legend.text = element_text(size = 11), strip.text = element_text(face = "bold", size = 12),
          panel.spacing = unit(0.8, "lines"))
  print(p_bar)
  ggsave(file.path(out_dir, paste0(file_slug, "_species_bar.png")), p_bar, width = 10, height = 6, dpi = 600)
  message("Saved: ", file_slug, "_species_bar.png")

  plot_genus_linear_presence(spp_df, spp_display_labels,
    title = paste0("Veillonella Species Relative Abundance & Presence - ", enrich_lbl, " (Reads)"),
    save_path = file.path(out_dir, paste0(file_slug, "_linear_presence.png")))
  plot_genus_presence(spp_df, spp_display_labels,
    title          = paste0("Veillonella Species Presence/Absence - ", enrich_lbl, " (Reads)"),
    save_path      = file.path(out_dir, paste0(file_slug, "_presence.png")))
}

```


## Tannerella Species

```{r bpm_tann_species, fig.height=7, fig.width=14}

for (enrich in levels(df$enrichment_type)) {
  spp_df <- df %>%
    filter(tax_level == "species",
           str_detect(taxon, "^Tannerella"),
           enrichment_type == enrich) %>%
    rename(name = taxon) %>%
    mutate(sample_type = factor(
      if_else(grepl("^EVI", as.character(group)), "Post-EVI", "Post-ASC"),
      levels = c("Post-EVI", "Post-ASC")
    ))
  if (nrow(spp_df) == 0) { message("No Tannerella data for: ", enrich); next }

  enrich_lbl   <- enrichment_labels[enrich]
  file_slug    <- paste0("tann_", gsub("[^A-Za-z0-9]", "_", enrich))
  tann_species <- sort(unique(spp_df$name))
  tann_fill    <- build_fill_vals(tann_species, species_colors_tannerella, pal = brewer.pal(8, "Set2"))

  bar_df <- spp_df %>%
    mutate(disp_label = factor(rename_vec(sample, spp_display_labels),
                               levels = unique(rename_vec(names(spp_display_labels), spp_display_labels))),
           name = factor(name, levels = rev(tann_species)))
  p_bar <- ggplot(bar_df, aes(x = disp_label, y = rel_abund, fill = name)) +
    geom_col(width = 0.7, colour = "white", linewidth = 0.3) +
    scale_fill_manual(values = tann_fill, name = "Species") +
    scale_y_continuous(labels = function(x) paste0(x, "%"), expand = expansion(mult = c(0, 0.02))) +
    facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +
    labs(title = paste0("Tannerella Species - ", enrich_lbl, " (Reads)"),
         subtitle = "Relative abundance of classified bases", x = NULL, y = "Relative abundance (%)") +
    theme_bw(base_size = 12) +
    theme(plot.title = element_text(face = "bold", size = 13),
          plot.subtitle = element_text(size = 10, colour = "grey40"),
          axis.text.x = element_text(angle = 45, hjust = 1),
          legend.text = element_text(size = 11), strip.text = element_text(face = "bold", size = 12),
          panel.spacing = unit(0.8, "lines"))
  print(p_bar)
  ggsave(file.path(out_dir, paste0(file_slug, "_species_bar.png")), p_bar, width = 10, height = 6, dpi = 600)
  message("Saved: ", file_slug, "_species_bar.png")

  plot_genus_linear_presence(spp_df, spp_display_labels,
    title = paste0("Tannerella Species Relative Abundance & Presence - ", enrich_lbl, " (Reads)"),
    save_path = file.path(out_dir, paste0(file_slug, "_linear_presence.png")))
  plot_genus_presence(spp_df, spp_display_labels,
    title          = paste0("Tannerella Species Presence/Absence - ", enrich_lbl, " (Reads)"),
    save_path      = file.path(out_dir, paste0(file_slug, "_presence.png")))
}

```


## Comamonas Species

```{r bpm_coma_species, fig.height=7, fig.width=14}

for (enrich in levels(df$enrichment_type)) {
  spp_df <- df %>%
    filter(tax_level == "species",
           str_detect(taxon, "^Comamonas"),
           enrichment_type == enrich) %>%
    rename(name = taxon) %>%
    mutate(sample_type = factor(
      if_else(grepl("^EVI", as.character(group)), "Post-EVI", "Post-ASC"),
      levels = c("Post-EVI", "Post-ASC")
    ))
  if (nrow(spp_df) == 0) { message("No Comamonas data for: ", enrich); next }

  enrich_lbl   <- enrichment_labels[enrich]
  file_slug    <- paste0("coma_", gsub("[^A-Za-z0-9]", "_", enrich))
  coma_species <- sort(unique(spp_df$name))
  coma_fill    <- build_fill_vals(coma_species, species_colors_comamonas, pal = brewer.pal(8, "Accent"))

  bar_df <- spp_df %>%
    mutate(disp_label = factor(rename_vec(sample, spp_display_labels),
                               levels = unique(rename_vec(names(spp_display_labels), spp_display_labels))),
           name = factor(name, levels = rev(coma_species)))
  p_bar <- ggplot(bar_df, aes(x = disp_label, y = rel_abund, fill = name)) +
    geom_col(width = 0.7, colour = "white", linewidth = 0.3) +
    scale_fill_manual(values = coma_fill, name = "Species") +
    scale_y_continuous(labels = function(x) paste0(x, "%"), expand = expansion(mult = c(0, 0.02))) +
    facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +
    labs(title = paste0("Comamonas Species - ", enrich_lbl, " (Reads)"),
         subtitle = "Relative abundance of classified bases", x = NULL, y = "Relative abundance (%)") +
    theme_bw(base_size = 12) +
    theme(plot.title = element_text(face = "bold", size = 13),
          plot.subtitle = element_text(size = 10, colour = "grey40"),
          axis.text.x = element_text(angle = 45, hjust = 1),
          legend.text = element_text(size = 11), strip.text = element_text(face = "bold", size = 12),
          panel.spacing = unit(0.8, "lines"))
  print(p_bar)
  ggsave(file.path(out_dir, paste0(file_slug, "_species_bar.png")), p_bar, width = 10, height = 6, dpi = 600)
  message("Saved: ", file_slug, "_species_bar.png")

  plot_genus_linear_presence(spp_df, spp_display_labels,
    title = paste0("Comamonas Species Relative Abundance & Presence - ", enrich_lbl, " (Reads)"),
    save_path = file.path(out_dir, paste0(file_slug, "_linear_presence.png")))
  plot_genus_presence(spp_df, spp_display_labels,
    title          = paste0("Comamonas Species Presence/Absence - ", enrich_lbl, " (Reads)"),
    save_path      = file.path(out_dir, paste0(file_slug, "_presence.png")))
}

```

---

# Non-bacterial Microbiome Reads - Archaea and Viruses (including Bacteriophage)

Metagenomics enables detection of all domains of life beyond bacteria. This section shows
archaeal and viral (including bacteriophage) reads detected across all sample sets,
illustrating the broader microbial community captured by an untargeted sequencing approach.
All eukaryotic reads (host/food-associated contamination: *Gallus gallus*, mammals, birds)
are excluded; only archaea and viruses are shown here as true microbiome members.

Relative abundance is expressed as a percentage of **all classified bases** in each sample.

```{r archaea_viral_reads, fig.height=10, fig.width=16}

if (!"superkingdom" %in% names(df_raw)) {
  message("NOTE: 'superkingdom' column not found - re-run kraken2_bpm.py to enable archaea/viral detection. Skipping.")
} else {

  # ── Colour palettes ────────────────────────────────────────────────────────
  archaea_pal <- c(
    "#2166AC", "#4393C3", "#92C5DE", "#D1E5F0",
    "#053061", "#67A9CF", "#1A6296", "#3A7DC9"
  )
  virus_pal <- c(
    "#762A83", "#9970AB", "#C2A5CF", "#E7D4E8",
    "#40004B", "#AF8DC3", "#5E1C7A", "#8856A7"
  )

  for (tax_lvl_v in c("genus", "species")) {
  lvl_lbl_v <- c(genus = "Genus", species = "Species")[tax_lvl_v]

  # ── Filter: Archaea + Viruses at genus / species level; exclude eukaryotes ──
  df_micro <- df_raw %>%
    filter(
      superkingdom %in% c("Archaea", "Viruses"),
      tax_level == tax_lvl_v,
      rel_abund > 0
    ) %>%
    mutate(
      sample_type = factor(
        if_else(grepl("^EVI", as.character(group)), "Post-EVI", "Post-ASC"),
        levels = c("Post-EVI", "Post-ASC")
      ),
      kingdom = factor(superkingdom, levels = c("Archaea", "Viruses"))
    )

  if (nrow(df_micro) == 0) {
    message("No archaea or viral ", tolower(lvl_lbl_v),
            "-level reads detected across any sample.")
  } else {

    message("Non-bacterial microbiome reads detected (", lvl_lbl_v, " level):")
    message("  Archaea: ",
            paste(sort(unique(df_micro$taxon[df_micro$superkingdom == "Archaea"])), collapse = ", "))
    message("  Viruses: ",
            paste(sort(unique(df_micro$taxon[df_micro$superkingdom == "Viruses"])), collapse = ", "))

    for (enrich in levels(df$enrichment_type)) {

      micro_sub <- df_micro %>% filter(enrichment_type == enrich)
      if (nrow(micro_sub) == 0) {
        message("No archaea/viral reads for: ", enrich); next
      }

      enrich_lbl <- enrichment_labels[enrich]
      file_slug  <- gsub("[^A-Za-z0-9]", "_", enrich)

      # Display labels: append (Post-EVI)/(Post-ASC) suffix to avoid duplicate CH IDs
      disp_map <- micro_sub %>%
        distinct(sample, sample_type) %>%
        mutate(disp_label = paste0(sample_labels[sample],
                                   " (", as.character(sample_type), ")")) %>%
        { setNames(.$disp_label, .$sample) }

      disp_order <- micro_sub %>%
        arrange(sample_type, sample_labels[sample]) %>%
        distinct(sample) %>%
        pull(sample) %>%
        { disp_map[.] }

      # ── Archaea panel ──────────────────────────────────────────────────────
      arch_sub <- micro_sub %>% filter(superkingdom == "Archaea")

      if (nrow(arch_sub) > 0) {
        arch_taxa <- arch_sub %>%
          group_by(taxon) %>% summarise(m = mean(rel_abund), .groups = "drop") %>%
          arrange(desc(m)) %>% pull(taxon)
        arch_fill <- setNames(
          colorRampPalette(archaea_pal)(length(arch_taxa)), arch_taxa
        )

        arch_bar_df <- arch_sub %>%
          mutate(
            disp_label = factor(disp_map[sample], levels = disp_order),
            taxon      = factor(taxon, levels = arch_taxa)
          )

        p_arch_bar <- ggplot(arch_bar_df,
                             aes(x = disp_label, y = rel_abund, fill = taxon)) +
          geom_col(width = 0.7, colour = "white", linewidth = 0.3) +
          scale_fill_manual(values = arch_fill, name = "Genus") +
          scale_y_continuous(
            labels = function(x) paste0(x, "%"),
            expand = expansion(mult = c(0, 0.05))
          ) +
          facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +
          labs(
            title    = paste0("Archaea - ", enrich_lbl, " (Reads)"),
            subtitle = paste0("Relative abundance of archaeal ", tolower(lvl_lbl_v),
                              "s (% of all classified bases)"),
            x = NULL, y = "Relative abundance (%)"
          ) +
          theme_bw(base_size = 12) +
          theme(
            plot.title    = element_text(face = "bold", size = 13),
            plot.subtitle = element_text(size = 10, colour = "grey40"),
            axis.text.x   = element_text(angle = 45, hjust = 1),
            legend.text   = element_text(size = 10, face = "italic"),
            strip.text    = element_text(face = "bold", size = 12),
            panel.spacing = unit(0.8, "lines")
          )

        arch_tile_df <- arch_sub %>%
          mutate(
            disp_label = factor(disp_map[sample], levels = disp_order),
            taxon      = factor(taxon, levels = rev(arch_taxa)),
            present    = rel_abund > 0
          )

        p_arch_tile <- ggplot(arch_tile_df,
                              aes(x = disp_label, y = taxon, fill = present)) +
          geom_tile(colour = "white", linewidth = 0.5) +
          scale_fill_manual(
            values = c("TRUE" = "#2166AC", "FALSE" = "#F7F7F7"),
            labels = c("TRUE" = "Present", "FALSE" = "Absent"),
            name   = NULL
          ) +
          facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +
          labs(x = NULL, y = NULL,
               subtitle = "Blue = present; grey = absent") +
          theme_bw(base_size = 11) +
          theme(
            axis.text.x     = element_text(angle = 45, hjust = 1),
            axis.text.y     = element_text(face = "italic"),
            strip.text      = element_text(face = "bold", size = 11),
            panel.spacing   = unit(0.8, "lines"),
            legend.position = "bottom"
          )

        p_arch <- p_arch_bar / p_arch_tile +
          plot_layout(heights = c(2, 1)) +
          plot_annotation(
            title = paste0("Archaea detected - ", enrich_lbl),
            theme = theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5))
          )
        print(p_arch)
        n_arch_taxa <- length(arch_taxa)
        h_arch <- max(5, min(20, n_arch_taxa * 0.4 + 3))
        ggsave(
          file.path(out_dir, paste0("archaea_", file_slug, "_", tax_lvl_v, "_reads.png")),
          p_arch, width = 12, height = h_arch, dpi = 600, limitsize = FALSE
        )
        message("Saved: archaea_", file_slug, "_", tax_lvl_v, "_reads.png")

      } else {
        message("No archaeal reads detected for: ", enrich)
      }

      # ── Viral / bacteriophage panel ────────────────────────────────────────
      virus_sub <- micro_sub %>% filter(superkingdom == "Viruses")

      if (nrow(virus_sub) > 0) {
        virus_taxa <- virus_sub %>%
          group_by(taxon) %>% summarise(m = mean(rel_abund), .groups = "drop") %>%
          arrange(desc(m)) %>% pull(taxon)
        virus_fill <- setNames(
          colorRampPalette(virus_pal)(length(virus_taxa)), virus_taxa
        )

        virus_bar_df <- virus_sub %>%
          mutate(
            disp_label = factor(disp_map[sample], levels = disp_order),
            taxon      = factor(taxon, levels = virus_taxa)
          )

        p_virus_bar <- ggplot(virus_bar_df,
                              aes(x = disp_label, y = rel_abund, fill = taxon)) +
          geom_col(width = 0.7, colour = "white", linewidth = 0.3) +
          scale_fill_manual(values = virus_fill, name = "Genus") +
          scale_y_continuous(
            labels = function(x) paste0(x, "%"),
            expand = expansion(mult = c(0, 0.05))
          ) +
          facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +
          labs(
            title    = paste0("Viruses (including bacteriophage) - ", enrich_lbl, " (Reads)"),
            subtitle = paste0("Relative abundance of viral ", tolower(lvl_lbl_v),
                              "s (% of all classified bases)"),
            x = NULL, y = "Relative abundance (%)"
          ) +
          theme_bw(base_size = 12) +
          theme(
            plot.title    = element_text(face = "bold", size = 13),
            plot.subtitle = element_text(size = 10, colour = "grey40"),
            axis.text.x   = element_text(angle = 45, hjust = 1),
            legend.text   = element_text(size = 10, face = "italic"),
            strip.text    = element_text(face = "bold", size = 12),
            panel.spacing = unit(0.8, "lines")
          )

        virus_tile_df <- virus_sub %>%
          mutate(
            disp_label = factor(disp_map[sample], levels = disp_order),
            taxon      = factor(taxon, levels = rev(virus_taxa)),
            present    = rel_abund > 0
          )

        p_virus_tile <- ggplot(virus_tile_df,
                               aes(x = disp_label, y = taxon, fill = present)) +
          geom_tile(colour = "white", linewidth = 0.5) +
          scale_fill_manual(
            values = c("TRUE" = "#762A83", "FALSE" = "#F7F7F7"),
            labels = c("TRUE" = "Present", "FALSE" = "Absent"),
            name   = NULL
          ) +
          facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +
          labs(x = NULL, y = NULL,
               subtitle = "Purple = present; grey = absent") +
          theme_bw(base_size = 11) +
          theme(
            axis.text.x     = element_text(angle = 45, hjust = 1),
            axis.text.y     = element_text(face = "italic"),
            strip.text      = element_text(face = "bold", size = 11),
            panel.spacing   = unit(0.8, "lines"),
            legend.position = "bottom"
          )

        p_virus <- p_virus_bar / p_virus_tile +
          plot_layout(heights = c(2, 1)) +
          plot_annotation(
            title = paste0("Viruses (including bacteriophage) detected - ", enrich_lbl),
            theme = theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5))
          )
        print(p_virus)
        n_virus_taxa <- length(virus_taxa)
        h_virus <- if (n_virus_taxa > 80) max(8, n_virus_taxa * 0.22 + 3) else
                   if (n_virus_taxa > 40) max(8, n_virus_taxa * 0.32 + 3) else
                   max(8, n_virus_taxa * 0.45 + 3)
        ggsave(
          file.path(out_dir, paste0("viral_", file_slug, "_", tax_lvl_v, "_reads.png")),
          p_virus, width = 14, height = h_virus, dpi = 600, limitsize = FALSE
        )
        message("Saved: viral_", file_slug, "_", tax_lvl_v, "_reads.png")

      } else {
        message("No viral reads detected for: ", enrich)
      }
    }

    # ── Per-enrichment summary bubble plots (one per enrichment type) ─────────
    # Samples (CH01-CH06) on x-axis; Post-EVI/Post-ASC as facet rows.
    # Produces three focused, portrait-format plots rather than one huge figure.
    for (enrich_s in levels(df$enrichment_type)) {

      sum_sub <- df_micro %>%
        filter(enrichment_type == enrich_s) %>%
        mutate(disp_label = sample_labels[sample]) %>%
        group_by(superkingdom, taxon, sample_type, disp_label) %>%
        summarise(rel_abund = mean(rel_abund), .groups = "drop") %>%
        mutate(
          taxon_label = paste0(taxon, " (", superkingdom, ")"),
          kingdom     = factor(superkingdom, levels = c("Archaea", "Viruses")),
          disp_label  = factor(disp_label, levels = paste0("CH0", 1:6))
        )

      if (nrow(sum_sub) == 0) {
        message("No non-bacterial reads for: ", enrich_s); next
      }

      enrich_lbl_s <- enrichment_labels[enrich_s]
      file_slug_s  <- gsub("[^A-Za-z0-9]", "_", enrich_s)
      n_taxa_s     <- length(unique(sum_sub$taxon_label))

      # Count total columns across panels: with free_x, a taxon in both panels = 2 columns
      n_cols_total <- sum_sub %>%
        group_by(sample_type, taxon_label) %>%
        summarise(.groups = "drop") %>%
        nrow()

      # Bottom margin: longest label length drives the 45-degree drop-down height
      max_label_chars <- max(nchar(as.character(unique(sum_sub$taxon_label))), na.rm = TRUE)
      # at 10pt font (~3.5mm/char), 45-degree projection ≈ nchar * 2.5 mm; add 10 mm buffer
      bottom_mm <- max_label_chars * 2.5 + 10

      p_sum_s <- ggplot(
        sum_sub,
        aes(x = reorder(taxon_label, desc(taxon_label)),
            y = disp_label,
            size   = rel_abund,
            colour = kingdom)
      ) +
        geom_point(alpha = 0.85) +
        scale_size_continuous(
          trans  = "log10",
          range  = c(1, 6),
          name   = "Rel. abund (%)",
          breaks = c(0.001, 0.01, 0.1, 1),
          labels = c("0.001", "0.010", "0.100", "1.000")
        ) +
        scale_colour_manual(
          values = c("Archaea" = "#2166AC", "Viruses" = "#762A83"),
          name   = "Kingdom"
        ) +
        facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +
        scale_y_discrete(drop = FALSE) +
        labs(
          title    = paste0("Non-bacterial microbiome reads: Archaea and Viruses\n",
                            enrich_lbl_s),
          subtitle = "Bubble area proportional to relative abundance (% of all classified bases)",
          x = NULL, y = NULL
        ) +
        theme_bw(base_size = 12) +
        theme(
          plot.title       = element_text(face = "bold", size = 13, hjust = 0.5),
          plot.subtitle    = element_text(size = 10, colour = "grey40", hjust = 0.5),
          axis.text.x      = element_text(angle = 45, hjust = 1, face = "italic", size = 10),
          axis.text.y      = element_text(size = 11),
          strip.text       = element_text(face = "bold", size = 12),
          strip.background = element_rect(fill = "grey92", colour = "grey70"),
          panel.grid.major = element_line(colour = "grey88"),
          legend.position  = "right",
          legend.text      = element_text(size = 11),
          legend.title     = element_text(size = 12, face = "bold"),
          panel.spacing    = unit(1, "lines"),
          plot.margin      = unit(c(5, 15, bottom_mm, 5), "mm")
        )

      print(p_sum_s)
      # 0.45 inches per column (each panel column counted separately)
      w_sum <- n_cols_total * 0.45 + 4
      ggsave(
        file.path(out_dir, paste0("nonbacterial_microbiome_", file_slug_s,
                                  "_", tax_lvl_v, "_reads.png")),
        p_sum_s,
        width     = max(8, w_sum),
        height    = 9,
        dpi       = 600,
        limitsize = FALSE
      )
      message("Saved: nonbacterial_microbiome_", file_slug_s, "_", tax_lvl_v, "_reads.png")
    }
  }
  } # end for tax_lvl_v (genus / species)
}

```


---

# Avian Virus Reads

Bubble plot showing reads classified to known avian-associated virus genera across all sample sets
and enrichment types. These include major poultry pathogens (avian influenza, Newcastle disease,
infectious bronchitis, Marek's disease, Gumboro, etc.) and other viruses commonly detected in
poultry metagenomics. Bubble area is proportional to relative abundance on a log10 scale.

```{r avian_virus_reads, fig.height=8, fig.width=10}

if (!"superkingdom" %in% names(df_raw)) {
  message("NOTE: 'superkingdom' column not found - re-run kraken2_bpm.py. Skipping avian virus plots.")
} else {

  # ── Known avian-associated virus genera ──────────────────────────────────────
  avian_virus_genera <- c(
    "Alphainfluenzavirus",  # Avian influenza (H5N1, H7N9, etc.)
    "Gammacoronavirus",     # Infectious Bronchitis Virus (IBV)
    "Deltacoronavirus",     # Turkey / duck deltacoronaviruses
    "Orthoavulavirus",      # Newcastle Disease Virus (current classification)
    "Avulavirus",           # Newcastle Disease Virus (older classification)
    "Mardivirus",           # Marek's Disease Virus (MDV)
    "Avibirnavirus",        # Infectious Bursal Disease Virus (Gumboro)
    "Avipoxvirus",          # Fowlpox virus
    "Gyrovirus",            # Chicken Anaemia Virus (CAV)
    "Aviadenovirus",        # Fowl adenovirus (FAdV)
    "Siadenovirus",         # Turkey Hemorrhagic Enteritis Virus
    "Atadenovirus",         # Egg Drop Syndrome virus
    "Orthoreovirus",        # Avian reovirus (viral tenosynovitis)
    "Avastrovirus",         # Avian nephritis / turkey astrovirus
    "Aveparvovirus",        # Chicken / turkey parvovirus
    "Circovirus",           # Duck circovirus
    "Alpharetrovirus",      # Avian Leukosis Virus (ALV)
    "Gammaretrovirus",      # Reticuloendotheliosis virus
    "Dependoparvovirus",    # Avian dependoparvovirus
    "Rotavirus",            # Avian rotavirus (enteric disease)
    "Lagovirus",            # Reported in poultry-associated environments
    "Picobirnavirus"        # Enteric, detected in poultry microbiome studies
  )

  for (tax_lvl_v in c("genus", "species")) {
  lvl_lbl_v <- c(genus = "Genus", species = "Species")[tax_lvl_v]

  # Primary filter: exact genus / species match
  df_avian <- df_raw %>%
    filter(
      superkingdom == "Viruses",
      tax_level    == tax_lvl_v,
      taxon        %in% avian_virus_genera,
      rel_abund    > 0
    ) %>%
    mutate(
      sample_type = factor(
        if_else(grepl("^EVI", as.character(group)), "Post-EVI", "Post-ASC"),
        levels = c("Post-EVI", "Post-ASC")
      )
    )

  # Fallback: keyword search if exact genera not found (handles NCBI name variation)
  if (nrow(df_avian) == 0) {
    message("Exact genus names not matched - trying keyword fallback...")
    df_avian <- df_raw %>%
      filter(
        superkingdom == "Viruses",
        tax_level    == tax_lvl_v,
        str_detect(taxon, regex(
          "influenza|avi[a-z]|mardi|gyro|birna|avulo|reovirus|astrovirus|circovirus|parvovirus|corona|retrovirus|poxvirus",
          ignore_case = TRUE
        )),
        rel_abund > 0
      ) %>%
      mutate(
        sample_type = factor(
          if_else(grepl("^EVI", as.character(group)), "Post-EVI", "Post-ASC"),
          levels = c("Post-EVI", "Post-ASC")
        )
      )
  }

  if (nrow(df_avian) == 0) {
    message("No avian virus ", tolower(lvl_lbl_v), "s detected across any sample.")
  } else {

    message("Avian virus genera detected: ",
            paste(sort(unique(df_avian$taxon)), collapse = ", "))

    for (enrich_av in levels(df$enrichment_type)) {

      av_sub <- df_avian %>% filter(enrichment_type == enrich_av)
      if (nrow(av_sub) == 0) {
        message("No avian virus reads for: ", enrich_av); next
      }

      enrich_lbl_av <- enrichment_labels[enrich_av]
      file_slug_av  <- gsub("[^A-Za-z0-9]", "_", enrich_av)
      n_taxa_av     <- length(unique(av_sub$taxon))

      av_plot_df <- av_sub %>%
        mutate(disp_label = factor(sample_labels[sample], levels = paste0("CH0", 1:6))) %>%
        group_by(taxon, sample_type, disp_label) %>%
        summarise(rel_abund = mean(rel_abund), .groups = "drop")

      # Count columns per panel separately (free_x: same taxon in both panels = 2 columns)
      n_av_cols_total <- av_plot_df %>%
        group_by(sample_type, taxon) %>%
        summarise(.groups = "drop") %>%
        nrow()

      max_av_chars    <- max(nchar(as.character(unique(av_plot_df$taxon))), na.rm = TRUE)
      av_bottom_mm    <- max_av_chars * 2.5 + 10

      p_av <- ggplot(
        av_plot_df,
        aes(x    = reorder(taxon, desc(taxon)),
            y    = disp_label,
            size = rel_abund)
      ) +
        geom_point(alpha = 0.85, colour = "#D95F02") +
        scale_size_continuous(
          trans  = "log10",
          range  = c(1, 6),
          name   = "Rel. abund (%)",
          breaks = c(0.001, 0.01, 0.1, 1),
          labels = c("0.001", "0.010", "0.100", "1.000")
        ) +
        facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +
        scale_y_discrete(drop = FALSE) +
        labs(
          title    = paste0("Avian Virus Reads - ", enrich_lbl_av),
          subtitle = "Known avian-associated virus genera; bubble area proportional to relative abundance (log10 scale)",
          x = NULL, y = NULL
        ) +
        theme_bw(base_size = 12) +
        theme(
          plot.title       = element_text(face = "bold", size = 13, hjust = 0.5),
          plot.subtitle    = element_text(size = 10, colour = "grey40", hjust = 0.5),
          axis.text.x      = element_text(angle = 45, hjust = 1, face = "italic", size = 10),
          axis.text.y      = element_text(size = 11),
          strip.text       = element_text(face = "bold", size = 12),
          strip.background = element_rect(fill = "grey92", colour = "grey70"),
          panel.grid.major = element_line(colour = "grey88"),
          legend.position  = "right",
          legend.text      = element_text(size = 11),
          legend.title     = element_text(size = 12, face = "bold"),
          panel.spacing    = unit(1, "lines"),
          plot.margin      = unit(c(5, 15, av_bottom_mm, 5), "mm")
        )

      print(p_av)
      w_av <- n_av_cols_total * 0.45 + 4
      ggsave(
        file.path(out_dir, paste0("avian_virus_", file_slug_av, "_", tax_lvl_v, "_reads.png")),
        p_av,
        width     = max(8, w_av),
        height    = 9,
        dpi       = 600,
        limitsize = FALSE
      )
      message("Saved: avian_virus_", file_slug_av, "_", tax_lvl_v, "_reads.png")

      # Presence/absence tile (orange to match dot colour)
      av_sub_rename <- av_sub %>% rename(name = taxon)
      plot_genus_presence(
        av_sub_rename, spp_display_labels,
        title          = paste0("Avian Virus Presence/Absence (", lvl_lbl_v, ") - ",
                                enrich_lbl_av, " (Reads)"),
        subtitle       = "Orange = detected in sample; grey = absent",
        present_colour = "#D95F02",
        save_path      = file.path(out_dir, paste0("avian_virus_", file_slug_av,
                                                    "_", tax_lvl_v, "_presence.png"))
      )
    }
  }
  } # end for tax_lvl_v (genus / species)
}

```
