---
title: "Antimicrobial Resistance Gene and Taxonomy Associations"
subtitle: "Poultry Processing Plant: AMR Read-Level Analysis"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 12,
  fig.height = 12
)

# ── Output directory for saved plots ─────────────────────────────────────────
plot_dir <- "/Volumes/Alice/2_POULTRY_PROCESSING/Analysis/chord_plots_read"
dir.create(plot_dir, showWarnings = FALSE)

# Helper: save a circlize/base-R plot to PNG and PDF
save_chord <- function(filename, plot_fn, width = 14, height = 14, dpi = 300) {
  png(file.path(plot_dir, paste0(filename, ".png")),
      width = width, height = height, units = "in", res = dpi)
  plot_fn()
  dev.off()

  png(file.path(plot_dir, paste0(filename, "_nobg.png")),
      width = width, height = height, units = "in", res = dpi,
      bg = "transparent")
  plot_fn()
  dev.off()

  pdf(file.path(plot_dir, paste0(filename, ".pdf")),
      width = width, height = height)
  plot_fn()
  dev.off()
}
```

## Overview

Each row in the source TSV files is a single nanopore read with:
- a Kraken2 taxonomic classification, and
- an ABRicate resistance gene hit (across 7 databases: resfinder, card, ncbi,
  megares, ecoli_vf, vfdb, plasmidfinder).

The chord diagram connects **resistance gene classes** (left arc) to
**bacterial genera** (right arc). Chord width = number of reads carrying that
gene-class/genus combination. Genes are coloured by antibiotic class; genera
are coloured by phylum.

**Sample key:**

| Treatment | Description |
|---|---|
| EVI_RVS | Post-evisceration, RVS broth enrichment |
| EVI_CAT | Post-evisceration, BF Bolton + CAT enrichment |
| ASC_RVS | Post-acidified sodium chlorite (ASC) wash, RVS broth enrichment |
| ASC_CAT | Post-acidified sodium chlorite (ASC) wash, BF Bolton + CAT enrichment |

---

## 1. Setup

```{r packages}
library(tidyverse)
library(circlize)
library(RColorBrewer)
library(viridis)
library(knitr)
library(kableExtra)
library(pheatmap)
library(reshape2)
library(ggrepel)
```

---

## 2. Load and Combine Data

```{r load-data}
# ── Path to the folder containing per-treatment sub-directories ──────────────
data_dir <- "/Volumes/Alice/2_POULTRY_PROCESSING/Analysis/Read_level_kraken-to-abricate_associations"

# Auto-detect all TSV files recursively
tsv_files <- list.files(data_dir, pattern = "\\.tsv$",
                         full.names = TRUE, recursive = TRUE)

cat("Found", length(tsv_files), "TSV files:\n")
cat(paste(" ", basename(tsv_files)), sep = "\n")

# Extract treatment and sample ID from file name
# Expected pattern: CH01_ASC_CAT_all_databases_with_taxonomy.tsv
parse_filename <- function(path) {
  base <- basename(path)
  # Extract sample (e.g. CH01) and treatment (e.g. ASC_CAT)
  sample_id  <- str_extract(base, "^[A-Z]+[0-9]+")
  treatment  <- str_extract(base, "(?<=_)[A-Z]+_[A-Z]+(?=_all)")
  data.frame(file = path, sample_id = sample_id, treatment = treatment,
             stringsAsFactors = FALSE)
}

file_meta <- map_dfr(tsv_files, parse_filename)

raw <- map2_dfr(
  file_meta$file, seq_len(nrow(file_meta)),
  function(path, i) {
    df <- read_tsv(path, col_types = cols(.default = "c"),
                   show_col_types = FALSE)
    df$sample_id  <- file_meta$sample_id[i]
    df$treatment  <- file_meta$treatment[i]
    df
  }
) %>%
  mutate(
    `%COVERAGE` = as.numeric(`%COVERAGE`),
    `%IDENTITY` = as.numeric(`%IDENTITY`)
  )

cat("\nTotal rows loaded:", nrow(raw), "\n")
cat("Treatments found:", paste(unique(raw$treatment), collapse = ", "), "\n")
```

---

## 3. Filter to NCBI

**NCBI AMRFinderPlus** is used as the single source database for resistance gene
annotations. NCBI gene names are already in standard literature nomenclature
(e.g. `tet(A)`, `erm(B)`, `blaCTX-M-1`) and the RESISTANCE column is consistently
populated, making no further name mapping or class inference necessary.

Each read is used as-is - one row per read. Allele-version suffixes (e.g. `-1`, `-15`)
are retained as they carry biologically meaningful allele information.

```{r dedup}
primary_db <- "ncbi"

cat("Databases present in raw data:", paste(sort(unique(raw$DATABASE)), collapse = ", "), "\n")

dedup <- raw %>%
  filter(DATABASE == primary_db) %>%
  filter(!is.na(GENE), GENE != "") %>%
  mutate(
    gene_norm  = str_trim(GENE),   # preserve original case - NCBI names are already conventional
    res_source = coalesce(RESISTANCE, PRODUCT, "")
  )

cat("\nDatabase used:", primary_db, "\n")
cat("Rows loaded:", nrow(dedup), "\n")
cat("Unique genes:", n_distinct(dedup$gene_norm), "\n")
cat("Unique reads with a hit:", n_distinct(dedup$READ_ID), "\n")
```

---

## 4. Normalise Resistance Classes

The RESISTANCE field has inconsistent naming across databases. Map all
variants to standard antibiotic class names.

```{r resistance-normalise}
normalise_resistance <- function(res) {
  res <- str_to_lower(res)
  case_when(
    # beta-lactam: "beta.lactam" matches hyphenated/spaced forms; "betalactam" (no separator)
    # matches MEGARes PRODUCT hierarchy (e.g. "Drugs:betalactams:Class_A_betalactamases:CTX")
    str_detect(res, "beta.?lactam|betalactam|cephalosporin|penam|carbapenem|penicillin|monobactam|cephamycin|blaec|blac|blad|ctxm|shv|tem:|oxacillin") ~ "beta-lactam",
    str_detect(res, "tetracycline|doxycycline|minocycline|tigecycline")                                ~ "tetracycline",
    str_detect(res, "aminoglycoside|amikacin|gentamicin|kanamycin|streptomycin|tobramycin")            ~ "aminoglycoside",
    str_detect(res, "fluoroquinolone|quinolone|ciprofloxacin|nalidixic")                               ~ "quinolone",
    # MLS = Macrolide-Lincosamide-Streptogramin (MEGARes uses "MLS" as a combined category)
    str_detect(res, "macrolide|erythromycin|azithromycin|clarithromycin|mls:|:mls")                   ~ "macrolide",
    str_detect(res, "phenicol|chloramphenicol")                                                        ~ "phenicol",
    str_detect(res, "lincosamide|lincomycin|clindamycin")                                              ~ "lincosamide",
    str_detect(res, "streptogramin|dalfopristin|virginiamycin|pristinamycin")                          ~ "streptogramin",
    str_detect(res, "sulfonamide|sulphonamide|sulph|sulf")                                             ~ "sulphonamide",
    str_detect(res, "trimethoprim")                                                                    ~ "trimethoprim",
    str_detect(res, "glycopeptide|vancomycin|teicoplanin")                                             ~ "glycopeptide",
    str_detect(res, "oxazolidinone|linezolid")                                                         ~ "oxazolidinone",
    str_detect(res, "pleuromutilin|tiamulin")                                                          ~ "pleuromutilin",
    str_detect(res, "rifamycin|rifampicin")                                                            ~ "rifamycin",
    str_detect(res, "fosfomycin")                                                                      ~ "fosfomycin",
    str_detect(res, "fusidic")                                                                         ~ "fusidic acid",
    str_detect(res, "disinfectant|biocide|qac|quaternary")                                            ~ "disinfectant/biocide",
    # Cationic antimicrobial peptides (colistin/polymyxin resistance via lipid A modification)
    # MEGARes PRODUCT: "Drugs:Cationic_antimicrobial_peptides:Lipid_A_modification:PMRF/EPTA/UGD"
    str_detect(res, "cationic|polymyxin|colistin|lipid_a|pmrf|epta|arnb|peta")                       ~ "peptide antibiotic",
    # Bacitracin resistance (undecaprenyl pyrophosphate phosphatase genes - BacA)
    str_detect(res, "bacitracin|baca|undecaprenyl")                                                   ~ "bacitracin",
    str_detect(res, "streptothricin|sat:")                                                             ~ "streptothricin",
    str_detect(res, "multi|efflux")                                                                    ~ "multidrug efflux",
    TRUE ~ "other"
  )
}

dedup <- dedup %>%
  # MEGARes does not populate the RESISTANCE column in ABRicate output.
  # res_source was set in Section 3 to PRODUCT (the resistance description field)
  # when RESISTANCE is blank, so normalise_resistance works correctly either way.
  mutate(res_class = normalise_resistance(res_source))

cat("Resistance class distribution:\n")
print(sort(table(dedup$res_class), decreasing = TRUE))
```

---

## 5. Clean Taxonomy and Assign Genus + Phylum

Reads are annotated at three taxonomic levels:

- **Family** - broadest resolution; all genus/species reads are mapped up to family
- **Genus** - first word of the Kraken2 taxonomy string
- **Species** - reads with a two-word taxonomy (genus + species epithet)

Reads classified only to root, phylum, class, order, or family level by Kraken2
cannot be resolved to genus and are grouped as **"Unclassified"** in each chord diagram.

```{r taxonomy-clean}
# ── Genus → Phylum lookup ─────────────────────────────────────────────────────
phylum_lookup <- c(
  "Campylobacter"          = "Campylobacterota",
  "Arcobacter"             = "Campylobacterota",
  "Helicobacter"           = "Campylobacterota",
  "Escherichia"            = "Proteobacteria",
  "Salmonella"             = "Proteobacteria",
  "Comamonas"              = "Proteobacteria",
  "Pseudomonas"            = "Proteobacteria",
  "Acinetobacter"          = "Proteobacteria",
  "Klebsiella"             = "Proteobacteria",
  "Enterobacter"           = "Proteobacteria",
  "Citrobacter"            = "Proteobacteria",
  "Hafnia"                 = "Proteobacteria",
  "Serratia"               = "Proteobacteria",
  "Morganella"             = "Proteobacteria",
  "Vibrio"                 = "Proteobacteria",
  "Candidatus"             = "Proteobacteria",
  "Lactobacillus"          = "Firmicutes",
  "Lactococcus"            = "Firmicutes",
  "Megamonas"              = "Firmicutes",
  "Enterococcus"           = "Firmicutes",
  "Staphylococcus"         = "Firmicutes",
  "Listeria"               = "Firmicutes",
  "Clostridium"            = "Firmicutes",
  "Ruminococcus"           = "Firmicutes",
  "Faecalibacterium"       = "Firmicutes",
  "Blautia"                = "Firmicutes",
  "Erysipelatoclostridium" = "Firmicutes",
  "Streptococcus"          = "Firmicutes",
  "Bacillus"               = "Firmicutes",
  "Ligilactobacillus"      = "Firmicutes",
  "Limosilactobacillus"    = "Firmicutes",
  "Bacteroides"            = "Bacteroidota",
  "Prevotella"             = "Bacteroidota",
  "Parabacteroides"        = "Bacteroidota",
  "Alistipes"              = "Bacteroidota",
  "Bifidobacterium"        = "Actinomycetota",
  "Corynebacterium"        = "Actinomycetota",
  "Propionibacterium"      = "Actinomycetota",
  "Cutibacterium"          = "Actinomycetota"
)

# ── Genus → Family lookup ─────────────────────────────────────────────────────
family_lookup <- c(
  "Campylobacter"          = "Campylobacteraceae",
  "Arcobacter"             = "Arcobacteraceae",
  "Helicobacter"           = "Helicobacteraceae",
  "Escherichia"            = "Enterobacteriaceae",
  "Salmonella"             = "Enterobacteriaceae",
  "Klebsiella"             = "Enterobacteriaceae",
  "Enterobacter"           = "Enterobacteriaceae",
  "Citrobacter"            = "Enterobacteriaceae",
  "Hafnia"                 = "Enterobacteriaceae",
  "Serratia"               = "Enterobacteriaceae",
  "Morganella"             = "Enterobacteriaceae",
  "Comamonas"              = "Comamonadaceae",
  "Pseudomonas"            = "Pseudomonadaceae",
  "Acinetobacter"          = "Moraxellaceae",
  "Vibrio"                 = "Vibrionaceae",
  "Candidatus"             = "Unclassified",
  "Lactobacillus"          = "Lactobacillaceae",
  "Ligilactobacillus"      = "Lactobacillaceae",
  "Limosilactobacillus"    = "Lactobacillaceae",
  "Lactococcus"            = "Streptococcaceae",
  "Streptococcus"          = "Streptococcaceae",
  "Enterococcus"           = "Enterococcaceae",
  "Staphylococcus"         = "Staphylococcaceae",
  "Listeria"               = "Listeriaceae",
  "Bacillus"               = "Bacillaceae",
  "Megamonas"              = "Veillonellaceae",
  "Clostridium"            = "Clostridiaceae",
  "Faecalibacterium"       = "Oscillospiraceae",
  "Ruminococcus"           = "Oscillospiraceae",
  "Blautia"                = "Lachnospiraceae",
  "Erysipelatoclostridium" = "Erysipelotrichaceae",
  "Bacteroides"            = "Bacteroidaceae",
  "Prevotella"             = "Prevotellaceae",
  "Parabacteroides"        = "Tannerellaceae",
  "Alistipes"              = "Rikenellaceae",
  "Bifidobacterium"        = "Bifidobacteriaceae",
  "Corynebacterium"        = "Corynebacteriaceae",
  "Propionibacterium"      = "Propionibacteriaceae",
  "Cutibacterium"          = "Propionibacteriaceae"
)

# ── Reverse lookup: family name → phylum (for reads classified at family level)
# Kraken2 can classify a read to any rank. A read labelled "Enterobacteriaceae"
# (family level) has word(1) = "Enterobacteriaceae", which is not in phylum_lookup
# (genus keys only). Without this reverse map, those reads fall through to
# "Unclassified" even in the family chord - losing real signal.
family_to_phylum <- family_lookup %>%
  enframe(name = "genus_key", value = "family_name") %>%
  mutate(phylum_val = phylum_lookup[genus_key]) %>%
  distinct(family_name, phylum_val) %>%
  { setNames(.$phylum_val, .$family_name) }

known_families <- names(family_to_phylum)   # e.g. "Enterobacteriaceae", etc.

# ── Assign all three levels ───────────────────────────────────────────────────
# Rollup logic:
#   Species read  ("Escherichia coli")   → genus = "Escherichia", family via lookup ✓
#   Genus read    ("Escherichia")         → genus = "Escherichia", family via lookup ✓
#   Family read   ("Enterobacteriaceae")  → genus = "Unclassified", family = name  ✓ (fixed)
#   Higher / root                         → Unclassified at all levels              ✓
taxonomy_clean <- dedup %>%
  filter(!is.na(TAXONOMY), TAXONOMY != "") %>%
  mutate(
    tax1 = word(TAXONOMY, 1),   # candidate genus (or family/higher for coarse reads)

    # Genus: recognised only when first word is a known genus
    genus  = if_else(tax1 %in% names(phylum_lookup), tax1, "Unclassified"),
    phylum = if_else(genus != "Unclassified", phylum_lookup[genus], "Unclassified"),

    # Family: from resolved genus OR, when genus is unknown, from a direct family match
    family = case_when(
      genus != "Unclassified" ~ coalesce(family_lookup[genus], "Unclassified"),
      tax1  %in% known_families ~ tax1,   # Kraken2 stopped at family level
      TRUE                      ~ "Unclassified"
    ),

    # Phylum for family-only reads (needed to colour family sectors correctly)
    phylum = case_when(
      phylum != "Unclassified"  ~ phylum,
      family %in% known_families ~ coalesce(family_to_phylum[family], "Unclassified"),
      TRUE                       ~ "Unclassified"
    ),

    # Species: genus + epithet when the read has two words and genus is resolved
    species = case_when(
      genus == "Unclassified" ~ "Unclassified",
      str_count(str_trim(TAXONOMY), "\\S+") >= 2 ~
        paste(word(TAXONOMY, 1), word(TAXONOMY, 2)),
      TRUE ~ "Unclassified"
    )
  ) %>%
  select(-tax1)

# ── Resolution summary across levels ─────────────────────────────────────────
res_summary <- taxonomy_clean %>%
  group_by(treatment) %>%
  summarise(
    total          = n(),
    family_resolved  = sum(family  != "Unclassified"),
    genus_resolved   = sum(genus   != "Unclassified"),
    species_resolved = sum(species != "Unclassified"),
    .groups = "drop"
  ) %>%
  mutate(across(ends_with("resolved"),
                ~ paste0(.x, " (", round(100 * .x / total, 1), "%)"),
                .names = "{.col}_pct")) %>%
  select(Treatment = treatment,
         `Family resolved`  = family_resolved_pct,
         `Genus resolved`   = genus_resolved_pct,
         `Species resolved` = species_resolved_pct)

kable(res_summary, caption = "Read counts resolved at each taxonomic level")
```

---

## 6. Build Count Matrices

```{r build-matrices}
# taxon_col: "family", "genus", or "species"
# Resistance-class level (rows = res_class)
build_mat <- function(df, taxon_col = "genus", min_reads = 3) {
  mat <- df %>%
    count(res_class, taxon = .data[[taxon_col]]) %>%
    pivot_wider(names_from = taxon, values_from = n, values_fill = 0L) %>%
    column_to_rownames("res_class") %>%
    as.matrix()
  mat <- mat[rowSums(mat) >= min_reads, colSums(mat) >= min_reads, drop = FALSE]
  mat
}

# Gene level (rows = individual gene_norm names)
# min_reads filters to genes with enough reads to be visible in the diagram
build_mat_gene <- function(df, taxon_col = "genus", min_reads = 5, top_n = NULL) {
  mat <- df %>%
    count(gene_norm, taxon = .data[[taxon_col]]) %>%
    pivot_wider(names_from = taxon, values_from = n, values_fill = 0L) %>%
    column_to_rownames("gene_norm") %>%
    as.matrix()
  mat <- mat[rowSums(mat) >= min_reads, colSums(mat) >= min_reads, drop = FALSE]
  # Optionally keep only the top N genes by total read count (keeps diagrams readable)
  if (!is.null(top_n) && nrow(mat) > top_n) {
    top_genes <- names(sort(rowSums(mat), decreasing = TRUE)[seq_len(top_n)])
    mat <- mat[top_genes, , drop = FALSE]
  }
  mat
}

# Class-level matrices
mat_family  <- build_mat(taxonomy_clean, "family")
mat_genus   <- build_mat(taxonomy_clean, "genus")
mat_species <- build_mat(taxonomy_clean, "species")

# Gene-level matrices
mat_gene_family  <- build_mat_gene(taxonomy_clean, "family")
mat_gene_genus   <- build_mat_gene(taxonomy_clean, "genus")
mat_gene_species <- build_mat_gene(taxonomy_clean, "species")

cat("── Class-level matrices ──\n")
cat("Family  matrix:", nrow(mat_family),  "×", ncol(mat_family),  "\n")
cat("Genus   matrix:", nrow(mat_genus),   "×", ncol(mat_genus),   "\n")
cat("Species matrix:", nrow(mat_species), "×", ncol(mat_species), "\n")
cat("── Gene-level matrices ──\n")
cat("Family  matrix:", nrow(mat_gene_family),  "×", ncol(mat_gene_family),  "\n")
cat("Genus   matrix:", nrow(mat_gene_genus),   "×", ncol(mat_gene_genus),   "\n")
cat("Species matrix:", nrow(mat_gene_species), "×", ncol(mat_gene_species), "\n")
```

---

## 7. Colour Palettes

```{r colours}
# ── Resistance class colours ──────────────────────────────────────────────────
res_class_colours <- c(
  "beta-lactam"         = "#5B7FBF",
  "tetracycline"        = "#3A9E4D",
  "aminoglycoside"      = "#E8821A",
  "quinolone"           = "#1F3A7A",
  "macrolide"           = "#8DB4E2",
  "phenicol"            = "#404040",
  "lincosamide"         = "#E8A0C8",
  "streptogramin"       = "#D46FA0",
  "sulphonamide"        = "#5BC8D2",
  "trimethoprim"        = "#1B6B2A",
  "glycopeptide"        = "#6B3FA0",
  "oxazolidinone"       = "#A07850",
  "pleuromutilin"       = "#E0B040",
  "rifamycin"           = "#B05030",
  "fosfomycin"          = "#E060C0",
  "fusidic acid"        = "#D08050",
  "disinfectant/biocide"= "#505050",
  "multidrug efflux"    = "#909090",
  "peptide antibiotic"  = "#2E8B8B",
  "bacitracin"          = "#8B6914",
  "streptothricin"      = "#7A3F00",
  "other"               = "#D0D0D0"
)

# ── Phylum colours (shared across all taxonomic levels) ──────────────────────
phylum_colours <- c(
  "Campylobacterota" = "#2E9E9E",
  "Proteobacteria"   = "#E07820",
  "Firmicutes"       = "#C83878",
  "Bacteroidota"     = "#D4B840",
  "Actinomycetota"   = "#7B4FCF",
  "Unclassified"     = "#C8C8C8"
)

# ── Per-level taxon → phylum → colour lookups ─────────────────────────────────
# Family
family_phylum_tbl <- taxonomy_clean %>%
  distinct(family, phylum) %>%
  filter(family != "Unclassified") %>%
  bind_rows(tibble(family = "Unclassified", phylum = "Unclassified"))

family_colours  <- setNames(phylum_colours[family_phylum_tbl$phylum],
                             family_phylum_tbl$family)
family_phylum_v <- setNames(family_phylum_tbl$phylum, family_phylum_tbl$family)

# Genus
genus_phylum_tbl <- taxonomy_clean %>% distinct(genus, phylum)
genus_colours    <- setNames(phylum_colours[genus_phylum_tbl$phylum],
                              genus_phylum_tbl$genus)
genus_phylum_v   <- setNames(genus_phylum_tbl$phylum, genus_phylum_tbl$genus)

# Supplement with ALL genera from phylum_lookup so that feature-specific databases
# (VFDB, PlasmidFinder, etc.) whose genera may not appear in the MEGARes data still
# get correct phylum assignments in draw_chord's bottomright legend.
missing_genera <- setdiff(names(phylum_lookup), names(genus_phylum_v))
if (length(missing_genera) > 0) {
  genus_phylum_v <- c(genus_phylum_v, phylum_lookup[missing_genera])
  genus_colours  <- c(genus_colours,
                      setNames(phylum_colours[phylum_lookup[missing_genera]],
                               missing_genera))
}

# Species
species_phylum_tbl <- taxonomy_clean %>%
  distinct(species, phylum) %>%
  filter(species != "Unclassified") %>%
  bind_rows(tibble(species = "Unclassified", phylum = "Unclassified"))

species_colours  <- setNames(phylum_colours[species_phylum_tbl$phylum],
                              species_phylum_tbl$species)
species_phylum_v <- setNames(species_phylum_tbl$phylum, species_phylum_tbl$species)

# ── Gene → colour (genes are coloured by their resistance class) ──────────────
gene_class_lut <- taxonomy_clean %>% distinct(gene_norm, res_class)
gene_row_colours <- setNames(res_class_colours[gene_class_lut$res_class],
                              gene_class_lut$gene_norm)
gene_row_colours[is.na(gene_row_colours)] <- "#AAAAAA"

# ── Sector colour builder (works for any level) ───────────────────────────────
# row_colours: named colour vector for row sectors (res_class_colours or gene_row_colours)
make_sector_colours <- function(mat, taxon_colours, row_colours = res_class_colours) {
  rc <- row_colours[rownames(mat)]
  tc <- taxon_colours[colnames(mat)]
  rc[is.na(rc)] <- "#AAAAAA"
  tc[is.na(tc)] <- "#C8C8C8"
  c(rc, tc)
}
```

---

## 8. Chord Diagrams

```{r draw-chord-function}
# taxon_colours  : named vector of colours for each taxon sector
# phylum_of_taxon: named vector mapping taxon name → phylum (for legend)
# level_label    : string shown in the phylum legend title ("Family" / "Genus" / "Species")
# row_colours    : named vector for row sectors; defaults to res_class_colours (class-level);
#                  pass gene_row_colours for gene-level diagrams
draw_chord <- function(mat, title = "",
                        taxon_colours, phylum_of_taxon,
                        level_label = "Taxon",
                        row_colours = res_class_colours,
                        link.sort = TRUE,
                        scale_fn = sqrt,
                        row_legend_cols  = NULL,   # named colour vector for custom legend
                        row_legend_title = "Antibiotic resistance\ngene class",
                        label_cex          = 1.1,   # sector label size
                        legend_cex         = 1.1,   # legend text size
                        label_track        = 30,    # label track height in mm
                        label_y_frac       = 0.5,   # 0 = inner edge (close to circle), 0.5 = middle
                        canvas_expand      = 1.0,   # expand canvas beyond default c(-1,1)
                        show_phylum_legend = TRUE) {# set FALSE to hide phylum legend

  # Drop any rows/columns whose name is NA or empty - these would cause
  # "length of dimnames not equal to array extent" inside chordDiagram
  mat <- mat[!is.na(rownames(mat)) & rownames(mat) != "", ,  drop = FALSE]
  mat <- mat[ , !is.na(colnames(mat)) & colnames(mat) != "", drop = FALSE]
  if (nrow(mat) < 2 || ncol(mat) < 2) {
    message("draw_chord: fewer than 2 valid rows or columns after NA removal - skipping.")
    return(invisible(NULL))
  }

  sector_col <- make_sector_colours(mat, taxon_colours, row_colours)

  row_ord <- rownames(mat)[order(rowSums(mat), decreasing = TRUE)]
  col_ord <- colnames(mat)[order(colSums(mat), decreasing = TRUE)]
  mat_ord <- mat[row_ord, col_ord, drop = FALSE]

  # Apply scale transformation to compress dynamic range so low-count chords
  # remain visible. Defaults to sqrt(); pass scale_fn = NULL to use raw counts.
  if (!is.null(scale_fn)) {
    mat_plot <- scale_fn(mat_ord + 1) - scale_fn(1)  # shift so 0 stays 0
  } else {
    mat_plot <- mat_ord
  }

  circos.clear()
  # Expand margins to give labels breathing room (scaled with canvas_expand)
  mar_outer <- 2 + 3 * (canvas_expand - 1)
  par(mar = c(mar_outer, mar_outer, mar_outer + 2, mar_outer))

  # Adaptive gap: cap total gap space at ~120° so many-sector diagrams don't overflow.
  # circos.par() must be called AFTER circos.clear() to take effect.
  n_sectors <- nrow(mat_ord) + ncol(mat_ord)
  gap_deg   <- max(0.3, min(2, 120 / n_sectors))
  circos.par(
    gap.after    = gap_deg,
    canvas.xlim  = c(-canvas_expand, canvas_expand),
    canvas.ylim  = c(-canvas_expand, canvas_expand)
  )

  chordDiagram(
    mat_plot,
    grid.col          = sector_col,
    transparency      = 0.35,
    link.sort         = link.sort,
    link.decreasing   = TRUE,
    annotationTrack   = NULL,
    preAllocateTracks = list(
      list(track.height = mm_h(label_track)),  # track 1 (outermost): labels
      list(track.height = mm_h(7))             # track 2: coloured blocks
    ),
    directional = 0
  )

  # Track 2: coloured sector blocks
  circos.trackPlotRegion(
    track.index = 2,
    panel.fun   = function(x, y) {
      sec  <- get.cell.meta.data("sector.index")
      xlim <- get.cell.meta.data("xlim")
      ylim <- get.cell.meta.data("ylim")
      circos.rect(xlim[1], ylim[1], xlim[2], ylim[2],
                  col = sector_col[sec], border = "white")
    },
    bg.border = NA
  )

  # Track 1: labels outside coloured blocks
  circos.trackPlotRegion(
    track.index = 1,
    panel.fun   = function(x, y) {
      sec  <- get.cell.meta.data("sector.index")
      xlim <- get.cell.meta.data("xlim")
      ylim <- get.cell.meta.data("ylim")
      y_pos <- ylim[1] + (ylim[2] - ylim[1]) * label_y_frac
      circos.text(
        x = mean(xlim), y = y_pos, labels = sec,
        facing = "clockwise", niceFacing = TRUE,
        cex = label_cex, col = "black", adj = c(0, 0.5)
      )
    },
    bg.border = NA
  )

  title(title, cex.main = 1.2, line = 0.5)

  # Legends
  phyla_present <- unique(na.omit(phylum_of_taxon[colnames(mat_ord)]))

  # Row legend: use supplied palette when provided; otherwise reverse-match
  # colours back to res_class_colours (works for class- and gene-level AMR diagrams)
  if (!is.null(row_legend_cols)) {
    leg_cols  <- row_legend_cols
    leg_names <- names(row_legend_cols)
  } else {
    row_cols_used <- unname(row_colours[rownames(mat_ord)])
    row_cols_used <- row_cols_used[!is.na(row_cols_used)]
    leg_names     <- names(res_class_colours)[res_class_colours %in% row_cols_used]
    leg_cols      <- res_class_colours[leg_names]
  }

  legend("bottomleft",
    legend = leg_names, fill = leg_cols,
    title  = row_legend_title,
    bty = "n", cex = legend_cex, x.intersp = 0.8)

  if (show_phylum_legend && length(phyla_present) > 0) {
    legend("bottomright",
      legend = phyla_present, fill = phylum_colours[phyla_present],
      title  = paste("Phylum\n(", level_label, "level)"),
      bty = "n", cex = legend_cex, x.intersp = 0.8)
  }
}
```

### 8.1 Family Level - All Samples

```{r chord-family-all, fig.width=14, fig.height=14}
draw_chord(mat_family, "AMR Resistance Class by Family (All Samples)",
           family_colours, family_phylum_v, "Family")

save_chord("chord_class_family_all",
           function() draw_chord(mat_family,
             "AMR Resistance Class by Family (All Samples)",
             family_colours, family_phylum_v, "Family"))
```

### 8.2 Genus Level - All Samples

```{r chord-genus-all, fig.width=14, fig.height=14}
draw_chord(mat_genus, "AMR Resistance Class by Genus (All Samples)",
           genus_colours, genus_phylum_v, "Genus")

save_chord("chord_class_genus_all",
           function() draw_chord(mat_genus,
             "AMR Resistance Class by Genus (All Samples)",
             genus_colours, genus_phylum_v, "Genus"))
```

### 8.3 Species Level - All Samples

```{r chord-species-all, fig.width=14, fig.height=14}
if (ncol(mat_species) >= 2 && nrow(mat_species) >= 2) {
  draw_chord(mat_species, "AMR Resistance Class by Species (All Samples)",
             species_colours, species_phylum_v, "Species")

  save_chord("chord_class_species_all",
             function() draw_chord(mat_species,
               "AMR Resistance Class by Species (All Samples)",
               species_colours, species_phylum_v, "Species"))
} else {
  cat("Insufficient species-level data for combined chord diagram.\n")
}
```

### 8.4 Per-treatment Chord Diagrams - Resistance Class Level

```{r chord-per-treatment, fig.width=13, fig.height=13, results='asis'}
treatments <- sort(unique(taxonomy_clean$treatment))

for (trt in treatments) {
  df_trt   <- taxonomy_clean %>% filter(treatment == trt)
  trt_safe <- gsub("[^A-Za-z0-9]", "_", trt)

  for (lvl in c("family", "genus", "species")) {
    tc <- switch(lvl,
                 family  = family_colours,
                 genus   = genus_colours,
                 species = species_colours)
    pv <- switch(lvl,
                 family  = family_phylum_v,
                 genus   = genus_phylum_v,
                 species = species_phylum_v)

    mat_trt <- build_mat(df_trt, taxon_col = lvl, min_reads = 2)

    if (nrow(mat_trt) < 2 || ncol(mat_trt) < 2) next

    heading <- paste0("\n#### ", trt, " - ", str_to_title(lvl), " level\n\n")
    cat(heading)

    ttl   <- paste0("AMR Resistance Class by ", str_to_title(lvl), ": ", trt)
    fname <- paste0("chord_class_", trt_safe, "_", lvl)

    draw_chord(mat_trt, ttl, tc, pv, str_to_title(lvl))

    local({
      m <- mat_trt; t <- ttl; colours <- tc; phyla <- pv; lbl <- str_to_title(lvl); fn <- fname
      save_chord(fn, function() draw_chord(m, t, colours, phyla, lbl))
    })

    cat("\n\n")
  }
}
```

---

## Gene-level Chord Diagrams

Individual resistance genes (coloured by their resistance class) linked to taxonomy.
Genes with fewer than 5 reads across all samples are excluded for readability.

### 8.5 Gene Level - Family - All Samples

```{r chord-gene-family-all, fig.width=16, fig.height=16}
if (nrow(mat_gene_family) >= 2 && ncol(mat_gene_family) >= 2) {
  draw_chord(mat_gene_family,
             "AMR Resistance Gene by Family (All Samples)",
             family_colours, family_phylum_v, "Family",
             row_colours = gene_row_colours)

  save_chord("chord_gene_family_all",
             function() draw_chord(mat_gene_family,
               "AMR Resistance Gene by Family (All Samples)",
               family_colours, family_phylum_v, "Family",
               row_colours = gene_row_colours),
             width = 16, height = 16)
} else {
  cat("Insufficient gene-level family data for combined chord diagram.\n")
}
```

### 8.5a Gene by Family - Graphical Abstract Version

Clinically concerning acquired resistance genes only; intrinsic/chromosomal
genes are excluded. Remaining genes are ranked by read count and the top 20
are shown. Larger labels and legends ensure legibility at reduced scale.
Saved as `chord_gene_family_graphical_abstract_nobg.png` (transparent background).

**Excluded gene groups (intrinsic/constitutive — not acquired):**

| Pattern | Gene / organism | Reason excluded |
|---|---|---|
| `blaEC-*` | Chromosomal AmpC | Intrinsic in *Escherichia coli* |
| `cepA` | *Bacteroides* β-lactamase | Intrinsic in *Bacteroides* spp. |
| `cblA` | *Bacteroides* β-lactamase | Intrinsic in *Bacteroides* spp. |
| `oqxA`, `oqxB*` | OqxAB efflux | Chromosomal in *Klebsiella pneumoniae* |
| `fosA*` | Fosfomycin resistance | Chromosomal in *Klebsiella* / *Pseudomonas* |

```{r chord-gene-family-graphical-abstract, fig.width=20, fig.height=20}
# ── Intrinsic/constitutive gene patterns to exclude ───────────────────────────
# These are chromosomally encoded in their primary host and do not represent
# acquired, mobile resistance of clinical concern.
intrinsic_regex <- paste(
  "^blaEC",      # chromosomal AmpC — intrinsic in E. coli
  "^cepA$",      # intrinsic beta-lactamase in Bacteroides spp.
  "^cblA$",      # intrinsic beta-lactamase in Bacteroides spp.
  "^oqxA",       # OqxAB efflux pump — chromosomal in K. pneumoniae
  "^oqxB",       # OqxAB efflux pump — chromosomal in K. pneumoniae
  "^fosA",       # fosfomycin resistance — chromosomal in Klebsiella / Pseudomonas
  sep = "|"
)

mat_ga_full <- build_mat_gene(taxonomy_clean, "family", min_reads = 3)

# Remove intrinsic gene rows
mat_ga_concern <- mat_ga_full[!grepl(intrinsic_regex, rownames(mat_ga_full),
                                      ignore.case = TRUE), , drop = FALSE]

# Drop columns that become empty after gene removal, then keep top 20 genes
mat_ga_concern <- mat_ga_concern[, colSums(mat_ga_concern) > 0, drop = FALSE]
top_genes <- names(sort(rowSums(mat_ga_concern), decreasing = TRUE)
                   [seq_len(min(20, nrow(mat_ga_concern)))])
mat_ga <- mat_ga_concern[top_genes, , drop = FALSE]

cat("Genes excluded (intrinsic):\n")
cat(paste(" ", grep(intrinsic_regex, rownames(mat_ga_full),
                     value = TRUE, ignore.case = TRUE)), sep = "\n")
cat("\nGenes retained (acquired / concerning):\n")
cat(paste(" ", rownames(mat_ga)), sep = "\n")

if (nrow(mat_ga) >= 2 && ncol(mat_ga) >= 2) {
  draw_chord(mat_ga,
             "Acquired AMR Genes by Family (All Samples)",
             family_colours, family_phylum_v, "Family",
             row_colours         = gene_row_colours,
             label_cex           = 1.8,
             legend_cex          = 2.0,
             label_track         = 35,
             label_y_frac        = 0.1,
             canvas_expand       = 1.15,
             show_phylum_legend  = FALSE)

  save_chord("chord_gene_family_graphical_abstract",
             function() draw_chord(mat_ga,
               "Acquired AMR Genes by Family (All Samples)",
               family_colours, family_phylum_v, "Family",
               row_colours        = gene_row_colours,
               label_cex          = 1.8,
               legend_cex         = 2.0,
               label_track        = 35,
               label_y_frac       = 0.1,
               canvas_expand      = 1.15,
               show_phylum_legend = FALSE),
             width = 20, height = 20)
} else {
  cat("Insufficient data after intrinsic gene removal.\n")
}
```

### 8.6 Gene Level - Genus - All Samples

```{r chord-gene-genus-all, fig.width=16, fig.height=16}
if (nrow(mat_gene_genus) >= 2 && ncol(mat_gene_genus) >= 2) {
  draw_chord(mat_gene_genus,
             "AMR Resistance Gene by Genus (All Samples)",
             genus_colours, genus_phylum_v, "Genus",
             row_colours = gene_row_colours)

  save_chord("chord_gene_genus_all",
             function() draw_chord(mat_gene_genus,
               "AMR Resistance Gene by Genus (All Samples)",
               genus_colours, genus_phylum_v, "Genus",
               row_colours = gene_row_colours),
             width = 16, height = 16)
} else {
  cat("Insufficient gene-level genus data for combined chord diagram.\n")
}
```

### 8.7 Gene Level - Species - All Samples

```{r chord-gene-species-all, fig.width=16, fig.height=16}
if (nrow(mat_gene_species) >= 2 && ncol(mat_gene_species) >= 2) {
  draw_chord(mat_gene_species,
             "AMR Resistance Gene by Species (All Samples)",
             species_colours, species_phylum_v, "Species",
             row_colours = gene_row_colours)

  save_chord("chord_gene_species_all",
             function() draw_chord(mat_gene_species,
               "AMR Resistance Gene by Species (All Samples)",
               species_colours, species_phylum_v, "Species",
               row_colours = gene_row_colours),
             width = 16, height = 16)
} else {
  cat("Insufficient gene-level species data for combined chord diagram.\n")
}
```

### 8.8 Per-treatment Chord Diagrams - Gene by Genus

```{r chord-gene-per-treatment, fig.width=16, fig.height=16, results='asis'}
treatment_labels <- c(
  "EVI_RVS" = "Post-Evisceration (RVS Enrichment)",
  "EVI_CAT" = "Post-Evisceration (BF Bolton + CAT Enrichment)",
  "ASC_RVS" = "Post-ASC Wash (RVS Enrichment)",
  "ASC_CAT" = "Post-ASC Wash (BF Bolton + CAT Enrichment)"
)

for (trt in treatments) {
  df_trt   <- taxonomy_clean %>% filter(treatment == trt)
  trt_safe <- gsub("[^A-Za-z0-9]", "_", trt)
  trt_long <- treatment_labels[trt]

  mat_trt <- build_mat_gene(df_trt, taxon_col = "genus", min_reads = 3)

  if (nrow(mat_trt) < 2 || ncol(mat_trt) < 2) {
    cat("\n#### ", trt_long, "\n\nInsufficient data.\n\n")
    next
  }

  cat("\n#### ", trt_long, "\n\n")

  ttl   <- paste0("AMR Resistance Gene by Genus: ", trt_long)
  fname <- paste0("chord_gene_genus_", trt_safe)

  draw_chord(mat_trt, ttl, genus_colours, genus_phylum_v, "Genus",
             row_colours = gene_row_colours)

  local({
    m <- mat_trt; t <- ttl; fn <- fname
    save_chord(fn, function() draw_chord(m, t, genus_colours, genus_phylum_v,
                                          "Genus", row_colours = gene_row_colours),
               width = 16, height = 16)
  })

  cat("\n\n")
}
```

---

## 9. Summary Tables

### 9.1 Read counts per resistance class by genus

```{r table-all}
taxonomy_clean %>%
  count(treatment, res_class, genus) %>%
  arrange(treatment, desc(n)) %>%
  kable(caption = "Read counts per treatment, resistance class, and genus") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

### 9.2 MEGARes hit summary

```{r dedup-summary}
dedup %>%
  count(treatment, sample_id, name = "megares_hits") %>%
  left_join(
    raw %>% count(treatment, sample_id, name = "total_raw_hits"),
    by = c("treatment", "sample_id")
  ) %>%
  mutate(pct_megares = round(100 * megares_hits / total_raw_hits, 1)) %>%
  kable(caption = "MEGARes hits per sample (all reads, no deduplication)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

## 10. Gene-level Co-occurrence Analysis

Gene co-occurrence is measured as **Jaccard similarity**: J(A,B) = |A∩B| / |A∪B|,
where A and B are the sets of samples in which each gene was detected.
A score of 1 means the two genes always appear together; 0 means never.

A gene is counted as *present* in a sample if at least one de-duplicated read
from that sample was assigned to it.

```{r cooc-setup}
heatmap_colors <- colorRampPalette(c("#FFFFFF", "#FFF3CD", "#FF9B00", "#D62728"))(100)

# ── Gene presence/absence: gene detected in ≥1 read per sample ───────────────
build_read_pres_mat <- function(df) {
  df %>%
    distinct(gene_norm, sample_id) %>%
    mutate(present = 1L) %>%
    pivot_wider(names_from  = sample_id,
                values_from = present, values_fill = 0L) %>%
    column_to_rownames("gene_norm") %>%
    as.matrix()
}

# ── Pairwise Jaccard similarity matrix ───────────────────────────────────────
jaccard_matrix <- function(mat) {
  n     <- nrow(mat)
  genes <- rownames(mat)
  jac   <- matrix(0, n, n, dimnames = list(genes, genes))
  for (i in seq_len(n)) {
    for (j in seq_len(n)) {
      a <- mat[i, ]; b <- mat[j, ]
      inter <- sum(a == 1L & b == 1L)
      uni   <- sum(a == 1L | b == 1L)
      jac[i, j] <- if (uni == 0) 0 else inter / uni
    }
  }
  jac
}

# ── Co-occurrence heatmap ─────────────────────────────────────────────────────
plot_cooccurrence <- function(jac_mat, gene_class_df, title,
                               max_genes = 60, save_name = NULL) {
  if (nrow(jac_mat) > max_genes) {
    top_genes <- names(sort(rowMeans(jac_mat), decreasing = TRUE)[seq_len(max_genes)])
    jac_mat   <- jac_mat[top_genes, top_genes]
    title     <- paste0(title, "\n(top ", max_genes, " genes by mean co-occurrence)")
  }

  ann_df <- gene_class_df %>%
    filter(gene_norm %in% rownames(jac_mat)) %>%
    distinct(gene_norm, .keep_all = TRUE) %>%
    select(gene_norm, res_class) %>%
    column_to_rownames("gene_norm")
  ann_df <- ann_df[rownames(jac_mat), , drop = FALSE]

  p <- pheatmap(
    jac_mat,
    color                    = heatmap_colors,
    clustering_distance_rows = "euclidean",
    clustering_distance_cols = "euclidean",
    clustering_method        = "ward.D2",
    annotation_row           = ann_df,
    annotation_col           = ann_df,
    show_rownames            = nrow(jac_mat) <= 60,
    show_colnames            = ncol(jac_mat) <= 60,
    fontsize_row             = if (nrow(jac_mat) > 40) 5.5 else 7,
    fontsize_col             = if (ncol(jac_mat) > 40) 5.5 else 7,
    main                     = title,
    border_color             = NA,
    legend_breaks            = c(0, 0.25, 0.5, 0.75, 1),
    legend_labels            = c("0.00", "0.25", "0.50", "0.75", "1.00"),
    breaks                   = seq(0, 1, length.out = 101),
    silent                   = TRUE
  )

  if (!is.null(save_name)) {
    ggsave(file.path(plot_dir, paste0(save_name, ".png")),
           plot = p$gtable, width = 13, height = 11, dpi = 300)
    ggsave(file.path(plot_dir, paste0(save_name, ".pdf")),
           plot = p$gtable, width = 13, height = 11)
  }
  print(p)
}

```

### 10.1 Per-treatment Co-occurrence Heatmaps

```{r cooc-heatmaps, fig.width=13, fig.height=11, results='asis'}
# Gene metadata: one row per (treatment, gene_norm).
# Use the most-detected res_class to avoid duplicated vertex names in igraph
# when the same gene appears under multiple resistance classes across databases.
gene_meta <- taxonomy_clean %>%
  group_by(treatment, gene_norm, res_class) %>%
  summarise(sample_count = n_distinct(sample_id), .groups = "drop") %>%
  group_by(treatment, gene_norm) %>%
  slice_max(sample_count, n = 1, with_ties = FALSE) %>%
  ungroup()

pres_mats <- list()
jac_mats  <- list()

for (trt in treatments) {
  df_trt  <- taxonomy_clean %>% filter(treatment == trt)
  gm_trt  <- gene_meta    %>% filter(treatment == trt)

  pres_mats[[trt]] <- build_read_pres_mat(df_trt)
  jac_mats[[trt]]  <- jaccard_matrix(pres_mats[[trt]])

  trt_safe <- gsub("[^A-Za-z0-9]", "_", trt)
  cat("\n#### ", trt, "\n\n")

  plot_cooccurrence(
    jac_mats[[trt]], gm_trt,
    paste0("AMR Gene Co-occurrence: ", trt),
    save_name = paste0("cooc_heatmap_", trt_safe)
  )
  cat("\n\n")
}
```

### 10.2 Top Co-occurring Gene Pairs

```{r cooc-top-pairs}
top_gene_pairs <- function(jac_mat, trt, threshold = 0.75) {
  melt(jac_mat, varnames = c("Gene1", "Gene2"), value.name = "Jaccard") %>%
    filter(as.character(Gene1) < as.character(Gene2), Jaccard >= threshold) %>%
    arrange(desc(Jaccard)) %>%
    mutate(Treatment = trt) %>%
    select(Treatment, Gene1, Gene2, Jaccard)
}

# bind_rows preserves column names even when all component frames are empty,
# unlike map_dfr which can collapse to a zero-column tibble.
pairs_all <- map(treatments, function(trt) top_gene_pairs(jac_mats[[trt]], trt)) %>%
  bind_rows()

if (nrow(pairs_all) > 0) {
  kable(pairs_all, digits = 3,
        caption = "Gene pairs with Jaccard co-occurrence ≥ 0.75") %>%
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
} else {
  cat("No gene pairs exceed Jaccard 0.75. Showing top 20 across all treatments:\n")
  top20 <- map(treatments, function(trt) top_gene_pairs(jac_mats[[trt]], trt, 0)) %>%
    bind_rows()
  if (nrow(top20) > 0) {
    top20 %>%
      arrange(desc(Jaccard)) %>%
      slice_head(n = 20) %>%
      kable(digits = 3, caption = "Top 20 gene pairs by co-occurrence") %>%
      kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
  } else {
    cat("No gene pairs found across any treatment (all matrices have fewer than 2 genes).\n")
  }
}
```

### 10.3 All-treatments Combined Co-occurrence Heatmap

Genes detected in ≥ 2 samples across all treatments are pooled into a single
presence/absence matrix. The resulting Jaccard co-occurrence heatmap gives an
overview of gene pairs that consistently appear together across the full dataset,
independent of treatment group. Resistance-class annotation is shown on both axes.

```{r cooc-combined, fig.width=14, fig.height=13}
# Pool all samples across treatments; each column = one sample
pres_all <- taxonomy_clean %>%
  distinct(gene_norm, sample_id) %>%
  mutate(present = 1L) %>%
  pivot_wider(names_from = sample_id, values_from = present, values_fill = 0L) %>%
  column_to_rownames("gene_norm") %>%
  as.matrix()

# Determine the minimum sample threshold adaptively:
# prefer genes in ≥2 samples, but fall back to ≥1 if too few genes survive.
n_samples   <- ncol(pres_all)
min_samples <- if (n_samples >= 4) 2L else 1L
pres_all    <- pres_all[rowSums(pres_all) >= min_samples, , drop = FALSE]

cat("Genes retained (present in ≥", min_samples, "samples):", nrow(pres_all), "\n")

if (nrow(pres_all) < 2) {
  cat("Fewer than 2 genes passed the filter - combined heatmap cannot be drawn.\n",
      "Consider lowering min_reads in build_mat_gene or checking MEGARes gene coverage.\n")
} else {

  jac_all <- jaccard_matrix(pres_all)

  # Resistance-class annotation for rows/cols
  ann_all <- gene_meta %>%
    distinct(gene_norm, .keep_all = TRUE) %>%
    filter(gene_norm %in% rownames(jac_all)) %>%
    select(gene_norm, res_class) %>%
    column_to_rownames("gene_norm")
  ann_all <- ann_all[rownames(jac_all), , drop = FALSE]
  names(ann_all) <- "Resistance class"

  ann_colours <- list(
    `Resistance class` = res_class_colours[intersect(names(res_class_colours),
                                                      unique(ann_all[["Resistance class"]]))]
  )

  n    <- nrow(jac_all)
  fnt  <- if (n > 40) 5.5 else 7
  w    <- max(12, min(22, n * 0.3 + 6))
  h    <- max(10, min(20, n * 0.3 + 4))

  p_all <- pheatmap(
    jac_all,
    color                    = heatmap_colors,
    clustering_distance_rows = "euclidean",
    clustering_distance_cols = "euclidean",
    clustering_method        = "ward.D2",
    annotation_row           = ann_all,
    annotation_col           = ann_all,
    annotation_colors        = ann_colours,
    show_rownames            = TRUE,
    show_colnames            = TRUE,
    fontsize_row             = fnt,
    fontsize_col             = fnt,
    main                     = "AMR Gene Co-occurrence (All Samples)",
    border_color             = NA,
    legend_breaks            = c(0, 0.25, 0.5, 0.75, 1),
    legend_labels            = c("0.00", "0.25", "0.50", "0.75", "1.00"),
    breaks                   = seq(0, 1, length.out = 101),
    silent                   = TRUE
  )

  ggsave(file.path(plot_dir, "cooc_heatmap_all_combined.png"),
         plot = p_all$gtable, width = w, height = h, dpi = 300)
  ggsave(file.path(plot_dir, "cooc_heatmap_all_combined.pdf"),
         plot = p_all$gtable, width = w, height = h)
  print(p_all)
}
```

---

## 12. Feature-specific Chord Diagrams

Chord diagrams for three additional gene categories using ABRicate databases
beyond ResFinder. Each read is deduplicated per READ_ID by normalised gene name
(highest %IDENTITY kept). Genus assignment uses the same Kraken2 taxonomy and
phylum lookup as the main analysis.

```{r feature-helpers}
# ── Build deduplicated, genus-annotated data frame from selected databases ─────
build_feature_data <- function(databases, resistance_pattern = NULL,
                                min_identity = 80) {
  avail <- intersect(databases, unique(raw$DATABASE))
  if (length(avail) == 0) return(tibble())

  d <- raw %>%
    filter(DATABASE %in% avail,
           !is.na(GENE), GENE != "",
           !is.na(TAXONOMY), TAXONOMY != "",
           as.numeric(`%IDENTITY`) >= min_identity)

  if (!is.null(resistance_pattern))
    d <- filter(d, str_detect(str_to_lower(coalesce(RESISTANCE, "")),
                              resistance_pattern))

  d %>%
    mutate(gene_norm = str_to_lower(GENE) %>%
             str_remove("_[0-9]+$") %>% str_trim()) %>%
    group_by(sample_id, treatment, READ_ID, gene_norm) %>%
    slice_max(as.numeric(`%IDENTITY`), n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(
      genus  = word(TAXONOMY, 1),
      phylum = phylum_lookup[genus],
      genus  = if_else(is.na(phylum), "Unclassified", genus)
    )
}

# ── Assign colours to gene_norm based on functional category (RESISTANCE field)
build_gene_cat_colours <- function(df, palette_fn) {
  cat_lut <- df %>%
    distinct(gene_norm, RESISTANCE) %>%
    mutate(category = str_to_lower(str_trim(
      str_extract(coalesce(RESISTANCE, "unknown"), "^[^;,/]+")))
    ) %>%
    group_by(gene_norm) %>%
    slice(1) %>%
    ungroup()

  unique_cats <- sort(unique(cat_lut$category))
  cat_pal     <- setNames(palette_fn(length(unique_cats)), unique_cats)

  list(
    gene_colours = setNames(cat_pal[cat_lut$category], cat_lut$gene_norm),
    cat_colours  = cat_pal
  )
}
```

---

### 12.1 Efflux Pump Genes by Genus

Efflux pump genes are extracted from CARD, NCBI, and MEGARes databases by
filtering for "efflux", "pump", or "multidrug" in the RESISTANCE annotation.
Genes are coloured by efflux category as annotated by the database.

```{r efflux-data}
efflux_data <- build_feature_data(
  databases          = c("card", "ncbi", "megares"),
  resistance_pattern = "efflux|pump|multidrug|mdr"
)

cat("Databases used:", paste(intersect(c("card","ncbi","megares"), unique(raw$DATABASE)),
                              collapse = ", "), "\n")
cat("Reads:", nrow(efflux_data), "| Unique genes:", n_distinct(efflux_data$gene_norm),
    "| Genera:", n_distinct(efflux_data$genus[efflux_data$genus != "Unclassified"]), "\n")
```

```{r efflux-chord-all, fig.width=16, fig.height=16}
if (nrow(efflux_data) > 0) {
  efflux_cols <- build_gene_cat_colours(
    efflux_data,
    colorRampPalette(c("#FF6600", "#D62728", "#8B1A1A", "#FF9944", "#CC3300"))
  )

  mat_efflux <- build_mat_gene(efflux_data, "genus", min_reads = 3, top_n = 40)

  if (nrow(mat_efflux) >= 2 && ncol(mat_efflux) >= 2) {
    ec <- efflux_cols$gene_colours[rownames(mat_efflux)]
    ec[is.na(ec)] <- "#AAAAAA"

    draw_chord(mat_efflux,
               "Efflux Pump Genes by Genus (All Samples)",
               genus_colours, genus_phylum_v, "Genus",
               row_colours      = ec,
               row_legend_cols  = efflux_cols$cat_colours,
               row_legend_title = "Efflux category")

    save_chord("chord_efflux_genus_all",
               function() {
                 ec2 <- efflux_cols$gene_colours[rownames(mat_efflux)]
                 ec2[is.na(ec2)] <- "#AAAAAA"
                 draw_chord(mat_efflux,
                            "Efflux Pump Genes by Genus (All Samples)",
                            genus_colours, genus_phylum_v, "Genus",
                            row_colours = ec2,
                            row_legend_cols = efflux_cols$cat_colours,
                            row_legend_title = "Efflux category")
               }, width = 16, height = 16)
  } else {
    cat("Insufficient efflux data (need ≥2 genes and ≥2 genera after min_reads filter).\n")
  }
} else {
  cat("No efflux pump reads found in the loaded databases.\n")
}
```

```{r efflux-chord-per-treatment, fig.width=16, fig.height=16, results='asis'}
if (nrow(efflux_data) > 0) {
  for (trt in treatments) {
    df_trt   <- efflux_data %>% filter(treatment == trt)
    trt_safe <- gsub("[^A-Za-z0-9]", "_", trt)
    trt_long <- treatment_labels[trt]
    mat_trt  <- build_mat_gene(df_trt, "genus", min_reads = 2, top_n = 40)

    if (nrow(mat_trt) < 2 || ncol(mat_trt) < 2) next

    cat("\n####", trt_long, "\n\n")

    ec <- efflux_cols$gene_colours[rownames(mat_trt)]
    ec[is.na(ec)] <- "#AAAAAA"

    draw_chord(mat_trt, paste0("Efflux Pump Genes by Genus: ", trt_long),
               genus_colours, genus_phylum_v, "Genus",
               row_colours = ec, row_legend_cols = efflux_cols$cat_colours,
               row_legend_title = "Efflux category")

    local({
      m  <- mat_trt; fn <- paste0("chord_efflux_genus_", trt_safe)
      t  <- paste0("Efflux Pump Genes by Genus: ", trt_long)
      lc <- efflux_cols$cat_colours
      rc <- ec
      save_chord(fn, function() draw_chord(m, t, genus_colours, genus_phylum_v,
                                            "Genus", row_colours = rc,
                                            row_legend_cols = lc,
                                            row_legend_title = "Efflux category"),
                 width = 16, height = 16)
    })
    cat("\n\n")
  }
}
```

---

### 12.2 Virulence Factor Genes by Genus

Virulence factor genes are extracted from the VFDB and ecoli_vf databases.
Genes are coloured by virulence factor category (e.g. toxin, adhesin, iron
acquisition) as annotated in the RESISTANCE/function field.

```{r virulence-data}
vf_data <- build_feature_data(
  databases     = c("vfdb", "ecoli_vf"),
  min_identity  = 80
)

cat("Databases used:", paste(intersect(c("vfdb","ecoli_vf"), unique(raw$DATABASE)),
                              collapse = ", "), "\n")
cat("Reads:", nrow(vf_data), "| Unique genes:", n_distinct(vf_data$gene_norm),
    "| Genera:", n_distinct(vf_data$genus[vf_data$genus != "Unclassified"]), "\n")
```

```{r virulence-chord-all, fig.width=16, fig.height=16}
if (nrow(vf_data) > 0) {
  vf_cols <- build_gene_cat_colours(
    vf_data,
    colorRampPalette(c("#7B2D8B", "#C463D7", "#E8A0F0", "#9E3BAB",
                       "#5C1A6B", "#D966E8", "#B34FCC", "#802B99"))
  )

  mat_vf <- build_mat_gene(vf_data, "genus", min_reads = 3, top_n = 40)

  if (nrow(mat_vf) >= 2 && ncol(mat_vf) >= 2) {
    vc <- vf_cols$gene_colours[rownames(mat_vf)]
    vc[is.na(vc)] <- "#AAAAAA"

    draw_chord(mat_vf,
               "Virulence Factor Genes by Genus (All Samples)",
               genus_colours, genus_phylum_v, "Genus",
               row_colours      = vc,
               row_legend_cols  = vf_cols$cat_colours,
               row_legend_title = "Virulence factor\ncategory")

    save_chord("chord_virulence_genus_all",
               function() {
                 vc2 <- vf_cols$gene_colours[rownames(mat_vf)]
                 vc2[is.na(vc2)] <- "#AAAAAA"
                 draw_chord(mat_vf,
                            "Virulence Factor Genes by Genus (All Samples)",
                            genus_colours, genus_phylum_v, "Genus",
                            row_colours = vc2,
                            row_legend_cols = vf_cols$cat_colours,
                            row_legend_title = "Virulence factor\ncategory")
               }, width = 16, height = 16)
  } else {
    cat("Insufficient virulence data (need ≥2 genes and ≥2 genera after min_reads filter).\n")
  }
} else {
  cat("No virulence factor reads found in the loaded databases.\n")
}
```

```{r virulence-chord-per-treatment, fig.width=16, fig.height=16, results='asis'}
if (nrow(vf_data) > 0) {
  for (trt in treatments) {
    df_trt   <- vf_data %>% filter(treatment == trt)
    trt_safe <- gsub("[^A-Za-z0-9]", "_", trt)
    trt_long <- treatment_labels[trt]
    mat_trt  <- build_mat_gene(df_trt, "genus", min_reads = 2, top_n = 40)

    if (nrow(mat_trt) < 2 || ncol(mat_trt) < 2) next

    cat("\n####", trt_long, "\n\n")

    vc <- vf_cols$gene_colours[rownames(mat_trt)]
    vc[is.na(vc)] <- "#AAAAAA"

    draw_chord(mat_trt, paste0("Virulence Factor Genes by Genus: ", trt_long),
               genus_colours, genus_phylum_v, "Genus",
               row_colours = vc, row_legend_cols = vf_cols$cat_colours,
               row_legend_title = "Virulence factor\ncategory")

    local({
      m  <- mat_trt; fn <- paste0("chord_virulence_genus_", trt_safe)
      t  <- paste0("Virulence Factor Genes by Genus: ", trt_long)
      lc <- vf_cols$cat_colours; rc <- vc
      save_chord(fn, function() draw_chord(m, t, genus_colours, genus_phylum_v,
                                            "Genus", row_colours = rc,
                                            row_legend_cols = lc,
                                            row_legend_title = "Virulence factor\ncategory"),
                 width = 16, height = 16)
    })
    cat("\n\n")
  }
}
```

---

### 12.3 Plasmid Replicons by Genus

Plasmid replicon sequences are extracted from the PlasmidFinder database.
Replicons (row sectors) are coloured by incompatibility group family; genera
(column sectors) are coloured by phylum as in all other diagrams.

```{r plasmid-data}
plasmid_data <- build_feature_data(
  databases    = "plasmidfinder",
  min_identity = 80
)

cat("Databases used:", paste(intersect("plasmidfinder", unique(raw$DATABASE)),
                              collapse = ", "), "\n")
cat("Reads:", nrow(plasmid_data), "| Unique replicons:", n_distinct(plasmid_data$gene_norm),
    "| Genera:", n_distinct(plasmid_data$genus[plasmid_data$genus != "Unclassified"]), "\n")
```

```{r plasmid-chord-all, fig.width=16, fig.height=16}
if (nrow(plasmid_data) > 0) {
  # Colour replicons by incompatibility group family (prefix before underscore or digit)
  plasmid_data <- plasmid_data %>%
    mutate(inc_group = str_extract(gene_norm, "^[a-z]+"))

  inc_groups  <- sort(unique(plasmid_data$inc_group))
  inc_pal     <- setNames(
    colorRampPalette(c("#006D2C", "#31A354", "#74C476", "#BAE4B3",
                       "#00441B", "#238B45", "#41AE76", "#66C2A4"))(length(inc_groups)),
    inc_groups
  )

  plasmid_gene_cols <- plasmid_data %>%
    distinct(gene_norm, inc_group) %>%
    mutate(colour = inc_pal[inc_group]) %>%
    { setNames(.$colour, .$gene_norm) }

  mat_plasmid <- build_mat_gene(plasmid_data, "genus", min_reads = 3, top_n = 40)

  if (nrow(mat_plasmid) >= 2 && ncol(mat_plasmid) >= 2) {
    pc <- plasmid_gene_cols[rownames(mat_plasmid)]
    pc[is.na(pc)] <- "#AAAAAA"

    draw_chord(mat_plasmid,
               "Plasmid Replicons by Genus (All Samples)",
               genus_colours, genus_phylum_v, "Genus",
               row_colours      = pc,
               row_legend_cols  = inc_pal,
               row_legend_title = "Incompatibility\ngroup family")

    save_chord("chord_plasmid_genus_all",
               function() {
                 pc2 <- plasmid_gene_cols[rownames(mat_plasmid)]
                 pc2[is.na(pc2)] <- "#AAAAAA"
                 draw_chord(mat_plasmid,
                            "Plasmid Replicons by Genus (All Samples)",
                            genus_colours, genus_phylum_v, "Genus",
                            row_colours = pc2,
                            row_legend_cols = inc_pal,
                            row_legend_title = "Incompatibility\ngroup family")
               }, width = 16, height = 16)
  } else {
    cat("Insufficient plasmid data (need ≥2 replicons and ≥2 genera after min_reads filter).\n")
  }
} else {
  cat("No plasmid replicon reads found in the loaded databases.\n")
}
```

```{r plasmid-chord-per-treatment, fig.width=16, fig.height=16, results='asis'}
if (nrow(plasmid_data) > 0) {
  for (trt in treatments) {
    df_trt   <- plasmid_data %>% filter(treatment == trt)
    trt_safe <- gsub("[^A-Za-z0-9]", "_", trt)
    trt_long <- treatment_labels[trt]
    mat_trt  <- build_mat_gene(df_trt, "genus", min_reads = 2)

    if (nrow(mat_trt) < 2 || ncol(mat_trt) < 2) next

    cat("\n####", trt_long, "\n\n")

    pc <- plasmid_gene_cols[rownames(mat_trt)]
    pc[is.na(pc)] <- "#AAAAAA"

    draw_chord(mat_trt, paste0("Plasmid Replicons by Genus: ", trt_long),
               genus_colours, genus_phylum_v, "Genus",
               row_colours = pc, row_legend_cols = inc_pal,
               row_legend_title = "Incompatibility\ngroup family")

    local({
      m  <- mat_trt; fn <- paste0("chord_plasmid_genus_", trt_safe)
      t  <- paste0("Plasmid Replicons by Genus: ", trt_long)
      rc <- pc
      save_chord(fn, function() draw_chord(m, t, genus_colours, genus_phylum_v,
                                            "Genus", row_colours = rc,
                                            row_legend_cols = inc_pal,
                                            row_legend_title = "Incompatibility\ngroup family"),
                 width = 16, height = 16)
    })
    cat("\n\n")
  }
}
```

---

---

## 13. *Campylobacter*-only Chord Diagrams

Reads where `genus == "Campylobacter"` are filtered from `taxonomy_clean`.
Two chord diagrams are produced for each subset - one at **resistance class** level
and one at **resistance gene** level. Both diagrams show resistance genes on one arc
and resistance classes on the other, revealing the AMR profile specific to
*Campylobacter*. Diagrams are generated for all samples combined and for each
sample type separately.

```{r campy-setup}
tc_campy <- taxonomy_clean %>%
  filter(genus == "Campylobacter",
         res_class != "disinfectant/biocide")

cat("Campylobacter reads:", nrow(tc_campy), "\n")
cat("Treatments present:", paste(sort(unique(tc_campy$treatment)), collapse = ", "), "\n")
cat("Species found:",
    paste(sort(unique(tc_campy$species[tc_campy$species != "Unclassified"])), collapse = ", "),
    "\n")
cat("Resistance classes:", paste(sort(unique(tc_campy$res_class)), collapse = ", "), "\n")
cat("Unique genes:",       n_distinct(tc_campy$gene_norm), "\n")
```

```{r campy-matrix-helper}
# Build a gene_norm by res_class count matrix (Campylobacter-specific)
# Used for both the class-level and gene-level Campy chords.
build_campy_mat <- function(df, min_reads = 1) {
  mat <- df %>%
    count(gene_norm, res_class) %>%
    pivot_wider(names_from = res_class, values_from = n, values_fill = 0L) %>%
    column_to_rownames("gene_norm") %>%
    as.matrix()
  mat[rowSums(mat) >= min_reads, colSums(mat) >= min_reads, drop = FALSE]
}

# Draw a Campylobacter gene by class chord using draw_chord.
# Row sectors = genes (coloured by resistance class).
# Column sectors = resistance classes (coloured by res_class_colours).
draw_campy_chord <- function(df, title, min_reads = 1, save_name = NULL,
                              width = 16, height = 16) {
  mat <- build_campy_mat(df, min_reads)

  if (nrow(mat) < 2 || ncol(mat) < 2) {
    message(title, ": insufficient data - skipping.")
    return(invisible(NULL))
  }

  # Column sectors are resistance classes; colour them directly
  tc <- res_class_colours[colnames(mat)]
  tc[is.na(tc)] <- "#AAAAAA"

  # phylum_of_taxon drives the bottomright phylum legend - pass NA vector to suppress it
  po <- setNames(rep(NA_character_, ncol(mat)), colnames(mat))

  draw_chord(mat, title,
             taxon_colours    = tc,
             phylum_of_taxon  = po,
             level_label      = "Resistance class",
             row_colours      = gene_row_colours,
             row_legend_title = "Resistance gene\n(by class colour)")

  if (!is.null(save_name)) {
    local({
      m <- mat; t <- title; tc2 <- tc; po2 <- po
      save_chord(save_name,
                 function() draw_chord(m, t,
                                       taxon_colours   = tc2,
                                       phylum_of_taxon = po2,
                                       level_label     = "Resistance class",
                                       row_colours     = gene_row_colours,
                                       row_legend_title = "Resistance gene\n(by class colour)"),
                 width = width, height = height)
    })
  }
}
```

### 13.1 *Campylobacter* - Resistance Class Level - All Samples

```{r campy-class-all, fig.width=16, fig.height=16}
if (nrow(tc_campy) > 0) {
  draw_campy_chord(
    tc_campy,
    title     = expression(italic("Campylobacter") ~ "AMR Resistance Gene by Class (All Samples)"),
    min_reads = 1,
    save_name = "chord_campy_class_all"
  )
} else {
  cat("No Campylobacter reads found in the dataset.\n")
}
```

### 13.2 *Campylobacter* - Gene Level - All Samples

```{r campy-gene-all, fig.width=16, fig.height=16}
if (nrow(tc_campy) > 0) {
  draw_campy_chord(
    tc_campy,
    title     = expression(italic("Campylobacter") ~ "AMR Resistance Gene by Class (All Samples, Gene-Level View)"),
    min_reads = 1,
    save_name = "chord_campy_gene_all"
  )
}
```

### 13.3 *Campylobacter* - Resistance Class Level - Per Treatment

```{r campy-class-per-treatment, fig.width=16, fig.height=16, results='asis'}
if (nrow(tc_campy) > 0) {
  for (trt in sort(unique(tc_campy$treatment))) {
    df_trt   <- tc_campy %>% filter(treatment == trt)
    trt_safe <- gsub("[^A-Za-z0-9]", "_", trt)
    trt_long <- treatment_labels[trt]

    mat_check <- build_campy_mat(df_trt, min_reads = 1)
    if (nrow(mat_check) < 2 || ncol(mat_check) < 2) {
      cat("\n####", trt_long, "\n\nInsufficient Campylobacter data.\n\n")
      next
    }

    cat("\n####", trt_long, "\n\n")

    draw_campy_chord(
      df_trt,
      title     = bquote(italic("Campylobacter") ~ "AMR Resistance Gene by Class:" ~ .(trt_long)),
      min_reads = 1,
      save_name = paste0("chord_campy_class_", trt_safe)
    )

    cat("\n\n")
  }
}
```

### 13.4 *Campylobacter* - Gene Level - Per Treatment

```{r campy-gene-per-treatment, fig.width=16, fig.height=16, results='asis'}
if (nrow(tc_campy) > 0) {
  for (trt in sort(unique(tc_campy$treatment))) {
    df_trt   <- tc_campy %>% filter(treatment == trt)
    trt_safe <- gsub("[^A-Za-z0-9]", "_", trt)
    trt_long <- treatment_labels[trt]

    mat_check <- build_campy_mat(df_trt, min_reads = 1)
    if (nrow(mat_check) < 2 || ncol(mat_check) < 2) next

    cat("\n####", trt_long, "\n\n")

    draw_campy_chord(
      df_trt,
      title     = bquote(italic("Campylobacter") ~ "AMR Resistance Gene by Class (Gene-Level View):" ~ .(trt_long)),
      min_reads = 1,
      save_name = paste0("chord_campy_gene_", trt_safe)
    )

    cat("\n\n")
  }
}
```

---

## 14. *Escherichia*-only Chord Diagrams

```{r ecoli-setup}
tc_ecoli <- taxonomy_clean %>%
  filter(genus == "Escherichia",
         res_class != "disinfectant/biocide")

cat("Escherichia reads:", nrow(tc_ecoli), "\n")
cat("Treatments present:", paste(sort(unique(tc_ecoli$treatment)), collapse = ", "), "\n")
cat("Resistance classes:", paste(sort(unique(tc_ecoli$res_class)), collapse = ", "), "\n")
cat("Unique genes:",       n_distinct(tc_ecoli$gene_norm), "\n")
```

### 14.1 *Escherichia* - Resistance Class Level - All Samples

```{r ecoli-class-all, fig.width=16, fig.height=16}
if (nrow(tc_ecoli) > 0) {
  draw_campy_chord(
    tc_ecoli,
    title     = expression(italic("Escherichia") ~ "AMR Resistance Gene by Class (All Samples)"),
    min_reads = 1,
    save_name = "chord_ecoli_class_all"
  )
} else {
  cat("No Escherichia reads found in the dataset.\n")
}
```

### 14.2 *Escherichia* - Gene Level - All Samples

```{r ecoli-gene-all, fig.width=16, fig.height=16}
if (nrow(tc_ecoli) > 0) {
  draw_campy_chord(
    tc_ecoli,
    title     = expression(italic("Escherichia") ~ "AMR Resistance Gene by Class (All Samples, Gene-Level View)"),
    min_reads = 1,
    save_name = "chord_ecoli_gene_all"
  )
}
```

### 14.3 *Escherichia* - Resistance Class Level - Per Treatment

```{r ecoli-class-per-treatment, fig.width=16, fig.height=16, results='asis'}
if (nrow(tc_ecoli) > 0) {
  for (trt in sort(unique(tc_ecoli$treatment))) {
    df_trt   <- tc_ecoli %>% filter(treatment == trt)
    trt_safe <- gsub("[^A-Za-z0-9]", "_", trt)
    trt_long <- treatment_labels[trt]

    mat_check <- build_campy_mat(df_trt, min_reads = 1)
    if (nrow(mat_check) < 2 || ncol(mat_check) < 2) {
      cat("\n####", trt_long, "\n\nInsufficient Escherichia data.\n\n")
      next
    }

    cat("\n####", trt_long, "\n\n")

    draw_campy_chord(
      df_trt,
      title     = bquote(italic("Escherichia") ~ "AMR Resistance Gene by Class:" ~ .(trt_long)),
      min_reads = 1,
      save_name = paste0("chord_ecoli_class_", trt_safe)
    )

    cat("\n\n")
  }
}
```

### 14.4 *Escherichia* - Gene Level - Per Treatment

```{r ecoli-gene-per-treatment, fig.width=16, fig.height=16, results='asis'}
if (nrow(tc_ecoli) > 0) {
  for (trt in sort(unique(tc_ecoli$treatment))) {
    df_trt   <- tc_ecoli %>% filter(treatment == trt)
    trt_safe <- gsub("[^A-Za-z0-9]", "_", trt)
    trt_long <- treatment_labels[trt]

    mat_check <- build_campy_mat(df_trt, min_reads = 1)
    if (nrow(mat_check) < 2 || ncol(mat_check) < 2) next

    cat("\n####", trt_long, "\n\n")

    draw_campy_chord(
      df_trt,
      title     = bquote(italic("Escherichia") ~ "AMR Resistance Gene by Class (Gene-Level View):" ~ .(trt_long)),
      min_reads = 1,
      save_name = paste0("chord_ecoli_gene_", trt_safe)
    )

    cat("\n\n")
  }
}
```

---

## 15. *Klebsiella pneumoniae*-only Chord Diagrams

Reads where `genus == "Klebsiella"` are filtered from `taxonomy_clean`.
Reads classified only to genus level (no species epithet) are included alongside
confirmed *K. pneumoniae* reads to maximise data; the majority of
*Klebsiella* reads in this dataset are expected to be *K. pneumoniae*.

```{r kpneu-setup}
tc_kpneu <- taxonomy_clean %>%
  filter(genus == "Klebsiella",
         res_class != "disinfectant/biocide")

cat("Klebsiella reads:", nrow(tc_kpneu), "\n")
cat("Treatments present:", paste(sort(unique(tc_kpneu$treatment)), collapse = ", "), "\n")
cat("Species found:",
    paste(sort(unique(tc_kpneu$species[tc_kpneu$species != "Unclassified"])), collapse = ", "),
    "\n")
cat("Resistance classes:", paste(sort(unique(tc_kpneu$res_class)), collapse = ", "), "\n")
cat("Unique genes:",       n_distinct(tc_kpneu$gene_norm), "\n")
```

### 15.1 *Klebsiella pneumoniae* - Resistance Class Level - All Samples

```{r kpneu-class-all, fig.width=16, fig.height=16}
if (nrow(tc_kpneu) > 0) {
  draw_campy_chord(
    tc_kpneu,
    title     = expression(italic("Klebsiella pneumoniae") ~ "AMR Resistance Gene by Class (All Samples)"),
    min_reads = 1,
    save_name = "chord_kpneu_class_all"
  )
} else {
  cat("No Klebsiella reads found in the dataset.\n")
}
```

### 15.2 *Klebsiella pneumoniae* - Gene Level - All Samples

```{r kpneu-gene-all, fig.width=16, fig.height=16}
if (nrow(tc_kpneu) > 0) {
  draw_campy_chord(
    tc_kpneu,
    title     = expression(italic("Klebsiella pneumoniae") ~ "AMR Resistance Gene by Class (All Samples, Gene-Level View)"),
    min_reads = 1,
    save_name = "chord_kpneu_gene_all"
  )
}
```

### 15.3 *Klebsiella pneumoniae* - Resistance Class Level - Per Treatment

```{r kpneu-class-per-treatment, fig.width=16, fig.height=16, results='asis'}
if (nrow(tc_kpneu) > 0) {
  for (trt in sort(unique(tc_kpneu$treatment))) {
    df_trt   <- tc_kpneu %>% filter(treatment == trt)
    trt_safe <- gsub("[^A-Za-z0-9]", "_", trt)
    trt_long <- treatment_labels[trt]

    mat_check <- build_campy_mat(df_trt, min_reads = 1)
    if (nrow(mat_check) < 2 || ncol(mat_check) < 2) {
      cat("\n####", trt_long, "\n\nInsufficient *Klebsiella* data.\n\n")
      next
    }

    cat("\n####", trt_long, "\n\n")

    draw_campy_chord(
      df_trt,
      title     = bquote(italic("Klebsiella pneumoniae") ~ "AMR Resistance Gene by Class:" ~ .(trt_long)),
      min_reads = 1,
      save_name = paste0("chord_kpneu_class_", trt_safe)
    )

    cat("\n\n")
  }
}
```

### 15.4 *Klebsiella pneumoniae* - Gene Level - Per Treatment

```{r kpneu-gene-per-treatment, fig.width=16, fig.height=16, results='asis'}
if (nrow(tc_kpneu) > 0) {
  for (trt in sort(unique(tc_kpneu$treatment))) {
    df_trt   <- tc_kpneu %>% filter(treatment == trt)
    trt_safe <- gsub("[^A-Za-z0-9]", "_", trt)
    trt_long <- treatment_labels[trt]

    mat_check <- build_campy_mat(df_trt, min_reads = 1)
    if (nrow(mat_check) < 2 || ncol(mat_check) < 2) next

    cat("\n####", trt_long, "\n\n")

    draw_campy_chord(
      df_trt,
      title     = bquote(italic("Klebsiella pneumoniae") ~ "AMR Resistance Gene by Class (Gene-Level View):" ~ .(trt_long)),
      min_reads = 1,
      save_name = paste0("chord_kpneu_gene_", trt_safe)
    )

    cat("\n\n")
  }
}
```

---

## 11. Session Info

```{r session-info, results='markup'}
sessionInfo()
```
