---
title: "Resistance Gene Co-occurrence Analysis"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 10
)

# ── Output directory for saved plots ─────────────────────────────────────────
plot_dir <- "/Users/alicebradbury/Desktop/cooccurrence_plots"
dir.create(plot_dir, showWarnings = FALSE)
```

## Overview

This analysis examines the co-occurrence of antimicrobial resistance (AMR) genes
across samples from a poultry processing plant. Two treatment sites are compared:

- **EVI** – Post manual evisceration
- **ASC** – Post acidified sodium chlorite (ASC) wash

Two enrichment broths per site:

- **RVS broth enriched**
- **Bolton broth + CAT enriched**

Co-occurrence is calculated as the Jaccard similarity index (0 = never co-occur,
1 = always co-occur) between gene presence/absence profiles across samples.
The network graphs show genes as nodes (sized by number of detections, coloured
by resistance class) with edges weighted by co-occurrence strength.

---

## 1. Setup

```{r packages}
library(readxl)
library(tidyverse)
library(pheatmap)
library(RColorBrewer)
library(viridis)
library(ggplot2)
library(stringr)
library(reshape2)
library(grid)
library(igraph)
library(ggraph)
library(tidygraph)
library(ggrepel)
library(patchwork)
library(graphlayouts)
```

---

## 2. Load and Combine Data

```{r load-data}
base_dir <- "/Users/alicebradbury/Desktop/ABRicate_dereplicated"

# ── Descriptive dataset labels ────────────────────────────────────────────────
dataset_labels <- c(
  "EVI_RVS" = "Post Evisceration – RVS Broth Enriched",
  "EVI_CAT" = "Post Evisceration – Bolton Broth + CAT Enriched",
  "ASC_RVS" = "Post ASC Wash – RVS Broth Enriched",
  "ASC_CAT" = "Post ASC Wash – Bolton Broth + CAT Enriched"
)

read_abricate <- function(path, dataset_label) {
  df <- read_excel(path)
  names(df) <- make.unique(names(df))
  df$Dataset <- dataset_label
  df
}

evi_rvs <- read_abricate(file.path(base_dir, "COMBINED_EVI_RVS.xlsx"), "EVI_RVS")
evi_cat <- read_abricate(file.path(base_dir, "COMBINED_EVI_CAT.xlsx"), "EVI_CAT")
asc_rvs <- read_abricate(file.path(base_dir, "COMBINED_ASC_RVS.xlsx"), "ASC_RVS")
asc_cat <- read_abricate(file.path(base_dir, "COMBINED_ASC_CAT.xlsx"), "ASC_CAT")

all_data <- bind_rows(evi_rvs, evi_cat, asc_rvs, asc_cat) %>%
  mutate(RESISTANCE = str_to_lower(RESISTANCE))

cat("Total gene-dataset rows:", nrow(all_data), "\n")
cat("Unique genes across all datasets:", n_distinct(all_data$GENE), "\n")

# Report databases present if that column exists
if ("DATABASE" %in% names(all_data)) {
  cat("Databases present:", paste(sort(unique(all_data$DATABASE)), collapse = ", "), "\n")
}
```

---

## 3. Normalise Gene Names

The COMBINED Excel files are pre-aggregated summaries across all 7 ABRicate
databases. Each row gives a gene name, the samples it was detected in, and a
RESISTANCE annotation. Because these files are already summarised (not
read-level), **filtering to a single database would discard the majority of
detections and leave too few genes for clustering**.

Instead, deduplication is handled at the matrix-building stage (Section 4):
after gene-name normalisation, `distinct(GENE, Samples)` ensures the same gene
is counted only once per sample, even if it appeared in multiple databases.

Gene-name normalisation collapses allele variant suffixes so that, e.g.,
`blaTEM-1b_1` and `blaTEM-1b_2` are treated as the same gene (`blatem-1b`).

```{r normalise-genes}
normalise_gene <- function(x) {
  str_to_lower(x) %>%
    str_remove("_[0-9]+$") %>%
    str_trim()
}

evi_rvs <- evi_rvs %>% mutate(GENE = normalise_gene(GENE))
evi_cat <- evi_cat %>% mutate(GENE = normalise_gene(GENE))
asc_rvs <- asc_rvs %>% mutate(GENE = normalise_gene(GENE))
asc_cat <- asc_cat %>% mutate(GENE = normalise_gene(GENE))

cat("EVI_RVS genes after normalisation:", n_distinct(evi_rvs$GENE), "\n")
cat("EVI_CAT genes after normalisation:", n_distinct(evi_cat$GENE), "\n")
cat("ASC_RVS genes after normalisation:", n_distinct(asc_rvs$GENE), "\n")
cat("ASC_CAT genes after normalisation:", n_distinct(asc_cat$GENE), "\n")
```

---

## 4. Build Presence/Absence Matrices

```{r presence-absence-function}
# distinct(GENE, Samples) after normalisation ensures allele variants collapsed
# above don't create duplicate rows in the matrix.
build_presence_matrix <- function(df) {
  df %>%
    select(GENE, Samples) %>%
    mutate(Samples = str_split(Samples, ",\\s*")) %>%
    unnest(Samples) %>%
    mutate(Samples = str_trim(Samples)) %>%
    filter(Samples != "", !is.na(Samples)) %>%
    distinct(GENE, Samples) %>%
    mutate(present = 1L) %>%
    pivot_wider(names_from = Samples, values_from = present, values_fill = 0L) %>%
    column_to_rownames("GENE") %>%
    as.matrix()
}

mat_evi_rvs <- build_presence_matrix(evi_rvs)
mat_evi_cat <- build_presence_matrix(evi_cat)
mat_asc_rvs <- build_presence_matrix(asc_rvs)
mat_asc_cat <- build_presence_matrix(asc_cat)

cat("EVI RVS matrix:", nrow(mat_evi_rvs), "genes ×", ncol(mat_evi_rvs), "samples\n")
cat("EVI CAT matrix:", nrow(mat_evi_cat), "genes ×", ncol(mat_evi_cat), "samples\n")
cat("ASC RVS matrix:", nrow(mat_asc_rvs), "genes ×", ncol(mat_asc_rvs), "samples\n")
cat("ASC CAT matrix:", nrow(mat_asc_cat), "genes ×", ncol(mat_asc_cat), "samples\n")
```

---

## 5. Calculate Co-occurrence (Jaccard Similarity)

```{r jaccard-function}
jaccard_matrix <- function(mat) {
  n    <- nrow(mat)
  genes <- rownames(mat)
  jac  <- matrix(0, nrow = n, ncol = n, dimnames = list(genes, genes))
  for (i in seq_len(n)) {
    for (j in seq_len(n)) {
      a <- mat[i, ]; b <- mat[j, ]
      inter <- sum(a == 1 & b == 1)
      uni   <- sum(a == 1 | b == 1)
      jac[i, j] <- if (uni == 0) 0 else inter / uni
    }
  }
  jac
}

jac_evi_rvs <- jaccard_matrix(mat_evi_rvs)
jac_evi_cat <- jaccard_matrix(mat_evi_cat)
jac_asc_rvs <- jaccard_matrix(mat_asc_rvs)
jac_asc_cat <- jaccard_matrix(mat_asc_cat)
```

---

## 6. Co-occurrence Heatmaps

```{r heatmap-function}
make_gene_annotation <- function(df) {
  df %>%
    select(GENE, RESISTANCE) %>%
    filter(!is.na(RESISTANCE), RESISTANCE != "") %>%
    mutate(RESISTANCE_CLASS = str_to_lower(str_extract(RESISTANCE, "^[^;]+"))) %>%
    group_by(GENE) %>%
    slice(1) %>%
    ungroup() %>%
    select(GENE, RESISTANCE_CLASS) %>%
    column_to_rownames("GENE")
}

ann_evi_rvs <- make_gene_annotation(evi_rvs)
ann_evi_cat <- make_gene_annotation(evi_cat)
ann_asc_rvs <- make_gene_annotation(asc_rvs)
ann_asc_cat <- make_gene_annotation(asc_cat)

heatmap_colors <- colorRampPalette(c("#FFFFFF", "#FFF3CD", "#FF9B00", "#D62728"))(100)

plot_cooccurrence <- function(jac_mat, ann_df, title, max_genes = 60,
                               show_names = TRUE, save_name = NULL) {
  if (nrow(jac_mat) > max_genes) {
    top_genes <- names(sort(rowMeans(jac_mat), decreasing = TRUE)[seq_len(max_genes)])
    jac_mat   <- jac_mat[top_genes, top_genes]
    ann_df    <- ann_df[rownames(ann_df) %in% top_genes, , drop = FALSE]
    title     <- paste0(title, "\n(top ", max_genes, " genes by mean co-occurrence)")
  }
  ann_plot <- ann_df[rownames(jac_mat), , drop = FALSE]

  # Scale font sizes for number display
  n        <- nrow(jac_mat)
  fnt_cell <- if (n > 40) 4 else if (n > 20) 6 else 7
  fnt_lab  <- if (n > 40) 5.5 else 7

  p <- pheatmap(
    jac_mat,
    color                    = heatmap_colors,
    clustering_distance_rows = "euclidean",
    clustering_distance_cols = "euclidean",
    clustering_method        = "ward.D2",
    annotation_row           = ann_plot,
    annotation_col           = ann_plot,
    display_numbers          = matrix(sprintf("%.2f", round(jac_mat, 2)),
                                      nrow = nrow(jac_mat)),
    number_color             = "black",
    fontsize_number          = fnt_cell,
    show_rownames            = show_names,
    show_colnames            = show_names,
    fontsize_row             = fnt_lab,
    fontsize_col             = fnt_lab,
    main                     = title,
    border_color             = "grey80",
    legend_breaks            = c(0, 0.25, 0.5, 0.75, 1),
    legend_labels            = c("0.00", "0.25", "0.50", "0.75", "1.00"),
    breaks                   = seq(0, 1, length.out = 101),
    silent                   = TRUE
  )

  # Widen output for gene labels + annotation sidebar
  w <- max(11, min(20, n * 0.35 + 6))
  h <- max(9,  min(18, n * 0.35 + 4))

  if (!is.null(save_name)) {
    ggsave(file.path(plot_dir, paste0(save_name, ".png")),
           plot = p$gtable, width = w, height = h, dpi = 300)
    ggsave(file.path(plot_dir, paste0(save_name, ".pdf")),
           plot = p$gtable, width = w, height = h)
  }
  print(p)
}
```

### 6.1 Post Evisceration – RVS Broth Enriched

```{r heatmap-evi-rvs}
plot_cooccurrence(jac_evi_rvs, ann_evi_rvs,
                  "Gene Co-occurrence – Post Evisceration, RVS Broth Enriched",
                  save_name = "gene_cooc_EVI_RVS")
```

### 6.2 Post Evisceration – Bolton Broth + CAT Enriched

```{r heatmap-evi-cat}
plot_cooccurrence(jac_evi_cat, ann_evi_cat,
                  "Gene Co-occurrence – Post Evisceration, Bolton Broth + CAT Enriched",
                  save_name = "gene_cooc_EVI_CAT")
```

### 6.3 Post ASC Wash – RVS Broth Enriched

```{r heatmap-asc-rvs}
plot_cooccurrence(jac_asc_rvs, ann_asc_rvs,
                  "Gene Co-occurrence – Post ASC Wash, RVS Broth Enriched",
                  save_name = "gene_cooc_ASC_RVS")
```

### 6.4 Post ASC Wash – Bolton Broth + CAT Enriched

```{r heatmap-asc-cat}
plot_cooccurrence(jac_asc_cat, ann_asc_cat,
                  "Gene Co-occurrence – Post ASC Wash, Bolton Broth + CAT Enriched",
                  save_name = "gene_cooc_ASC_CAT")
```

---

## 7. Cross-dataset Co-occurrence (EVI vs ASC, shared genes)

```{r shared-genes}
shared_rvs <- intersect(rownames(mat_evi_rvs), rownames(mat_asc_rvs))
shared_cat <- intersect(rownames(mat_evi_cat), rownames(mat_asc_cat))
cat("Shared genes – RVS broth:", length(shared_rvs), "\n")
cat("Shared genes – CAT broth:", length(shared_cat), "\n")
```

```{r combined-rvs, fig.height=10, fig.width=12}
if (length(shared_rvs) >= 3) {
  rvs_combined     <- cbind(mat_evi_rvs[shared_rvs, ], mat_asc_rvs[shared_rvs, ])
  jac_rvs_combined <- jaccard_matrix(rvs_combined)
  gene_ann_rvs     <- make_gene_annotation(bind_rows(evi_rvs, asc_rvs))
  gene_ann_rvs     <- gene_ann_rvs[rownames(gene_ann_rvs) %in% shared_rvs, , drop = FALSE]

  p <- pheatmap(
    jac_rvs_combined,
    color            = heatmap_colors,
    clustering_distance_rows = "euclidean",
    clustering_distance_cols = "euclidean",
    clustering_method        = "ward.D2",
    annotation_row   = gene_ann_rvs[rownames(jac_rvs_combined), , drop = FALSE],
    show_rownames    = TRUE, show_colnames    = TRUE,
    fontsize_row     = if (nrow(jac_rvs_combined) > 40) 5.5 else 7,
    fontsize_col     = if (ncol(jac_rvs_combined) > 40) 5.5 else 7,
    main             = "Co-occurrence – Shared RVS Broth Genes (EVI + ASC combined)",
    border_color     = NA, breaks = seq(0, 1, length.out = 101),
    silent           = TRUE
  )
  ggsave(file.path(plot_dir, "heatmap_shared_RVS.png"),
         plot = p$gtable, width = 12, height = 10, dpi = 300)
  ggsave(file.path(plot_dir, "heatmap_shared_RVS.pdf"),
         plot = p$gtable, width = 12, height = 10)
  print(p)
}
```

```{r combined-cat, fig.height=10, fig.width=12}
if (length(shared_cat) >= 3) {
  cat_combined     <- cbind(mat_evi_cat[shared_cat, ], mat_asc_cat[shared_cat, ])
  jac_cat_combined <- jaccard_matrix(cat_combined)
  gene_ann_cat     <- make_gene_annotation(bind_rows(evi_cat, asc_cat))
  gene_ann_cat     <- gene_ann_cat[rownames(gene_ann_cat) %in% shared_cat, , drop = FALSE]

  p <- pheatmap(
    jac_cat_combined,
    color            = heatmap_colors,
    clustering_distance_rows = "euclidean",
    clustering_distance_cols = "euclidean",
    clustering_method        = "ward.D2",
    annotation_row   = gene_ann_cat[rownames(jac_cat_combined), , drop = FALSE],
    show_rownames    = TRUE, show_colnames    = TRUE,
    fontsize_row     = if (nrow(jac_cat_combined) > 40) 5.5 else 7,
    fontsize_col     = if (ncol(jac_cat_combined) > 40) 5.5 else 7,
    main             = "Co-occurrence – Shared CAT Broth Genes (EVI + ASC combined)",
    border_color     = NA, breaks = seq(0, 1, length.out = 101),
    silent           = TRUE
  )
  ggsave(file.path(plot_dir, "heatmap_shared_CAT.png"),
         plot = p$gtable, width = 12, height = 10, dpi = 300)
  ggsave(file.path(plot_dir, "heatmap_shared_CAT.pdf"),
         plot = p$gtable, width = 12, height = 10)
  print(p)
}
```

---

## 8. Top Co-occurring Gene Pairs

```{r top-pairs-function}
top_pairs <- function(jac_mat, dataset_label, threshold = 0.75) {
  melt(jac_mat, varnames = c("Gene1", "Gene2"), value.name = "Jaccard") %>%
    filter(as.character(Gene1) < as.character(Gene2), Jaccard >= threshold) %>%
    arrange(desc(Jaccard)) %>%
    mutate(Dataset = dataset_labels[dataset_label]) %>%
    select(Dataset, Gene1, Gene2, Jaccard)
}

pairs_all <- bind_rows(
  top_pairs(jac_evi_rvs, "EVI_RVS"),
  top_pairs(jac_evi_cat, "EVI_CAT"),
  top_pairs(jac_asc_rvs, "ASC_RVS"),
  top_pairs(jac_asc_cat, "ASC_CAT")
)

if (nrow(pairs_all) > 0) {
  knitr::kable(pairs_all, digits = 3,
               caption = "Gene pairs with Jaccard co-occurrence ≥ 0.75")
} else {
  cat("No gene pairs exceed 0.75. Showing top 20:\n")
  bind_rows(
    top_pairs(jac_evi_rvs, "EVI_RVS", 0), top_pairs(jac_evi_cat, "EVI_CAT", 0),
    top_pairs(jac_asc_rvs, "ASC_RVS", 0), top_pairs(jac_asc_cat, "ASC_CAT", 0)
  ) %>% arrange(desc(Jaccard)) %>% slice_head(n = 20) %>%
    knitr::kable(digits = 3, caption = "Top 20 gene pairs by co-occurrence")
}
```

---

## 9. Resistance Class Co-occurrence

```{r class-cooccurrence-function}
class_cooccurrence <- function(pres_mat, df) {
  gene_class <- df %>%
    select(GENE, RESISTANCE) %>%
    distinct(GENE, .keep_all = TRUE) %>%
    mutate(RESISTANCE_CLASS = str_to_lower(str_extract(RESISTANCE, "^[^;]+"))) %>%
    filter(!is.na(RESISTANCE_CLASS), RESISTANCE_CLASS != "na",
           GENE %in% rownames(pres_mat))

  classes <- unique(gene_class$RESISTANCE_CLASS)
  samples <- colnames(pres_mat)

  class_mat <- sapply(classes, function(cls) {
    g <- gene_class$GENE[gene_class$RESISTANCE_CLASS == cls]
    g <- g[g %in% rownames(pres_mat)]
    if (!length(g)) return(rep(0L, length(samples)))
    apply(pres_mat[g, , drop = FALSE], 2, function(x) as.integer(any(x == 1)))
  })
  rownames(class_mat) <- samples
  t(class_mat) %*% class_mat
}

plot_class_cooc <- function(cooc_mat, title, save_name = NULL) {
  n <- nrow(cooc_mat)
  jac_class <- matrix(0, n, n, dimnames = dimnames(cooc_mat))
  for (i in seq_len(n)) for (j in seq_len(n)) {
    u <- cooc_mat[i,i] + cooc_mat[j,j] - cooc_mat[i,j]
    jac_class[i,j] <- if (u == 0) 0 else cooc_mat[i,j] / u
  }

  p <- pheatmap(
    jac_class,
    color            = heatmap_colors,
    clustering_method = "ward.D2",
    display_numbers  = round(jac_class, 2),
    number_color     = "black",
    fontsize_number  = 7,
    main             = title,
    border_color     = "grey80",
    breaks           = seq(0, 1, length.out = 101),
    silent           = TRUE
  )
  if (!is.null(save_name)) {
    ggsave(file.path(plot_dir, paste0(save_name, ".png")),
           plot = p$gtable, width = 10, height = 8, dpi = 300)
    ggsave(file.path(plot_dir, paste0(save_name, ".pdf")),
           plot = p$gtable, width = 10, height = 8)
  }
  print(p)
}
```

### 9.1 Post Evisceration – RVS Broth Enriched

```{r class-evi-rvs, fig.height=8, fig.width=10}
cooc_evi_rvs <- class_cooccurrence(mat_evi_rvs, evi_rvs)
plot_class_cooc(cooc_evi_rvs,
                "Resistance Class Co-occurrence – Post Evisceration, RVS Broth Enriched",
                save_name = "class_cooc_EVI_RVS")
```

### 9.2 Post Evisceration – Bolton Broth + CAT Enriched

```{r class-evi-cat, fig.height=8, fig.width=10}
cooc_evi_cat <- class_cooccurrence(mat_evi_cat, evi_cat)
plot_class_cooc(cooc_evi_cat,
                "Resistance Class Co-occurrence – Post Evisceration, Bolton Broth + CAT Enriched",
                save_name = "class_cooc_EVI_CAT")
```

### 9.3 Post ASC Wash – RVS Broth Enriched

```{r class-asc-rvs, fig.height=8, fig.width=10}
cooc_asc_rvs <- class_cooccurrence(mat_asc_rvs, asc_rvs)
plot_class_cooc(cooc_asc_rvs,
                "Resistance Class Co-occurrence – Post ASC Wash, RVS Broth Enriched",
                save_name = "class_cooc_ASC_RVS")
```

### 9.4 Post ASC Wash – Bolton Broth + CAT Enriched

```{r class-asc-cat, fig.height=8, fig.width=10}
cooc_asc_cat <- class_cooccurrence(mat_asc_cat, asc_cat)
plot_class_cooc(cooc_asc_cat,
                "Resistance Class Co-occurrence – Post ASC Wash, Bolton Broth + CAT Enriched",
                save_name = "class_cooc_ASC_CAT")
```

---

## 10. Co-occurrence Network Graphs

Resistance genes are shown as nodes, sized by **number of detections** (the
number of samples a gene was found in) and coloured by resistance class.
Edges connect gene pairs that co-occur; edge style reflects Jaccard distance (JD = 1 − Jaccard similarity):

- **Thick solid** — JD < 0.15 (strong co-occurrence)
- **Thin solid**  — JD 0.15–0.30 (moderate co-occurrence)
- **Dashed**      — JD 0.30–0.50 (weak co-occurrence; only shown when `jd_display_max` > 0.30)

Genes are clustered by the Louvain community detection algorithm.
Edges with JD ≥ `jd_display_max` (default 0.30) are hidden to reduce clutter.
Stress layout (graphlayouts) is used in place of FR for better node separation.

```{r network-setup}
# ── Resistance class colour palette ──────────────────────────────────────────
# Classes are mapped to canonical names; synonyms from ABRicate databases are
# collapsed so the legend never contains duplicate-colour entries or NA.
res_class_pal <- c(
  "beta-lactam"          = "#5B7FBF",
  "cephalosporin"        = "#5B7FBF",
  "penam"                = "#5B7FBF",
  "carbapenem"           = "#3A5FA0",
  "tetracycline"         = "#3A9E4D",
  "aminoglycoside"       = "#B0B0B0",
  "quinolone"            = "#1F3A7A",
  "fluoroquinolone"      = "#1F3A7A",
  "macrolide"            = "#8DB4E2",
  "phenicol"             = "#404040",
  "lincosamide"          = "#E8A0C8",
  "streptogramin"        = "#D46FA0",
  "sulphonamide"         = "#5BC8D2",
  "sulfonamide"          = "#5BC8D2",
  "trimethoprim"         = "#1B6B2A",
  "glycopeptide"         = "#6B3FA0",
  "oxazolidinone"        = "#A07850",
  "pleuromutilin"        = "#E0B040",
  "rifamycin"            = "#B05030",
  "fosfomycin"           = "#E060C0",
  "fusidic acid"         = "#D08050",
  "fusidic_acid"         = "#D08050",
  "disinfectant/biocide" = "#505050",
  "disinfectant"         = "#505050",
  "multidrug efflux"     = "#909090",
  "multidrug"            = "#909090",
  "efflux"               = "#909090",
  "other"                = "#D0D0D0"
)

# Canonical display names for the legend (deduplicated palette)
res_class_pal_display <- c(
  "beta-lactam"          = "#5B7FBF",
  "tetracycline"         = "#3A9E4D",
  "aminoglycoside"       = "#B0B0B0",
  "quinolone"            = "#1F3A7A",
  "macrolide"            = "#8DB4E2",
  "phenicol"             = "#404040",
  "lincosamide"          = "#E8A0C8",
  "streptogramin"        = "#D46FA0",
  "sulphonamide"         = "#5BC8D2",
  "trimethoprim"         = "#1B6B2A",
  "glycopeptide"         = "#6B3FA0",
  "oxazolidinone"        = "#A07850",
  "pleuromutilin"        = "#E0B040",
  "rifamycin"            = "#B05030",
  "fosfomycin"           = "#E060C0",
  "fusidic acid"         = "#D08050",
  "disinfectant/biocide" = "#505050",
  "multidrug efflux"     = "#909090",
  "other"                = "#D0D0D0"
)

# ── Build network from Jaccard matrix + gene metadata ────────────────────────
# jd_max: edges with JD ≥ jd_max excluded from graph (affects Louvain clustering)
build_gene_network <- function(jac_mat, df, jd_max = 0.4, max_clusters = 6) {

  nodes <- df %>%
    select(GENE, RESISTANCE, Count) %>%
    distinct(GENE, .keep_all = TRUE) %>%
    filter(GENE %in% rownames(jac_mat)) %>%
    mutate(
      res_class = str_to_lower(str_extract(RESISTANCE, "^[^;]+")) %>% str_trim(),
      res_class = if_else(is.na(res_class) | !res_class %in% names(res_class_pal),
                          "other", res_class),
      count_n   = as.integer(Count)
    ) %>%
    rename(name = GENE)

  jac_sub <- jac_mat[nodes$name, nodes$name]

  edges <- melt(jac_sub, varnames = c("from", "to"), value.name = "jaccard") %>%
    filter(as.character(from) < as.character(to)) %>%
    mutate(jd = 1 - jaccard) %>%
    filter(jd > 0, jd < jd_max) %>%
    mutate(edge_type = case_when(
      jd < 0.15 ~ "strong",
      jd < 0.30 ~ "moderate",
      TRUE      ~ "weak"
    ))

  if (nrow(edges) == 0) return(NULL)

  g    <- graph_from_data_frame(edges, directed = FALSE, vertices = nodes)
  comm <- cluster_louvain(g, weights = E(g)$jaccard)
  memb <- membership(comm)
  sizes <- table(memb)

  multi     <- names(sizes)[sizes > 1]
  size_rank <- rank(-sizes[multi], ties.method = "first")

  V(g)$cluster_label <- sapply(as.character(memb), function(m) {
    if (!m %in% multi) return("Isolated genes")
    r <- size_rank[m]
    if (r <= max_clusters) paste("Cluster", r) else "Isolated genes"
  })

  as_tbl_graph(g)
}

# ── Plot one cluster subgraph ─────────────────────────────────────────────────
.plot_cluster <- function(g_ig, node_df, cl, count_range,
                           jd_display_max, with_legend) {

  idx   <- which(node_df$cluster_label == cl)
  g_sub <- igraph::induced_subgraph(g_ig, idx)

  bad_e <- which(igraph::edge_attr(g_sub, "jd") >= jd_display_max)
  if (length(bad_e)) g_sub <- igraph::delete_edges(g_sub, bad_e)

  n_nodes <- igraph::vcount(g_sub)
  n_edges <- igraph::ecount(g_sub)

  # Layout: stress (graphlayouts) for connected clusters — better separation
  # than Fruchterman-Reingold; manual grid for no-edge panels.
  if (n_edges == 0) {
    nc  <- max(1, ceiling(sqrt(n_nodes)))
    x_g <- ((seq_len(n_nodes) - 1) %% nc) * 1.5
    y_g <- -((seq_len(n_nodes) - 1) %/% nc) * 1.5
    p_base <- ggraph(as_tbl_graph(g_sub), layout = "manual", x = x_g, y = y_g)
  } else {
    set.seed(42)
    p_base <- ggraph(as_tbl_graph(g_sub), layout = "stress")
  }

  # Adaptive label threshold: in dense panels only label genes detected in ≥2 samples
  v_df            <- igraph::as_data_frame(g_sub, what = "vertices")
  label_threshold <- if (n_nodes > 15) 2L else 1L

  # Colours limited to classes actually present in this subgraph (clean legend)
  classes_here <- unique(v_df$res_class)
  pal_here     <- res_class_pal_display[names(res_class_pal_display) %in% classes_here]

  p <- p_base +
    geom_node_point(
      aes(size = count_n, fill = res_class),
      shape = 21, colour = "white", stroke = 0.6
    ) +
    scale_size_continuous(
      name   = "No. of\ndetections",
      range  = c(3, 14),
      breaks = unique(round(seq(count_range[1], count_range[2], length.out = 4))),
      limits = count_range
    ) +
    scale_fill_manual(
      values       = res_class_pal,
      name         = "Resistance class",
      breaks       = names(pal_here),   # only classes present → no NA row
      labels       = names(pal_here),
      na.value     = "#C8C8C8",
      na.translate = FALSE
    ) +
    geom_node_text(
      aes(label = if_else(count_n >= label_threshold, name, "")),
      repel              = TRUE,
      size               = 2.5,
      fontface           = "italic",
      colour             = "grey10",
      max.overlaps       = 30,
      box.padding        = 0.4,
      point.padding      = 0.3,
      force              = 5,
      force_pull         = 0.8,
      min.segment.length = 0.1,
      segment.size       = 0.25,
      segment.color      = "grey50",
      segment.alpha      = 0.7
    ) +
    labs(title = cl) +
    theme_graph(base_size = 10) +
    theme(
      plot.title      = element_text(face = "bold", size = 11, hjust = 0.5,
                                     margin = margin(b = 4)),
      legend.position = if (with_legend) "right" else "none",
      legend.box      = "vertical",
      legend.key.size = unit(0.45, "cm"),
      legend.text     = element_text(size = 8),
      legend.title    = element_text(size = 8.5, face = "bold"),
      plot.margin     = margin(8, 8, 8, 8)
    )

  if (n_edges > 0) {
    p <- p +
      geom_edge_link(
        aes(edge_width = edge_type, edge_linetype = edge_type,
            edge_alpha  = edge_type),
        colour = "grey25"
      ) +
      scale_edge_width_manual(
        name   = "Co-occurrence\n(Jaccard distance)",
        values = c("strong" = 1.8, "moderate" = 0.8, "weak" = 0.35),
        labels = c("strong" = "JD < 0.15", "moderate" = "JD 0.15–0.30",
                   "weak"   = "JD 0.30–0.50"),
        limits = c("strong", "moderate", "weak"), drop = FALSE
      ) +
      scale_edge_linetype_manual(
        name   = "Co-occurrence\n(Jaccard distance)",
        values = c("strong" = "solid", "moderate" = "solid", "weak" = "dashed"),
        labels = c("strong" = "JD < 0.15", "moderate" = "JD 0.15–0.30",
                   "weak"   = "JD 0.30–0.50"),
        limits = c("strong", "moderate", "weak"), drop = FALSE
      ) +
      scale_edge_alpha_manual(
        values = c("strong" = 0.90, "moderate" = 0.60, "weak" = 0.30),
        guide  = "none"
      )
  }
  p
}

# ── Main plot function (patchwork of per-cluster panels) ──────────────────────
plot_gene_network <- function(tg, title, save_name = NULL,
                               jd_display_max = 0.30,
                               max_clusters   = 6,
                               width = 22, height = 16) {
  if (is.null(tg)) {
    message("No edges: ", title); return(invisible(NULL))
  }

  g_ig    <- as.igraph(tg)
  node_df <- igraph::as_data_frame(g_ig, what = "vertices")

  named_cls <- sort(unique(node_df$cluster_label[node_df$cluster_label != "Isolated genes"]))
  has_iso   <- "Isolated genes" %in% node_df$cluster_label
  all_cls   <- c(named_cls, if (has_iso) "Isolated genes")

  count_range <- range(node_df$count_n, na.rm = TRUE)

  panels <- lapply(seq_along(all_cls), function(i) {
    .plot_cluster(g_ig, node_df, all_cls[i], count_range,
                  jd_display_max,
                  with_legend = (i == length(named_cls)))
  })

  # Column count: 2 for ≤2 named clusters, 3 for more
  ncols <- if (length(named_cls) <= 2) 2L else 3L

  p_out <- wrap_plots(panels, ncol = ncols, guides = "collect") +
    plot_annotation(
      title   = title,
      caption = paste0(
        "Nodes sized by number of sample detections; coloured by resistance class.  ",
        "Edges shown where Jaccard distance < ", jd_display_max,
        " (thick solid: JD < 0.15; thin solid: JD 0.15–0.30; dashed: JD 0.30–0.50).  ",
        "Louvain community detection; singletons grouped as 'Isolated genes'."
      ),
      theme = theme(
        plot.title   = element_text(face = "bold", size = 14, margin = margin(b = 6)),
        plot.caption = element_text(size = 7, colour = "grey45", hjust = 0,
                                    margin = margin(t = 6))
      )
    )

  if (!is.null(save_name)) {
    ggsave(file.path(plot_dir, paste0(save_name, ".png")),
           plot = p_out, width = width, height = height, dpi = 300)
    ggsave(file.path(plot_dir, paste0(save_name, ".pdf")),
           plot = p_out, width = width, height = height)
  }
  p_out
}
```

### 9.1 Post Evisceration – RVS Broth Enriched

```{r network-evi-rvs, fig.width=22, fig.height=16}
net_evi_rvs <- build_gene_network(jac_evi_rvs, evi_rvs)
plot_gene_network(net_evi_rvs,
                  "AMR Gene Co-occurrence Network – Post Evisceration, RVS Broth Enriched",
                  save_name = "network_EVI_RVS")
```

### 9.2 Post Evisceration – Bolton Broth + CAT Enriched

```{r network-evi-cat, fig.width=22, fig.height=16}
net_evi_cat <- build_gene_network(jac_evi_cat, evi_cat)
plot_gene_network(net_evi_cat,
                  "AMR Gene Co-occurrence Network – Post Evisceration, Bolton Broth + CAT Enriched",
                  save_name = "network_EVI_CAT")
```

### 9.3 Post ASC Wash – RVS Broth Enriched

```{r network-asc-rvs, fig.width=22, fig.height=16}
net_asc_rvs <- build_gene_network(jac_asc_rvs, asc_rvs)
plot_gene_network(net_asc_rvs,
                  "AMR Gene Co-occurrence Network – Post ASC Wash, RVS Broth Enriched",
                  save_name = "network_ASC_RVS")
```

### 9.4 Post ASC Wash – Bolton Broth + CAT Enriched

```{r network-asc-cat, fig.width=22, fig.height=16}
net_asc_cat <- build_gene_network(jac_asc_cat, asc_cat)
plot_gene_network(net_asc_cat,
                  "AMR Gene Co-occurrence Network – Post ASC Wash, Bolton Broth + CAT Enriched",
                  save_name = "network_ASC_CAT")
```

---

## 11. Session Info

```{r session-info}
sessionInfo()
```
