---
title: "Read Level Normalised Bacterial Diversity - Chapter 3"
output: html_notebook
---

- EVI vs ASC samples 

*Sample normalisation:*
- Bases per million reads = (bases assigned to taxon / total bases in sample) × 1,000,000
- Then presented as relative abundance of classified samples

```{r Install packages}

library(tidyverse)
library(ggplot2)
library(patchwork)
library(scales)
library(vegan)
library(gridExtra)
library(ggtext)

```

```{r Label data}
# Enrichment type labels 

enrichment_labels <- c(
  "No enrichment"        = "No enrichment",
  "BPW then RVS enrichment" = "BPW then RVS enrichment",
  "BF Bolton Base + CAT" = "BF Bolton Base + CAT"
)

# Sample labels 

sample_labels <- c(
  "CH01_ASC"      = "CH01",
  "CH02_ASC"      = "CH02",
  "CH03_ASC"      = "CH03",
  "CH04_ASC"      = "CH04",
  "CH05_ASC"      = "CH05",
  "CH06_ASC"      = "CH06",
  "CH01_ASC_RVS"  = "CH01",
  "CH02_ASC_RVS"  = "CH02",
  "CH03_ASC_RVS"  = "CH03",
  "CH04_ASC_RVS"  = "CH04",
  "CH05_ASC_RVS"  = "CH05",
  "CH06_ASC_RVS"  = "CH06",
  "CH01_ASC_BFBB" = "CH01",
  "CH02_ASC_BFBB" = "CH02",
  "CH03_ASC_BFBB" = "CH03",
  "CH04_ASC_BFBB" = "CH04",
  "CH05_ASC_BFBB" = "CH05",
  "CH06_ASC_BFBB" = "CH06",
  "CH01_EVI"      = "CH01",
  "CH02_EVI"      = "CH02",
  "CH03_EVI"      = "CH03",
  "CH04_EVI"      = "CH04",
  "CH05_EVI"      = "CH05",
  "CH06_EVI"      = "CH06",
  "CH01_EVI_RVS"  = "CH01",
  "CH02_EVI_RVS"  = "CH02",
  "CH03_EVI_RVS"  = "CH03",
  "CH04_EVI_RVS"  = "CH04",
  "CH05_EVI_RVS"  = "CH05",
  "CH06_EVI_RVS"  = "CH06",
  "CH01_EVI_BFBB" = "CH01",
  "CH02_EVI_BFBB" = "CH02",
  "CH03_EVI_BFBB" = "CH03",
  "CH04_EVI_BFBB" = "CH04",
  "CH05_EVI_BFBB" = "CH05",
  "CH06_EVI_BFBB" = "CH06"
)

```


```{r Data import and set-up}

bpm_file <- "/Volumes/Alice/2_POULTRY_PROCESSING/Analysis/all_samples_bpm.tsv"
out_dir  <- "/Volumes/Alice/2_POULTRY_PROCESSING/Analysis/figures_read"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# Map group -> enrichment type
enrich_map <- c(
  "ASC"      = "No enrichment",
  "EVI"      = "No enrichment",
  "ASC_RVS"  = "BPW then RVS enrichment",
  "EVI_RVS"  = "BPW then RVS enrichment",
  "ASC_BFBB" = "BF Bolton Base + CAT",
  "EVI_BFBB" = "BF Bolton Base + CAT"
)

add_factors <- function(d) {
  d %>% mutate(
    group           = factor(group, levels = c("ASC", "ASC_RVS", "ASC_BFBB",
                                               "EVI", "EVI_RVS", "EVI_BFBB")),
    enrichment_type = factor(enrich_map[as.character(group)],
                             levels = names(enrichment_labels)),
    sample_label    = factor(sample_labels[sample],
                             levels = unique(sample_labels[unique(sample[order(group, sample)])]))
  )
}

# Full dataset (all superkingdoms) - used for the host proportion chart
df_raw <- read.delim(bpm_file, header = TRUE, sep = "\t", stringsAsFactors = FALSE) %>%
  add_factors()

# Bacteria-only dataset - used for all composition plots
# If superkingdom column is present (requires re-running kraken2_bpm.py), filter by it.
# Otherwise fall back to name-based exclusion of known eukaryotes.
if ("superkingdom" %in% names(df_raw)) {
  df <- df_raw %>% filter(superkingdom == "Bacteria")
} else {
  message("NOTE: 'superkingdom' column not found in TSV. ",
          "Re-run kraken2_bpm.py to enable automatic eukaryote filtering. ",
          "Using name-based fallback filter for now.")
  df <- df_raw %>%
    filter(!grepl(paste0("^Homo|Hominidae|Homininae|Hominini|Primates|",
                         "^Gallus|Phasianidae|Galliformes|Aves|",
                         "^Meleagris|^Sus |^Bos |^Felis|^Canis"),
                  taxon))
}

# Per-sample totals (denominators for classified/unclassified chart)
# Drawn from df_raw so total_bases reflects all sequenced bases
sample_totals <- df_raw %>%
  filter(tax_level == "genus") %>%
  group_by(enrichment_type, group, sample, sample_label) %>%
  summarise(
    total_bases      = first(total_bases),
    classified_bases = first(classified_bases),
    .groups = "drop"
  )

palette_bar <- c("#1F78B4", "#33A02C", "#E31A1C", "#FF7F00",
                 "#6A3D9A", "#B15928", "#A6CEE3", "#B2DF8A",
                 "#FB9A99", "#FDBF6F", "#CAB2D6", "#FFFF99")

```


```{r Plots}

tax_level_labels <- c(genus = "Genus", family = "Family", species = "Species")

for (tax_lvl in c("genus", "family", "species")) {
for (enrich in levels(df$enrichment_type)) {

  enrich_df <- df %>% filter(enrichment_type == enrich, tax_level == tax_lvl)

  # Skip if no data for this combination
  if (nrow(enrich_df) == 0) {
    message("Skipping (no data): ", enrich, " | ", tax_lvl)
    next
  }

  evi_samples <- sort(unique(enrich_df$sample[grepl("^EVI", as.character(enrich_df$group))]))
  asc_samples <- sort(unique(enrich_df$sample[grepl("^ASC", as.character(enrich_df$group))]))
  ordered_labels <- unique(sample_labels[c(evi_samples, asc_samples)])

  enrich_df <- enrich_df %>%
    mutate(
      sample_label = factor(sample_labels[sample], levels = ordered_labels),
      sample_type  = factor(if_else(grepl("^EVI", as.character(group)), "Post-EVI", "Post-ASC"),
                            levels = c("Post-EVI", "Post-ASC"))
    ) %>%
    droplevels()

  # Skip if sample_type has no levels after droplevels (all filtered out)
  if (nlevels(enrich_df$sample_type) == 0 || nrow(enrich_df) == 0) {
    message("Skipping (empty after filter): ", enrich, " | ", tax_lvl)
    next
  }

  enrich_label <- enrichment_labels[enrich]
  lvl_label    <- tax_level_labels[tax_lvl]
  lvl_plural   <- c(genus = "genera", family = "families", species = "species")[tax_lvl]
  file_slug    <- paste0(gsub("[^A-Za-z0-9]", "_", enrich), "_", tax_lvl)

  # Top 12 taxa for this enrichment type × taxonomic level
  top12 <- enrich_df %>%
    group_by(taxon) %>%
    summarise(mean_ra = mean(rel_abund), .groups = "drop") %>%
    slice_max(mean_ra, n = 12) %>%
    arrange(desc(mean_ra)) %>%
    pull(taxon)

  stack_colours <- c(setNames(palette_bar, top12), "Other" = "#CCCCCC")

  # Stacked bar
  other_df <- enrich_df %>%
    group_by(group, sample, sample_label) %>%
    summarise(
      rel_abund = 100 - sum(rel_abund[taxon %in% top12]),
      .groups   = "drop"
    ) %>%
    mutate(taxon = "Other")

  sample_type_map <- enrich_df %>% distinct(sample, sample_type)

  stack_df <- enrich_df %>%
    filter(taxon %in% top12) %>%
    select(group, sample, sample_label, sample_type, taxon, rel_abund) %>%
    bind_rows(other_df %>% left_join(sample_type_map, by = "sample")) %>%
    mutate(taxon = fct_reorder(taxon, rel_abund, .fun = mean))

  p_stack <- ggplot(stack_df, aes(x = sample_label, y = rel_abund, fill = taxon)) +
    geom_col(width = 0.7, colour = "white", linewidth = 0.3) +
    scale_fill_manual(values = stack_colours,
                      labels = function(b) ifelse(b == "Other", b, paste0("*", b, "*"))) +
    scale_y_continuous(labels = function(x) paste0(x, "%"),
                       expand = expansion(mult = c(0, 0.02))) +
    facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +
    labs(
      title    = paste0(enrich_label, " - ", lvl_label, " level"),
      subtitle = paste0("Top 12 ", lvl_plural, " by mean relative abundance"),
      x        = NULL,
      y        = "Relative abundance (% of classified bases)",
      fill     = lvl_label
    ) +
    theme_bw(base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(size = 10, colour = "grey40"),
      axis.text.x   = element_text(angle = 45, hjust = 1),
      legend.text   = element_markdown(size = 12),
      strip.text    = element_text(face = "bold", size = 12),
      panel.spacing = unit(0.8, "lines")
    )

  # Classified vs Unclassified (same for all tax levels - use genus-level totals)
  class_df <- sample_totals %>%
    filter(enrichment_type == enrich) %>%
    left_join(enrich_df %>% distinct(sample, sample_type), by = "sample") %>%
    filter(!is.na(sample_type)) %>%   # drop samples absent from enrich_df
    mutate(
      sample_label = factor(sample_labels[sample], levels = levels(enrich_df$sample_label)),
      Classified   = classified_bases / total_bases * 100,
      Unclassified = 100 - Classified
    ) %>%
    pivot_longer(cols = c(Classified, Unclassified),
                 names_to = "status", values_to = "pct")

  p_class <- ggplot(class_df, aes(x = sample_label, y = pct, fill = status)) +
    geom_col(width = 0.6, colour = "white", linewidth = 0.4) +
    scale_fill_manual(values = c("Classified" = "#1F78B4", "Unclassified" = "#B2DF8A")) +
    scale_y_continuous(labels = function(x) paste0(x, "%"),
                       expand = expansion(mult = c(0, 0.02))) +
    facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +
    labs(
      title = NULL,
      x     = NULL,
      y     = "% of total bases",
      fill  = NULL
    ) +
    theme_bw(base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold", size = 13),
      axis.text.x   = element_text(angle = 45, hjust = 1),
      strip.text    = element_text(face = "bold", size = 12),
      panel.spacing = unit(0.8, "lines")
    )

  # Combined panel
  p_combined <- p_stack / p_class +
    plot_layout(heights = c(2, 1))

  fname <- file.path(out_dir, paste0(file_slug, "_bpm_combined.png"))
  ggsave(fname, p_combined, width = 12, height = 10, dpi = 600)
  message("Saved: ", basename(fname))

  fname2 <- file.path(out_dir, paste0(file_slug, "_bpm_stacked.png"))
  ggsave(fname2, p_stack, width = 10, height = 7, dpi = 600)
  message("Saved: ", basename(fname2))

  # Extra plot for no-enrichment group only: heatmap of top 20 taxa
  if (enrich == "No enrichment") {

    # Top 20 taxa ordered ascending so the highest-abundance taxon sits at top of y-axis
    top20 <- enrich_df %>%
      group_by(taxon) %>%
      summarise(mean_ra = mean(rel_abund), .groups = "drop") %>%
      slice_max(mean_ra, n = 20, with_ties = FALSE) %>%
      arrange(mean_ra) %>%
      pull(taxon)

    sample_meta20 <- enrich_df %>% distinct(sample, sample_label, sample_type)

    grid20 <- expand.grid(
      taxon  = top20,
      sample = unique(enrich_df$sample),
      stringsAsFactors = FALSE
    ) %>%
      left_join(enrich_df %>% filter(taxon %in% top20) %>%
                  select(sample, taxon, rel_abund),
                by = c("taxon", "sample")) %>%
      left_join(sample_meta20, by = "sample") %>%
      mutate(
        rel_abund    = replace_na(rel_abund, 0),
        rel_plot     = ifelse(rel_abund > 0, rel_abund, NA_real_),
        taxon        = factor(taxon, levels = top20),
        sample_label = factor(as.character(sample_label),
                              levels = levels(enrich_df$sample_label))
      )

    p_heat <- ggplot(grid20, aes(x = sample_label, y = taxon, fill = rel_plot)) +
      geom_tile(colour = "white", linewidth = 0.6) +
      scale_fill_gradient(
        low      = "#D9F0E3",
        high     = "#00441B",
        trans    = "log10",
        labels   = function(x) paste0(x, "%"),
        na.value = "#EEEEEE",
        name     = "Rel. abundance\n(% of classified bases)"
      ) +
      scale_y_discrete(expand = expansion(add = 0.5)) +
      facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +
      labs(
        title    = paste0(enrich_label, " - Top 20 ", lvl_plural,
                          " by mean relative abundance (", lvl_label, " level)"),
        subtitle = "Colour intensity = relative abundance (% of classified bases); grey = not detected",
        x        = NULL,
        y        = NULL
      ) +
      theme_light(base_size = 13) +
      theme(
        axis.text.x     = element_text(angle = 30, hjust = 1, size = 11),
        axis.text.y     = element_text(size = 10, face = "italic"),
        plot.title      = element_text(face = "bold", size = 13),
        plot.subtitle   = element_text(size = 10, colour = "grey40"),
        strip.text      = element_text(face = "bold", size = 12),
        panel.spacing   = unit(0.8, "lines"),
        legend.position = "right",
        panel.grid      = element_blank()
      )

    fname3 <- file.path(out_dir, paste0("No_enrichment_bpm_top20_heatmap_", tax_lvl, ".png"))
    ggsave(fname3, p_heat, width = 12, height = 8, dpi = 600)
    message("Saved: ", basename(fname3))
  }

} # end enrich loop
} # end tax_lvl loop

message("\nAll done. Figures saved to: ", out_dir)

```


```{r Option 2 - threshold-based stacked bars}

# Option 2: show all taxa with mean rel_abund >= threshold; collapse rest into "Other"
# Adjust threshold_pct to taste (1 = 1%)
threshold_pct <- 1

for (tax_lvl in c("genus", "family", "species")) {
for (enrich in levels(df$enrichment_type)) {

  enrich_df <- df %>% filter(enrichment_type == enrich, tax_level == tax_lvl)
  if (nrow(enrich_df) == 0) next

  evi_samples    <- sort(unique(enrich_df$sample[grepl("^EVI", as.character(enrich_df$group))]))
  asc_samples    <- sort(unique(enrich_df$sample[grepl("^ASC", as.character(enrich_df$group))]))
  ordered_labels <- unique(sample_labels[c(evi_samples, asc_samples)])

  enrich_df <- enrich_df %>%
    mutate(
      sample_label = factor(sample_labels[sample], levels = ordered_labels),
      sample_type  = factor(if_else(grepl("^EVI", as.character(group)), "Post-EVI", "Post-ASC"),
                            levels = c("Post-EVI", "Post-ASC"))
    ) %>%
    droplevels()

  if (nlevels(enrich_df$sample_type) == 0 || nrow(enrich_df) == 0) next

  enrich_label <- enrichment_labels[enrich]
  lvl_label    <- tax_level_labels[tax_lvl]
  lvl_plural   <- c(genus = "genera", family = "families", species = "species")[tax_lvl]
  file_slug    <- paste0(gsub("[^A-Za-z0-9]", "_", enrich), "_", tax_lvl)

  # Select taxa that exceed the threshold on average
  above_thresh <- enrich_df %>%
    group_by(taxon) %>%
    summarise(mean_ra = mean(rel_abund), .groups = "drop") %>%
    filter(mean_ra >= threshold_pct) %>%
    arrange(desc(mean_ra)) %>%
    pull(taxon)

  n_above       <- length(above_thresh)
  palette_thr   <- colorRampPalette(palette_bar)(max(n_above, 1))
  thresh_colours <- c(setNames(palette_thr, above_thresh), "Other" = "#CCCCCC")

  sample_type_map <- enrich_df %>% distinct(sample, sample_type)

  other_thr_df <- enrich_df %>%
    group_by(group, sample, sample_label) %>%
    summarise(
      rel_abund = 100 - sum(rel_abund[taxon %in% above_thresh]),
      .groups   = "drop"
    ) %>%
    mutate(taxon = "Other")

  stack_thr_df <- enrich_df %>%
    filter(taxon %in% above_thresh) %>%
    select(group, sample, sample_label, sample_type, taxon, rel_abund) %>%
    bind_rows(other_thr_df %>% left_join(sample_type_map, by = "sample")) %>%
    mutate(taxon = fct_reorder(taxon, rel_abund, .fun = mean))

  p_thresh <- ggplot(stack_thr_df, aes(x = sample_label, y = rel_abund, fill = taxon)) +
    geom_col(width = 0.7, colour = "white", linewidth = 0.3) +
    scale_fill_manual(values = thresh_colours,
                      labels = function(b) ifelse(b == "Other", b, paste0("*", b, "*"))) +
    scale_y_continuous(labels = function(x) paste0(x, "%"),
                       expand = expansion(mult = c(0, 0.02))) +
    facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +
    labs(
      title    = paste0(enrich_label, " - ", lvl_label, " level"),
      subtitle = paste0(n_above, " ", lvl_plural, " with mean relative abundance ≥ ",
                        threshold_pct, "% shown individually"),
      x        = NULL,
      y        = "Relative abundance (% of classified bases)",
      fill     = lvl_label
    ) +
    theme_bw(base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(size = 10, colour = "grey40"),
      axis.text.x   = element_text(angle = 45, hjust = 1),
      legend.text   = element_markdown(size = 10),
      strip.text    = element_text(face = "bold", size = 12),
      panel.spacing = unit(0.8, "lines")
    )

  fname_thr <- file.path(out_dir, paste0(file_slug, "_threshold_bpm_stacked.png"))
  ggsave(fname_thr, p_thresh, width = 10, height = 7, dpi = 600)
  message("Saved: ", basename(fname_thr))

} # end enrich loop
} # end tax_lvl loop

message("\nAll done. Option 2 (", threshold_pct, "% threshold) figures saved to: ", out_dir)
# Note: Option 3 (unfiltered all-taxa) is already generated above for the No enrichment group only.

```


---

# *Klebsiella pneumoniae* relative abundance - BPW then RVS enrichment

Paired plot showing *Klebsiella pneumoniae* relative abundance in Post-EVI vs Post-ASC samples enriched with BPW then RVS. Each line connects the matched pair from the same chicken (CH01–CH06).

```{r Klebsiella pneumoniae BPW+RVS, fig.height=8, fig.width=6}

kp_taxon  <- "Klebsiella pneumoniae"   # species-level name in Kraken2 output
kp_enrich <- "BPW then RVS enrichment"

# Try species level first; fall back to genus if species not found
kp_df_sp <- df %>%
  filter(enrichment_type == kp_enrich,
         tax_level       == "species",
         taxon           == kp_taxon)

kp_df_ge <- df %>%
  filter(enrichment_type == kp_enrich,
         tax_level       == "genus",
         taxon           == "Klebsiella")

if (nrow(kp_df_sp) > 0) {
  kp_df     <- kp_df_sp
  kp_label  <- kp_taxon
  kp_lvl    <- "species"
} else if (nrow(kp_df_ge) > 0) {
  message("NOTE: '", kp_taxon, "' not found at species level; using genus 'Klebsiella'.")
  kp_df     <- kp_df_ge
  kp_label  <- "Klebsiella (genus)"
  kp_lvl    <- "genus"
} else {
  message("WARNING: ", kp_taxon, " not found in BPW then RVS data at species or genus level.")
  kp_df <- NULL
}

if (!is.null(kp_df) && nrow(kp_df) > 0) {

  kp_plot_df <- kp_df %>%
    mutate(
      sample_type  = factor(if_else(grepl("^EVI", as.character(group)),
                                    "Post-EVI", "Post-ASC"),
                            levels = c("Post-EVI", "Post-ASC")),
      chicken_id   = sub("^CH(\\d+)_.*", "CH\\1", sample)
    ) %>%
    select(sample, chicken_id, sample_type, rel_abund)

  # Wilcoxon test
  evi_vals <- kp_plot_df$rel_abund[kp_plot_df$sample_type == "Post-EVI"]
  asc_vals <- kp_plot_df$rel_abund[kp_plot_df$sample_type == "Post-ASC"]
  wx <- tryCatch(
    wilcox.test(evi_vals, asc_vals, paired = TRUE, exact = FALSE),
    error = function(e) NULL
  )
  p_label <- if (!is.null(wx)) {
    pv <- wx$p.value
    if (pv < 0.001) "paired Wilcoxon: ***" else if (pv < 0.01) "paired Wilcoxon: **" else
    if (pv < 0.05)  "paired Wilcoxon: *"   else paste0("paired Wilcoxon: p = ", round(pv, 3))
  } else ""

  p_kp <- ggplot(kp_plot_df, aes(x = sample_type, y = rel_abund)) +
    geom_jitter(aes(colour = sample_type), size = 3.5, alpha = 0.9, position = position_jitter(width = 0.05, height = 0, seed = 42)) +
    scale_colour_manual(values = c("Post-EVI" = "#377EB8", "Post-ASC" = "#E41A1C")) +
    scale_y_continuous(labels = function(x) paste0(x, "%"),
                       limits = c(0, NA),
                       expand = expansion(mult = c(0.02, 0.1))) +
    labs(
      title    = paste0("<b><i>", kp_label, "</i></b> - BPW then RVS enrichment"),
      subtitle = paste0("Relative abundance (% of classified bases)  ", p_label),
      x        = NULL,
      y        = "Relative abundance (%)",
      colour   = NULL
    ) +
    theme_bw(base_size = 13) +
    theme(
      plot.title      = element_markdown(size = 13),
      plot.subtitle   = element_text(size = 10, colour = "grey40"),
      axis.text.x     = element_text(size = 12),
      legend.position = "none"
    )

  # Log10 version - add small pseudocount so zero values are visible
  pseudocount <- 0.01
  p_kp_log <- p_kp +
    scale_y_log10(labels = function(x) paste0(x, "%"),
                  expand = expansion(mult = c(0.05, 0.1))) +
    geom_jitter(data = kp_plot_df %>% mutate(rel_abund = rel_abund + pseudocount),
                aes(colour = sample_type), width = 0.05, height = 0,
                size = 3.5, alpha = 0.9, seed = 42) +
    labs(subtitle = paste0("Log10 scale (+ ", pseudocount, "% pseudocount)  ", p_label))

  kp_table <- kp_plot_df %>%
    arrange(sample_type, chicken_id) %>%
    select(chicken_id, sample_type, rel_abund) %>%
    mutate(rel_abund = round(rel_abund, 4)) %>%
    rename(`Chicken` = chicken_id, `Group` = sample_type, `Rel. abundance (%)` = rel_abund)
  print(kp_table)

  print(p_kp)
  print(p_kp_log)

  kp_fname <- file.path(out_dir, paste0("Klebsiella_pneumoniae_BPW_RVS_", kp_lvl, ".png"))
  ggsave(kp_fname, p_kp, width = 6, height = 8, dpi = 600)
  message("Saved: ", basename(kp_fname))

  kp_fname_log <- file.path(out_dir, paste0("Klebsiella_pneumoniae_BPW_RVS_", kp_lvl, "_log10.png"))
  ggsave(kp_fname_log, p_kp_log, width = 6, height = 8, dpi = 600)
  message("Saved: ", basename(kp_fname_log))
}

```


---

# *Salmonella enterica* Relative Abundance - All Enrichment Types

Boxplot with individual sample points (Post-EVI vs Post-ASC, unpaired) showing *Salmonella enterica* relative abundance across all enrichment types, including non-enriched samples. Samples with no detected reads shown as zero.

```{r Salmonella enterica all enrichments, fig.height=6, fig.width=10}

se_taxon <- "Salmonella enterica"

# Try species level first; fall back to genus
se_df_sp <- df %>%
  filter(tax_level == "species", taxon == se_taxon)

se_df_ge <- df %>%
  filter(tax_level == "genus", taxon == "Salmonella")

if (nrow(se_df_sp) > 0) {
  se_df    <- se_df_sp
  se_label <- se_taxon
  se_lvl   <- "species"
} else if (nrow(se_df_ge) > 0) {
  message("NOTE: '", se_taxon, "' not found at species level; using genus 'Salmonella'.")
  se_df    <- se_df_ge
  se_label <- "Salmonella (genus)"
  se_lvl   <- "genus"
} else {
  message("WARNING: ", se_taxon, " not found in data at species or genus level.")
  se_df <- NULL
}

if (!is.null(se_df) && nrow(se_df) > 0) {

  # Build complete sample scaffold so samples with 0 reads still appear
  all_samples_se <- df %>%
    filter(tax_level == "species") %>%
    distinct(group, sample, enrichment_type)

  se_plot_df <- all_samples_se %>%
    left_join(se_df %>% select(sample, rel_abund),
              by = "sample") %>%
    replace_na(list(rel_abund = 0)) %>%
    mutate(
      sample_type  = factor(if_else(grepl("^EVI", as.character(group)),
                                    "Post-EVI", "Post-ASC"),
                            levels = c("Post-EVI", "Post-ASC")),
      enrich_label = factor(enrichment_labels[as.character(enrichment_type)],
                            levels = unname(enrichment_labels))
    )

  se_detect_df <- se_plot_df %>% filter(rel_abund > 0)

  p_se <- ggplot(se_detect_df,
                 aes(x = sample_type, y = rel_abund, colour = sample_type, fill = sample_type)) +
    geom_boxplot(width = 0.45, alpha = 0.3, outlier.shape = NA, colour = "grey30") +
    geom_jitter(size = 3, alpha = 0.85, position = position_jitter(width = 0.1, height = 0, seed = 42)) +
    scale_colour_manual(values = c("Post-EVI" = "#377EB8", "Post-ASC" = "#E41A1C"),
                        drop = FALSE) +
    scale_fill_manual(values   = c("Post-EVI" = "#377EB8", "Post-ASC" = "#E41A1C"),
                      drop = FALSE) +
    scale_x_discrete(drop = FALSE) +
    scale_y_continuous(labels = function(x) paste0(x, "%"),
                       limits = c(0, NA),
                       expand = expansion(mult = c(0.02, 0.08))) +
    facet_wrap(~ enrich_label, nrow = 1) +
    labs(
      title    = paste0("<b><i>", se_label, "</i></b> - relative abundance by sample type"),
      subtitle = "Relative abundance (% of classified bases). Data points absent where taxon not detected in sample.",
      x        = NULL,
      y        = "Relative abundance (%)",
      colour   = NULL
    ) +
    theme_bw(base_size = 13) +
    theme(
      plot.title      = element_markdown(size = 13),
      plot.subtitle   = element_text(size = 10, colour = "grey40"),
      axis.text.x     = element_text(size = 12),
      strip.text      = element_text(face = "bold", size = 11),
      panel.spacing   = unit(0.8, "lines"),
      legend.position = "none"
    )

  print(p_se)

  se_fname <- file.path(out_dir, paste0("Salmonella_enterica_all_enrichments_",
                                        se_lvl, ".png"))
  ggsave(se_fname, p_se, width = 10, height = 6, dpi = 600)
  message("Saved: ", basename(se_fname))

  # Log10 version - only detected samples, pseudocount added for log scale
  pseudocount <- 0.01
  se_log_df <- se_detect_df %>% mutate(rel_abund = rel_abund + pseudocount)
  p_se_log <- ggplot(se_log_df,
                     aes(x = sample_type, y = rel_abund,
                         colour = sample_type, fill = sample_type)) +
    geom_boxplot(width = 0.45, alpha = 0.3, outlier.shape = NA, colour = "grey30") +
    geom_jitter(size = 3, alpha = 0.85, position = position_jitter(width = 0.1, height = 0, seed = 42)) +
    scale_colour_manual(values = c("Post-EVI" = "#377EB8", "Post-ASC" = "#E41A1C"),
                        drop = FALSE) +
    scale_fill_manual(values   = c("Post-EVI" = "#377EB8", "Post-ASC" = "#E41A1C"),
                      drop = FALSE) +
    scale_x_discrete(drop = FALSE) +
    scale_y_log10(labels = function(x) paste0(x, "%"),
                  expand = expansion(mult = c(0.05, 0.08))) +
    facet_wrap(~ enrich_label, nrow = 1) +
    labs(
      title    = paste0("<b><i>", se_label, "</i></b> - relative abundance by sample type"),
      subtitle = paste0("Log10 scale (+ ", pseudocount,
                        "% pseudocount). Data points absent where taxon not detected in sample."),
      x = NULL, y = "Relative abundance (%)", colour = NULL, fill = NULL
    ) +
    theme_bw(base_size = 13) +
    theme(
      plot.title      = element_markdown(size = 13),
      plot.subtitle   = element_text(size = 10, colour = "grey40"),
      axis.text.x     = element_text(size = 12),
      strip.text      = element_text(face = "bold", size = 11),
      panel.spacing   = unit(0.8, "lines"),
      legend.position = "none"
    )

  print(p_se_log)

  se_fname_log <- file.path(out_dir, paste0("Salmonella_enterica_all_enrichments_",
                                             se_lvl, "_log10.png"))
  ggsave(se_fname_log, p_se_log, width = 10, height = 6, dpi = 600)
  message("Saved: ", basename(se_fname_log))

  # Per-sample relative abundance table
  se_table <- se_plot_df %>%
    mutate(
      Sample      = sample_labels[sample],
      `Sample type` = as.character(sample_type),
      Enrichment  = as.character(enrich_label),
      `Relative abundance (%)` = ifelse(rel_abund == 0, "Not detected",
                                         paste0(round(rel_abund, 4), "%"))
    ) %>%
    arrange(enrich_label, sample_type, Sample) %>%
    select(Enrichment, `Sample type`, Sample, `Relative abundance (%)`)

  print(knitr::kable(se_table,
                     caption = paste0("*", se_label, "* - relative abundance per sample (read-level)"),
                     align   = c("l", "l", "l", "r")))

} else {
  message("No Salmonella data found - skipping figure.")
}

```


---

# *Campylobacter* Relative Abundance - All Enrichment Types

Boxplot with individual sample points (Post-EVI vs Post-ASC, unpaired) showing *Campylobacter* relative abundance at genus level across all enrichment types. Mean relative abundance per group is printed in the summary table. Samples with no detected reads are shown as zero.

```{r Campylobacter genus all enrichments, fig.height=6, fig.width=10}

camp_taxon <- "Campylobacter"

camp_df <- df %>%
  filter(tax_level == "genus", taxon == camp_taxon)

if (nrow(camp_df) > 0) {

  # Complete scaffold so samples with 0 reads still appear
  all_samples_camp <- df %>%
    filter(tax_level == "genus") %>%
    distinct(group, sample, enrichment_type)

  camp_plot_df <- all_samples_camp %>%
    left_join(camp_df %>% select(sample, rel_abund), by = "sample") %>%
    replace_na(list(rel_abund = 0)) %>%
    mutate(
      sample_type  = factor(if_else(grepl("^EVI", as.character(group)),
                                    "Post-EVI", "Post-ASC"),
                            levels = c("Post-EVI", "Post-ASC")),
      enrich_label = factor(enrichment_labels[as.character(enrichment_type)],
                            levels = unname(enrichment_labels))
    )

  # Mean relative abundance summary table (wide format)
  camp_summary <- camp_plot_df %>%
    group_by(enrich_label, sample_type) %>%
    summarise(mean_rel_abund = round(mean(rel_abund), 4), .groups = "drop") %>%
    pivot_wider(names_from = sample_type, values_from = mean_rel_abund) %>%
    rename(Enrichment = enrich_label)

  print(knitr::kable(camp_summary,
                     caption = paste0("*", camp_taxon, "* (genus) - mean relative abundance (%) across six samples per group"),
                     col.names = c("Enrichment", "Post-EVI mean (%)", "Post-ASC mean (%)"),
                     align     = c("l", "r", "r")))

  # Overall mean pooled across all enrichment types
  camp_overall <- camp_plot_df %>%
    group_by(sample_type) %>%
    summarise(mean_rel_abund = round(mean(rel_abund), 4), .groups = "drop") %>%
    pivot_wider(names_from = sample_type, values_from = mean_rel_abund)

  print(knitr::kable(camp_overall,
                     caption = paste0("*", camp_taxon, "* (genus) - overall mean relative abundance (%) pooled across all enrichment types"),
                     col.names = c("Post-EVI mean (%)", "Post-ASC mean (%)"),
                     align     = c("r", "r")))

  # Detect samples only for boxplot (zero-reads shown if all samples zero)
  camp_detect_df <- camp_plot_df

  p_camp <- ggplot(camp_detect_df,
                   aes(x = sample_type, y = rel_abund, colour = sample_type, fill = sample_type)) +
    geom_boxplot(width = 0.45, alpha = 0.3, outlier.shape = NA, colour = "grey30") +
    geom_jitter(size = 3, alpha = 0.85, position = position_jitter(width = 0.1, height = 0, seed = 42)) +
    scale_colour_manual(values = c("Post-EVI" = "#377EB8", "Post-ASC" = "#E41A1C"), drop = FALSE) +
    scale_fill_manual(values   = c("Post-EVI" = "#377EB8", "Post-ASC" = "#E41A1C"), drop = FALSE) +
    scale_x_discrete(drop = FALSE) +
    scale_y_continuous(labels = function(x) paste0(x, "%"),
                       limits = c(0, NA),
                       expand = expansion(mult = c(0.02, 0.08))) +
    facet_wrap(~ enrich_label, nrow = 1) +
    labs(
      title    = paste0("<b><i>", camp_taxon, "</i></b> (genus) - relative abundance by sample type"),
      subtitle = "Relative abundance (% of classified bases). Samples with no detected reads shown as zero.",
      x        = NULL,
      y        = "Relative abundance (%)",
      colour   = NULL
    ) +
    theme_bw(base_size = 13) +
    theme(
      plot.title      = element_markdown(size = 13),
      plot.subtitle   = element_text(size = 10, colour = "grey40"),
      axis.text.x     = element_text(size = 12),
      strip.text      = element_text(face = "bold", size = 11),
      panel.spacing   = unit(0.8, "lines"),
      legend.position = "none"
    )

  print(p_camp)

  camp_fname <- file.path(out_dir, "Campylobacter_genus_all_enrichments.png")
  ggsave(camp_fname, p_camp, width = 10, height = 6, dpi = 600)
  message("Saved: ", basename(camp_fname))

  # Log10 version
  pseudocount <- 0.01
  camp_log_df <- camp_detect_df %>% mutate(rel_abund = rel_abund + pseudocount)
  p_camp_log <- ggplot(camp_log_df,
                       aes(x = sample_type, y = rel_abund,
                           colour = sample_type, fill = sample_type)) +
    geom_boxplot(width = 0.45, alpha = 0.3, outlier.shape = NA, colour = "grey30") +
    geom_jitter(size = 3, alpha = 0.85, position = position_jitter(width = 0.1, height = 0, seed = 42)) +
    scale_colour_manual(values = c("Post-EVI" = "#377EB8", "Post-ASC" = "#E41A1C"), drop = FALSE) +
    scale_fill_manual(values   = c("Post-EVI" = "#377EB8", "Post-ASC" = "#E41A1C"), drop = FALSE) +
    scale_x_discrete(drop = FALSE) +
    scale_y_log10(labels = function(x) paste0(x, "%"),
                  expand = expansion(mult = c(0.05, 0.08))) +
    facet_wrap(~ enrich_label, nrow = 1) +
    labs(
      title    = paste0("<b><i>", camp_taxon, "</i></b> (genus) - relative abundance by sample type"),
      subtitle = paste0("Log10 scale (+ ", pseudocount, "% pseudocount)."),
      x = NULL, y = "Relative abundance (%)", colour = NULL, fill = NULL
    ) +
    theme_bw(base_size = 13) +
    theme(
      plot.title      = element_markdown(size = 13),
      plot.subtitle   = element_text(size = 10, colour = "grey40"),
      axis.text.x     = element_text(size = 12),
      strip.text      = element_text(face = "bold", size = 11),
      panel.spacing   = unit(0.8, "lines"),
      legend.position = "none"
    )

  print(p_camp_log)

  camp_fname_log <- file.path(out_dir, "Campylobacter_genus_all_enrichments_log10.png")
  ggsave(camp_fname_log, p_camp_log, width = 10, height = 6, dpi = 600)
  message("Saved: ", basename(camp_fname_log))

} else {
  message("No Campylobacter data found at genus level - skipping figure.")
}

```


---

# Top 10 Taxa by Relative Abundance — Per Sample Breakdown

For each enrichment type and taxonomic level (genus and species), the top 10 taxa are selected by mean relative abundance across all samples in that group. Values are shown per individual sample, with mean relative abundance for Post-EVI and Post-ASC groups. TSV files are saved to `out_dir`.

```{r Top 10 per-sample breakdown, results='asis'}

for (tax_lvl in c("genus", "species")) {
  for (enrich in levels(df$enrichment_type)) {

    enrich_df <- df %>%
      filter(enrichment_type == enrich, tax_level == tax_lvl) %>%
      mutate(
        sample_type  = if_else(grepl("^EVI", as.character(group)), "Post-EVI", "Post-ASC"),
        sample_label = sample_labels[sample],
        col_name     = paste0(if_else(sample_type == "Post-EVI", "EVI", "ASC"),
                              " ", sample_label)
      )

    if (nrow(enrich_df) == 0) {
      message("Skipping (no data): ", enrich, " | ", tax_lvl)
      next
    }

    # Top 10 taxa by mean relative abundance across all samples in this enrichment type
    top10 <- enrich_df %>%
      group_by(taxon) %>%
      summarise(mean_ra = mean(rel_abund), .groups = "drop") %>%
      slice_max(mean_ra, n = 10, with_ties = FALSE) %>%
      arrange(desc(mean_ra)) %>%
      pull(taxon)

    if (length(top10) == 0) next

    top10_df <- enrich_df %>% filter(taxon %in% top10)

    # Wide table: one row per taxon, one column per sample
    per_sample <- top10_df %>%
      select(taxon, col_name, rel_abund) %>%
      pivot_wider(names_from = col_name, values_from = rel_abund, values_fill = 0) %>%
      mutate(across(-taxon, ~ round(.x, 4)))

    # Group means (Post-EVI and Post-ASC separately)
    group_means <- top10_df %>%
      group_by(taxon, sample_type) %>%
      summarise(mean_ra = round(mean(rel_abund), 4), .groups = "drop") %>%
      pivot_wider(names_from = sample_type, values_from = mean_ra, values_fill = 0)

    if (!"Post-EVI" %in% names(group_means)) group_means[["Post-EVI"]] <- 0
    if (!"Post-ASC" %in% names(group_means)) group_means[["Post-ASC"]] <- 0

    group_means <- group_means %>%
      rename(`Mean Post-EVI (%)` = `Post-EVI`,
             `Mean Post-ASC (%)` = `Post-ASC`)

    # Column order: EVI samples → ASC samples → group means
    evi_cols <- sort(grep("^EVI ", names(per_sample), value = TRUE))
    asc_cols <- sort(grep("^ASC ", names(per_sample), value = TRUE))

    tbl <- per_sample %>%
      left_join(group_means, by = "taxon") %>%
      mutate(taxon = factor(taxon, levels = top10)) %>%
      arrange(taxon) %>%
      select(Taxon = taxon,
             all_of(evi_cols),
             all_of(asc_cols),
             `Mean Post-EVI (%)`,
             `Mean Post-ASC (%)`)

    lvl_label  <- tax_level_labels[tax_lvl]
    lvl_plural <- c(genus = "genera", family = "families", species = "species")[tax_lvl]
    enrich_lbl <- enrichment_labels[enrich]
    file_slug  <- paste0(gsub("[^A-Za-z0-9]", "_", enrich), "_", tax_lvl)

    cat("\n\n###", lvl_label, "level —", enrich_lbl, "\n\n")

    print(knitr::kable(tbl,
                       caption = paste0("Top 10 ", lvl_plural,
                                        " by mean relative abundance (%) — ", enrich_lbl),
                       digits  = 4,
                       align   = c("l", rep("r", ncol(tbl) - 1))))

    write.table(tbl,
                file.path(out_dir, paste0("top10_", file_slug, "_per_sample.tsv")),
                sep = "\t", row.names = FALSE, quote = FALSE)
    message("Saved: top10_", file_slug, "_per_sample.tsv")
  }
}

message("\nTop-10 per-sample breakdown tables done.")

```


---

# *Campylobacter* spp., *Salmonella* spp. & *Escherichia* spp. - Species Relative Abundance Heatmap (Reads)

Heatmap showing relative abundance of each species within each genus, per sample, by enrichment type. Tile colour intensity reflects relative abundance (% of classified bases); grey tiles indicate the species was not detected.

```{r read_genus_heatmap_fn}

plot_read_genus_heatmap <- function(spp_df, title, subtitle = NULL, save_path = NULL) {
  if (nrow(spp_df) == 0) { message("No data — skipping."); return(invisible(NULL)) }

  species_order <- spp_df %>%
    group_by(name) %>%
    summarise(m = mean(rel_abund, na.rm = TRUE), .groups = "drop") %>%
    arrange(m) %>% pull(name)

  sample_meta <- spp_df %>% distinct(sample, sample_label, sample_type)

  grid <- expand.grid(name   = species_order,
                      sample = unique(spp_df$sample),
                      stringsAsFactors = FALSE) %>%
    left_join(spp_df %>% select(sample, name, rel_abund), by = c("sample", "name")) %>%
    left_join(sample_meta, by = "sample") %>%
    mutate(
      rel_abund    = replace_na(rel_abund, 0),
      # Zero (not detected) is mapped to NA so it renders as a distinct grey tile
      # rather than the bottom of the colour gradient
      rel_plot     = ifelse(rel_abund > 0, rel_abund, NA_real_),
      name         = factor(name, levels = species_order),
      sample_label = factor(sample_label, levels = sort(unique(sample_meta$sample_label)))
    )

  # Wrap the title onto its own lines using an explicit break rather than
  # relying on auto text-wrapping, which can overlap the subtitle below
  title_wrapped <- gsub(" - ", "<br>", title)

  p <- ggplot(grid, aes(x = sample_label, y = name, fill = rel_plot)) +
    geom_tile(colour = "white", linewidth = 0.6) +
    scale_fill_gradient(
      low      = "#D9F0E3",
      high     = "#00441B",
      trans    = "log10",
      labels   = function(x) paste0(x, "%"),
      na.value = "#EEEEEE",
      name     = "Rel. abundance\n(% of classified bases)"
    ) +
    scale_y_discrete(expand = expansion(add = 0.5)) +
    facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +
    labs(title = title_wrapped, subtitle = subtitle, x = NULL, y = NULL) +
    theme_light(base_size = 13) +
    theme(
      axis.text.x     = element_text(angle = 30, hjust = 1, size = 11),
      axis.text.y     = element_text(size = 10, face = "italic"),
      plot.title      = element_markdown(face = "bold", size = 14, lineheight = 1.2,
                                         margin = margin(b = 8)),
      plot.subtitle   = element_text(size = 10, colour = "grey40", margin = margin(b = 4)),
      plot.margin     = margin(t = 10, r = 10, b = 8, l = 10),
      strip.text      = element_text(face = "bold", size = 12),
      legend.position = "right",
      panel.grid      = element_blank()
    )

  print(p)
  if (!is.null(save_path)) ggsave(save_path, p, width = 10, height = 7, dpi = 600)
  invisible(p)
}

```

```{r pathogen_spp_heatmap_reads, fig.height=7, fig.width=12}

pathogen_genera_read <- list(
  list(genus = "Campylobacter", slug = "campylobacter"),
  list(genus = "Salmonella",    slug = "salmonella"),
  list(genus = "Escherichia",   slug = "escherichia")
)

for (pg in pathogen_genera_read) {

  # Species detected anywhere for this genus — defines the y-axis for all enrichment plots
  global_spp <- df %>%
    filter(tax_level == "species",
           str_detect(taxon, paste0("^", pg$genus)),
           rel_abund > 0) %>%
    pull(taxon) %>% unique()

  if (length(global_spp) == 0) {
    message("No ", pg$genus, " species detected in any sample — skipping.")
    next
  }

  for (enrich in levels(df$enrichment_type)) {

    spp_df <- df %>%
      filter(tax_level == "species",
             taxon %in% global_spp,
             enrichment_type == enrich) %>%
      mutate(
        name         = taxon,
        sample_type  = factor(if_else(grepl("^EVI", as.character(group)),
                                      "Post-EVI", "Post-ASC"),
                              levels = c("Post-EVI", "Post-ASC")),
        sample_label = sample_labels[sample]
      )

    # If genus not recorded at all for this enrichment, build zero-abundance rows
    # so the plot still renders (all grey tiles)
    if (nrow(spp_df) == 0) {
      enrich_meta <- df %>%
        filter(enrichment_type == enrich) %>%
        distinct(sample, group)
      if (nrow(enrich_meta) == 0) {
        message("No samples for: ", enrich, " — skipping.")
        next
      }
      spp_df <- expand.grid(name   = global_spp,
                            sample = enrich_meta$sample,
                            stringsAsFactors = FALSE) %>%
        left_join(enrich_meta, by = "sample") %>%
        mutate(
          rel_abund    = 0,
          sample_type  = factor(if_else(grepl("^EVI", as.character(group)),
                                        "Post-EVI", "Post-ASC"),
                                levels = c("Post-EVI", "Post-ASC")),
          sample_label = sample_labels[sample]
        )
    }

    file_slug <- gsub("[^A-Za-z0-9]", "_", enrich)

    plot_read_genus_heatmap(
      spp_df,
      title     = paste0("<i>", pg$genus, "</i> Species Relative Abundance - ",
                         enrichment_labels[enrich], " (Reads)"),
      subtitle  = "Colour intensity = relative abundance (% of classified bases); grey = not detected",
      save_path = file.path(out_dir,
                            paste0(pg$slug, "_", file_slug, "_spp_heatmap_reads.png"))
    )

    message(pg$genus, " heatmap done: ", enrich)
  }
}

```


---

# Host DNA Proportion - Bacteria vs Eukaryota vs Unclassified

Proportion of total sequenced bases attributable to host/eukaryotic DNA (e.g. *Gallus gallus*), bacterial reads, other classified (Archaea, Viruses), and unclassified reads. Calculated per sample at genus level.

```{r Host proportion chart, fig.height=6, fig.width=12}

if (!"superkingdom" %in% names(df_raw)) {
  message("NOTE: 'superkingdom' column missing - host proportion chart requires re-running kraken2_bpm.py. Skipping.")
  knitr::knit_exit()
}

host_df <- df_raw %>%
  filter(tax_level == "genus") %>%
  group_by(enrichment_type, group, sample) %>%
  summarise(
    total_bases     = first(total_bases),
    bacteria_bases  = sum(bases[superkingdom == "Bacteria"],                          na.rm = TRUE),
    gallus_bases    = sum(bases[taxon == "Gallus"],                                   na.rm = TRUE),
    other_euk_bases = sum(bases[superkingdom == "Eukaryota" & taxon != "Gallus"],     na.rm = TRUE),
    other_bases     = sum(bases[!superkingdom %in% c("Bacteria", "Eukaryota", "Unknown")],
                          na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    unclassified_bases = total_bases - bacteria_bases - gallus_bases -
                         other_euk_bases - other_bases,
    `Bacteria`         = bacteria_bases    / total_bases * 100,
    `Gallus gallus`    = gallus_bases      / total_bases * 100,
    `Other Eukaryota`  = other_euk_bases   / total_bases * 100,
    `Other classified` = other_bases       / total_bases * 100,
    `Unclassified`     = unclassified_bases / total_bases * 100,
    sample_label = factor(sample_labels[sample],
                          levels = unique(sample_labels[unique(sample[order(group, sample)])])),
    sample_type  = factor(if_else(grepl("^EVI", as.character(group)), "Post-EVI", "Post-ASC"),
                          levels = c("Post-EVI", "Post-ASC"))
  ) %>%
  pivot_longer(
    cols      = c(`Bacteria`, `Gallus gallus`, `Other Eukaryota`,
                  `Other classified`, `Unclassified`),
    names_to  = "category",
    values_to = "pct"
  ) %>%
  mutate(category = factor(category,
                           levels = c("Gallus gallus", "Other Eukaryota",
                                      "Bacteria", "Other classified", "Unclassified")))

host_colours <- c(
  "Gallus gallus"    = "#FF7F00",
  "Other Eukaryota"  = "#E31A1C",
  "Bacteria"         = "#1F78B4",
  "Other classified" = "#6A3D9A",
  "Unclassified"     = "#B2DF8A"
)

# Combined 3-row × 2-column plot: rows = enrichment type, columns = Post-EVI / Post-ASC
host_plot_df <- host_df %>%
  mutate(
    sample_label    = factor(sample_labels[sample], levels = paste0("CH0", 1:6)),
    sample_type     = factor(if_else(grepl("^EVI", as.character(group)),
                                     "Post-EVI", "Post-ASC"),
                             levels = c("Post-EVI", "Post-ASC")),
    enrichment_type = factor(as.character(enrichment_type),
                             levels = levels(df_raw$enrichment_type))
  )

p_host_all <- ggplot(host_plot_df, aes(x = sample_label, y = pct, fill = category)) +
  geom_col(width = 0.7, colour = "white", linewidth = 0.3) +
  scale_fill_manual(
    values = host_colours,
    labels = function(b) ifelse(b == "Gallus gallus", "*Gallus gallus*", b)
  ) +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.02))) +
  facet_grid(enrichment_type ~ sample_type, scales = "free_x", space = "free_x") +
  labs(
    title    = "Host vs Bacterial vs Unclassified bases - All enrichment types",
    subtitle = "Proportion of total sequenced bases; *Gallus gallus* shown separately from other Eukaryota",
    x        = NULL,
    y        = "% of total bases",
    fill     = NULL
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 14),
    plot.subtitle   = element_markdown(size = 10, colour = "grey40"),
    axis.text.x     = element_text(angle = 45, hjust = 1),
    strip.text.x    = element_text(face = "bold", size = 12),
    strip.text.y    = element_text(face = "bold", size = 10),
    panel.spacing   = unit(0.6, "lines"),
    legend.text     = element_markdown(),
    legend.position = "right"
  )

print(p_host_all)
ggsave(file.path(out_dir, "host_proportion_all_enrichments.png"),
       p_host_all, width = 12, height = 10, dpi = 600)
message("Saved: host_proportion_all_enrichments.png")

```


---

# Non-bacterial Microbiome - Archaea and Viruses

Bubble plot of Archaea and Virus genera detected per sample. Bubble area is proportional to relative abundance (% of all classified bases). Faceted by Post-EVI vs Post-ASC, one plot per enrichment type.

```{r nonbact_bubble, fig.height=6, fig.width=16}

if (!"superkingdom" %in% names(df_raw)) {
  message("NOTE: 'superkingdom' column missing - non-bacterial bubble plots require re-running kraken2_bpm.py. Skipping.")
} else {

  nonbact_df <- df_raw %>%
    filter(tax_level == "genus",
           superkingdom %in% c("Archaea", "Viruses"),
           rel_abund > 0) %>%
    mutate(
      taxon_label  = paste0("*", taxon, "* (", superkingdom, ")"),
      sample_type  = factor(if_else(grepl("^EVI", as.character(group)),
                                    "Post-EVI", "Post-ASC"),
                            levels = c("Post-EVI", "Post-ASC")),
      sample_label = factor(sample_labels[sample],
                            levels = paste0("CH0", 1:6))
    )

  for (enrich in levels(df_raw$enrichment_type)) {
    sub <- nonbact_df %>% filter(enrichment_type == enrich)
    if (nrow(sub) == 0) { message("No non-bacterial data for: ", enrich, " — skipping."); next }

    enrich_lbl <- enrichment_labels[enrich]
    file_slug  <- gsub("[^A-Za-z0-9]", "_", enrich)

    taxon_order <- sub %>%
      group_by(taxon_label, superkingdom) %>%
      summarise(m = mean(rel_abund), .groups = "drop") %>%
      arrange(superkingdom, m) %>%
      pull(taxon_label)

    sub <- sub %>% mutate(taxon_label = factor(taxon_label, levels = taxon_order))

    p <- ggplot(sub, aes(x = taxon_label, y = sample_label,
                         size = rel_abund, colour = superkingdom)) +
      geom_point(alpha = 0.85) +
      scale_size(range  = c(2, 10),
                 trans  = "sqrt",
                 breaks = c(0.001, 0.01, 0.1, 1),
                 labels = c("0.001", "0.010", "0.100", "1.000"),
                 name   = "Rel. abund (%)") +
      scale_colour_manual(values = c("Archaea" = "#33A02C", "Viruses" = "#6A3D9A"),
                          name   = "Kingdom") +
      facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +
      labs(
        title    = paste0("Non-bacterial microbiome reads: Archaea and Viruses\n", enrich_lbl),
        subtitle = "Bubble area proportional to relative abundance (% of all classified bases)",
        x = NULL, y = NULL
      ) +
      theme_bw(base_size = 12) +
      theme(
        axis.text.x      = element_markdown(angle = 45, hjust = 1, size = 9),
        axis.text.y      = element_text(size = 11),
        plot.title       = element_text(face = "bold", size = 13),
        plot.subtitle    = element_text(size = 10, colour = "grey40"),
        strip.text       = element_text(face = "bold", size = 12),
        panel.grid.major = element_line(colour = "grey90"),
        panel.grid.minor = element_blank()
      )

    print(p)
    ggsave(file.path(out_dir, paste0("nonbact_bubble_", file_slug, ".png")),
           p, width = 16, height = 6, dpi = 600)
    message("Saved non-bacterial bubble: ", enrich)
  }
}

```


---

# Diversity Analysis - Alpha, Beta & PERMANOVA

```{r Diversity helper functions}

# ── Alpha diversity ────────────────────────────────────────────────────────────
# Builds Shannon, Simpson, and observed richness from base counts per sample,
# then plots boxplots + jitter with Wilcoxon p-values per metric.
# data_sub : a data.frame with columns sample, taxon, bases, group (factor)
# label_a/b: group labels to compare (Post-EVI vs Post-ASC)

plot_alpha_bpm <- function(data_sub, label_a, label_b, title_str,
                            col_a = "#377EB8", col_b = "#E41A1C",
                            save_path = NULL) {

  make_tbl <- function(df, label) {
    wide <- df %>%
      filter(as.character(group) == label) %>%
      select(sample, taxon, bases) %>%
      pivot_wider(names_from = taxon, values_from = bases, values_fill = 0L)
    mat <- wide %>% column_to_rownames("sample") %>% as.matrix()
    tibble(
      sample   = rownames(mat),
      group    = label,
      Shannon  = diversity(mat, index = "shannon"),
      Simpson  = diversity(mat, index = "simpson"),
      Richness = rowSums(mat > 0)
    )
  }

  alpha_df <- bind_rows(
    make_tbl(data_sub, label_a),
    make_tbl(data_sub, label_b)
  ) %>%
    mutate(group = factor(group, levels = c(label_a, label_b)))

  cols <- setNames(c(col_a, col_b), c(label_a, label_b))

  alpha_long <- alpha_df %>%
    pivot_longer(c(Shannon, Simpson, Richness),
                 names_to = "Metric", values_to = "Value") %>%
    mutate(Metric = factor(Metric, levels = c("Shannon", "Simpson", "Richness")))

  pvals <- alpha_long %>%
    group_by(Metric) %>%
    summarise(
      p = tryCatch(
        wilcox.test(Value[group == label_a], Value[group == label_b],
                    exact = FALSE)$p.value,
        error = function(e) NA_real_),
      .groups = "drop"
    ) %>%
    mutate(label = case_when(
      p < 0.001   ~ "***",
      p < 0.01    ~ "**",
      p < 0.05    ~ "*",
      !is.na(p)   ~ paste0("p = ", round(p, 3)),
      TRUE        ~ ""
    ))

  p <- ggplot(alpha_long, aes(x = group, y = Value, fill = group, colour = group)) +
    geom_boxplot(width = 0.45, alpha = 0.6, outlier.shape = NA, colour = "grey30") +
    geom_jitter(width = 0.1, size = 2.2, alpha = 0.85) +
    geom_text(data = pvals, aes(x = 1.5, y = Inf, label = label),
              inherit.aes = FALSE, vjust = 1.6, size = 3.8, colour = "black") +
    facet_wrap(~ Metric, scales = "free_y", ncol = 3) +
    scale_fill_manual(values = cols) +
    scale_colour_manual(values = cols) +
    labs(title    = title_str,
         subtitle = "Wilcoxon rank-sum: *** p < 0.001, ** < 0.01, * < 0.05\nAlpha diversity calculated from base counts assigned per taxon",
         x = NULL, y = NULL) +
    theme_bw(base_size = 13) +
    theme(legend.position  = "none",
          axis.text        = element_text(size = 11),
          strip.text       = element_text(face = "bold", size = 13),
          plot.title       = element_text(face = "bold", size = 14),
          plot.subtitle    = element_text(size = 10, colour = "grey40"),
          strip.background = element_rect(fill = "lightgrey"))

  print(p)
  if (!is.null(save_path)) ggsave(save_path, p, width = 10, height = 5, dpi = 600)
  invisible(p)
}


# ── Beta diversity - Bray-Curtis PCoA + PERMANOVA ─────────────────────────────
# Uses relative abundance for Bray-Curtis dissimilarities.
# Plots PCoA with 95% ellipses; PERMANOVA R² and p-value shown in subtitle.

plot_beta_pcoa_bpm <- function(data_sub, label_a, label_b, title_str,
                                col_a = "#377EB8", col_b = "#E41A1C",
                                save_path = NULL) {

  wide <- data_sub %>%
    select(sample, taxon, rel_abund) %>%
    pivot_wider(names_from = taxon, values_from = rel_abund, values_fill = 0)
  mat <- wide %>% column_to_rownames("sample") %>% as.matrix()

  # Group vector aligned with matrix row order
  sample_grp <- data_sub %>%
    distinct(sample, group) %>%
    mutate(group = as.character(group))
  group_vec <- sample_grp$group[match(rownames(mat), sample_grp$sample)]

  dist_mat <- vegdist(mat, method = "bray")
  pcoa     <- cmdscale(dist_mat, k = 2, eig = TRUE)
  var_expl <- abs(pcoa$eig) / sum(abs(pcoa$eig)) * 100

  coords <- as.data.frame(pcoa$points) %>%
    setNames(c("PC1", "PC2")) %>%
    rownames_to_column("sample") %>%
    mutate(group = factor(group_vec, levels = c(label_a, label_b)))

  perm <- adonis2(dist_mat ~ group_vec, permutations = 999)
  r2   <- round(perm$R2[1], 3)
  p_v  <- perm$`Pr(>F)`[1]

  p <- ggplot(coords, aes(x = PC1, y = PC2, colour = group)) +
    geom_point(size = 3.5, alpha = 0.85) +
    stat_ellipse(level = 0.95, linewidth = 0.8) +
    scale_colour_manual(values = setNames(c(col_a, col_b), c(label_a, label_b))) +
    guides(colour = guide_legend(title = NULL, override.aes = list(size = 4))) +
    labs(title    = title_str,
         subtitle = paste0("Bray-Curtis PCoA | PERMANOVA: R² = ", r2,
                           ", p = ", round(p_v, 3), " (999 permutations)"),
         x = paste0("PC1 (", round(var_expl[1], 1), "% variance)"),
         y = paste0("PC2 (", round(var_expl[2], 1), "% variance)")) +
    theme_bw(base_size = 13) +
    theme(plot.title      = element_text(face = "bold", size = 14),
          plot.subtitle   = element_text(size = 11, colour = "grey40"),
          axis.text       = element_text(size = 12),
          axis.title      = element_text(size = 13),
          legend.text     = element_text(size = 12),
          legend.position = "bottom")

  print(p)
  if (!is.null(save_path)) ggsave(save_path, p, width = 8, height = 6, dpi = 600)
  invisible(p)
}


# ── Per-taxon Wilcoxon tests (Post-EVI vs Post-ASC) ───────────────────────────
# Returns a table with mean rel_abund per group, raw p-value, BH-adjusted p-value,
# and significance stars. Only taxa with mean >= min_mean in >= min_samples tested.

run_taxon_tests_bpm <- function(data_sub, label_a, label_b,
                                 min_mean = 0.1, min_samples = 2) {
  combined <- bind_rows(
    data_sub %>% filter(group == label_a) %>%
      select(sample, taxon, rel_abund, bpm) %>% mutate(grp = label_a),
    data_sub %>% filter(group == label_b) %>%
      select(sample, taxon, rel_abund, bpm) %>% mutate(grp = label_b)
  ) %>%
    complete(nesting(sample, grp), taxon, fill = list(rel_abund = 0, bpm = 0))

  taxa_keep <- combined %>%
    group_by(taxon) %>%
    summarise(mean_abund  = mean(rel_abund),
              n_nonzero   = sum(rel_abund > 0),
              .groups     = "drop") %>%
    filter(mean_abund >= min_mean, n_nonzero >= min_samples) %>%
    pull(taxon)

  results <- combined %>%
    filter(taxon %in% taxa_keep) %>%
    group_by(taxon) %>%
    summarise(
      mean_pct_a = mean(rel_abund[grp == label_a]),
      mean_pct_b = mean(rel_abund[grp == label_b]),
      p_value    = tryCatch(
        wilcox.test(rel_abund[grp == label_a],
                    rel_abund[grp == label_b],
                    exact = FALSE)$p.value,
        error = function(e) NA_real_),
      .groups = "drop"
    ) %>%
    mutate(
      p_adj     = p.adjust(p_value, method = "BH"),
      sig_unadj = case_when(p_value < 0.001 ~ "***", p_value < 0.01 ~ "**",
                             p_value < 0.05  ~ "*",   TRUE           ~ "ns"),
      sig       = case_when(p_adj   < 0.001 ~ "***", p_adj   < 0.01 ~ "**",
                             p_adj   < 0.05  ~ "*",   TRUE           ~ "ns")
    ) %>%
    arrange(p_value)

  names(results)[names(results) == "mean_pct_a"] <- paste0("Mean ", label_a, " (%)")
  names(results)[names(results) == "mean_pct_b"] <- paste0("Mean ", label_b, " (%)")
  results
}


# Format results table for display / PDF output
format_stats_display_bpm <- function(results, adjusted = TRUE) {
  pct_cols <- grep("\\(%\\)$", names(results), value = TRUE)
  df <- results
  df[pct_cols]  <- lapply(df[pct_cols], round, 2)
  df$p_value    <- formatC(df$p_value, format = "e", digits = 2)

  if (adjusted) {
    df$p_adj <- formatC(df$p_adj, format = "e", digits = 2)
    df <- df %>% select(taxon, all_of(pct_cols), p_value, p_adj, sig)
    names(df)[names(df) == "sig"]   <- "Sig. (BH)"
    names(df)[names(df) == "p_adj"] <- "p-adj (BH)"
  } else {
    df <- df %>% select(taxon, all_of(pct_cols), p_value, sig_unadj)
    names(df)[names(df) == "sig_unadj"] <- "Sig."
  }
  names(df)[names(df) == "taxon"]   <- "Taxon"
  names(df)[names(df) == "p_value"] <- "p-value"
  df
}


# Draw one table page to the active PDF device
draw_stats_table_bpm <- function(df, title_str) {
  sig_col  <- if ("Sig. (BH)" %in% names(df)) "Sig. (BH)" else "Sig."
  sig_vals <- df[[sig_col]]
  n_rows   <- nrow(df)
  n_cols   <- ncol(df)

  row_fill <- case_when(sig_vals == "***" ~ "#fd7e14",
                        sig_vals == "**"  ~ "#ffc107",
                        sig_vals == "*"   ~ "#fff3cd",
                        TRUE              ~ "#ffffff")
  fill_mat <- matrix(rep(row_fill, n_cols), nrow = n_rows, ncol = n_cols)

  # Italicise the Taxon column (scientific names), keep other columns plain
  face_mat <- matrix("plain", nrow = n_rows, ncol = n_cols)
  taxon_col <- which(names(df) == "Taxon")
  if (length(taxon_col) == 1) face_mat[, taxon_col] <- "italic"

  tbl <- tableGrob(
    df, rows = NULL,
    theme = ttheme_default(
      core    = list(bg_params = list(fill = fill_mat, col = "#dee2e6"),
                     fg_params = list(fontsize = 7, fontface = face_mat)),
      colhead = list(bg_params = list(fill = "#495057", col = "#dee2e6"),
                     fg_params = list(fontsize = 7.5, col = "white",
                                      fontface = "bold"))
    )
  )

  title_grob <- grid::textGrob(title_str,
                                gp    = grid::gpar(fontsize = 10, fontface = "bold"),
                                hjust = 0.5, x = 0.5)
  combined <- arrangeGrob(title_grob, tbl, nrow = 2, heights = c(0.06, 0.94))
  grid::grid.draw(combined)
}


# Save a 2-page PDF: page 1 = unadjusted p-values, page 2 = BH-adjusted
save_stats_pdf_bpm <- function(results, title_str, path) {
  unadj_df <- format_stats_display_bpm(results, adjusted = FALSE)
  adj_df   <- format_stats_display_bpm(results, adjusted = TRUE)
  page_h   <- max(5, nrow(results) * 0.26 + 1.8)
  pdf(path, width = 16, height = page_h)
  draw_stats_table_bpm(unadj_df, paste0(title_str, "  |  Unadjusted p-values"))
  grid::grid.newpage()
  draw_stats_table_bpm(adj_df,   paste0(title_str, "  |  BH-adjusted p-values (FDR)"))
  dev.off()
  message("Saved: ", basename(path))
  invisible(path)
}

```


## Alpha Diversity - Post-EVI vs Post-ASC

Shannon entropy, Simpson index, and observed richness calculated from base counts assigned to each taxon per sample. Wilcoxon rank-sum p-value shown above each panel.

```{r Alpha diversity, fig.height=5, fig.width=10}

tax_level_labels <- c(genus = "Genus", family = "Family", species = "Species")

for (enrich in levels(df$enrichment_type)) {
  for (tax_lvl in c("genus", "family", "species")) {

    div_df <- df %>%
      filter(enrichment_type == enrich, tax_level == tax_lvl) %>%
      mutate(group = as.character(group),
             sample_type = if_else(grepl("^EVI", group), "Post-EVI", "Post-ASC"))

    # Need at least 2 samples per group
    n_evi <- n_distinct(div_df$sample[div_df$sample_type == "Post-EVI"])
    n_asc <- n_distinct(div_df$sample[div_df$sample_type == "Post-ASC"])
    if (n_evi < 2 || n_asc < 2) next

    enrich_lbl <- enrichment_labels[enrich]
    lvl_lbl    <- tax_level_labels[tax_lvl]
    file_slug  <- paste0(gsub("[^A-Za-z0-9]", "_", enrich), "_", tax_lvl)

    plot_alpha_bpm(
      data_sub   = div_df %>% select(-group) %>% rename(group = sample_type),
      label_a    = "Post-EVI",
      label_b    = "Post-ASC",
      title_str  = paste0("Alpha Diversity (", lvl_lbl, " level) - ", enrich_lbl),
      save_path  = file.path(out_dir, paste0(file_slug, "_alpha_diversity.png"))
    )
  }
}

```


## Beta Diversity - Bray-Curtis PCoA + PERMANOVA

Principal coordinates analysis of Bray-Curtis dissimilarities (relative abundance). Ellipses show 95% confidence intervals. PERMANOVA tests whether group membership explains community composition.

```{r Beta diversity PERMANOVA, fig.height=6, fig.width=8}

for (enrich in levels(df$enrichment_type)) {
  for (tax_lvl in c("genus", "family", "species")) {

    div_df <- df %>%
      filter(enrichment_type == enrich, tax_level == tax_lvl) %>%
      mutate(sample_type = if_else(grepl("^EVI", as.character(group)),
                                   "Post-EVI", "Post-ASC"))

    # Need at least 2 samples per group for PCoA
    n_evi <- n_distinct(div_df$sample[div_df$sample_type == "Post-EVI"])
    n_asc <- n_distinct(div_df$sample[div_df$sample_type == "Post-ASC"])
    if (n_evi < 2 || n_asc < 2) next

    enrich_lbl <- enrichment_labels[enrich]
    lvl_lbl    <- tax_level_labels[tax_lvl]
    file_slug  <- paste0(gsub("[^A-Za-z0-9]", "_", enrich), "_", tax_lvl)

    plot_beta_pcoa_bpm(
      data_sub  = div_df %>% select(-group) %>% rename(group = sample_type),
      label_a   = "Post-EVI",
      label_b   = "Post-ASC",
      title_str = paste0("Beta Diversity - Bray-Curtis PCoA (", lvl_lbl, " level)\n", enrich_lbl),
      save_path = file.path(out_dir, paste0(file_slug, "_beta_pcoa.png"))
    )
  }
}

```


## Beta Diversity - Bray-Curtis PCoA + Flock-Stratified PERMANOVA

Flock identity (CH01–CH06) is used as a blocking factor via `permute::how(blocks = flock)`. This restricts permutations within flocks to account for the shared flock origin of Post-EVI and Post-ASC samples, giving a more conservative and appropriate test than the unstratified version above.

```{r Beta diversity stratified PERMANOVA, fig.height=6, fig.width=8}

for (enrich in levels(df$enrichment_type)) {
  for (tax_lvl in c("genus", "family", "species")) {

    div_df <- df %>%
      filter(enrichment_type == enrich, tax_level == tax_lvl) %>%
      mutate(sample_type = if_else(grepl("^EVI", as.character(group)),
                                   "Post-EVI", "Post-ASC"))

    n_evi <- n_distinct(div_df$sample[div_df$sample_type == "Post-EVI"])
    n_asc <- n_distinct(div_df$sample[div_df$sample_type == "Post-ASC"])
    if (n_evi < 2 || n_asc < 2) next

    enrich_lbl <- enrichment_labels[enrich]
    lvl_lbl    <- tax_level_labels[tax_lvl]
    file_slug  <- paste0(gsub("[^A-Za-z0-9]", "_", enrich), "_", tax_lvl)

    wide <- div_df %>%
      select(sample, taxon, rel_abund) %>%
      pivot_wider(names_from = taxon, values_from = rel_abund, values_fill = 0)
    mat <- wide %>% column_to_rownames("sample") %>% as.matrix()

    sample_grp <- div_df %>% distinct(sample, sample_type)
    group_vec  <- sample_grp$sample_type[match(rownames(mat), sample_grp$sample)]

    # Extract flock ID (CH01, CH02 ...) from sample names for blocking
    flock_vec  <- sub("_(EVI|ASC).*", "", rownames(mat))

    dist_mat <- vegdist(mat, method = "bray")

    # Flock-stratified PERMANOVA
    perm_ctrl <- how(blocks = flock_vec, nperm = 999)
    perm_strat <- adonis2(dist_mat ~ group_vec, permutations = perm_ctrl)
    r2  <- round(perm_strat$R2[1], 3)
    p_v <- perm_strat$`Pr(>F)`[1]

    # PCoA coordinates (same as unstratified - visualisation unchanged)
    pcoa     <- cmdscale(dist_mat, k = 2, eig = TRUE)
    var_expl <- abs(pcoa$eig) / sum(abs(pcoa$eig)) * 100

    coords <- as.data.frame(pcoa$points) %>%
      setNames(c("PC1", "PC2")) %>%
      rownames_to_column("sample") %>%
      mutate(group = factor(group_vec, levels = c("Post-EVI", "Post-ASC")),
             flock = flock_vec)

    p_strat <- ggplot(coords, aes(x = PC1, y = PC2, colour = group)) +
      geom_point(size = 3.5, alpha = 0.85) +
      geom_text(aes(label = flock), nudge_y = 0.015, size = 3, colour = "grey30") +
      stat_ellipse(level = 0.95, linewidth = 0.8) +
      scale_colour_manual(values = c("Post-EVI" = "#377EB8", "Post-ASC" = "#E41A1C")) +
      guides(colour = guide_legend(title = NULL, override.aes = list(size = 4))) +
      labs(
        title    = paste0("Beta Diversity - Bray-Curtis PCoA (", lvl_lbl, " level)\n", enrich_lbl),
        subtitle = paste0("Flock-stratified PERMANOVA: R² = ", r2,
                          ", p = ", round(p_v, 3), " (999 permutations, blocked by flock)"),
        x = paste0("PC1 (", round(var_expl[1], 1), "% variance)"),
        y = paste0("PC2 (", round(var_expl[2], 1), "% variance)")
      ) +
      theme_bw(base_size = 13) +
      theme(
        plot.title      = element_text(face = "bold", size = 14),
        plot.subtitle   = element_text(size = 11, colour = "grey40"),
        axis.text       = element_text(size = 12),
        axis.title      = element_text(size = 13),
        legend.text     = element_text(size = 12),
        legend.position = "bottom"
      )

    print(p_strat)

    fname_strat <- file.path(out_dir, paste0(file_slug, "_beta_pcoa_stratified.png"))
    ggsave(fname_strat, p_strat, width = 8, height = 6, dpi = 600)
    message("Saved: ", basename(fname_strat))
  }
}

```


## Stratified PERMANOVA Summary Table

```{r Stratified PERMANOVA summary table}

perm_results_strat <- list()

for (enrich in levels(df$enrichment_type)) {
  for (tax_lvl in c("genus", "family", "species")) {

    div_df <- df %>%
      filter(enrichment_type == enrich, tax_level == tax_lvl) %>%
      mutate(sample_type = if_else(grepl("^EVI", as.character(group)),
                                   "Post-EVI", "Post-ASC"))

    n_evi <- n_distinct(div_df$sample[div_df$sample_type == "Post-EVI"])
    n_asc <- n_distinct(div_df$sample[div_df$sample_type == "Post-ASC"])
    if (n_evi < 2 || n_asc < 2) next

    wide <- div_df %>%
      select(sample, taxon, rel_abund) %>%
      pivot_wider(names_from = taxon, values_from = rel_abund, values_fill = 0)
    mat <- wide %>% column_to_rownames("sample") %>% as.matrix()

    sample_grp <- div_df %>% distinct(sample, sample_type)
    group_vec  <- sample_grp$sample_type[match(rownames(mat), sample_grp$sample)]
    flock_vec  <- sub("_(EVI|ASC).*", "", rownames(mat))

    dist_mat   <- vegdist(mat, method = "bray")
    perm_ctrl  <- how(blocks = flock_vec, nperm = 999)
    perm_strat <- adonis2(dist_mat ~ group_vec, permutations = perm_ctrl)

    perm_results_strat[[paste(enrich, tax_lvl, sep = " | ")]] <- tibble(
      Enrichment   = enrichment_labels[enrich],
      `Tax. level` = tax_level_labels[tax_lvl],
      `n Post-EVI` = n_evi,
      `n Post-ASC` = n_asc,
      `R²`         = round(perm_strat$R2[1], 3),
      `p-value`    = round(perm_strat$`Pr(>F)`[1], 3),
      Significant  = ifelse(perm_strat$`Pr(>F)`[1] < 0.05, "Yes", "No")
    )
  }
}

perm_table_strat <- bind_rows(perm_results_strat)
print(perm_table_strat)

write.table(perm_table_strat,
            file.path(out_dir, "permanova_stratified_summary.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)
message("Stratified PERMANOVA summary saved to: permanova_stratified_summary.tsv")

```


## PERMANOVA Summary Table

```{r PERMANOVA summary table}

perm_results <- list()

for (enrich in levels(df$enrichment_type)) {
  for (tax_lvl in c("genus", "family", "species")) {

    div_df <- df %>%
      filter(enrichment_type == enrich, tax_level == tax_lvl) %>%
      mutate(sample_type = if_else(grepl("^EVI", as.character(group)),
                                   "Post-EVI", "Post-ASC"))

    n_evi <- n_distinct(div_df$sample[div_df$sample_type == "Post-EVI"])
    n_asc <- n_distinct(div_df$sample[div_df$sample_type == "Post-ASC"])
    if (n_evi < 2 || n_asc < 2) next

    wide <- div_df %>%
      select(sample, taxon, rel_abund) %>%
      pivot_wider(names_from = taxon, values_from = rel_abund, values_fill = 0)
    mat <- wide %>% column_to_rownames("sample") %>% as.matrix()

    sample_grp <- div_df %>% distinct(sample, sample_type)
    group_vec  <- sample_grp$sample_type[match(rownames(mat), sample_grp$sample)]

    dist_mat <- vegdist(mat, method = "bray")
    perm     <- adonis2(dist_mat ~ group_vec, permutations = 999)

    perm_results[[paste(enrich, tax_lvl, sep = " | ")]] <- tibble(
      Enrichment     = enrichment_labels[enrich],
      `Tax. level`   = tax_level_labels[tax_lvl],
      `n Post-EVI`   = n_evi,
      `n Post-ASC`   = n_asc,
      `R²`           = round(perm$R2[1], 3),
      `p-value`      = round(perm$`Pr(>F)`[1], 3),
      Significant    = ifelse(perm$`Pr(>F)`[1] < 0.05, "Yes", "No")
    )
  }
}

perm_table <- bind_rows(perm_results)
print(perm_table)

# Save as TSV
write.table(perm_table,
            file.path(out_dir, "permanova_summary.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)
message("PERMANOVA summary saved to: permanova_summary.tsv")

```


---

## Per-Taxon Statistical Tests - Post-EVI vs Post-ASC

Wilcoxon rank-sum test (two-sided) per taxon at genus, family, and species level, run separately within each enrichment type. Only taxa present with mean relative abundance ≥ 0.1% in ≥ 2 samples are tested. Page 1 of each PDF shows unadjusted p-values; page 2 shows Benjamini-Hochberg FDR-corrected p-values. Rows highlighted orange (***), yellow (**), or pale yellow (*).

```{r Statistical tests per taxon}

tax_level_labels <- c(genus = "Genus", family = "Family", species = "Species")

for (tax_lvl in c("genus", "family", "species")) {
  for (enrich in levels(df$enrichment_type)) {

    test_df <- df %>%
      filter(enrichment_type == enrich, tax_level == tax_lvl) %>%
      mutate(group = if_else(grepl("^EVI", as.character(group)),
                             "Post-EVI", "Post-ASC"))

    n_evi <- n_distinct(test_df$sample[test_df$group == "Post-EVI"])
    n_asc <- n_distinct(test_df$sample[test_df$group == "Post-ASC"])
    if (n_evi < 2 || n_asc < 2) next

    results <- run_taxon_tests_bpm(test_df, "Post-EVI", "Post-ASC")

    if (nrow(results) == 0) {
      message("No taxa passed filters for: ", enrich, " | ", tax_lvl)
      next
    }

    enrich_lbl <- enrichment_labels[enrich]
    lvl_lbl    <- tax_level_labels[tax_lvl]
    file_slug  <- paste0(gsub("[^A-Za-z0-9]", "_", enrich), "_", tax_lvl)
    title_str  <- paste0("Post-EVI vs Post-ASC - ", lvl_lbl, " level | ", enrich_lbl)

    pdf_path <- file.path(out_dir, paste0("stats_", file_slug, "_EVI_vs_ASC.pdf"))
    save_stats_pdf_bpm(results, title_str, pdf_path)
  }
}

message("\nAll statistical PDFs saved to: ", out_dir)

```


---

# Sensitivity analysis - exclude CH04 Post-ASC (BF Bolton Base + CAT)

The original Post-ASC group for the BF Bolton Base + CAT enrichment shows large dispersion in both the PCoA and the alpha-diversity boxplots, with CH04 appearing as a likely outlier. This section drops `CH04_ASC_BFBB` and re-runs alpha diversity, Bray-Curtis PCoA + PERMANOVA, and the per-taxon Wilcoxon tests for that enrichment only, so we can check whether the EVI vs ASC differences are driven by that one sample. Original outputs are kept; new files are written with a `_no_CH04_ASC` suffix.

```{r Sensitivity - drop CH04 ASC BFBB, fig.height=5, fig.width=10}

drop_samples <- c("CH04_ASC_BFBB")
sens_enrich  <- "BF Bolton Base + CAT"
sens_slug    <- gsub("[^A-Za-z0-9]", "_", sens_enrich)

df_sens <- df %>%
  filter(enrichment_type == sens_enrich, !sample %in% drop_samples)

message("Sensitivity analysis - dropped samples: ",
        paste(drop_samples, collapse = ", "))
message("Remaining sample counts after exclusion:")
print(df_sens %>%
        filter(tax_level == "genus") %>%
        mutate(sample_type = if_else(grepl("^EVI", as.character(group)),
                                     "Post-EVI", "Post-ASC")) %>%
        distinct(sample, sample_type) %>%
        count(sample_type))

```


## Alpha diversity (CH04 Post-ASC excluded)

```{r Sensitivity alpha diversity, fig.height=5, fig.width=10}

for (tax_lvl in c("genus", "family", "species")) {

  div_df <- df_sens %>%
    filter(tax_level == tax_lvl) %>%
    mutate(group = as.character(group),
           sample_type = if_else(grepl("^EVI", group), "Post-EVI", "Post-ASC"))

  n_evi <- n_distinct(div_df$sample[div_df$sample_type == "Post-EVI"])
  n_asc <- n_distinct(div_df$sample[div_df$sample_type == "Post-ASC"])
  if (n_evi < 2 || n_asc < 2) next

  lvl_lbl   <- tax_level_labels[tax_lvl]
  file_slug <- paste0(sens_slug, "_", tax_lvl, "_no_CH04_ASC")

  plot_alpha_bpm(
    data_sub  = div_df %>% select(-group) %>% rename(group = sample_type),
    label_a   = "Post-EVI",
    label_b   = "Post-ASC",
    title_str = paste0("Alpha Diversity (", lvl_lbl, " level) - ", sens_enrich,
                       "\n(CH04 Post-ASC excluded)"),
    save_path = file.path(out_dir, paste0(file_slug, "_alpha_diversity.png"))
  )
}

```


## Beta diversity - Bray-Curtis PCoA (CH04 Post-ASC excluded)

```{r Sensitivity beta diversity, fig.height=6, fig.width=8}

for (tax_lvl in c("genus", "family", "species")) {

  div_df <- df_sens %>%
    filter(tax_level == tax_lvl) %>%
    mutate(sample_type = if_else(grepl("^EVI", as.character(group)),
                                 "Post-EVI", "Post-ASC"))

  n_evi <- n_distinct(div_df$sample[div_df$sample_type == "Post-EVI"])
  n_asc <- n_distinct(div_df$sample[div_df$sample_type == "Post-ASC"])
  if (n_evi < 2 || n_asc < 2) next

  lvl_lbl   <- tax_level_labels[tax_lvl]
  file_slug <- paste0(sens_slug, "_", tax_lvl, "_no_CH04_ASC")

  plot_beta_pcoa_bpm(
    data_sub  = div_df %>% select(-group) %>% rename(group = sample_type),
    label_a   = "Post-EVI",
    label_b   = "Post-ASC",
    title_str = paste0("Beta Diversity - Bray-Curtis PCoA (", lvl_lbl, " level)\n",
                       sens_enrich, " (CH04 Post-ASC excluded)"),
    save_path = file.path(out_dir, paste0(file_slug, "_beta_pcoa.png"))
  )
}

```


## PERMANOVA summary table (CH04 Post-ASC excluded)

```{r Sensitivity PERMANOVA summary}

perm_results_sens <- list()

for (tax_lvl in c("genus", "family", "species")) {

  div_df <- df_sens %>%
    filter(tax_level == tax_lvl) %>%
    mutate(sample_type = if_else(grepl("^EVI", as.character(group)),
                                 "Post-EVI", "Post-ASC"))

  n_evi <- n_distinct(div_df$sample[div_df$sample_type == "Post-EVI"])
  n_asc <- n_distinct(div_df$sample[div_df$sample_type == "Post-ASC"])
  if (n_evi < 2 || n_asc < 2) next

  wide <- div_df %>%
    select(sample, taxon, rel_abund) %>%
    pivot_wider(names_from = taxon, values_from = rel_abund, values_fill = 0)
  mat <- wide %>% column_to_rownames("sample") %>% as.matrix()

  sample_grp <- div_df %>% distinct(sample, sample_type)
  group_vec  <- sample_grp$sample_type[match(rownames(mat), sample_grp$sample)]

  dist_mat <- vegdist(mat, method = "bray")
  perm     <- adonis2(dist_mat ~ group_vec, permutations = 999)

  # PERMDISP - test homogeneity of group dispersions (helps interpret PERMANOVA)
  bd       <- betadisper(dist_mat, group_vec)
  bd_p     <- tryCatch(permutest(bd, permutations = 999)$tab[["Pr(>F)"]][1],
                       error = function(e) NA_real_)

  perm_results_sens[[paste(sens_enrich, tax_lvl, sep = " | ")]] <- tibble(
    Enrichment        = sens_enrich,
    `Tax. level`      = tax_level_labels[tax_lvl],
    `n Post-EVI`      = n_evi,
    `n Post-ASC`      = n_asc,
    `R²`              = round(perm$R2[1], 3),
    `p-value`         = round(perm$`Pr(>F)`[1], 3),
    Significant       = ifelse(perm$`Pr(>F)`[1] < 0.05, "Yes", "No"),
    `PERMDISP p`      = round(bd_p, 3),
    `Excluded sample` = paste(drop_samples, collapse = ", ")
  )
}

perm_table_sens <- bind_rows(perm_results_sens)
print(perm_table_sens)

write.table(perm_table_sens,
            file.path(out_dir, "permanova_summary_no_CH04_ASC.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)
message("Sensitivity PERMANOVA summary saved to: permanova_summary_no_CH04_ASC.tsv")


# Side-by-side comparison with the full-data PERMANOVA results, if available
if (exists("perm_table")) {
  comparison <- perm_table %>%
    filter(Enrichment == sens_enrich) %>%
    select(`Tax. level`, `R² (full)` = `R²`, `p (full)` = `p-value`) %>%
    left_join(
      perm_table_sens %>%
        select(`Tax. level`,
               `R² (no CH04)` = `R²`,
               `p (no CH04)`  = `p-value`,
               `PERMDISP p (no CH04)` = `PERMDISP p`),
      by = "Tax. level"
    )
  message("\nFull vs sensitivity comparison (BF Bolton Base + CAT):")
  print(comparison)

  write.table(comparison,
              file.path(out_dir, "permanova_comparison_full_vs_no_CH04_ASC.tsv"),
              sep = "\t", row.names = FALSE, quote = FALSE)
}

```


## Per-taxon statistical tests (CH04 Post-ASC excluded)

```{r Sensitivity per-taxon stats}

for (tax_lvl in c("genus", "family", "species")) {

  test_df <- df_sens %>%
    filter(tax_level == tax_lvl) %>%
    mutate(group = if_else(grepl("^EVI", as.character(group)),
                           "Post-EVI", "Post-ASC"))

  n_evi <- n_distinct(test_df$sample[test_df$group == "Post-EVI"])
  n_asc <- n_distinct(test_df$sample[test_df$group == "Post-ASC"])
  if (n_evi < 2 || n_asc < 2) next

  results <- run_taxon_tests_bpm(test_df, "Post-EVI", "Post-ASC")

  if (nrow(results) == 0) {
    message("No taxa passed filters for: ", sens_enrich, " | ", tax_lvl,
            " (CH04 Post-ASC excluded)")
    next
  }

  lvl_lbl   <- tax_level_labels[tax_lvl]
  file_slug <- paste0(sens_slug, "_", tax_lvl, "_no_CH04_ASC")
  title_str <- paste0("Post-EVI vs Post-ASC - ", lvl_lbl, " level | ",
                      sens_enrich, " (CH04 Post-ASC excluded)")

  pdf_path <- file.path(out_dir, paste0("stats_", file_slug, "_EVI_vs_ASC.pdf"))
  save_stats_pdf_bpm(results, title_str, pdf_path)
}

message("\nSensitivity-analysis outputs saved to: ", out_dir)

```

