---
title: "qPCR Enrichment Analysis: Campylobacter and Salmonella"
author: "Analysis Report"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)
```

# Setup and Data Loading

```{r libraries}
# Load required libraries
library(tidyverse)
library(readxl)
library(broom)
library(knitr)
library(kableExtra)
library(scales)

# Set theme for all plots
theme_set(theme_minimal(base_size = 12))
```

```{r helper_functions}
# Helper function for Wilcoxon test
safe_wilcox_test <- function(df, value_col, time_col = "Time_hr", baseline_time, test_time = 24) {
  baseline_data <- df[[value_col]][df[[time_col]] == baseline_time]
  test_data <- df[[value_col]][df[[time_col]] == test_time]
  
  if (length(baseline_data) == 0 || length(test_data) == 0) return(NA_real_)
  
  tryCatch(
    wilcox.test(test_data, baseline_data, exact = FALSE)$p.value,
    error = function(e) NA_real_
  )
}

# Helper function for p-value interpretation
interpret_p_value <- function(p) {
  case_when(
    is.na(p) ~ "Insufficient data",
    p < 0.001 ~ "Highly significant (p < 0.001)",
    p < 0.01  ~ "Very significant (p < 0.01)",
    p < 0.05  ~ "Significant (p < 0.05)",
    TRUE ~ "Not significant (p >= 0.05)"
  )
}
```

---

# PART 1: CAMPYLOBACTER ANALYSIS

## 1.1 Data Loading and Cleaning

```{r load_campylobacter}
# Read Campylobacter data
campy_raw <- read_excel('/Volumes/Alice/1_METHOD_DEVELOPMENT/Enrichment_testing/Raw_data/qPCR_enrichment-samples_Campylobacter.xlsx')

# Clean and prepare data
campy_cleaned <- campy_raw %>%
  filter(Type == "Unknown") %>%
  mutate(
    Broth = str_extract(Name, "^[A-Z-]+"),
    Time_hr = as.numeric(str_extract(Name, "\\d+(?=hr)")),
    Ct = as.numeric(Ct),
    Concentration = as.numeric(`Calc Conc (ng/ul)`)
  ) %>%
  filter(Broth %in% c("BF-BB", "CAT", "BPW", "RVS")) %>%
  select(Name, Broth, Time_hr, Ct, Concentration, `Calc Conc (ng/ul)`)

# Rename broths to full names
campy_cleaned <- campy_cleaned %>%
  mutate(Broth = case_when(
    Broth == "BF-BB" ~ "BF Bolton Base",
    Broth == "CAT" ~ "BF Bolton Base + CAT",
    TRUE ~ Broth
  ))

# Define baseline times (All broths = 0hr)
campy_baseline_map <- tibble(
  Broth = c("BF Bolton Base", "BF Bolton Base + CAT", "BPW", "RVS"),
  baseline_hr = c(0, 0, 0, 0)
)
```

```{r campy_data_summary}
# Data summary
campy_summary <- campy_cleaned %>%
  group_by(Broth, Time_hr) %>%
  summarise(
    n_replicates = n(),
    n_Ct_detected = sum(!is.na(Ct)),
    n_Conc_detected = sum(!is.na(Concentration)),
    mean_Ct = mean(Ct, na.rm = TRUE),
    mean_Conc = mean(Concentration, na.rm = TRUE),
    .groups = "drop"
  )

kable(campy_summary, 
      caption = "Campylobacter Data Summary",
      digits = 4) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

---

## 1.2 Campylobacter Ct Value Analysis

### 1.2.1 Within-Broth Statistics (Ct)

```{r campy_ct_within_broth_stats}
# Prepare data for baseline vs 24hr comparison
campy_ct_within_data <- campy_cleaned %>%
  left_join(campy_baseline_map, by = "Broth") %>%
  filter(Time_hr %in% c(baseline_hr, 24)) %>%
  filter(!is.na(Ct))

# Calculate within-broth statistics
campy_ct_within_stats <- campy_ct_within_data %>%
  group_by(Broth, baseline_hr) %>%
  summarise(
    n_baseline = sum(Time_hr == baseline_hr),
    n_24hr = sum(Time_hr == 24),
    mean_Ct_baseline = mean(Ct[Time_hr == baseline_hr]),
    sd_Ct_baseline = sd(Ct[Time_hr == baseline_hr]),
    mean_Ct_24hr = mean(Ct[Time_hr == 24]),
    sd_Ct_24hr = sd(Ct[Time_hr == 24]),
    delta_Ct = mean_Ct_24hr - mean_Ct_baseline,
    fold_change = 2^(-delta_Ct),
    p_value = safe_wilcox_test(cur_data(), "Ct", "Time_hr", unique(baseline_hr), 24),
    .groups = "drop"
  ) %>%
  mutate(
    comparison = paste0(baseline_hr, "hr vs 24hr"),
    significant = if_else(!is.na(p_value) & p_value < 0.05, "Yes", "No"),
    interpretation = interpret_p_value(p_value)
  )

# Display results
campy_table1 <- kable(campy_ct_within_stats %>% 
        select(Broth, comparison, n_baseline, n_24hr, 
               mean_Ct_baseline, mean_Ct_24hr, delta_Ct, 
               fold_change, p_value, significant),
      caption = "Campylobacter Within-Broth Comparison (Ct Values)",
      digits = c(0, 0, 0, 0, 2, 2, 2, 1, 4, 0)) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(which(campy_ct_within_stats$significant == "Yes"), 
           bold = TRUE, color = "white", background = "#28a745")

campy_table1
```

**Detailed Statistical Results (Ct Analysis):**

```{r campy_ct_individual_broth_results, results='asis'}
# Print detailed results for each broth
for (i in 1:nrow(campy_ct_within_stats)) {
  broth_data <- campy_ct_within_stats[i, ]
  
  cat("\n#### **", broth_data$Broth, "Broth (", broth_data$comparison, ")**\n\n", sep = "")
  cat("- **Sample size:** n =", broth_data$n_baseline, "(baseline), n =", broth_data$n_24hr, "(24hr)\n")
  cat("- **Mean Ct at baseline:**", round(broth_data$mean_Ct_baseline, 2), 
      "± ", round(broth_data$sd_Ct_baseline, 2), "SD\n")
  cat("- **Mean Ct at 24hr:**", round(broth_data$mean_Ct_24hr, 2), 
      "± ", round(broth_data$sd_Ct_24hr, 2), "SD\n")
  cat("- **ΔCt:**", round(broth_data$delta_Ct, 2), "cycles\n")
  cat("- **Fold change (2^-ΔCt):**", format(round(broth_data$fold_change, 1), big.mark = ","), 
      "-fold increase\n")
  cat("- **Wilcoxon rank-sum test:** p =", format.pval(broth_data$p_value, digits = 4), "\n")
  cat("- **Significance:**", broth_data$significant, "-", broth_data$interpretation, "\n")
  
  # Add interpretation note
  if (broth_data$delta_Ct < 0) {
    cat("- **Interpretation:** Ct decreased by", abs(round(broth_data$delta_Ct, 2)), 
        "cycles, indicating increased target DNA\n")
  } else if (broth_data$delta_Ct > 0) {
    cat("- **Interpretation:** Ct increased by", round(broth_data$delta_Ct, 2), 
        "cycles, indicating decreased target DNA\n")
  }
  cat("\n")
}
```

**Key Findings (Ct Analysis):**

- **BF Bolton Base**: ΔCt = `r round(campy_ct_within_stats$delta_Ct[campy_ct_within_stats$Broth == "BF Bolton Base"], 2)` cycles → `r format(round(campy_ct_within_stats$fold_change[campy_ct_within_stats$Broth == "BF Bolton Base"], 0), big.mark = ",")`-fold increase (p = `r round(campy_ct_within_stats$p_value[campy_ct_within_stats$Broth == "BF Bolton Base"], 3)`)
- **BF Bolton Base + CAT**: ΔCt = `r round(campy_ct_within_stats$delta_Ct[campy_ct_within_stats$Broth == "BF Bolton Base + CAT"], 2)` cycles → `r format(round(campy_ct_within_stats$fold_change[campy_ct_within_stats$Broth == "BF Bolton Base + CAT"], 0), big.mark = ",")`-fold increase (p = `r round(campy_ct_within_stats$p_value[campy_ct_within_stats$Broth == "BF Bolton Base + CAT"], 3)`)
- **BPW**: `r ifelse("BPW" %in% campy_ct_within_stats$Broth, paste0("ΔCt = ", round(campy_ct_within_stats$delta_Ct[campy_ct_within_stats$Broth == "BPW"], 2), " cycles"), "Insufficient data for analysis")`
- **RVS**: `r ifelse("RVS" %in% campy_ct_within_stats$Broth, paste0("ΔCt = ", round(campy_ct_within_stats$delta_Ct[campy_ct_within_stats$Broth == "RVS"], 2), " cycles"), "No detectable signal")`
- *Note: Lower Ct = Higher target amount*

**Statistical Note:** Small sample sizes (n=3) limit statistical power to detect significance despite large fold changes.

### 1.2.2 Between-Broth Statistics at 24hr (Ct)

```{r campy_ct_between_broth_stats}
# Prepare data for between-broth comparison at 24hr
campy_ct_24hr <- campy_cleaned %>%
  filter(Time_hr == 24, !is.na(Ct))

# Summary by broth
campy_ct_24hr_summary <- campy_ct_24hr %>%
  group_by(Broth) %>%
  summarise(
    n = n(),
    mean_Ct = mean(Ct),
    sd_Ct = sd(Ct),
    median_Ct = median(Ct),
    .groups = "drop"
  ) %>%
  arrange(mean_Ct)

kable(campy_ct_24hr_summary,
      caption = "Campylobacter Ct Values at 24hr (Lower Ct = More target)",
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Kruskal-Wallis test
if (n_distinct(campy_ct_24hr$Broth) >= 2) {
  campy_ct_kw <- kruskal.test(Ct ~ Broth, data = campy_ct_24hr)
  cat("\n**Kruskal-Wallis Test:**\n")
  cat("χ² =", round(campy_ct_kw$statistic, 3), 
      ", p =", format.pval(campy_ct_kw$p.value, digits = 3), "\n")
  
  if (campy_ct_kw$p.value < 0.05) {
    cat("**Result: Significant differences exist between broths (p < 0.05)**\n")
  }
}
```

### 1.2.3 Campylobacter Ct Plots

```{r campy_ct_baseline_vs_24hr_plot, fig.height=8}
# Plot: Baseline vs 24hr (Ct values)
campy_ct_plot_baseline_24 <- ggplot(campy_ct_within_data, 
                                     aes(x = factor(Time_hr), y = Ct, 
                                         fill = factor(Time_hr))) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.6, size = 2.5) +
  facet_wrap(~Broth, scales = "free_y", ncol = 2) +
  scale_y_reverse() +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = expression(paste(italic("Campylobacter"), " Ct Values: Baseline vs 24hr Comparison")),
    subtitle = "Lower Ct = Higher target amount (All broths baseline=0hr)",
    x = "Time Point (hours)",
    y = "Ct Value",
    fill = "Time Point"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, color = "gray30"),
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 11),
    strip.background = element_rect(fill = "gray90", color = NA)
  )

print(campy_ct_plot_baseline_24)
ggsave("campy_Ct_baseline_vs_24hr.png", 
       width = 10, height = 8, dpi = 600)
```

```{r campy_ct_broth_comparison_plot}
# Plot: Between-broth comparison at 24hr (Ct values)
campy_ct_plot_broths_24 <- ggplot(campy_ct_24hr, 
                                   aes(x = reorder(Broth, Ct), y = Ct, fill = Broth)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.6, size = 3) +
  scale_y_reverse() +
  scale_fill_manual(values = c("BF Bolton Base" = "#F8766D", 
                                "BF Bolton Base + CAT" = "#7CAE00",
                                "BPW" = "#00BFC4",
                                "RVS" = "#C77CFF")) +
  labs(
    title = expression(paste(italic("Campylobacter"), " Ct Values at 24hr: Between-Broth Comparison")),
    subtitle = "Lower Ct = Higher target amount",
    x = "Broth Type",
    y = "Ct Value"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11, face = "bold")
  )

print(campy_ct_plot_broths_24)
ggsave("campy_Ct_broths_24hr.png", 
       width = 10, height = 6, dpi = 600)
```

```{r campy_ct_time_course_plot}
# Plot: Time course (Ct values)
campy_ct_time_course_data <- campy_cleaned %>%
  filter(!is.na(Ct))

campy_ct_plot_time_course <- ggplot(campy_ct_time_course_data, 
                                     aes(x = Time_hr, y = Ct, 
                                         color = Broth, group = Broth)) +
  stat_summary(fun = mean, geom = "line", linewidth = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 1, linewidth = 0.8) +
  geom_jitter(alpha = 0.3, width = 0.3, height = 0, size = 2) +
  scale_y_reverse() +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 24),
                     labels = c("0", "4", "8", "12", "24")) +
  scale_color_manual(values = c("BF Bolton Base" = "#F8766D", 
                                "BF Bolton Base + CAT" = "#7CAE00",
                                "BPW" = "#00BFC4",
                                "RVS" = "#C77CFF")) +
  labs(
    title = expression(paste(italic("Campylobacter"), " Ct Values: Enrichment Testing")),
    subtitle = "Lower Ct = Higher target; Points = individual replicates, Lines = mean ± SE",
    x = "Time (hours)",
    y = "Ct Value",
    color = "Broth"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "right"
  )

print(campy_ct_plot_time_course)
ggsave("campy_Ct_time_course.png", 
       width = 10, height = 6, dpi = 600)
```

### 1.2.4 Campylobacter Dual-Axis and Panel Plots

```{r campy_dual_axis_plot}
# ============================================================================
# DUAL-AXIS PLOT: Ct + DNA Concentration
# ============================================================================
campy_conc_data <- campy_cleaned %>%
  filter(!is.na(Ct) & !is.na(`Calc Conc (ng/ul)`))

# Calculate the range for scaling (using log scale for concentration)
ct_min <- min(campy_conc_data$Ct, na.rm = TRUE)
ct_max <- max(campy_conc_data$Ct, na.rm = TRUE)
conc_min <- min(campy_conc_data$`Calc Conc (ng/ul)`, na.rm = TRUE)
conc_max <- max(campy_conc_data$`Calc Conc (ng/ul)`, na.rm = TRUE)

# Transformation function for concentration to Ct scale (using log scale)
conc_to_ct <- function(conc) {
  ct_max - (ct_max - ct_min) * (log10(conc) - log10(conc_min)) / (log10(conc_max) - log10(conc_min))
}

# Inverse transformation for secondary axis (from Ct to log concentration)
ct_to_conc <- function(ct) {
  10^(log10(conc_min) + (ct_max - ct) * (log10(conc_max) - log10(conc_min)) / (ct_max - ct_min))
}

campy_dual_axis_plot <- ggplot(campy_conc_data, 
                               aes(x = Time_hr, color = Broth, group = Broth)) +
  # Ct values (primary axis) - solid lines
  stat_summary(aes(y = Ct), fun = mean, geom = "line", 
               linewidth = 1.2, linetype = "solid") +
  stat_summary(aes(y = Ct), fun = mean, geom = "point", 
               size = 3, shape = 16) +
  geom_jitter(aes(y = Ct), alpha = 0.3, width = 0.3, height = 0, 
              size = 2, shape = 16) +
  
  # DNA concentration (secondary axis) - dashed lines with triangles
  stat_summary(aes(y = conc_to_ct(`Calc Conc (ng/ul)`)), 
               fun = mean, geom = "line", 
               linewidth = 1.0, linetype = "dashed", alpha = 0.6) +
  stat_summary(aes(y = conc_to_ct(`Calc Conc (ng/ul)`)), 
               fun = mean, geom = "point", 
               size = 2.5, shape = 17, alpha = 0.6) +
  geom_jitter(aes(y = conc_to_ct(`Calc Conc (ng/ul)`)), 
              alpha = 0.2, width = 0.3, height = 0, 
              size = 1.5, shape = 17) +
  
  scale_y_reverse(
    name = "Ct Value (Lower = More Target)",
    sec.axis = sec_axis(
      trans = ~ ct_to_conc(.),
      name = "DNA Concentration (ng/µl)"
    )
  ) +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 24),
                     labels = c("0", "4", "8", "12", "24")) +
  scale_color_manual(values = c("BF Bolton Base" = "#F8766D", 
                                "BF Bolton Base + CAT" = "#7CAE00",
                                "BPW" = "#00BFC4",
                                "RVS" = "#C77CFF")) +
  labs(
    title = expression(paste(italic("Campylobacter"), " Enrichment: Ct Values & DNA Concentration")),
    subtitle = "Solid circles/lines = Ct values | Dashed triangles/lines = DNA concentration (log scale)",
    x = "Time (hours)",
    color = "Broth"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10),
    axis.title.y.left = element_text(color = "black", face = "bold"),
    axis.title.y.right = element_text(color = "gray40", face = "bold"),
    legend.position = "right"
  )

print(campy_dual_axis_plot)
ggsave("campy_dual_axis_time_course.png", 
       width = 11, height = 6, dpi = 600)
```

```{r campy_panel_plots, fig.height=8}
# ============================================================================
# SEPARATE PANELS PLOT (RECOMMENDED)
# ============================================================================
campy_long <- campy_conc_data %>%
  select(Time_hr, Broth, Ct, `Calc Conc (ng/ul)`) %>%
  pivot_longer(
    cols = c(Ct, `Calc Conc (ng/ul)`),
    names_to = "Measurement",
    values_to = "Value"
  ) %>%
  mutate(
    Measurement = factor(Measurement, 
                        levels = c("Ct", "Calc Conc (ng/ul)"),
                        labels = c("Ct Value", 
                                  "DNA Concentration (ng/µl)"))
  )

campy_panel_plot <- ggplot(campy_long, 
                          aes(x = Time_hr, y = Value, 
                              color = Broth, group = Broth)) +
  stat_summary(fun = mean, geom = "line", linewidth = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 1, linewidth = 0.8) +
  geom_jitter(alpha = 0.3, width = 0.3, height = 0, size = 2) +
  facet_wrap(~ Measurement, ncol = 1, scales = "free_y", strip.position = "left") +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 24),
                     labels = c("0", "4", "8", "12", "24")) +
  scale_color_manual(values = c("BF Bolton Base" = "#F8766D", 
                                "BF Bolton Base + CAT" = "#7CAE00",
                                "BPW" = "#00BFC4",
                                "RVS" = "#C77CFF")) +
  labs(
    title = expression(paste(italic("Campylobacter"), " Enrichment Testing: Ct Values & DNA Concentration")),
    subtitle = "Each point = qPCR run replicates | Lines = mean ± SE",
    x = "Time (hours)",
    y = NULL,
    color = "Broth"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10),
    strip.background = element_rect(fill = "gray90"),
    strip.text = element_text(face = "bold", size = 10),
    strip.placement = "outside",
    legend.position = "right"
  )

print(campy_panel_plot)
ggsave("campy_panels_time_course.png", 
       width = 10, height = 8, dpi = 600)
```

```{r campy_panel_sidebyside}
# ============================================================================
# SIDE-BY-SIDE PANELS (Alternative layout)
# ============================================================================
campy_panel_sidebyside <- ggplot(campy_long, 
                                aes(x = Time_hr, y = Value, 
                                    color = Broth, group = Broth)) +
  stat_summary(fun = mean, geom = "line", linewidth = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 1, linewidth = 0.8) +
  geom_jitter(alpha = 0.3, width = 0.3, height = 0, size = 2) +
  facet_wrap(~ Measurement, ncol = 2, scales = "free_y") +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 24),
                     labels = c("0", "4", "8", "12", "24")) +
  scale_color_manual(values = c("BF Bolton Base" = "#F8766D", 
                                "BF Bolton Base + CAT" = "#7CAE00",
                                "BPW" = "#00BFC4",
                                "RVS" = "#C77CFF")) +
  labs(
    title = expression(paste(italic("Campylobacter"), " Enrichment: Ct Values & DNA Concentration")),
    subtitle = "Points = individual replicates, Lines = mean ± SE | Y-axes on log scale",
    x = "Time (hours)",
    y = NULL,
    color = "Broth"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10),
    strip.background = element_rect(fill = "gray90"),
    strip.text = element_text(face = "bold", size = 10),
    legend.position = "bottom"
  )

campy_long

print(campy_panel_sidebyside)
ggsave("campy_panels_sidebyside_time_course.png", 
       width = 12, height = 6, dpi = 600)
```

---

## 1.3 Campylobacter Concentration Analysis

### 1.3.1 Within-Broth Statistics (Concentration)

```{r campy_conc_within_broth_stats}
# Prepare data for baseline vs 24hr comparison
campy_conc_within_data <- campy_cleaned %>%
  left_join(campy_baseline_map, by = "Broth") %>%
  filter(Time_hr %in% c(baseline_hr, 24)) %>%
  filter(!is.na(Concentration))

# Calculate within-broth statistics
campy_conc_within_stats <- campy_conc_within_data %>%
  group_by(Broth, baseline_hr) %>%
  summarise(
    n_baseline = sum(Time_hr == baseline_hr),
    n_24hr = sum(Time_hr == 24),
    mean_conc_baseline = mean(Concentration[Time_hr == baseline_hr]),
    sd_conc_baseline = sd(Concentration[Time_hr == baseline_hr]),
    mean_conc_24hr = mean(Concentration[Time_hr == 24]),
    sd_conc_24hr = sd(Concentration[Time_hr == 24]),
    fold_change = mean_conc_24hr / mean_conc_baseline,
    log2_fold_change = log2(fold_change),
    p_value = safe_wilcox_test(cur_data(), "Concentration", "Time_hr", unique(baseline_hr), 24),
    .groups = "drop"
  ) %>%
  mutate(
    comparison = paste0(baseline_hr, "hr vs 24hr"),
    significant = if_else(!is.na(p_value) & p_value < 0.05, "Yes", "No"),
    interpretation = interpret_p_value(p_value)
  )

# Display results
kable(campy_conc_within_stats %>% 
        select(Broth, comparison, n_baseline, n_24hr, 
               mean_conc_baseline, mean_conc_24hr, 
               fold_change, p_value, significant),
      caption = "Campylobacter Within-Broth Comparison (Concentration)",
      digits = c(0, 0, 0, 0, 4, 2, 1, 4, 0)) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(which(campy_conc_within_stats$significant == "Yes"), 
           bold = TRUE, color = "white", background = "#28a745")
```

**Detailed Statistical Results (Concentration Analysis):**

```{r campy_conc_individual_broth_results, results='asis'}
# Print detailed results for each broth
for (i in 1:nrow(campy_conc_within_stats)) {
  broth_data <- campy_conc_within_stats[i, ]
  
  cat("\n#### **", broth_data$Broth, "Broth (", broth_data$comparison, ")**\n\n", sep = "")
  cat("- **Sample size:** n =", broth_data$n_baseline, "(baseline), n =", broth_data$n_24hr, "(24hr)\n")
  cat("- **Mean concentration at baseline:**", format(broth_data$mean_conc_baseline, scientific = TRUE, digits = 3), 
      "± ", format(broth_data$sd_conc_baseline, scientific = TRUE, digits = 2), "SD ng/µl\n")
  cat("- **Mean concentration at 24hr:**", format(broth_data$mean_conc_24hr, scientific = TRUE, digits = 3), 
      "± ", format(broth_data$sd_conc_24hr, scientific = TRUE, digits = 2), "SD ng/µl\n")
  cat("- **Fold change:**", format(round(broth_data$fold_change, 1), big.mark = ","), 
      "-fold increase\n")
  cat("- **Log2 fold change:**", round(broth_data$log2_fold_change, 2), "\n")
  cat("- **Wilcoxon rank-sum test:** p =", format.pval(broth_data$p_value, digits = 4), "\n")
  cat("- **Significance:**", broth_data$significant, "-", broth_data$interpretation, "\n")
  
  # Add interpretation note
  if (!is.na(broth_data$fold_change) && broth_data$fold_change > 1) {
    cat("- **Interpretation:** Concentration increased", 
        format(round(broth_data$fold_change, 1), big.mark = ","), 
        "times from baseline to 24hr\n")
  }
  cat("\n")
}
```

**Key Findings (Concentration Analysis):**

- **BF Bolton Base**: `r format(round(campy_conc_within_stats$fold_change[campy_conc_within_stats$Broth == "BF Bolton Base"], 1), big.mark = ",")`-fold increase (p = `r round(campy_conc_within_stats$p_value[campy_conc_within_stats$Broth == "BF Bolton Base"], 3)`)
- **BF Bolton Base + CAT**: `r format(round(campy_conc_within_stats$fold_change[campy_conc_within_stats$Broth == "BF Bolton Base + CAT"], 1), big.mark = ",")`-fold increase (p = `r round(campy_conc_within_stats$p_value[campy_conc_within_stats$Broth == "BF Bolton Base + CAT"], 3)`)
- **BPW**: `r ifelse("BPW" %in% campy_conc_within_stats$Broth, paste0(format(round(campy_conc_within_stats$fold_change[campy_conc_within_stats$Broth == "BPW"], 1), big.mark = ","), "-fold change"), "Insufficient data for analysis")`
- **RVS**: `r ifelse("RVS" %in% campy_conc_within_stats$Broth, paste0(format(round(campy_conc_within_stats$fold_change[campy_conc_within_stats$Broth == "RVS"], 1), big.mark = ","), "-fold change"), "No detectable signal")`

**Statistical Note:** Small sample sizes (n=3) limit statistical power. Large fold changes indicate biological significance despite p > 0.05.

### 1.3.2 Between-Broth Statistics at 24hr (Concentration)

```{r campy_conc_between_broth_stats}
# Prepare data for between-broth comparison at 24hr
campy_conc_24hr <- campy_cleaned %>%
  filter(Time_hr == 24, !is.na(Concentration))

# Summary by broth
campy_conc_24hr_summary <- campy_conc_24hr %>%
  group_by(Broth) %>%
  summarise(
    n = n(),
    mean_conc = mean(Concentration),
    sd_conc = sd(Concentration),
    median_conc = median(Concentration),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_conc))

kable(campy_conc_24hr_summary,
      caption = "Campylobacter Concentration at 24hr (ng/µl)",
      digits = 4) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Kruskal-Wallis test
if (n_distinct(campy_conc_24hr$Broth) >= 2) {
  campy_conc_kw <- kruskal.test(Concentration ~ Broth, data = campy_conc_24hr)
  cat("\n**Kruskal-Wallis Test:**\n")
  cat("χ² =", round(campy_conc_kw$statistic, 3), 
      ", p =", format.pval(campy_conc_kw$p.value, digits = 3), "\n")
  
  if (campy_conc_kw$p.value < 0.05) {
    cat("**Result: Significant differences exist between broths (p < 0.05)**\n")
  }
}
```

---

## 1.4 Campylobacter Summary

### Overall Comparison: All Broths (0hr vs 24hr)

```{r campy_summary_comparison}
# Comparison table: Ct vs Concentration methods
campy_comparison <- campy_ct_within_stats %>%
  select(Broth, 
         n_Ct = n_24hr,
         mean_Ct_0hr = mean_Ct_baseline, 
         mean_Ct_24hr, 
         delta_Ct,
         fold_change_ct = fold_change, 
         p_value_ct = p_value) %>%
  left_join(
    campy_conc_within_stats %>%
      select(Broth, 
             n_Conc = n_24hr,
             mean_conc_0hr = mean_conc_baseline,
             mean_conc_24hr, 
             fold_change_conc = fold_change, 
             p_value_conc = p_value),
    by = "Broth"
  )

kable(campy_comparison %>%
        mutate(
          mean_Ct_0hr = round(mean_Ct_0hr, 2),
          mean_Ct_24hr = round(mean_Ct_24hr, 2),
          delta_Ct = round(delta_Ct, 2),
          fold_change_ct = format(round(fold_change_ct, 0), big.mark = ","),
          p_value_ct = format.pval(p_value_ct, digits = 3),
          mean_conc_0hr = format(mean_conc_0hr, scientific = TRUE, digits = 2),
          mean_conc_24hr = format(mean_conc_24hr, scientific = TRUE, digits = 2),
          fold_change_conc = format(round(fold_change_conc, 0), big.mark = ","),
          p_value_conc = format.pval(p_value_conc, digits = 3)
        ),
      caption = "Campylobacter: Complete Comparison (Ct and Concentration)",
      col.names = c("Broth", "n", "Ct 0hr", "Ct 24hr", "ΔCt", "Fold↑ (Ct)", "p-value",
                    "n", "Conc 0hr", "Conc 24hr", "Fold↑ (Conc)", "p-value")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  add_header_above(c(" " = 1, "Ct Value Analysis" = 6, "Concentration Analysis" = 5))
```

### Statistical Interpretation by Broth

```{r campy_interpretation_table, results='asis'}
cat("\n")
cat("| Broth | Growth Status | ΔCt (cycles) | Fold Change | Significance | Recommendation |\n")
cat("|-------|---------------|--------------|-------------|--------------|----------------|\n")

for (i in 1:nrow(campy_ct_within_stats)) {
  broth <- campy_ct_within_stats$Broth[i]
  delta_ct <- campy_ct_within_stats$delta_Ct[i]
  fold <- campy_ct_within_stats$fold_change[i]
  pval <- campy_ct_within_stats$p_value[i]
  
  # Determine growth status
  if (is.na(delta_ct)) {
    status <- "No detection"
    recommendation <- "Not suitable"
  } else if (delta_ct < -15) {
    status <- "Excellent growth"
    recommendation <- "**Highly recommended**"
  } else if (delta_ct < -10) {
    status <- "Good growth"
    recommendation <- "**Recommended**"
  } else if (delta_ct < -5) {
    status <- "Moderate growth"
    recommendation <- "Acceptable"
  } else if (delta_ct < 0) {
    status <- "Minimal growth"
    recommendation <- "Limited suitability"
  } else {
    status <- "No growth"
    recommendation <- "Not suitable"
  }
  
  # Significance
  sig <- ifelse(is.na(pval), "N/A", 
                ifelse(pval < 0.05, "Yes (p<0.05)", 
                       paste0("No (p=", round(pval, 2), ")")))
  
  cat("| ", broth, " | ", status, " | ", 
      ifelse(is.na(delta_ct), "N/A", round(delta_ct, 2)), " | ",
      ifelse(is.na(fold), "N/A", format(round(fold, 0), big.mark = ",")), " | ",
      sig, " | ", recommendation, " |\n", sep = "")
}
```

**Campylobacter Conclusions:**

1. **BF Bolton Base + CAT** shows the best performance for *Campylobacter* enrichment at 24hr
   - ΔCt: `r round(campy_ct_within_stats$delta_Ct[campy_ct_within_stats$Broth == "BF Bolton Base + CAT"], 2)` cycles (most negative)
   - `r format(round(campy_ct_within_stats$fold_change[campy_ct_within_stats$Broth == "BF Bolton Base + CAT"], 0), big.mark = ",")` fold increase
   - **RECOMMENDED for *Campylobacter* enrichment**

2. **BF Bolton Base** shows substantial growth
   - ΔCt: `r round(campy_ct_within_stats$delta_Ct[campy_ct_within_stats$Broth == "BF Bolton Base"], 2)` cycles
   - `r format(round(campy_ct_within_stats$fold_change[campy_ct_within_stats$Broth == "BF Bolton Base"], 0), big.mark = ",")` fold increase
   - **Acceptable alternative**

3. **BPW and RVS** show minimal/no detectable growth
   - Not suitable for *Campylobacter* enrichment

4. Both Ct and concentration analyses reach the same biological conclusions

5. **Statistical caveat:** Small sample sizes (n=3) limit power to detect significance in within-broth comparisons despite large effect sizes

---

# PART 2: SALMONELLA ANALYSIS

## 2.1 Data Loading and Cleaning

```{r load_salmonella}
# Read Salmonella data
salm_raw <- read_excel('/Volumes/Alice/1_METHOD_DEVELOPMENT/Enrichment_testing/Raw_data/qPCR_Salmonella_enrichment-testing.xlsx')


# Clean and prepare data
salm_cleaned <- salm_raw %>%
  filter(Type == "Unknown") %>%
  mutate(
    Broth = str_extract(Name, "^[A-Z-]+"),
    Time_hr = as.numeric(str_extract(Name, "\\d+(?=hr)")),
    Ct = as.numeric(Ct),
    Concentration = as.numeric(`Calc Conc (ng/ul)`)
  ) %>%
  filter(Broth %in% c("BF-BB", "CAT", "BPW", "RVS")) %>%
  select(Name, Broth, Time_hr, Ct, Concentration, `Calc Conc (ng/ul)`)

# Rename broths to full names
salm_cleaned <- salm_cleaned %>%
  mutate(Broth = case_when(
    Broth == "BF-BB" ~ "BF Bolton Base",
    Broth == "CAT" ~ "BF Bolton Base + CAT",
    TRUE ~ Broth
  ))

# Define baseline times for Salmonella
salm_baseline_map <- tibble(
  Broth = c("BF Bolton Base", "BF Bolton Base + CAT", "BPW", "RVS"),
  baseline_hr = c(0, 0, 0, 0)
)
```

```{r salm_data_summary}
# Data summary
salm_summary <- salm_cleaned %>%
  group_by(Broth, Time_hr) %>%
  summarise(
    n_replicates = n(),
    n_Ct_detected = sum(!is.na(Ct)),
    n_Conc_detected = sum(!is.na(Concentration)),
    mean_Ct = mean(Ct, na.rm = TRUE),
    mean_Conc = mean(Concentration, na.rm = TRUE),
    .groups = "drop"
  )

kable(salm_summary, 
      caption = "Salmonella Data Summary",
      digits = 4) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

---

## 2.2 Salmonella Ct Value Analysis

### 2.2.1 Within-Broth Statistics (Ct)

```{r salm_ct_within_broth_stats}
# Prepare data for baseline vs 24hr comparison
salm_ct_within_data <- salm_cleaned %>%
  left_join(salm_baseline_map, by = "Broth") %>%
  filter(Time_hr %in% c(baseline_hr, 24)) %>%
  filter(!is.na(Ct))

# Calculate within-broth statistics
salm_ct_within_stats <- salm_ct_within_data %>%
  group_by(Broth, baseline_hr) %>%
  summarise(
    n_baseline = sum(Time_hr == baseline_hr),
    n_24hr = sum(Time_hr == 24),
    mean_Ct_baseline = mean(Ct[Time_hr == baseline_hr]),
    sd_Ct_baseline = sd(Ct[Time_hr == baseline_hr]),
    mean_Ct_24hr = mean(Ct[Time_hr == 24]),
    sd_Ct_24hr = sd(Ct[Time_hr == 24]),
    delta_Ct = mean_Ct_24hr - mean_Ct_baseline,
    fold_change = 2^(-delta_Ct),
    p_value = safe_wilcox_test(cur_data(), "Ct", "Time_hr", unique(baseline_hr), 24),
    .groups = "drop"
  ) %>%
  mutate(
    comparison = paste0(baseline_hr, "hr vs 24hr"),
    significant = if_else(!is.na(p_value) & p_value < 0.05, "Yes", "No"),
    interpretation = interpret_p_value(p_value)
  )

# Display results
kable(salm_ct_within_stats %>% 
        select(Broth, comparison, n_baseline, n_24hr, 
               mean_Ct_baseline, mean_Ct_24hr, delta_Ct, 
               fold_change, p_value, significant),
      caption = "Salmonella Within-Broth Comparison (Ct Values)",
      digits = c(0, 0, 0, 0, 2, 2, 2, 1, 4, 0)) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(which(salm_ct_within_stats$significant == "Yes"), 
           bold = TRUE, color = "white", background = "#28a745")
```

### 2.2.2 Between-Broth Statistics at 24hr (Ct)

```{r salm_ct_between_broth_stats}
# Prepare data for between-broth comparison at 24hr
salm_ct_24hr <- salm_cleaned %>%
  filter(Time_hr == 24, !is.na(Ct))

# Summary by broth
salm_ct_24hr_summary <- salm_ct_24hr %>%
  group_by(Broth) %>%
  summarise(
    n = n(),
    mean_Ct = mean(Ct),
    sd_Ct = sd(Ct),
    median_Ct = median(Ct),
    .groups = "drop"
  ) %>%
  arrange(mean_Ct)

kable(salm_ct_24hr_summary,
      caption = "Salmonella Ct Values at 24hr (Lower Ct = More target)",
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Kruskal-Wallis test
if (n_distinct(salm_ct_24hr$Broth) >= 2) {
  salm_ct_kw <- kruskal.test(Ct ~ Broth, data = salm_ct_24hr)
  cat("\n**Kruskal-Wallis Test:**\n")
  cat("χ² =", round(salm_ct_kw$statistic, 3), 
      ", p =", format.pval(salm_ct_kw$p.value, digits = 3), "\n")
  
  if (salm_ct_kw$p.value < 0.05) {
    cat("**Result: Significant differences exist between broths (p < 0.05)**\n")
  }
}
```

### 2.2.3 Salmonella Ct Plots

```{r salm_ct_baseline_vs_24hr_plot, fig.height=8}
# Plot: Baseline vs 24hr (Ct values)
salm_ct_plot_baseline_24 <- ggplot(salm_ct_within_data, 
                                     aes(x = factor(Time_hr), y = Ct, 
                                         fill = factor(Time_hr))) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.6, size = 2.5) +
  facet_wrap(~Broth, scales = "free_y", ncol = 2) +
  scale_y_reverse() +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = expression(paste(italic("Salmonella"), " Ct Values: Baseline vs 24hr Comparison")),
    subtitle = "Lower Ct = Higher target amount",
    x = "Time Point (hours)",
    y = "Ct Value",
    fill = "Time Point"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, color = "gray30"),
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 11),
    strip.background = element_rect(fill = "gray90", color = NA)
  )

print(salm_ct_plot_baseline_24)
ggsave("salm_Ct_baseline_vs_24hr.png", 
       width = 10, height = 8, dpi = 600)
```

```{r salm_ct_broth_comparison_plot}
# Plot: Between-broth comparison at 24hr (Ct values)
salm_ct_plot_broths_24 <- ggplot(salm_ct_24hr, 
                                   aes(x = reorder(Broth, Ct), y = Ct, fill = Broth)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.6, size = 3) +
  scale_y_reverse() +
  scale_fill_manual(values = c("BF Bolton Base" = "#F8766D", 
                                "BF Bolton Base + CAT" = "#7CAE00",
                                "BPW" = "#00BFC4",
                                "RVS" = "#C77CFF")) +
  labs(
    title = expression(paste(italic("Salmonella"), " Ct Values at 24hr: Between-Broth Comparison")),
    subtitle = "Lower Ct = Higher target amount",
    x = "Broth Type",
    y = "Ct Value"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11, face = "bold")
  )

print(salm_ct_plot_broths_24)
ggsave("salm_Ct_broths_24hr.png", 
       width = 10, height = 6, dpi = 600)
```

```{r salm_ct_time_course_plot}
# Plot: Time course (Ct values)
salm_ct_time_course_data <- salm_cleaned %>%
  filter(!is.na(Ct))

salm_ct_plot_time_course <- ggplot(salm_ct_time_course_data, 
                                     aes(x = Time_hr, y = Ct, 
                                         color = Broth, group = Broth)) +
  stat_summary(fun = mean, geom = "line", linewidth = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 1, linewidth = 0.8) +
  geom_jitter(alpha = 0.3, width = 0.3, height = 0, size = 2) +
  scale_y_reverse() +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 24),
                     labels = c("0", "4", "8", "12", "24")) +
  scale_color_manual(values = c("BF Bolton Base" = "#F8766D", 
                                "BF Bolton Base + CAT" = "#7CAE00",
                                "BPW" = "#00BFC4",
                                "RVS" = "#C77CFF")) +
  labs(
    title = expression(paste(italic("Salmonella"), " Ct Values: Enrichment Testing")),
    subtitle = "Lower Ct = Higher target; Points = individual replicates, Lines = mean ± SE",
    x = "Time (hours)",
    y = "Ct Value",
    color = "Broth"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "right"
  )

print(salm_ct_plot_time_course)
ggsave("salm_Ct_time_course.png", 
       width = 10, height = 6, dpi = 600)
```

### 2.2.4 Salmonella Dual-Axis and Panel Plots

```{r salm_dual_axis_plot}
# ============================================================================
# DUAL-AXIS PLOT: Ct + DNA Concentration
# ============================================================================
salm_conc_data <- salm_cleaned %>%
  filter(!is.na(Ct) & !is.na(`Calc Conc (ng/ul)`))

# Calculate the range for scaling (using log scale for concentration)
ct_min <- min(salm_conc_data$Ct, na.rm = TRUE)
ct_max <- max(salm_conc_data$Ct, na.rm = TRUE)
conc_min <- min(salm_conc_data$`Calc Conc (ng/ul)`, na.rm = TRUE)
conc_max <- max(salm_conc_data$`Calc Conc (ng/ul)`, na.rm = TRUE)

# Transformation function for concentration to Ct scale (using log scale)
conc_to_ct <- function(conc) {
  ct_max - (ct_max - ct_min) * (log10(conc) - log10(conc_min)) / (log10(conc_max) - log10(conc_min))
}

# Inverse transformation for secondary axis (from Ct to log concentration)
ct_to_conc <- function(ct) {
  10^(log10(conc_min) + (ct_max - ct) * (log10(conc_max) - log10(conc_min)) / (ct_max - ct_min))
}

salm_dual_axis_plot <- ggplot(salm_conc_data, 
                               aes(x = Time_hr, color = Broth, group = Broth)) +
  # Ct values (primary axis) - solid lines
  stat_summary(aes(y = Ct), fun = mean, geom = "line", 
               linewidth = 1.2, linetype = "solid") +
  stat_summary(aes(y = Ct), fun = mean, geom = "point", 
               size = 3, shape = 16) +
  geom_jitter(aes(y = Ct), alpha = 0.3, width = 0.3, height = 0, 
              size = 2, shape = 16) +
  
  # DNA concentration (secondary axis) - dashed lines with triangles
  stat_summary(aes(y = conc_to_ct(`Calc Conc (ng/ul)`)), 
               fun = mean, geom = "line", 
               linewidth = 1.0, linetype = "dashed", alpha = 0.6) +
  stat_summary(aes(y = conc_to_ct(`Calc Conc (ng/ul)`)), 
               fun = mean, geom = "point", 
               size = 2.5, shape = 17, alpha = 0.6) +
  geom_jitter(aes(y = conc_to_ct(`Calc Conc (ng/ul)`)), 
              alpha = 0.2, width = 0.3, height = 0, 
              size = 1.5, shape = 17) +
  
  scale_y_reverse(
    name = "Ct Value (Lower = More Target)",
    sec.axis = sec_axis(
      trans = ~ ct_to_conc(.),
      name = "DNA Concentration (ng/µl)"
    )
  ) +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 24),
                     labels = c("0", "4", "8", "12", "24")) +
  scale_color_manual(values = c("BF Bolton Base" = "#F8766D", 
                                "BF Bolton Base + CAT" = "#7CAE00",
                                "BPW" = "#00BFC4",
                                "RVS" = "#C77CFF")) +
  labs(
    title = expression(paste(italic("Salmonella"), " Enrichment: Ct Values & DNA Concentration")),
    subtitle = "Solid circles/lines = Ct values | Dashed triangles/lines = DNA concentration (log scale)",
    x = "Time (hours)",
    color = "Broth"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10),
    axis.title.y.left = element_text(color = "black", face = "bold"),
    axis.title.y.right = element_text(color = "gray40", face = "bold"),
    legend.position = "right"
  )

print(salm_dual_axis_plot)
ggsave("salm_dual_axis_time_course.png", 
       width = 11, height = 6, dpi = 600)
```

```{r salm_panel_plots, fig.height=8}
# ============================================================================
# SEPARATE PANELS PLOT (RECOMMENDED)
# ============================================================================
salm_long <- salm_conc_data %>%
  select(Time_hr, Broth, Ct, `Calc Conc (ng/ul)`) %>%
  pivot_longer(
    cols = c(Ct, `Calc Conc (ng/ul)`),
    names_to = "Measurement",
    values_to = "Value"
  ) %>%
  mutate(
    Measurement = factor(Measurement, 
                        levels = c("Ct", "Calc Conc (ng/ul)"),
                        labels = c("Ct Value", 
                                  "DNA Concentration (ng/µl)(log10 scale)"))
  )

salm_panel_plot <- ggplot(salm_long, 
                          aes(x = Time_hr, y = Value, 
                              color = Broth, group = Broth)) +
  stat_summary(fun = mean, geom = "line", linewidth = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 1, linewidth = 0.8) +
  geom_jitter(alpha = 0.3, width = 0.3, height = 0, size = 2) +
  facet_wrap(~ Measurement, ncol = 1, scales = "free_y", strip.position = "left") +
  scale_y_log10() +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 24),
                     labels = c("0", "4", "8", "12", "24")) +
  scale_color_manual(values = c("BF Bolton Base" = "#F8766D", 
                                "BF Bolton Base + CAT" = "#7CAE00",
                                "BPW" = "#00BFC4",
                                "RVS" = "#C77CFF")) +
  labs(
    title = expression(paste(italic("Salmonella"), " Enrichment Testing: Ct Values & DNA Concentration")),
    subtitle = "Each point = qPCR run replicates | Lines = mean ± SE | DNA concentration on log scale",
    x = "Time (hours)",
    y = NULL,
    color = "Broth"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10),
    strip.background = element_rect(fill = "gray90"),
    strip.text = element_text(face = "bold", size = 10),
    strip.placement = "outside",
    legend.position = "right"
  )

print(salm_panel_plot)
ggsave("salm_panels_time_course.png", 
       width = 10, height = 8, dpi = 600)
```

```{r salm_panel_sidebyside}
# ============================================================================
# SIDE-BY-SIDE PANELS (Alternative layout)
# ============================================================================
salm_panel_sidebyside <- ggplot(salm_long, 
                                aes(x = Time_hr, y = Value, 
                                    color = Broth, group = Broth)) +
  stat_summary(fun = mean, geom = "line", linewidth = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 1, linewidth = 0.8) +
  geom_jitter(alpha = 0.3, width = 0.3, height = 0, size = 2) +
  facet_wrap(~ Measurement, ncol = 2, scales = "free_y") +
  scale_y_log10() +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 24),
                     labels = c("0", "4", "8", "12", "24")) +
  scale_color_manual(values = c("BF Bolton Base" = "#F8766D", 
                                "BF Bolton Base + CAT" = "#7CAE00",
                                "BPW" = "#00BFC4",
                                "RVS" = "#C77CFF")) +
  labs(
    title = expression(paste(italic("Salmonella"), " Enrichment: Ct Values & DNA Concentration")),
    subtitle = "Points = individual replicates, Lines = mean ± SE | Y-axes on log scale",
    x = "Time (hours)",
    y = NULL,
    color = "Broth"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10),
    strip.background = element_rect(fill = "gray90"),
    strip.text = element_text(face = "bold", size = 10),
    legend.position = "bottom"
  )

print(salm_panel_sidebyside)
ggsave("salm_panels_sidebyside_time_course.png", 
       width = 12, height = 6, dpi = 600)
```

---

## 2.3 Salmonella Concentration Analysis

### 2.3.1 Within-Broth Statistics (Concentration)

```{r salm_conc_within_broth_stats}
# Prepare data for baseline vs 24hr comparison
salm_conc_within_data <- salm_cleaned %>%
  left_join(salm_baseline_map, by = "Broth") %>%
  filter(Time_hr %in% c(baseline_hr, 24)) %>%
  filter(!is.na(Concentration))

# Calculate within-broth statistics
salm_conc_within_stats <- salm_conc_within_data %>%
  group_by(Broth, baseline_hr) %>%
  summarise(
    n_baseline = sum(Time_hr == baseline_hr),
    n_24hr = sum(Time_hr == 24),
    mean_conc_baseline = mean(Concentration[Time_hr == baseline_hr]),
    sd_conc_baseline = sd(Concentration[Time_hr == baseline_hr]),
    mean_conc_24hr = mean(Concentration[Time_hr == 24]),
    sd_conc_24hr = sd(Concentration[Time_hr == 24]),
    fold_change = mean_conc_24hr / mean_conc_baseline,
    log2_fold_change = log2(fold_change),
    p_value = safe_wilcox_test(cur_data(), "Concentration", "Time_hr", unique(baseline_hr), 24),
    .groups = "drop"
  ) %>%
  mutate(
    comparison = paste0(baseline_hr, "hr vs 24hr"),
    significant = if_else(!is.na(p_value) & p_value < 0.05, "Yes", "No"),
    interpretation = interpret_p_value(p_value)
  )

# Display results
kable(salm_conc_within_stats %>% 
        select(Broth, comparison, n_baseline, n_24hr, 
               mean_conc_baseline, mean_conc_24hr, 
               fold_change, p_value, significant),
      caption = "Salmonella Within-Broth Comparison (Concentration)",
      digits = c(0, 0, 0, 0, 4, 2, 1, 4, 0)) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(which(salm_conc_within_stats$significant == "Yes"), 
           bold = TRUE, color = "white", background = "#28a745")
```

### 2.3.2 Between-Broth Statistics at 24hr (Concentration)

```{r salm_conc_between_broth_stats}
# Prepare data for between-broth comparison at 24hr
salm_conc_24hr <- salm_cleaned %>%
  filter(Time_hr == 24, !is.na(Concentration))

# Summary by broth
salm_conc_24hr_summary <- salm_conc_24hr %>%
  group_by(Broth) %>%
  summarise(
    n = n(),
    mean_conc = mean(Concentration),
    sd_conc = sd(Concentration),
    median_conc = median(Concentration),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_conc))

kable(salm_conc_24hr_summary,
      caption = "Salmonella Concentration at 24hr (ng/µl)",
      digits = 4) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Kruskal-Wallis test
if (n_distinct(salm_conc_24hr$Broth) >= 2) {
  salm_conc_kw <- kruskal.test(Concentration ~ Broth, data = salm_conc_24hr)
  cat("\n**Kruskal-Wallis Test:**\n")
  cat("χ² =", round(salm_conc_kw$statistic, 3), 
      ", p =", format.pval(salm_conc_kw$p.value, digits = 3), "\n")
  
  if (salm_conc_kw$p.value < 0.05) {
    cat("**Result: Significant differences exist between broths (p < 0.05)**\n")
  }
}
```

---

## 2.4 Salmonella Summary

### Overall Comparison: All Broths (0hr vs 24hr)

```{r salm_summary_comparison}
# Comparison table: Ct vs Concentration methods
salm_comparison <- salm_ct_within_stats %>%
  select(Broth, 
         n_Ct = n_24hr,
         mean_Ct_0hr = mean_Ct_baseline, 
         mean_Ct_24hr, 
         delta_Ct,
         fold_change_ct = fold_change, 
         p_value_ct = p_value) %>%
  left_join(
    salm_conc_within_stats %>%
      select(Broth, 
             n_Conc = n_24hr,
             mean_conc_0hr = mean_conc_baseline,
             mean_conc_24hr, 
             fold_change_conc = fold_change, 
             p_value_conc = p_value),
    by = "Broth"
  )

kable(salm_comparison %>%
        mutate(
          mean_Ct_0hr = round(mean_Ct_0hr, 2),
          mean_Ct_24hr = round(mean_Ct_24hr, 2),
          delta_Ct = round(delta_Ct, 2),
          fold_change_ct = format(round(fold_change_ct, 0), big.mark = ","),
          p_value_ct = format.pval(p_value_ct, digits = 3),
          mean_conc_0hr = format(mean_conc_0hr, scientific = TRUE, digits = 2),
          mean_conc_24hr = format(mean_conc_24hr, scientific = TRUE, digits = 2),
          fold_change_conc = format(round(fold_change_conc, 0), big.mark = ","),
          p_value_conc = format.pval(p_value_conc, digits = 3)
        ),
      caption = "Salmonella: Complete Comparison (Ct and Concentration)",
      col.names = c("Broth", "n", "Ct 0hr", "Ct 24hr", "ΔCt", "Fold↑ (Ct)", "p-value",
                    "n", "Conc 0hr", "Conc 24hr", "Fold↑ (Conc)", "p-value")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  add_header_above(c(" " = 1, "Ct Value Analysis" = 6, "Concentration Analysis" = 5))
```

### Statistical Interpretation by Broth

```{r salm_interpretation_table, results='asis'}
cat("\n")
cat("| Broth | Growth Status | ΔCt (cycles) | Fold Change | Significance | Recommendation |\n")
cat("|-------|---------------|--------------|-------------|--------------|----------------|\n")

for (i in 1:nrow(salm_ct_within_stats)) {
  broth <- salm_ct_within_stats$Broth[i]
  delta_ct <- salm_ct_within_stats$delta_Ct[i]
  fold <- salm_ct_within_stats$fold_change[i]
  pval <- salm_ct_within_stats$p_value[i]
  
  # Determine growth status
  if (is.na(delta_ct)) {
    status <- "No detection"
    recommendation <- "Not suitable"
  } else if (delta_ct < -15) {
    status <- "Excellent growth"
    recommendation <- "**Highly recommended**"
  } else if (delta_ct < -10) {
    status <- "Good growth"
    recommendation <- "**Recommended**"
  } else if (delta_ct < -5) {
    status <- "Moderate growth"
    recommendation <- "Acceptable"
  } else if (delta_ct < 0) {
    status <- "Minimal growth"
    recommendation <- "Limited suitability"
  } else {
    status <- "No growth"
    recommendation <- "Not suitable"
  }
  
  # Significance
  sig <- ifelse(is.na(pval), "N/A", 
                ifelse(pval < 0.05, "Yes (p<0.05)", 
                       paste0("No (p=", round(pval, 2), ")")))
  
  cat("| ", broth, " | ", status, " | ", 
      ifelse(is.na(delta_ct), "N/A", round(delta_ct, 2)), " | ",
      ifelse(is.na(fold), "N/A", format(round(fold, 0), big.mark = ",")), " | ",
      sig, " | ", recommendation, " |\n", sep = "")
}
```
---

# Session Information

```{r session_info}
sessionInfo()
```
