---
title: "Read Tracking Across Processing Steps"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
dir.create("figures", showWarnings = FALSE)
library(tidyverse)
library(knitr)
library(kableExtra)
library(scales)
```

## Overview

Reads were tracked across three processing stages:

1. **Raw** — reads as sequenced
2. **Post-QC** — after adapter trimming (Porechop) and quality/length filtering (NanoFilt)
3. **Post-host depletion** — after bioinformatic removal of host (chicken) DNA

---

## Data

```{r data}
# QC data (raw → post-QC) from table across all sample types
qc_data <- tribble(
  ~sample_type,                    ~replicate, ~raw_reads, ~post_qc_reads,
  "Post-ASC",                      "CH01",      562337,     391752,
  "Post-ASC",                      "CH02",     1080072,     849899,
  "Post-ASC",                      "CH03",      98895,      69014,
  "Post-ASC",                      "CH04",      4532,       2960,
  "Post-ASC",                      "CH05",     1345828,    1073510,
  "Post-ASC",                      "CH06",      70431,      42563,
  "Post-ASC BF Bolton base + CAT", "CH01",     1780424,    1321890,
  "Post-ASC BF Bolton base + CAT", "CH02",     5385771,    4163882,
  "Post-ASC BF Bolton base + CAT", "CH03",     4136393,    3109141,
  "Post-ASC BF Bolton base + CAT", "CH04",      71146,      61211,
  "Post-ASC BF Bolton base + CAT", "CH05",     5234812,    4353808,
  "Post-ASC BF Bolton base + CAT", "CH06",     1281858,    1012131,
  "Post-ASC BPW then RVS",            "CH01",     6452280,    4877878,
  "Post-ASC BPW then RVS",            "CH02",     4380026,    2805742,
  "Post-ASC BPW then RVS",            "CH03",      28402,      20713,
  "Post-ASC BPW then RVS",            "CH04",      90856,      54875,
  "Post-ASC BPW then RVS",            "CH05",     3149323,    2281601,
  "Post-ASC BPW then RVS",            "CH06",     2833572,    2010907,
  "Post-EVI",                      "CH01",      868926,     624396,
  "Post-EVI",                      "CH02",      769659,     588592,
  "Post-EVI",                      "CH03",      401658,     348438,
  "Post-EVI",                      "CH04",      603448,     538739,
  "Post-EVI",                      "CH05",      65099,      33611,
  "Post-EVI",                      "CH06",      771002,     670622,
  "Post-EVI BF Bolton base + CAT", "CH01",     3213560,    1894168,
  "Post-EVI BF Bolton base + CAT", "CH02",     8046910,    5489116,
  "Post-EVI BF Bolton base + CAT", "CH03",     4393790,    2798460,
  "Post-EVI BF Bolton base + CAT", "CH04",     7064274,    4834243,
  "Post-EVI BF Bolton base + CAT", "CH05",     2811819,    2038763,
  "Post-EVI BF Bolton base + CAT", "CH06",     4965836,    3942476,
  "Post-EVI BPW then RVS",            "CH01",     1655061,    1185896,
  "Post-EVI BPW then RVS",            "CH02",     4352260,    3364055,
  "Post-EVI BPW then RVS",            "CH03",     2657706,    1878954,
  "Post-EVI BPW then RVS",            "CH04",     2285651,    1673254,
  "Post-EVI BPW then RVS",            "CH05",     3487325,    2483961,
  "Post-EVI BPW then RVS",            "CH06",     1570155,    1119562
)

# Fix the one typo in sample_type
qc_data <- qc_data %>%
  mutate(sample_type = str_replace(sample_type, "bpw", "BPW"))

# Host-depletion data (post-QC → post-host depletion) from NanoPlot HTML reports
host_data <- tribble(
  ~sample_type,                    ~replicate, ~post_qc_reads, ~post_host_reads, ~pct_removed,
  "Post-ASC",                      "CH01",      391752,          36119,           90.78,
  "Post-ASC",                      "CH02",      849899,          32675,           96.16,
  "Post-ASC",                      "CH03",       69014,           3716,           94.62,
  "Post-ASC",                      "CH04",        2960,            157,           94.70,
  "Post-ASC",                      "CH05",     1073510,          16688,           98.45,
  "Post-ASC",                      "CH06",       42563,           2504,           94.12,
  "Post-ASC BF Bolton base + CAT", "CH01",     1321890,        1013671,          23.32,
  "Post-ASC BF Bolton base + CAT", "CH02",     4163882,        2104950,          49.45,
  "Post-ASC BF Bolton base + CAT", "CH03",     3109141,        2565216,          17.49,
  "Post-ASC BF Bolton base + CAT", "CH04",      61211,          14214,           76.78,
  "Post-ASC BF Bolton base + CAT", "CH05",     4353808,        3814604,          12.38,
  "Post-ASC BF Bolton base + CAT", "CH06",     1012131,         766111,          24.31,
  "Post-ASC BPW then RVS",            "CH01",     4877878,        4630234,           5.08,
  "Post-ASC BPW then RVS",            "CH02",     2805742,        2626224,           6.40,
  "Post-ASC BPW then RVS",            "CH03",      20713,           4294,           79.27,
  "Post-ASC BPW then RVS",            "CH04",      54875,           6012,           89.04,
  "Post-ASC BPW then RVS",            "CH05",     2281601,        2185023,           4.23,
  "Post-ASC BPW then RVS",            "CH06",     2010907,        1847583,           8.12,
  "Post-EVI",                      "CH01",      624396,          15381,           97.54,
  "Post-EVI",                      "CH02",      588592,          15565,           97.36,
  "Post-EVI",                      "CH03",      348438,           4162,           98.81,
  "Post-EVI",                      "CH04",      538739,          11867,           97.80,
  "Post-EVI",                      "CH05",       33611,            301,           99.10,
  "Post-EVI",                      "CH06",      670622,          12092,           98.20,
  "Post-EVI BF Bolton base + CAT", "CH01",     1894168,        1723406,           9.02,
  "Post-EVI BF Bolton base + CAT", "CH02",     5489116,        5289232,           3.64,
  "Post-EVI BF Bolton base + CAT", "CH03",     2798460,        2488681,          11.07,
  "Post-EVI BF Bolton base + CAT", "CH04",     4834243,        4475906,           7.41,
  "Post-EVI BF Bolton base + CAT", "CH05",     2038763,        1689198,          17.15,
  "Post-EVI BF Bolton base + CAT", "CH06",     3942476,        3777871,           4.18,
  "Post-EVI BPW then RVS",            "CH01",     1185896,        1043198,          12.03,
  "Post-EVI BPW then RVS",            "CH02",     3364055,        3088263,           8.20,
  "Post-EVI BPW then RVS",            "CH03",     1878954,        1601298,          14.78,
  "Post-EVI BPW then RVS",            "CH04",     1673254,        1414998,          15.43,
  "Post-EVI BPW then RVS",            "CH05",     2483961,        1941906,          21.82,
  "Post-EVI BPW then RVS",            "CH06",     1119562,         952845,          14.89
)

# Merge all data
full_data <- qc_data %>%
  left_join(host_data %>% select(sample_type, replicate, post_host_reads, pct_removed),
            by = c("sample_type", "replicate")) %>%
  mutate(
    pct_retained_qc    = round(post_qc_reads   / raw_reads     * 100, 1),
    pct_retained_host  = round(post_host_reads / post_qc_reads * 100, 1),
    pct_retained_total = round(post_host_reads / raw_reads     * 100, 1)
  )
```

---

## Read tracking tables

### All sample types — QC step

```{r table-qc}
full_data %>%
  select(sample_type, replicate, raw_reads, post_qc_reads, pct_retained_qc) %>%
  mutate(
    raw_reads     = comma(raw_reads),
    post_qc_reads = comma(post_qc_reads),
    pct_retained_qc = paste0(pct_retained_qc, "%")
  ) %>%
  rename(
    `Sample type`       = sample_type,
    `Replicate`         = replicate,
    `Raw reads`         = raw_reads,
    `Post-QC reads`     = post_qc_reads,
    `% retained (QC)`   = pct_retained_qc
  ) %>%
  kable(align = c("l", "c", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  pack_rows("Post-ASC", 1, 6) %>%
  pack_rows("Post-ASC BF Bolton base + CAT", 7, 12) %>%
  pack_rows("Post-ASC BPW then RVS", 13, 18) %>%
  pack_rows("Post-EVI", 19, 24) %>%
  pack_rows("Post-EVI BF Bolton base + CAT", 25, 30) %>%
  pack_rows("Post-EVI BPW then RVS", 31, 36)
```

### All samples — host-DNA depletion step

```{r table-host}
full_data %>%
  select(sample_type, replicate, post_qc_reads, post_host_reads, pct_retained_host) %>%
  mutate(
    post_qc_reads     = comma(post_qc_reads),
    post_host_reads   = comma(post_host_reads),
    pct_retained_host = paste0(pct_retained_host, "%")
  ) %>%
  rename(
    `Sample type`               = sample_type,
    `Replicate`                 = replicate,
    `Post-QC reads`             = post_qc_reads,
    `Post-host depletion reads` = post_host_reads,
    `% retained (host removal)` = pct_retained_host
  ) %>%
  kable(align = c("l", "c", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  pack_rows("Post-ASC", 1, 6) %>%
  pack_rows("Post-ASC BF Bolton base + CAT", 7, 12) %>%
  pack_rows("Post-ASC BPW then RVS", 13, 18) %>%
  pack_rows("Post-EVI", 19, 24) %>%
  pack_rows("Post-EVI BF Bolton base + CAT", 25, 30) %>%
  pack_rows("Post-EVI BPW then RVS", 31, 36)
```

### All samples — full pipeline summary

```{r table-full}
full_data %>%
  select(sample_type, replicate, raw_reads, post_qc_reads, post_host_reads, pct_retained_total) %>%
  mutate(
    raw_reads          = comma(raw_reads),
    post_qc_reads      = comma(post_qc_reads),
    post_host_reads    = comma(post_host_reads),
    pct_retained_total = paste0(pct_retained_total, "%")
  ) %>%
  rename(
    `Sample type`          = sample_type,
    `Replicate`            = replicate,
    `Raw reads`            = raw_reads,
    `Post-QC`              = post_qc_reads,
    `Post-host depletion`  = post_host_reads,
    `% retained (overall)` = pct_retained_total
  ) %>%
  kable(align = c("l", "c", "r", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  pack_rows("Post-ASC", 1, 6) %>%
  pack_rows("Post-ASC BF Bolton base + CAT", 7, 12) %>%
  pack_rows("Post-ASC BPW then RVS", 13, 18) %>%
  pack_rows("Post-EVI", 19, 24) %>%
  pack_rows("Post-EVI BF Bolton base + CAT", 25, 30) %>%
  pack_rows("Post-EVI BPW then RVS", 31, 36)
```

---

## Visualisations

### Read counts at each processing step

```{r plot-tracking, fig.width=10, fig.height=9}
# Facet order: col 1 = Post-ASC, col 2 = Post-EVI, rows = enrichment type
facet_order <- c(
  "Post-EVI",                      "Post-ASC",
  "Post-EVI BF Bolton base + CAT", "Post-ASC BF Bolton base + CAT",
  "Post-EVI BPW then RVS",            "Post-ASC BPW then RVS"
)

plot_data <- full_data %>%
  select(sample_type, replicate, raw_reads, post_qc_reads, post_host_reads) %>%
  pivot_longer(cols = c(raw_reads, post_qc_reads, post_host_reads),
               names_to = "stage", values_to = "reads") %>%
  mutate(
    stage = factor(stage,
                   levels = c("raw_reads", "post_qc_reads", "post_host_reads"),
                   labels = c("Raw", "Post-QC", "Post-host\ndepletion")),
    sample_type = factor(sample_type, levels = facet_order)
  )

p1 <- ggplot(plot_data, aes(x = stage, y = reads, colour = replicate, group = replicate)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  facet_wrap(~sample_type, scales = "free_y", ncol = 2) +
  scale_y_continuous(labels = label_comma()) +
  scale_colour_brewer(palette = "Dark2") +
  labs(
    title = "Read count changes during bioinformatic processing steps",
    subtitle = "1. Raw reads | 2. After quality control (Porechop & Nanofilt) | 3. Removal of chicken (host) & contamination reads",
    x      = "Processing stage",
    y      = "Read count",
    colour = "Sample Number"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title            = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(face = "plain", size = 11),
    strip.background = element_rect(fill = "steelblue4"),
    strip.text       = element_text(colour = "white", face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  ) +
  guides(colour = guide_legend(nrow = 1))

ggsave("figures/read_tracking_counts.png", plot = p1,
       width = 10, height = 9, units = "in", dpi = 600)

p1
```

### Read counts at each processing step — fixed y-axis (raw scale)

```{r plot-tracking-fixed, fig.width=10, fig.height=9}
p1_fixed <- ggplot(plot_data, aes(x = stage, y = reads, colour = replicate, group = replicate)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  facet_wrap(~sample_type, scales = "fixed", ncol = 2) +
  scale_y_continuous(labels = label_comma(), breaks = pretty_breaks(n = 12)) +
  scale_colour_brewer(palette = "Dark2") +
  labs(
    title = "Read count changes during bioinformatic processing steps (fixed y-axis)",
    subtitle = "1. Raw reads | 2. After quality control (Porechop & Nanofilt) | 3. Removal of chicken (host) & contamination reads",
    x      = "Processing stage",
    y      = "Read count",
    colour = "Sample Number"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(face = "plain", size = 11),
    strip.background = element_rect(fill = "steelblue4"),
    strip.text       = element_text(colour = "white", face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  ) +
  guides(colour = guide_legend(nrow = 1))

ggsave("figures/read_tracking_counts_fixed.png", plot = p1_fixed,
       width = 10, height = 9, units = "in", dpi = 300)

p1_fixed
```

### Read counts at each processing step — fixed log₁₀ y-axis

```{r plot-tracking-log10, fig.width=10, fig.height=9}
p1_log <- ggplot(plot_data, aes(x = stage, y = reads, colour = replicate, group = replicate)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  facet_wrap(~sample_type, scales = "fixed", ncol = 2) +
  scale_y_log10(labels = label_comma()) +
  scale_colour_brewer(palette = "Dark2") +
  labs(
    title = "Read count changes during bioinformatic processing steps (log10 scale, fixed y-axis)",
    subtitle = "1. Raw reads | 2. After quality control (Porechop & Nanofilt) | 3. Removal of chicken (host) & contamination reads",
    x      = "Processing stage",
    y      = "Read count (log₁₀)",
    colour = "Sample Number"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(face = "plain", size = 11),
    strip.background = element_rect(fill = "steelblue4"),
    strip.text       = element_text(colour = "white", face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  ) +
  guides(colour = guide_legend(nrow = 1))

ggsave("figures/read_tracking_counts_log10.png", plot = p1_log,
       width = 10, height = 9, units = "in", dpi = 600)

p1_log
```

### Read counts at each processing step — row-shared y-axis (lower-read samples visible)

```{r plot-tracking-row-scales, fig.width=10, fig.height=9}
grid_data <- plot_data %>%
  mutate(
    processing = if_else(str_detect(sample_type, "EVI"), "Post-EVI", "Post-ASC"),
    enrichment = case_when(
      str_detect(sample_type, "BF Bolton") ~ "BF Bolton base + CAT",
      str_detect(sample_type, "BPW")       ~ "BPW then RVS",
      TRUE                                 ~ "No enrichment"
    ),
    enrichment = factor(enrichment,
                        levels = c("No enrichment",
                                   "BF Bolton base + CAT",
                                   "BPW then RVS"))
  )

p1_grid <- ggplot(grid_data, aes(x = stage, y = reads, colour = replicate, group = replicate)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  facet_grid(enrichment ~ processing, scales = "free_y") +
  scale_y_continuous(labels = label_comma(), breaks = pretty_breaks(n = 8)) +
  scale_colour_brewer(palette = "Dark2") +
  labs(
    title    = "Read count changes during bioinformatic processing steps (row-shared y-axis)",
    subtitle = "1. Raw reads | 2. After quality control (Porechop & Nanofilt) | 3. Removal of chicken (host) & contamination reads",
    x        = "Processing stage",
    y        = "Read count",
    colour   = "Sample Number"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(face = "plain", size = 11),
    strip.background = element_rect(fill = "steelblue4"),
    strip.text       = element_text(colour = "white", face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  ) +
  guides(colour = guide_legend(nrow = 1))

ggsave("figures/read_tracking_counts_row_scales.png", plot = p1_grid,
       width = 10, height = 9, units = "in", dpi = 600)

p1_grid
```

### Heatmap of read counts across processing steps

```{r plot-heatmap, fig.width=12, fig.height=10}
heatmap_data <- full_data %>%
  select(sample_type, replicate, raw_reads, post_qc_reads, post_host_reads) %>%
  pivot_longer(cols = c(raw_reads, post_qc_reads, post_host_reads),
               names_to = "stage", values_to = "reads") %>%
  mutate(
    stage = factor(stage,
                   levels = c("raw_reads", "post_qc_reads", "post_host_reads"),
                   labels = c("Raw", "Post-QC", "Post-host\ndepletion")),
    sample_type = factor(sample_type, levels = facet_order),
    replicate   = factor(replicate, levels = rev(paste0("CH0", 1:6)))
  )

p_heat <- ggplot(heatmap_data, aes(x = stage, y = replicate, fill = reads)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  geom_text(aes(label = comma(reads, accuracy = 1)), size = 2.4, colour = "black") +
  facet_wrap(~sample_type, ncol = 2) +
  scale_fill_distiller(palette = "Blues", direction = 1,
                       trans   = "log10",
                       labels  = label_comma(),
                       name    = "Read count\n(log₁₀ scale)") +
  labs(
    title = "Heatmap of read counts across bioinformatic processing steps",
    x     = "Processing stage",
    y     = NULL
  ) +
  theme_bw(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    strip.background = element_rect(fill = "steelblue4"),
    strip.text       = element_text(colour = "white", face = "bold"),
    panel.grid       = element_blank(),
    legend.position  = "right"
  )

ggsave("figures/read_tracking_heatmap.png", plot = p_heat,
       width = 10, height = 8, units = "in", dpi = 300)

p_heat
```

### Percentage of reads retained at each step

```{r plot-pct, fig.width=10, fig.height=7}
pct_data <- full_data %>%
  select(sample_type, replicate, pct_retained_qc, pct_retained_host) %>%
  pivot_longer(cols = c(pct_retained_qc, pct_retained_host),
               names_to = "stage", values_to = "pct_retained") %>%
  mutate(
    stage = factor(stage,
                   levels = c("pct_retained_qc", "pct_retained_host"),
                   labels = c("QC filtering", "Host-DNA depletion"))
  )

p2 <- ggplot(pct_data, aes(x = replicate, y = pct_retained, fill = stage)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65) +
  geom_hline(yintercept = c(50, 75), linetype = "dashed", colour = "grey50", linewidth = 0.4) +
  facet_wrap(~sample_type, ncol = 2) +
  scale_fill_manual(values = c("QC filtering" = "steelblue3", "Host-DNA depletion" = "coral3")) +
  scale_y_continuous(limits = c(0, 105), labels = function(x) paste0(x, "%")) +
  labs(
    x    = "Replicate",
    y    = "% reads retained",
    fill = "Processing step"
  ) +
  theme_bw(base_size = 12) +
  theme(
    strip.background = element_rect(fill = "steelblue4"),
    strip.text       = element_text(colour = "white", face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  )

ggsave("figures/read_tracking_pct_retained.png", plot = p2,
       width = 10, height = 7, units = "in", dpi = 300)

p2
```

---

## Summary statistics

```{r summary-stats}
full_data %>%
  group_by(sample_type) %>%
  summarise(
    `Mean raw reads`            = comma(round(mean(raw_reads))),
    `Mean post-QC reads`        = comma(round(mean(post_qc_reads))),
    `Mean post-host reads`      = comma(round(mean(post_host_reads))),
    `Mean % retained (QC)`      = paste0(round(mean(pct_retained_qc), 1), "%"),
    `Mean % retained (host dep)`= paste0(round(mean(pct_retained_host), 1), "%"),
    `Mean % retained (overall)` = paste0(round(mean(pct_retained_total), 1), "%"),
    .groups = "drop"
  ) %>%
  rename(`Sample type` = sample_type) %>%
  kable(align = c("l", rep("r", 6))) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```

---

## Base tracking

### Data

```{r base-data}
# Helper: parse "Total bases" from a NanoPlot HTML report
extract_total_bases <- function(path) {
  html <- readr::read_file(path)
  m    <- regmatches(html, regexpr("Total bases</th>\\s*<td>([^<]+)</td>", html))
  val  <- sub(".*<td>([^<]+)</td>", "\\1", m)
  as.numeric(gsub(",", "", val))   # strip commas; trailing .0 handled by as.numeric
}

nanoplot_base <- "/Volumes/Alice/2_POULTRY_PROCESSING/Analysis/Nanoplot_poultry_processing"

groups <- tribble(
  ~sample_type,                    ~folder_prefix, ~file_prefix,
  "Post-ASC",                      "ASC",          "ASC",
  "Post-ASC BF Bolton base + CAT", "ASC_CAT",      "ASC_CAT",
  "Post-ASC BPW then RVS",            "ASC_RVS",      "ASC_RVS",
  "Post-EVI",                      "EVI",          "EVI",
  "Post-EVI BF Bolton base + CAT", "EVI_CAT",      "EVI_CAT",
  "Post-EVI BPW then RVS",            "EVI_RVS",      "EVI_RVS"
)

# Raw and post-QC bases from the _raw and _filtered NanoPlot reports
base_qc <- crossing(groups, replicate = paste0("CH0", 1:6), stage = c("raw", "filtered")) %>%
  mutate(
    path  = file.path(nanoplot_base,
                      paste0(folder_prefix, "_Nanoplot"),
                      paste0("NanoPlot-report_", file_prefix, "_", replicate, "_", stage, ".html")),
    bases = map_dbl(path, extract_total_bases),
    col   = if_else(stage == "raw", "raw_bases", "post_qc_bases")
  ) %>%
  select(sample_type, replicate, col, bases) %>%
  pivot_wider(names_from = col, values_from = bases)

# Post-host-depletion bases from the host-dep NanoPlot reports
base_host <- crossing(groups, replicate = paste0("CH0", 1:6)) %>%
  mutate(
    path            = file.path(nanoplot_base,
                                paste0(folder_prefix, "_Host-dep_Nanoplot"),
                                paste0(replicate, "_", file_prefix, "_host.html")),
    post_host_bases = map_dbl(path, extract_total_bases)
  ) %>%
  select(sample_type, replicate, post_host_bases)

base_full <- base_qc %>%
  left_join(base_host, by = c("sample_type", "replicate")) %>%
  mutate(
    pct_retained_qc    = round(post_qc_bases   / raw_bases      * 100, 1),
    pct_retained_host  = round(post_host_bases / post_qc_bases  * 100, 1),
    pct_retained_total = round(post_host_bases / raw_bases      * 100, 1)
  )
```

---

### Base tracking tables

#### All sample types — QC step

```{r base-table-qc}
base_full %>%
  select(sample_type, replicate, raw_bases, post_qc_bases, pct_retained_qc) %>%
  mutate(
    raw_bases       = comma(raw_bases),
    post_qc_bases   = comma(post_qc_bases),
    pct_retained_qc = paste0(pct_retained_qc, "%")
  ) %>%
  rename(
    `Sample type`      = sample_type,
    `Replicate`        = replicate,
    `Raw bases`        = raw_bases,
    `Post-QC bases`    = post_qc_bases,
    `% retained (QC)`  = pct_retained_qc
  ) %>%
  kable(align = c("l", "c", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  pack_rows("Post-ASC", 1, 6) %>%
  pack_rows("Post-ASC BF Bolton base + CAT", 7, 12) %>%
  pack_rows("Post-ASC BPW then RVS", 13, 18) %>%
  pack_rows("Post-EVI", 19, 24) %>%
  pack_rows("Post-EVI BF Bolton base + CAT", 25, 30) %>%
  pack_rows("Post-EVI BPW then RVS", 31, 36)
```

#### All samples — host-DNA depletion step

```{r base-table-host}
base_full %>%
  select(sample_type, replicate, post_qc_bases, post_host_bases, pct_retained_host) %>%
  mutate(
    post_qc_bases     = comma(post_qc_bases),
    post_host_bases   = comma(post_host_bases),
    pct_retained_host = paste0(pct_retained_host, "%")
  ) %>%
  rename(
    `Sample type`               = sample_type,
    `Replicate`                 = replicate,
    `Post-QC bases`             = post_qc_bases,
    `Post-host depletion bases` = post_host_bases,
    `% retained (host removal)` = pct_retained_host
  ) %>%
  kable(align = c("l", "c", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  pack_rows("Post-ASC", 1, 6) %>%
  pack_rows("Post-ASC BF Bolton base + CAT", 7, 12) %>%
  pack_rows("Post-ASC BPW then RVS", 13, 18) %>%
  pack_rows("Post-EVI", 19, 24) %>%
  pack_rows("Post-EVI BF Bolton base + CAT", 25, 30) %>%
  pack_rows("Post-EVI BPW then RVS", 31, 36)
```

#### All samples — full pipeline summary

```{r base-table-full}
base_full %>%
  select(sample_type, replicate, raw_bases, post_qc_bases, post_host_bases, pct_retained_total) %>%
  mutate(
    raw_bases          = comma(raw_bases),
    post_qc_bases      = comma(post_qc_bases),
    post_host_bases    = comma(post_host_bases),
    pct_retained_total = paste0(pct_retained_total, "%")
  ) %>%
  rename(
    `Sample type`          = sample_type,
    `Replicate`            = replicate,
    `Raw bases`            = raw_bases,
    `Post-QC`              = post_qc_bases,
    `Post-host depletion`  = post_host_bases,
    `% retained (overall)` = pct_retained_total
  ) %>%
  kable(align = c("l", "c", "r", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  pack_rows("Post-ASC", 1, 6) %>%
  pack_rows("Post-ASC BF Bolton base + CAT", 7, 12) %>%
  pack_rows("Post-ASC BPW then RVS", 13, 18) %>%
  pack_rows("Post-EVI", 19, 24) %>%
  pack_rows("Post-EVI BF Bolton base + CAT", 25, 30) %>%
  pack_rows("Post-EVI BPW then RVS", 31, 36)
```

---

### Visualisations

```{r base-plot-data}
base_plot_data <- base_full %>%
  select(sample_type, replicate, raw_bases, post_qc_bases, post_host_bases) %>%
  pivot_longer(cols = c(raw_bases, post_qc_bases, post_host_bases),
               names_to = "stage", values_to = "bases") %>%
  mutate(
    stage = factor(stage,
                   levels = c("raw_bases", "post_qc_bases", "post_host_bases"),
                   labels = c("Raw", "Post-QC", "Post-host\ndepletion")),
    sample_type = factor(sample_type, levels = facet_order)
  )
```

#### Base counts at each processing step — fixed log₁₀ y-axis

```{r base-plot-log10, fig.width=10, fig.height=9}
pb_log <- ggplot(base_plot_data, aes(x = stage, y = bases, colour = replicate, group = replicate)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  facet_wrap(~sample_type, scales = "fixed", ncol = 2) +
  scale_y_log10(labels = label_comma()) +
  scale_colour_brewer(palette = "Dark2") +
  labs(
    title    = "Base count changes during bioinformatic processing steps",
    subtitle = "1. Raw bases | 2. After quality control (Porechop & Nanofilt) | 3. Removal of chicken (host) & contamination reads",
    x        = "Processing stage",
    y        = "Base count (log₁₀)",
    colour   = "Sample Number"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(face = "plain", size = 11),
    strip.background = element_rect(fill = "steelblue4"),
    strip.text       = element_text(colour = "white", face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  ) +
  guides(colour = guide_legend(nrow = 1))

ggsave("figures/base_tracking_counts_log10.png", plot = pb_log,
       width = 10, height = 9, units = "in", dpi = 600)

pb_log
```

#### Base counts at each processing step — free y-axis

```{r base-plot-free, fig.width=10, fig.height=9}
pb_free <- ggplot(base_plot_data, aes(x = stage, y = bases, colour = replicate, group = replicate)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  facet_wrap(~sample_type, scales = "free_y", ncol = 2) +
  scale_y_continuous(labels = label_comma()) +
  scale_colour_brewer(palette = "Dark2") +
  labs(
    title    = "Base count changes during bioinformatic processing steps",
    subtitle = "1. Raw bases | 2. After quality control (Porechop & Nanofilt) | 3. Removal of chicken (host) & contamination reads",
    x        = "Processing stage",
    y        = "Base count",
    colour   = "Sample Number"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(face = "plain", size = 11),
    strip.background = element_rect(fill = "steelblue4"),
    strip.text       = element_text(colour = "white", face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  ) +
  guides(colour = guide_legend(nrow = 1))

ggsave("figures/base_tracking_counts.png", plot = pb_free,
       width = 10, height = 9, units = "in", dpi = 600)

pb_free
```

#### Base counts at each processing step — row-shared y-axis

```{r base-plot-grid, fig.width=10, fig.height=9}
base_grid_data <- base_plot_data %>%
  mutate(
    processing = if_else(str_detect(sample_type, "EVI"), "Post-EVI", "Post-ASC"),
    enrichment = case_when(
      str_detect(sample_type, "BF Bolton") ~ "BF Bolton base + CAT",
      str_detect(sample_type, "BPW")       ~ "BPW then RVS",
      TRUE                                 ~ "No enrichment"
    ),
    enrichment = factor(enrichment,
                        levels = c("No enrichment", "BF Bolton base + CAT", "BPW then RVS"))
  )

pb_grid <- ggplot(base_grid_data, aes(x = stage, y = bases, colour = replicate, group = replicate)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  facet_grid(enrichment ~ processing, scales = "free_y") +
  scale_y_continuous(labels = label_comma(), breaks = pretty_breaks(n = 8)) +
  scale_colour_brewer(palette = "Dark2") +
  labs(
    title    = "Base count changes during bioinformatic processing steps (row-shared y-axis)",
    subtitle = "1. Raw bases | 2. After quality control (Porechop & Nanofilt) | 3. Removal of chicken (host) & contamination reads",
    x        = "Processing stage",
    y        = "Base count",
    colour   = "Sample Number"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(face = "plain", size = 11),
    strip.background = element_rect(fill = "steelblue4"),
    strip.text       = element_text(colour = "white", face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  ) +
  guides(colour = guide_legend(nrow = 1))

ggsave("figures/base_tracking_counts_row_scales.png", plot = pb_grid,
       width = 10, height = 9, units = "in", dpi = 600)

pb_grid
```

#### Heatmap of base counts across processing steps

```{r base-plot-heatmap, fig.width=12, fig.height=10}
base_heat_data <- base_full %>%
  select(sample_type, replicate, raw_bases, post_qc_bases, post_host_bases) %>%
  pivot_longer(cols = c(raw_bases, post_qc_bases, post_host_bases),
               names_to = "stage", values_to = "bases") %>%
  mutate(
    stage = factor(stage,
                   levels = c("raw_bases", "post_qc_bases", "post_host_bases"),
                   labels = c("Raw", "Post-QC", "Post-host\ndepletion")),
    sample_type = factor(sample_type, levels = facet_order),
    replicate   = factor(replicate, levels = rev(paste0("CH0", 1:6)))
  )

pb_heat <- ggplot(base_heat_data, aes(x = stage, y = replicate, fill = bases)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  geom_text(aes(label = comma(bases, accuracy = 1)), size = 2.4, colour = "black") +
  facet_wrap(~sample_type, ncol = 2) +
  scale_fill_distiller(palette = "Blues", direction = 1,
                       trans   = "log10",
                       labels  = label_comma(),
                       name    = "Base count\n(log₁₀ scale)") +
  labs(
    title = "Heatmap of base counts across bioinformatic processing steps",
    x     = "Processing stage",
    y     = NULL
  ) +
  theme_bw(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    strip.background = element_rect(fill = "steelblue4"),
    strip.text       = element_text(colour = "white", face = "bold"),
    panel.grid       = element_blank(),
    legend.position  = "right"
  )

ggsave("figures/base_tracking_heatmap.png", plot = pb_heat,
       width = 10, height = 8, units = "in", dpi = 300)

pb_heat
```

---

## Base tracking summary statistics

```{r base-summary-stats}
base_full %>%
  group_by(sample_type) %>%
  summarise(
    `Mean raw bases`            = comma(round(mean(raw_bases))),
    `Mean post-QC bases`        = comma(round(mean(post_qc_bases))),
    `Mean post-host bases`      = comma(round(mean(post_host_bases))),
    `Mean % retained (QC)`      = paste0(round(mean(pct_retained_qc), 1), "%"),
    `Mean % retained (host dep)`= paste0(round(mean(pct_retained_host), 1), "%"),
    `Mean % retained (overall)` = paste0(round(mean(pct_retained_total), 1), "%"),
    .groups = "drop"
  ) %>%
  rename(`Sample type` = sample_type) %>%
  kable(align = c("l", rep("r", 6))) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```
